---
title: "Morphoscape"
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Morphoscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  # eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.width=8, 
  fig.height=5, 
  fig.align = "center"
)
```


```{r setup}
library(Morphoscape)
```

The following provides a guide on the workflow of performing an adaptive landscape analysis, from the philosophies of morphospaces, and types of data required, to the details of how to use the functions in this "Morphoscape" package. This guide will cover the following topics:

    1. Defining a Morphospace
      - Phenotype
      - Performance data: Specimens or Warps
    2. Using Morphoscape
      - Creating Performance Surfaces
      - Calculate Landscapes
  
# Defining a morphospace
## Phenotype
Adaptive landscape analyses require two types of data to be married together: phenotypic data and performance data. Phenotype data can take many forms, and the only requirement of phenotype data to the construction of adaptive landscapes is the definition of a morphospace. The simplest type of morphospace can be a 2D plot with the axes defined by two numeric measurements of phenotype, such as the length and width of a skull, though any quantification of phenotype is valid. 

Geometric morphometrics has become the quintessential method for quantifying detailed shape variation in bony structures, and the final outcome in many studies is the visualization of a PCA of shape variation - which is a morphospace. However, how one defines the morphospace can drastically impact the outcomes of these adaptive landscape analyses. A PCA and between-group PCA are both valid methods to ordinate morphological data, yet will produce dramatically different morphospaces depending on the goals of the analysis. PCA will produce a (largely) unbiased ordination of the major axes of variation, while bgPCA will find axes of variation that maximize differences in groups. Caution should be applied when using constrained ordination like bgPCA, as they can create ecological patterns where none may actually exist (see Bookstein 2019, Cardini et al 2019 and Cardini & Polly 2020 for the cautionary debate). However, in many ways bgPCA, or other constrained ordinations are ideal for questions regarding functional and adaptive differences between ecological groups when actual morphological differences are present. 

For now the adaptive landscape methods in this package are limited to 2D morphospaces, with two axes of phenotypic variation. This largely done out of the complexities involved in analyzing and visualizing multivariate covariance in more than 3 dimensions. However, if more than two axes of phenotypic variation are desired to be analyzed, these can be done in separate analyses.

## Collecting performance data: Specimens or Warps
Once an ordination and 2D morphospace of the phenotypic data is defined, one must then collect data on performance of these phenotypes. There are two approaches one can take to collecting performance data: collect data directly from specimens, or collect data from phenotypic warps in morphospace. The former is the simplest and most direct as long as your dataset is not too large, and allows you to to know actual performance of actual specimens. However, error can creep in if the performance traits do not strongly covary with the axes of your morphospace, and can produce uneven and inconsistent surfaces. In addition, specimens may not evenly occupy morphospace resulting in regions of morphospace not defined by existing phenotypes, and thus will not have measured performance data and may produce erroneous interpolations or extrapolations of performance. Finally, collecting performance data can be time consuming, and may be impractical for extremely large datasets.  

The alternative is to collect performance data from hypothetical warps across morphospace, which eliminates these issues. As long as the method used to define your morphospace is reversible (such as `prcomp`, `geomorph::gm.prcomp`, `Morpho::prcompfast`, `Morpho::groupPCA`, `Morpho::CVA`), it is possible to extract the phenotype at any location in morphospace. In fact many of these ordination functions come with `predict` methods for this very task. As such, using prediction, it is possible to define phenotypes evenly across all of morphospace called warps. These warps are also useful in they can be defined to represent phenotypic variation for ONLY the axes of the morphospace, and ignore variation in other axes. These warps can however form biologically impossible phenotypes (such as the walls of a bone crossing over one another), which may or may not help your interpretations of why regions of morphospace might be occupied or not. 

Both these approaches are valid as long as morphospace is reasonably well covered. For an in-depth analysis of morphospace sampling see Smith et al (2021).

This package comes with the turtle humerus dataset from Dickson and Pierce (2019), which uses performance data collected from warps.

```{r}
data("turtles")
data("warps")

str(turtles)
str(warps)
```

`turtles` is a dataset of coordinate data for 40 turtle humerus specimens that have been ordinated in a bgPCA morphospace, the first two axes maximizing the differences between three ecological groups: Marine, Freshwater and Terrestrial turtles. This dataset also includes these and other ecological groupings.

`warps` is a dataset of 4x6 evenly spaced warps predicted from this morphospace and 4 performance metrics.

# Using Morphoscape

Once a morphospace is defined and performance data collected, the workflow of using Morphoscape is fairly straightforward. Using the `warp` and `turtles` datasets the first step is to make a functional dataframe using `as_fnc_df()`. The input to this function is a dataframe containing both coordinate data and performance data (and also grouping factors if desired). The first two columns must be coordinates, while the other columns can be defined as performance data, or as grouping factors. It is best to have your performance data named at this point to keep track.

```{r}
library(Morphoscape)

data("turtles")
data("warps")

str(turtles)
str(warps)

# Create an fnc_df object for downstream use
warps_fnc <- as_fnc_df(warps, func.names = c("hydro", "curve", "mech", "fea"))
str(warps_fnc)
```

## Creating Performance Surfaces

It is then a simple process to perform surface interpolation by automatic Kriging using the `krige_surf()` function. This will autofit a kriging function to the data. This is performed by the `automap::autoKrige()` function. For details on the fitting of variograms you should read the documentation of `automap`. All the `autoKrige` fitting data is kept in the `kriged_surfaces` object along with the output surface.

By default the `krige_surf()` function will interpolate within an alpha hull wrapped around the inputted datapoints. This is to avoid extrapolation beyond measured datapoints. This can be defined using the `resample_grid()` function, which will supply a grid object defining the points to interpolate, and optionally plot the area to be reconstructed. The strength of wrapping can be controlled using the `alpha` argument, with smaller values producing a stronger wrapping.

```{r}
# Create alpha-hulled grid for kriging
grid <- resample_grid(warps, hull = "concaveman", alpha = 3, plot = TRUE)
kr_surf <- krige_surf(warps_fnc, grid = grid)
kr_surf
plot(kr_surf)
```

However, if one wishes to also extrapolate to the full extent of morphospace, set `hull = NULL`. Because the `warps` dataset evenly samples morphospace, we can set `hull = NULL` and reconstruct the full rectangle of morphospace. When `hull = NULL` an amount of padding will also be applied to provide some space beyond the supplied datapoints and can be controlled using the `padding` argument.  Finally, we can also specify the density of interpolated points using the `resample` argument.


```{r}
# Create alpha-hulled grid for kriging
grid <- resample_grid(warps, hull = NULL, padding = 1.1)

# Do the kriging on the grid
kr_surf <- krige_surf(warps_fnc, grid = grid)
kr_surf
plot(kr_surf)
```

This reconstructed surface is missing actual specimen data points with associated ecological groupings, which are needed for later analyses. These can be added using the `krige_new_data()` function.

```{r}
# Do kriging on the sample dataset
kr_surf <- krige_new_data(kr_surf, new_data = turtles)
kr_surf
plot(kr_surf)
```

This of course all can all be done in one step. `krige_surf` will automatically call `resample_grid()` if no `grid` argument is supplied, and `new_data` can be supplied as the data by which later group optimums are calculated against.

```{r}
# Above steps all in one:
kr_surf <- krige_surf(warps_fnc, hull = NULL, padding = 1.1,
                      new_data = turtles)
kr_surf
plot(kr_surf)
```

## Calculate Landscapes

The next step is to calculate a distribution of adaptive landscapes. Each adaptive landscape is constructed as the summation of the performance surfaces in differing magnitudes, Each performance surface is multiplied by a weight ranging from 0-1, and the total sum of weights is equal to 1. For four equally weighted performance surfaces, the weights would be the vector `c(0.25, 0.25, 0.25, 0.25)`. To generate a distribution of different combinations of these weights, use the `generate_weights()` function. `generate_weights` must have either a `step` or `n` argument to determine how many allocations to generate, and `nvar` the number of variables. One can also provide either the `fnc_df` or `kr_surf` objects to the `data` argument. `step` determines the step size between weight values. `step = 0.1` will generate a vectors of `c(0, 0.1, 0.2 ... 1)`. Alternatively, one can set the number of values in the sequence set. `n = 10` will generate the same vectors of `c(0, 0.1, 0.2 ... 1)`. The function will then generate all combinations of `nvar` variables that sum to 1. 

For four variables, this will produce 286 combinations. A step size of 0.05 will produce 1771 rows. As the step size gets smaller, and the number of variables increases, the number of output rows will exponentially increase. It is recommended to start with large `step` sizes, or small `n` to ensure things are working correctly.

```{r}
# Generate weights to search for optimal landscapes
weights <- generate_weights(n = 10, nvar = 4)
weights <- generate_weights(step = 0.05 , data = kr_surf)

```

This `weights` matrix is then provided to the `calc_all_landscapes()` along with the `kr_surf` object to calculate the landscapes for each set of weights. For calculations with a large number of weights, outputs can take some time and can be very large in size, and it is recommended to utilize the `file` argument to save the output to file.

```{r}
# Calculate all landscapes; setting verbose = TRUE produces
# a progress bar
all_landscapes <- calc_all_lscps(kr_surf, grid_weights = weights)
```

With a distribution of landscapes, it is now possible to find landscapes that maximize 'fitness' for a given subset or group of your specimens in `new_data` using the `calcWprimeBy()` function. The `by` argument sets the grouping variable for the data provided in `new_data` from earlier. There are several ways `by` can be set: A one sided formula containing the name of a column in `new_data` or a vector containing a factor variable. 


```{r}
# Calculate optimal landscapes by Group

table(turtles$Ecology)

wprime_by_Group <- calcWprimeBy(all_landscapes, by = ~Ecology)
wprime_by_Group <- calcWprimeBy(all_landscapes, by = turtles$Ecology)

wprime_by_Group
summary(wprime_by_Group)
plot(wprime_by_Group, ncol = 2)

```

It is also possible to use `calcGrpWprime` to enumerate a single group or for the entire sample:

```{r}
# Calculate landscapes for one Group at a time

i <- which(turtles$Ecology == "T")
wprime_T <- calcGrpWprime(all_landscapes, index = i)
wprime_T
wprime_b <- calcGrpWprime(all_landscapes, Group == "box turtle")
wprime_b
plot(wprime_b)

wprime_all <- calcGrpWprime(all_landscapes)
wprime_all

```

Finally, it is possible to compare the landscapes for each group using `calcGrpWprime()`. This is done by comparing the number of landscape combinations are shared by the upper percentile between groups. 

```{r}

# Test for differences between Group landscapes
tests <- multi.lands.grp.test(wprime_by_Group)
tests

# Calculate landscapes for one Group at a time
wprime_b <- calcGrpWprime(all_landscapes, Group == "box turtle")
wprime_b
wprime_t <- calcGrpWprime(all_landscapes, Group == "tortoise")
wprime_t

# Test for differences between Group landscapes
lands.grp.test(wprime_b, wprime_t)
```

## References

Bookstein, F. L. (2019). Pathologies of between-groups principal components analysis in geometric morphometrics. Evolutionary Biology, 46(4), 271-302.

Cardini, A., O’Higgins, P., & Rohlf, F. J. (2019). Seeing distinct groups where there are none: spurious patterns from between-group PCA. Evolutionary Biology, 46(4), 303-316.

Cardini, A., & Polly, P. D. (2020). Cross-validated between group PCA scatterplots: A solution to spurious group separation?. Evolutionary Biology, 47(1), 85-95.

Dickson, B. V., & Pierce, S. E. (2019). Functional performance of turtle humerus shape across an ecological adaptive landscape. Evolution, 73(6), 1265-1277.

Smith, S. M., Stayton, C. T., & Angielczyk, K. D. (2021). How many trees to see the forest? Assessing the effects of morphospace coverage and sample size in performance surface analysis. Methods in Ecology and Evolution, 12(8), 1411-1424.
