---
title: "Complete Example"
author: "Guillermo Basulto-Elias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete example of glancedata}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Whenever I start working on a dataset, I need to take a glance at the data to see how the data are or if the format is the one that I am  expecting. I found myself coding similar lines over and over again with each data set I put my hands on. I decided to put that lines together in an R package so I and others can use them. I called it **`glancedata`**.

There are some already cool R packages to summarize information. Two of the best, in my opinion, are `skimr` and `GGally`. In this vignette, I provide examples of the functions in `glancedata` as well as some of the functions in these two packages.

Below is a table with the functions shown in this vignette.


```{r, echo = FALSE, message=FALSE}

library(dplyr)

pkg_list <- list(

    ## skim
  c(
    pkg = "`skimr`",
    fnct = "`skim`",
    des = "Alternative to `summary`. Friendly with `dplyr::group_by()`."
  ),
  
  ## glance_data
  c(
    pkg = "`glancedata`",
    fnct = "`glance_data`",
    des = "Alternative to `summary`. Emphasizes missing data and binary variables."
  ),
  
  ## glance_data_in_workbook
  c(
    pkg = "`glancedata`",
    fnct = "`glance_data_in_workbook`",
    des = "Similar to `glance_data`. Creates list of dataframes instead and saves XLSX file."
  ),
  
  ## plot_numerical_vars
  c(
    pkg = "`glancedata`",
    fnct = "`plot_numerical_vars`",
    des = "Creates a plot per numerical variable. It might be histogram, density plot, qqplot, violin plot or scatterplot."
  )
,
  
  ## plot_discrete_vars
  c(
    pkg = "`glancedata`",
    fnct = "`plot_discrete_vars`",
    des = "Creates a plot per variable with up to 20 different values. This limit can be adjusted.."
  )
,
  
  ## ggpairs
  c(
    pkg = "`GGally`",
    fnct = "`ggpairs`",
    des = "Creates plots for pairwise comparison of columns."
  )
) ## End list


pkg_list %>%
  purrr::map_df(~ tibble(Package = .x[1], 
                         Function = .x[2], 
                         Description = .x[3])) %>%
  knitr::kable()
  
```





## Example (`USArrests`)

I am going to use a built-in dataset in `R`. I added some columns to it so you we can see what happens with different type of columns. **You may load your own dataset** instead of `sample_data`.

The example we are going to use is `USArrests`, which contains statistics, in arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states in 1973. Also given is the percent of the population living in urban areas, according to the data set description (type `help("USArrests")` in the console to see more details).


```{r setup, message=FALSE, cache=TRUE}
library(dplyr)
library(tidyr)
library(knitr)

sample_data <- 
  tibble(State = state.name,
         Region = state.region) %>%
  bind_cols(as_tibble(state.x77)) %>%
  bind_cols(USArrests)

kable(head(sample_data))


```


## Use of `skimr`

There are many packages useful 


```{r, eval = FALSE, dependson= "setup"}
## Load package
library(skimr)

## Call main function
skim(sample_data)

```


```{r, dependson = "setup", echo = FALSE, results='asis', message=FALSE}

library(glancedata)
library(dplyr)
library(tidyr)
library(knitr)
library(skimr)

skim(sample_data)



```

## `glance_data`

```{r, dependson= "setup"}

library(glancedata)

glance_data(sample_data)

```

## `glance_data_in_workbook`

```{r}
library(glancedata)

glance_data_in_workbook(sample_data)

```


## Testing a date mode

```{r, eval = FALSE}

library(readr)
library(dplyr)

## Read data with dates from FiveThirtyEight
polls <- 
  read_csv("https://github.com/fivethirtyeight/data/tree/master/polls") 


```

## Future versions
