---
title: "Diversity and Inclusion Report"
subtitle: "This report uses randomly generated data and does not reflect the situation in a perticular company."
author: Dr. Philippe De Brouwer
date: "`r format(Sys.time(), '%d %B, %Y')`"
abstract: "This report informs about how diverse the team is and tries to indicate potential areas of bias. This report cannot prove or disprove bias, it can only direct the user to potential problems. Hence this report is in the first place an invitation to think. Much like feedback from other people is an invitation to think and not a diagnosis per se."
output:
  beamer_presentation:
    slide_level: 2
    toc: false
    keep_tex: true
classoption:
  - usenames
  - dvipsnames
header-includes:
 - \usetheme[progressbar=frametitle]{metropolis}
 - \usepackage{booktabs}
 - \usepackage{longtable}
 - \usepackage{array}
 - \usepackage{multirow}
 - \usepackage{wrapfig}
 - \usepackage{float}
 - \usepackage{colortbl}
 - \usepackage{pdflscape}
 - \usepackage{tabu}
 - \usepackage{threeparttable}
 - \usepackage{threeparttablex}
 - \usepackage[normalem]{ulem}
 - \usepackage{makecell}
 - \usepackage{xcolor}
bibliography: bibliography.bib
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Libraries
library(div)
library(gridExtra)
library(kableExtra)
library(grid)
library(lattice)

# Options and defaults:
options(knitr.table.format = "latex")
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(fig.width=12, fig.height=6) 

# The fake team:
teamSize       <- 400
pctFemales     <- 0.35
maleSalaryBias <- 1.035
d <- div_fake_team(seed = 18901228, 
                   N = teamSize, 
                   genders = c("F", "M"),
                   gender_prob = c(pctFemales, 1 - pctFemales),
                   gender_salaryBias = c(1, maleSalaryBias)
                   )


# Gender diversity tibble per grade:
tbl_gender_div <- table(d$gender, d$grade) %>%
 apply(2, diversity, prior = c(50.2, 49.8)) %>%
 tibble(value = ., label = paste("Grade", names(.)))

# Average gender diversity tibble:
tbl_gender_div_all <- table(d$gender) %>%
  as.vector %>%
  diversity(prior = c(50.2, 49.8)) %>%
 tibble(value = ., label = "Gender Diversity")

# Age diversity tibble:
ageRef <- c(15.94, 27.11, 23.99, 20.68, 12.28) / 100
ad <- table(cut(d$age, breaks = c(20, 30, 40, 50, 60, 65)))
ad <- ad / sum(ad)  # not needed, diversity() does this, bu useful for the table
tbl_age_div <- tibble(value = diversity(ad, prior = ageRef), label = "Age Diversity")

# Gender PayGap tibble:
pg_gender <- div_paygap(d)
               

# Citizenship PayGap tibble
pg_citizenship <- d                %>%
             div_paygap(x = "citizenship" , x_ctrl = "Polish")
             

```



## Table of Contents

\tableofcontents[hideallsubsections]


# The Particular Example Behind this Demo Report


The data used for this report is randomly generated with the following characteristics:

-   team size: `r teamSize`
-   percentage of females: `r pctFemales`
-   average male's salary / average female's salary: `r maleSalaryBias`
- no other biases are built in (so any other observations stem from random generation of data. This can be seen as "despite no bias against citizenship by the manager, some pay-gaps will be different from one." This means that being unbiased is not necessarily the same as having equal outcomes.)

### Equitable outcomes are not the same as equal outcomes!


# Overview

## Findings (in order of importance)

\small
```{r findings}
f1 <- tibble(Nbr = 1, 
             Area = "Gender", 
             Finding = "Where we can calculate the paygap between females and non-females, we find that the females generally earn less in similar roles and similar grades.", 
             Suggestion = "Check the gender-paygap table and identify the grade/role combinations where an the paygap has most stars. Check if the salary differences are justified.")
f2 <- tibble(Nbr = 2, 
             Area = "Age", 
             Finding = "The team is predominantly younger than the surrounding population (Poland).", 
             Suggestion = "Consider hiring older people to balance. Focus on retention.")
f3 <- tibble(Nbr = 3, 
             Area = "Gender", 
             Finding = "The diversity is good in grade 1 and 2, but under par in grade 3", 
             Suggestion = "Consider if females have barriers to apply to grade 3 jobs and remove the barriers.")
f4 <- tibble(Nbr = 4, 
             Area = "Gender", 
             Finding = "Males in Grade 2 seem to have been promoted faster.", 
             Suggestion = "Understand unconscious bias, coach everyone (and specially females), work on trust.")
findings <- rbind(f1, f2, f3, f4)
colnames(findings) <- c("Nbr", "Area", "Finding", "Suggestion")

knitr::kable(findings) %>% column_spec(3:4, width = "12em")
write_csv(findings,'findings.csv')
```


## Dashboard

```{r dash, fig.height=9, fig.width=12}
f_wrap <- function(x) {
  strwrap(x, width = 55, simplify = FALSE) %>%
  sapply(paste, collapse = "\n")
}

lay1 <- rbind(c(1,1,2),
              c(1,1,2)
              )
p1_1 <- div_gauge_plot(tbl_gender_div, ncol = 2, nbrSize = 4)
p1_2 <- div_gauge_plot(tbl_gender_div_all, ncol = 1)
p1 <- arrangeGrob(p1_1, p1_2, layout_matrix = lay1)
p2 <- textGrob(f_wrap(f3$Finding))

p3 <- div_gauge_plot(tbl_age_div, ncol = 1)
p4 <- textGrob(f_wrap(f2$Finding))

p5 <- div_plot_paygap_distribution(pg_gender$data$paygap, label = "Gender")
p6 <- textGrob(f_wrap(f1$Finding))

p7 <- div_plot_paygap_distribution(pg_citizenship$data$paygap, label = "Citizenship")
p8 <- textGrob(f_wrap("No Bias detetected for Citizenship (in salary) -- both team and unbiased distribution are virtually the same"))

lay <- rbind(c(2,2,4,4),
             c(1,1,3,3),
             c(1,1,3,3),
             c(1,1,3,3),
             c(6,6,8,8),
             c(5,5,7,7),
             c(5,5,7,7),
             c(5,5,7,7)
             )

grid.arrange(p1, p2, p3, p4, p5, p6, p7, p8, layout_matrix = lay)
```

# Diversity


## Gender diversity per grade


```{r, fig.cap="The diversity of the team with respect to gender per grade.\\label{fig:gender-gauge}"}
t <- table(d$gender, d$grade)
t1 <- apply(t, 2, diversity, prior = c(50.2, 49.8))
t2 <- tibble(value = t1, label = paste("Grade", names(t1)))
div_gauge_plot(t2, ncol = 3)
```

## Age diversity

```{r ageism, fig.cap="The diversity of the team with respect to age, assuming the age distribution of the country as reference.\\label{fig:ageism}"}
# ages in Poland
p1 <- div_gauge_plot(tbl_age_div)
t <- rbind(ad, ageRef)
rownames(t) <- c("team", "reference")
#knitr::kable(round(t,2), 
#             caption = "The percentages of different age groups in this team compared to the population of Poland. An equal distribution would yield a diversity of 1.\\label{tab:div:age}")

t2 <- t %>% t() %>% 
  as_tibble %>%
  mutate(age_group = rownames(t(t))) %>%
  as_tibble %>% gather("population", "percentage", 1:2)

p2 <- ggplot(t2, aes(x = age_group, y = percentage, fill = population)) +
  geom_bar(stat="identity",  position=position_dodge()) +
  geom_text(aes(label=round(percentage,2)), vjust=1.6, color="black", position = position_dodge(0.9), size=3.5) + 
  theme(legend.position="bottom")

grid.arrange(p1, p2, ncol=2, widths = 1:2)

```


## Diversity in nationalities (1/2)
\framesubtitle{Distribution for the team}
```{r nationalities, fig.height = 6,fig.cap = c("The barplot for the nationalities in the team over all grades.")}
#t <- table(d$citizenship, d$grade)
t <- d                     %>% 
     group_by(citizenship) %>% 
     summarise(n = n())    %>%
     arrange(desc(n))      %>%
     ggplot( aes(x = citizenship, y = n))    +
        geom_col(aes(fill = n), width = 0.7) +
        theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5),
              axis.title.x=element_blank())
              
t
```

## Diversity in nationalities (2/2)
\framesubtitle{Breakdown per Grade}
```{r nationalities2, fig.height = 6, fig.cap = c("The breakdown of each grade per nationalitiy.")}
t <- d                            %>% 
     group_by(citizenship, grade) %>% 
     summarise(n = n())           %>%
     ggplot( aes(x = grade, y = n))                +
        geom_col(aes(fill = n), width = 0.7)       +
        facet_wrap(~ citizenship, scales="free_y") +
        theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5),
                axis.title.x=element_blank())

t
```

# Inclusion

## The Gender PayGap

```{r parsePaygapGender, echo=FALSE}
div_parse_paygap(pg_gender)
```

## The Citizenship PayGap

```{r parsePaygapCitizenship, echo=FALSE}
div_parse_paygap(pg_citizenship)
```

## The Age Paygap
```{r agePaygap, echo=FALSE}
div_paygap(d, x = "age") %>% div_parse_paygap
```

## Time in firm paygap
```{r timeInTeamPaygap, echo=FALSE}
div_paygap(d, x = "tenure_firm") %>% div_parse_paygap
```

## Job Changes per Year per Gender

```{r TenureJob, echo=FALSE, fig.cap="Job changes per year indicate mobility and risk taking. They are a good indication for promotion (see Figure \\ref{fig:prom_p_gender})."}
ylim1 = boxplot.stats(1/d$tenure_job)$stats[c(1, 5)]
p <- ggplot(data = d, aes(x=gender, y=1/tenure_job, fill=gender)) + 
    geom_boxplot() +
    facet_wrap(~grade) + 
    coord_cartesian(ylim = ylim1*1.05)
p
```

## Promotions per Year per Gender

```{r TenureGrade, echo=FALSE, fig.cap="The number of promotions per year can show if a gender is more probable to be promoted.\\label{fig:prom_p_gender}"}
ylim1 = boxplot.stats(1/d$tenure_job)$stats[c(1, 5)]
p <- ggplot(data = d, aes(x=gender, y=1/tenure_grade, fill=gender)) + 
    geom_boxplot() +
    facet_wrap(~grade) + 
    coord_cartesian(ylim = ylim1*1.05)
p
```

# Conclusions

```{r conclusions}
conclusions <- rbind(tibble(Nbr = 0, 
                            Suggestion = 'Learn more by reading e.g. "The Essentials of Diversification & Inclusion", Dabrowska (2019)' ),
                     findings[,c(1,4)])
knitr::kable(conclusions) %>% column_spec(2, width = "22em")
```
---
nocite: | 
  @diversityhub2019
---

# Appendices

## Legend Paygap

-   `paygap` = the ratio of median salaries of one group divided by the median of the salaries of the other group
-   \colorbox{Gray}{`NA`} = numbers are too small, please look at individuals;
-   \colorbox{green}{nothing} = no bias detectable;
-   \colorbox{GreenYellow}{`.`} = maybe there is some bias, but the numbers are low, check individuals;
-   \colorbox{Yellow}{`*`} = you should check for bias;
-   \colorbox{BurntOrange}{`**`} = bias is probably there;
-   \colorbox{red}{`***`} = most certainly there is bias

So, there will be more stars if the probability of a bias is higher: this can be due to a higher bias and/or due to a larger sample size.

## Legend: Paygap Column Headers

\small

 - `grade` = the salary grade as used in the company
 - `jobID` = a unique identifier of the job category (can be abbreviated)
 - `sal_F` = the median salary of the females (F)
 - `sal_oth` = the median salary of the other groups (non F). The tool is open to use more than one gender.
 - `age_F` = the median age of the females (or `age_Pol` could be the median age of the team members with Polish citizenship)
 - `age_oth` = the median age of the other groups take together (e.g. the median age of non females)
 - `paygap` = the ratio of median salary earned by the selected group (e.g. females) divided by the median of the other people. If this is lower than $1$, then median female earns less than the median non-female.
 - `conf.` = the confidence level that this paygap is significant.


## The Diversity Index (1/2) {.shrink}

We express diversity as a number between zero and one. Our calculation is based on @debrouwer2020 and more in particular section 36.3.1 ``The Business Case: a Diversity Dashboard''. Details can be found in the book. The method is:

  * The diversity is $0$ if only one of the groups is present, and is $1$ if both groups are equitably present.
  * This calculation is similar to the established concept of entropy in physics. 
  * More than two categories can be used (e.g. one is not limited to two genders)
  * We calibrate the probabilities so that they show maximum entropy (or diversity) for the percentages that naturally occur (see next slide).
  
## The Diversity Index (2/2)

```{r diversityIllustrated, fig.cap = "The diversity index illustrated for the case where there are only two possible classes, and where the prior priorities are respectively 50/50 (top) and 70/30 (bottom). The index reaches a maximum at a distribution equal to the prior probabilities.\\label{fig:diversityIllustrated}"}
x <- seq(length.out = 500, from = 0, to = 1)
t <- cbind(x, 1-x)
tt <- tibble(diversity = apply(t, 1, FUN = diversity),
             x = x)
p1 <- ggplot(data=tt, aes(x=x, y=diversity, color = diversity)) +
        geom_line(lwd = 3) +
        ggtitle("Diversity index for binary variables and prior probabilities of 50/50") + 
        xlab("Proportion of one of the two classes")

tt <- tibble(diversity = apply(t, 
                               1, 
                               FUN = diversity, 
                               prior = c(.7,.3)),
             x = x)
p2 <- ggplot(data=tt, aes(x=x, y=diversity, color = diversity)) +
        geom_line(lwd = 3) +
        ggtitle("Diversity index for binary variables and prior probabilities of 70/30") + 
        xlab("Proportion of one of the two classes")
grid.arrange(p1, p2, ncol = 1)
```


## The confidence level and p-value
The p-value is the probability that we make a mistake by assuming that there is no paygap.

It is calculated by splitting the data on a variable in binary factors (e.g. Females and others) and then checking how likely it is that a random person from the first group earns less than a random person from the second group. This is done by a method known as Mann-Whitney U test: [see Wikipedia](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test)[^1]


[^1]: The Mann–Whitney U test (aka. Mann–Whitney–Wilcoxon (MWW), Wilcoxon rank-sum test, or Wilcoxon–Mann–Whitney test) is a nonparametric test of the null hypothesis that, for randomly selected values X and Y from two populations, the probability of X being greater than Y is equal to the probability of Y being greater than X. If we assume that the distributions are symmetric, it boils down to a test that the medians are different.

## Another view on the PayGap

```{r otherPayGapGender, fig.height = 6, echo=FALSE, fig.cap="Boxplots for each grade (over all job categories) per gender. This another view of the same data as in Table \\ref{tab:pg:gender}."}
# Transformation function for 2 decimals
scaleFUN <- function(x) sprintf("%.0f", x)

p <- ggplot(data = d, aes(x=gender, y=salary, fill=gender)) + 
    geom_boxplot() +
    facet_wrap(~grade + jobID, scale="free") + 
    scale_y_continuous(labels=scaleFUN)
p
```



## Bibliography
