---
title: "Aggregation over partitions"
author: "Ingo Rohlfing"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aggregation over partitions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(QCAcluster)
```

The functions `upset_conditions()` and `upset_conjunctions()` can be used after
one has derived partition-specific models with `partition_min()` or
`partition_min_inter()`. The functions take the models from the `solution` 
column of the dataframes produced with `partition_min()`/or 
`partition_min_inter()` as input to produce an UpSet plot. `upset_conditions()` 
and `upset_configurations()` are functions that draw on the `upset()`
function of the [UpSetR](https://CRAN.R-project.org/package=UpSetR/) 
package. We use the dataset by 
[Grauvogel and von Soest (2014)](https://doi.org/10.1017/S1755773910000251) 
for illustrating the meaning and interpretation of UpSet plots. 

### Aggregation over individual conditions
The following plot aggregates the occurrence and co-occurrence of single 
conditions over the partition-specific parsimonious solutions.
```{r, fig.width=6, fig.height=4}
data("Grauvogel2014")
# parsimonious solution for each type of Sender
GS_pars <- partition_min(
 dataset = Grauvogel2014,
 units = "Sender",
 cond = c("Comprehensiveness", "Linkage", "Vulnerability",
          "Repression", "Claims"),
 out = "Persistence",
 n_cut = 1, incl_cut = 0.75,
 solution = "P",
 BE_cons = rep(0.75, 3),
 BE_ncut = rep(1, 3))
# UpSet plot with three sets
upset_conditions(GS_pars, nsets = 4)
```

Each line (or horizontal bar) in an UpSet plot displays how often a single 
condition occurs over all `r nrow(GS_pars)` partition-specific models. 
The plot shows that `Claims` occurs as a single condition in 15 models out of 
`r nrow(GS_pars)`.

Each column or vertical bar in the plot shows how often single conditions 
occur together in a model. (This means two conditions that co-occur in a model
are not necessarily part of the same conjunction.) The plot shows that three 
different sets of conditions occur together in five models each and that 
three more combinations of conditions are specific to one model each.

### Aggregation over individual sufficient terms
The function `upset_configurations()` decomposes models into the constitutive
sufficient terms. A 'term' can be a configuration (aka as conjunction) of
conditions or single sufficient conditions. Nonetheless, we called this function
`upset_configurations()` for convenience and to distinguish it from the 
`upset_conditions()` function.
```{r, fig.width=6, fig.height=4}
upset_configurations(GS_pars, nsets = 3)
```

Over one pooled model and 15 within-models (there is extensive model ambiguity),
the individually sufficient condition `Vulnerability` is part of eight models
and is the only sufficient term in five models. In three additional models,
it coincides with the conjunction `~Repression * Claims`. In total,
`~Repression * Claims` occurs six times in the 16 models.