---
title: "Get LERI data over a region of interest and save a GeoTIFF"
author: "Maxwell B. Joseph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get LERI data over a region of interest and save a GeoTIFF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  fig.width = 8, 
  fig.height = 5, 
  purl = NOT_CRAN,
  eval = NOT_CRAN
)
```

The leri R package provides easy access to the Landscape Evaporative Response
Index (LERI) data - an experimental drought monitoring and early warning 
guidance tool produced by the National Oceanic and Atmospheric Administration.

The LERI product is available from the year 2000 to present at a 1 km spatial
resolution over the continental United States, at the following timescales:

- 1, 3, 7, and 12 month
- 8 day accumulated or non-accumulated from April - October

More information on the LERI product is available on the 
[NOAA LERI homepage](https://www.esrl.noaa.gov/psd/leri/).

This vignette covers a common use case acquiring data over a region of interest
defined by a shapefile, masking the LERI data to that region, and saving 
GeoTIFF files containing data for the region of interest.


## Defining a region of interest

By default, the leri package returns data for the continental United States, 
southern parks of Canada, and northern parts of Mexico. 
But, you may only be interested in a region of interest, as defined by a 
shapefile. 
Here, you will load a shapefile for the state of North Carolina that is 
distributed by default with the [sf](https://cran.r-project.org/package=sf)
package. 

```{r load-roi-shapefile}
library(sf)
library(raster)
library(viridis)
library(leri)
roi <- st_read(system.file("shape/nc.shp", package="sf"))
```

If you are using a different shapefile, replace
`system.file("shape/nc.shp", package="sf")` with its file path, e.g., 
`st_read("path/to/file.shp")`.

The `roi` object contains multiple columns of data, and a `geometry` column
that contains spatial information on the region of interest, which in this 
case consists of multiple counties. 

```{r print-roi-shapefile}
roi
```

Because you don't necessarily care about each county, but rather you want the
entire state (including all counties) you can use a spatial union to join data
from all counties:

```{r}
roi <- st_union(roi)
roi
```



## Acquiring LERI data

To acquire LERI data, you can use the `get_leri()` function. 
You will fetch the 8 day accumulated timescale data for the week of August
13, 2018:

```{r get-data}
leri_raster <- get_leri(date = "2018-08-13", product = "8 day ac")
```

The `leri_raster` object is a `RasterLayer`, and you can see
information on the spatial extent, resolution, and coordinate reference 
system by printing the object: 

```{r inspect-leri}
leri_raster
```

Plot the data with a custom color palette to see what the data look like:

```{r plot-leri-conus}
plot(leri_raster, col = cividis(255))
```



## Masking to the region of interest

Now you want to subset or mask the LERI data to the region of interest. 
First, you need to ensure that the raster data and the polygon for the region 
of interest have the same coordinate reference system. 

```{r reproject-to-same-crs}
roi_reprojected <- st_transform(roi, crs = projection(leri_raster))
```

Now, graphically verify that they align as expected: 

```{r plot-leri-with-shp}
plot(leri_raster, col = cividis(255))
plot(roi_reprojected, add = TRUE)
```

Now, you can crop the LERI data to match extents with the region of interest, 
then mask the raster set all values outside of the region of interest to `NA`. 
Because the raster package requires sp objects, rather than sf objects, you
will coerce our roi to a sp object first. 

```{r mask-leri}
roi_sp <- as(roi_reprojected, 'Spatial')
cropped_leri <- crop(leri_raster, roi_sp)
masked_leri <- mask(cropped_leri, roi_sp)
```

You can plot the masked raster along with the ROI to confirm:

```{r}
plot(masked_leri, col = cividis(255))
plot(roi_sp, add = TRUE)
```

## Saving GeoTIFF output

To write a GeoTIFF file of our `masked_leri` object, you can use `writeRaster`:

```{r write-tif, eval = FALSE}
writeRaster(masked_leri, 'leri-over-roi.tif')
```
