---
title: "An Introduction to the `distfreereg` Package"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{An Introduction to the `distfreereg` Package}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
linkcolor: blue
link-citations: true
csl: american-statistical-association.csl
bibliography: bibliography.bib
---

<!-- https://stackoverflow.com/questions/23957278/how-to-add-table-of-contents-in-rmarkdown -->

\def\eqdef{\mathrel{\raise0.4pt\hbox{$:$}\hskip-2pt=}}
\def\ev{{\rm E}}
\def\given{\mathrel|}

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, out.width = "75%",
                      fig.dim = c(6,5), fig.align = "center")
is_check <- ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(eval = !is_check)

library(distfreereg)
```

<!-- https://kbroman.org/pkg_primer/pages/vignettes.html -->


# Introduction

This vignette introduces the basic functionality of the `distfreereg` package.
The package has two main functions: `distfreereg()` and `compare()`. The former
is introduced below; `compare()` is introduced
[here](../doc/v2_compare.html).

The function `distfreereg()` implements the distribution-free regression testing
procedure introduced by @Khmaladze2021. Its purpose is to test whether or not a
specified function $Y=f(X;\theta)$ is the conditional mean, $\ev(Y\given X)$.
Specifically, the test is of the following type:
\begin{equation}
  H_0{:}\ \exists\theta\in\Theta\,|\,\ev(Y\given X)=f(X;\theta)
  \quad\hbox{against}\quad
  H_1{:}\ \forall\theta\in\Theta\,|\,\ev(Y\given X)\neq f(X;\theta).
  \label{eqn:hypo1}
\end{equation}

This vignette starts by discussing the [testing of a linear
model](#testing-a-linear-model) created using `lm()`. Following this is an
analogous discussion of [testing a non-linear
model](#testing-a-non-linear-model) created using `nls()`. Both scenarios can be
replicated using the [formulas directly](#means-specified-by-formulas). In the
[most general implementation](#means-specified-by-functions), the mean function
is specified by an `R` function. To use `distfreereg()` with a mean function
implemented in software other than `R`, see the section on the [default
method](#the-default-method-using-the-algorithm-with-external-functions).






# Testing a Linear Model

Suppose that a linear model object `m` is created using `lm()`, and we want to
test whether or not a linear model is appropriate; specifically, whether or not
the true mean function follows a specified form that is linear in its
parameters.

In some cases, a residual plot suffices to show that the specified form is
wrong. Below, the true mean function is $\ev(y|x)=x^{2.1}$, but the formula used
to build the model is linear in $x$. The residuals plot clearly indicates model
misspecification.

```{r lm_testing_with_plots}
set.seed(20240304)
n <- 3e2
form_lm <- y ~ x
data_lm <- data.frame(x = runif(n, min = 0, max = 3))
data_lm$y <- data_lm$x^2.1 + rnorm(n, sd = 0.3)
m_1 <- lm(form_lm, data = data_lm)
plot(m_1, which = 1)
```

If, however, we use a form that is quadratic in $x$, the residuals plot is
suggestive, but it leaves us without a clear conclusion. How certain can we be
that the specified form is wrong?

```{r create_lm}
form_lm_2 <- y ~ I(x^2)
m_2 <- lm(form_lm_2, data = data_lm)
plot(m_2, which = 1)
```

In this case, there appears to be evidence of a mis-specified mean, but we need
a formal test to quantify our (un)certainty. The function `distfreereg()`
implements such a test.

```{r lm_method}
set.seed(20240304)
(dfr_lm_2 <- distfreereg(test_mean = m_2))
```

In fact, tests based on two statistics are included by default: a
Kolmogorov--Smirnov (KS) statistic and a Cramér--von Mises (CvM) statistic,
defined as $\max_i|w_i|$ and ${1\over n}\sum_iw_i^2$, respectively, where
$w\eqdef(w_1,\ldots,w_n)$ is the partial sum process calculated by
`distfreereg()`. The preceding output shows a table summarizing the tests'
results, including p-values and Monte Carlo standard errors thereof. Both of the
p-values are marginal, supplying a measure of our intuitive conclusion based on
the residuals plot.






# Testing a Non-Linear Model

The procedure just illustrated applies to objects created with `nls()`, as well.
Here, the data generating function has the form being tested, so we should not
expect the null to be rejected. Note that the `weights` argument in `nls()` is
set equal to the inverse of the variance vector.

```{r create_nls}
set.seed(20240304)
n <- 3e2
sds <- runif(n, min = 0.5, max = 5)
data_nls <- data.frame(x = rnorm(n), y = rnorm(n))
data_nls$z <- exp(3*data_nls$x) - 2*data_nls$y^2 + rnorm(n, sd = sds)
form_nls <- z ~ exp(a*x) - b*y^2
m_2 <- nls(form_nls, data = data_nls, weights = 1/sds^2, start = c(a = 1, b = 1))
```

Testing this form of the mean structure is done just as with an `lm` object:

```{r nls_method}
set.seed(20240304)
(dfr_nls <- distfreereg(test_mean = m_2))
```

Both tests fail to reject the null.







# Means Specified by Formulas

If no `lm` or `nls` object has been created yet, the formula and data can be
supplied directly to `distfreereg()`. Below, we recreate the examples from above
without first creating the `lm` and `nls` objects.

The most notable practical difference in implementing the tests in this context
is that three arguments are required rather than just one:

1. `test_mean`: a `formula` object specifying the mean function;

1. `data`: a data frame containing the data;

1. `method`: specifies whether to use `lm()` or `nls()`. Defaults to "`lm`".



## Formulas with `lm()`

The following call recreates the setup behind `dfr_lm_2` above.

```{r formula_method_lm}
set.seed(20240304)
(dfr_form_lm_2 <- distfreereg(test_mean = form_lm_2, data = data_lm))
```

The key pieces of the output are identical to those of `dfr_lm`:

```{r comparison_lm}
identical(dfr_lm_2$observed_stats, dfr_form_lm_2$observed_stats)
identical(dfr_lm_2$p, dfr_form_lm_2$p)
```


## Formulas with `nls()`

Two additional arguments are used here:

1. `method`: a character string specifying the method to use to fit the model.
Its default value is "`lm`", so we must specify "`nls`" here.

1. `covariance`: a list containing a named object specifying the error
covariance structure. This ultimately supplies the weights values to `nls()`.
This can be omitted if weights are not being used. (See
[below](#specifying-the-covariance-structure) for more details.)

1. `theta_init`: a numeric vector specifying the starting values for parameter
estimation. (This is optional, since `nls()` will use default starting values
with a warning if not supplied. See `nls()` documentation for further details.)

The following call recreates the setup behind `dfr_nls`. The covariance
structure is specified by, `P`, the precision matrix. This produces results
identical, in the sense of `identical()`, to those from the example from above.
Using any other option (e.g., `Sigma = diag(sds^2)`) would result in numerically
equivalent results in the sense of `all.equal()`.

```{r formula_method_nls}
set.seed(20240304)
(dfr_form_nls <- distfreereg(test_mean = form_nls, data = data_nls, method = "nls",
                             covariance = list(P = diag(1/sds^2)),
                             theta_init = c(a = 1, b = 1)))
```

The key pieces of the output are identical to those of `dfr_nls`:

```{r comparison_nls}
identical(dfr_nls$observed_stats, dfr_form_nls$observed_stats)
identical(dfr_nls$p, dfr_form_nls$p)
```






# Means Specified by Functions

The most general case that can be handled entirely within `R` is that in which
the mean regression function is specified by an `R` function.

Suppose that we want to replicate the situation in the `nls` examples above, but
we want to account for errors that have a non-diagonal covariance matrix. (The
`weights` argument of `nls()` limits that function's applicability to diagonal
covariance matrices.) The example below illustrates how to do this.

The function being tested is specified by `test_mean`, which is a function with
two arguments, `X` and `theta`. Here, `X` represents the entire matrix of
covariates. Therefore, to replicate the references to `x` and `y` in [this
example](#testing-a-non-linear-model), we use `X[,1]` and `X[,2]`, respectively.
The arguments `X` and `Y` supply the covariate matrix and the response vector,
respectively, in lieu of the `data` argument.

```{r dfr_1}
set.seed(20240304)
n <- 3e2
true_mean <- function(X, theta) exp(theta[1]*X[,1]) - theta[2]*X[,2]^2
test_mean <- true_mean
theta <- c(3,-2)
Sigma <- rWishart(1, df = n, Sigma = diag(n))[,,1]
X <- matrix(rnorm(2*n), nrow = n)
Y <- distfreereg:::f2ftheta(true_mean, X)(theta) +
  distfreereg:::rmvnorm(n = n, reps = 1, SqrtSigma = distfreereg:::matsqrt(Sigma))

(dfr_1 <- distfreereg(test_mean = test_mean, Y = Y, X = X,
                      covariance = list(Sigma = Sigma),
                      theta_init = rep(1, length(theta))))
```





# The Default Method: Using the Algorithm with External Functions

All of the examples above illustrate how `distfreereg()` can be used when the
mean function being tested is defined in `R`. If the mean function is not
defined in `R`, `distfreereg()` can still be used as long as certain objects can
be imported into `R`. Specifically, `distfreereg()` needs the model's fitted
values and the Jacobian of the mean function, evaluated at the estimated
parameter vector for each observation's covariate values.

To illustrate this, we extract certain elements from the example
[above](#means-specified-by-functions), and pretend that these were created
using external software, and then imported into `R`.

```{r default_inputs}
J <- dfr_1$J
fitted_values <- fitted(dfr_1)
```

These can be supplied to `distfreereg()` to implement the test. Note that
`test_mean` is set to `NULL`, and since no parameter estimation is done here,
no value for `theta_init` is specified.

```{r default}
distfreereg(test_mean = NULL, Y = Y, X = X, fitted_values = fitted_values,
            J = J, covariance = list(Sigma = Sigma))
```






# Summary and Diagnostic Plots

Below is an introduction to the three types of plots available for `distfreereg`
objects. See the [plotting vignette](../doc/v3_plotting.html) for
more details regarding plot customization.

## A Summary Plot

A summary of the results of the test can be plotted easily.
```{r}
plot(dfr_1)
```

The density of the simulated statistic is plotted, as is a vertical line
showing the observed statistic. The upper tail is shaded, and the p-value (the
area of that shaded region) is shown. Finally, a 95% confidence band of the
density function is plotted, and the region is shaded in grey.

## Diagnostic Plots

Two diagnostic plots are also available. The first produces a time-series-like
plot of the residuals in the order specified by `res_order` in the `distfreereg`
object.

```{r residuals_plot}
plot(dfr_1, which = "residuals")
```

The second produces a plot of the empirical partial sum process.

```{r epsp_plot}
plot(dfr_1, which = "epsp")
```





# Specifying the Covariance Structure

The covariance structure of the errors, conditional on the covariates, must be
specified. For the `lm` and `nls` methods, the default behavior is to extract
the covariance structure from the supplied model object. In general, the
covariance structure is specified using the `covariance` argument. The value of
`covariance` is a list with at least one named element that specifies one of
four matrices:

1. `Sigma`, the covariance matrix;

1. `SqrtSigma`, the square root of `Sigma`;

1. `P`, the precision matrix (that is, the inverse of `Sigma`); and

1. `Q`, the square root of `P`.

Internally, the algorithm only needs `Q`, so some efficiency can be gained by
supplying this directly if it is known. Supplying more than one of the four
matrices is not forbidden, but is strongly discouraged. In this case, no
verification is done that the supplied matrices are consistent with each other,
and `Q` will be calculated using the most convenient supplied element.

Each of the four named elements can be one of three types of object:

1. A positive-definite matrix, the most general way to specify a covariance
structure;

1. a numeric vector of positive values whose length is the sample size; and

1. a length-1 numeric vector specifying a positive number.

Specifying a numeric vector is theoretically equivalent to specifying a diagonal
matrix with the given vector along the diagonal. Specifying a single number is
theoretically equivalent to specifying a diagonal matrix with that value along
the diagonal. Using the simplest possible expression in each case is recommended
for conceptual and computational simplicity.

The following code shows that supplying `Q` produces observed statistics
identical to those in [the previous example](#means-specified-by-functions).

```{r dfr_2}
Q <- distfreereg:::matsqrt(distfreereg:::matinv(Sigma, tol = .Machine$double.eps))
dfr_2 <- distfreereg(Y = Y, X = X, test_mean = test_mean,
                     covariance = list(Q = Q),
                     theta_init = rep(1, length(theta)))
identical(dfr_1$observed_stats, dfr_2$observed_stats)
```








# Methods

Several common generic functions have `distfreereg` methods, including `coef()`,
`predict()`, and `update()`. See the documentation for the complete list.



# Adding New Statistics

The `stat` argument of `distfreereg()` allows the user to specify other
statistics by specifying any function whose input is a numeric vector and whose
output is a numeric vector of length one.

```{r new_stat}
new_stat <- function(x) sum(abs(x))
update(dfr_1, stat = "new_stat")
```

If necessary, the default statistics can be selected using "`KS`" and "`CvM`" as
explicit values in the character vector supplied to `stat`.

```{r three_stats}
update(dfr_1, stat = c("new_stat", "KS", "CvM"))
```

Checks are done to verify that the functions provided calculate legitimate
statistics; that is, that they take a numeric vector as input and output a
single number.

```{r stats_checks}
bad_stat <- function(x) x[1,2]
try(update(dfr_1, stat = c("KS", "CvM", "bad_stat")))
try(update(dfr_1, stat = c("KS", "CvM", "undefined_stat")))
```



# Non-Normal Errors

One of the strengths of the test implemented by `distfreereg()` is that it makes
only weak assumptions about the distribution of the errors. Specifically, it
only requires that the error covariance matrix be finite and that the error
distribution be strongly mixing (also known as $\alpha$-mixing). Below is an
example illustrating that we get the expected results when the error
distribution is the $t$ distribution. Note that the $t$ distribution with three
degrees of freedom has variance 3.

```{r t_dist}
set.seed(20241003)
n <- 1e2
func <- function(X, theta) theta[1] + theta[2]*X[,1] + 0.5*X[,1]^2
theta <- c(2,5)
X <- matrix(rexp(n))
Y <- theta[1] + theta[2]*X[,1] + 0.5*X[,1]^2 + rt(n, df = 3)

dfr <- distfreereg(Y = Y, X = X, test_mean = func, theta_init = c(1,1),
                   covariance = list(Sigma = 3))
dfr

alt_func <- function(X, theta) theta[1] + theta[2]*X[,1]
update(dfr, test_mean = alt_func)
```







# References
