---
title: "Cellular assays using ‘bioassays’ package in R"
author: "Anwar Azad Palakkan, Jamie Davies"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bioassays-examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bioassays)
```


# Libraries 

Following libraries will be useful. If the libraries are not installed, please install them using **install.packages()** 

```{r libraries, message=FALSE, warning=FALSE, echo=TRUE, eval = FALSE}
library(tcltk)# for selecting the folder for analysis
library(dplyr)
library(ggplot2)# for plotting graphs
library(reshape2)
library(nplr)# for the standard curve fitting
```


# Introduction

In a cell culture lab various cellular assays are performed. The package "bioassays" will help to analyse the results of these experiments performed in multiwell plates. The functions in this package can be used to summarise data from any multiwell plate, and by incorporating them in a loop several plates can be analyzed automatically. Two examples are shown in this article. All the csv files used for the examples are availble in the folder exdata (inst/exdata).


# Example 1: Analysing result from a 96 well plate


## Input data

To set up a folder as working directory

```{r directory,echo=TRUE, eval = FALSE}
path1<-tk_choose.dir(getwd(), "Choose the folder for Analysis")
# A window will popup and ask you to select the folder containg the data  
setwd(path1)
```

Read files

Files can be read using the codes below. Files need to be in .csv format. In this examples file names reflect the type of assay used. For example  *"L_DIFF_P3_72HRS.csv"* :*L* is assay code (Lactate assay), *DIFF* differentiation (type of cells used), *p3*: Plate 3 (each plate represent specific compounds tested), *72HRS* 72hrs treatment with compound. These information will help to summarise the results.


```{r files}
filelist <-list("L_HEPG2_P3_72HRS.csv","L_HEPG2_P3_24HRS.csv") # list of files
fno<-1 # file number (in fileslist) that is going to be analyzed
result <- data.frame(stringsAsFactors= FALSE) ## An empty dataframe to dump result
zzz <- data.frame(stringsAsFactors= FALSE) ## An empty dataframe to dump result
```
 
 To read 1st file

```{r readcsv}
filename<-extract_filename(filelist[fno])[1]
filename
nickname<-extract_filename(filelist[fno], split="_",end=".csv",remove="",sep="")[2]
nickname
```


```{r readcsv2, eval=FALSE}
rawdata<-read.csv(filename,stringsAsFactors = FALSE, strip.white = TRUE, 
                  na.strings = c("NA",""),header = TRUE,skip=1)
head(rawdata)
```

```{r readcsv3, echo=FALSE}
data(rawdata96)
rawdata<-rawdata96
head(rawdata)
```

Reading metadata file

```{r metafile, eval=FALSE}
metadata<-read.csv("metafile.csv",stringsAsFactors = FALSE,strip.white = TRUE, 
                  na.strings = c("NA",""),header = TRUE)
head(metadata)
```

```{r metafile2, echo=FALSE}
data(metafile96)
metadata<-metafile96
head(metadata)
```

## Rearranging the data

96 well plates were used for the assay, so it is assumed that in data the rows being from A to G and columns 1 to 12 of a 96 well plate. data2plateformat function is used to label them correctly.


```{r rearrange-plate-reading}
rawdata<-data2plateformat(rawdata,platetype = 96)
head(rawdata)
```


## To convert data into a dataframe.

data2plateformat function uses the column name and row name of the above rawdata to format it as a data frame. So the data should be well formatted before using this function. 

```{r plate2df OD}
OD_df<- plate2df(rawdata)
head(OD_df)
```


## For an overview of the plate 

```{r heatmap, normalization, echo = TRUE, eval=TRUE, fig.width=4, fig.height=3.5}
data<-matrix96(OD_df,"value",rm="TRUE")
heatplate(data,"Plate 1", size=5)
```

## Filling metadata file using plate specific details

An example function is given below which will determine the compound,concentration, type and dilution from the file name

```{r joining metatadata}
plate_info<-function(file,i){
  
  file<-file[1]
  plate<- extract_filename(file,split = "_",end = ".csv", remove = " ", sep=" ")[5]

if(plate == "P2"){
  compound<-"CyclosporinA"   # Concentration of cyclosporinA used for experiment
  concentration<-c(0,1,5,10,15,20,25,50) # Concentration of cyclosporinA used for experiment
  type<-c("S1","S2","S3","S4","S5","S6","S7","S8") # sample names of corresponding concentration
  dilution<-5
  plate_meta<-list(compound=compound,concentration=round(concentration,2),type=type,
                   dilution=dilution)
}


if(plate == "P3"){
  compound<-"Taxol"
  concentration<-c(0,0.0125,.025,.05,0.1,1,5,10) 
  type<-c("S1","S2","S3","S4","S5","S6","S7","S8")
  dilution<-5
  plate_meta<-list(compound=compound,concentration=round(concentration,2),type=type,
                   dilution=dilution)
}


if(plate =="p4"){
  compound<-c("Cisplatin")
  concentration<-c(0,0.5,2,4,8,16,64,"") 
  type<-c("S1","S2","S3","S4","S5","S6","S7","") 
  dilution <- 5
  plate_meta<-list(compound=compound,concentration=round(concentration,2),type=type,
                   dilution=dilution)
}
  return(plate_meta)
}

plate_details<-plate_info(filelist,1)
plate_details
```

These 'plate details' can be used to fill metadata using plate_metadata function

```{r metadata compiling}
metadata1<-plate_metadata(plate_details,metadata,mergeby="type")
head(metadata1)
```


For joining metadata and platelayout 'inner_join' function is used

```{r metafile-joining, warning=FALSE}
data_DF<- dplyr::inner_join(OD_df,metadata1,by=c("row","col","position"))
assign(paste("data",sep="_",nickname),data_DF) # create a copy of data with platename
head(data_DF)
```


## Sorting blank wells and reducing blanks
 
 Blank can be reduced using 'reduceblank' function 

```{r reduce-blank}
data_DF<-reduceblank(data_DF, x_vector =c("All"),blank_vector = c("Blank"), "value")
head(data_DF)
assign(paste("Blkmin",sep="_",nickname),data_DF) # create a copy of data with platename
```

## Plotting standard curve

To filter standards

```{r standard-curve-calculation, warning=FALSE}
std<- dplyr::filter(data_DF, data_DF$id=="STD")  
std<- aggregate(std$blankminus ~ std$concentration, FUN = mean )
colnames (std) <-c("con", "OD")
head(std)
```

Calculations for standard curve : nonparametric logistic regression curve

```{r nprc graph, fig.width= 3, fig.height=3,warning=FALSE}

fit1<-nplr::nplr(std$con,std$OD,npars=3,useLog = FALSE) 
#npars = 3 for 3 parametric regression curve

#for graph
x1 <- nplr::getX(fit1); y1 <- nplr::getY(fit1)
x2 <- nplr::getXcurve(fit1); y2 <- nplr::getYcurve(fit1)
plot(x1, y1, pch=15, cex=1, col="red", xlab="Concentration",
     ylab="Mean OD", main=paste("Standard Curve: ", nickname), cex.main=1)
lines(x2, y2, lwd=3, col="seagreen4")
```


To evaluate nonparametric logistic regression fitting

```{r evaluate linear fitting}
params<-nplr::getPar(fit1)$params
nplr::getGoodness(fit1)
```


To estimate the values based on logistic regression fitting

```{r as a fxn estimate1, warning= FALSE}
estimated_nplr<-estimate(data_DF,colname="blankminus",fitformula=fit1,method="nplr")
head(estimated_nplr)
```


Calculations for standard curve : linear regression curve

```{r linear-regression-plot, fig.width= 3, fig.height=3}

fit2<-stats::lm(formula = con ~ OD,data = std)
ggplot2::ggplot(std, ggplot2::aes(x=OD,y=con))+
ggplot2::ggtitle(paste("Std Curve:", nickname))+
ggplot2::geom_point(color="red",size=2)+
ggplot2::geom_line(data = ggplot2::fortify(fit2),ggplot2::aes(x=OD,y=.fitted),
          colour="seagreen4",size=1)+
ggplot2::theme_bw()
```


To evaluate linear fitting


```{r linear-fit-summary, warning=FALSE}
conpred<-estimate(std,colname="OD",fitformula=fit2,method="linear")
compare<-conpred[,c(1,3)]
corAccuracy<-cor(compare)[1,2]
corAccuracy ## Correlation accuracy of regression line
summary(fit2)

```


To estimate the values based on linear curve

```{r as a fxn estimate3, warning= FALSE}
estimated_lr<-estimate(data_DF,colname="blankminus",fitformula=fit2,method="linear")
head(estimated_lr)
```


For multiply estimated by dilution

```{r multiply by dilution1, warning=FALSE}
estimated_lr$estimated2 <- estimated_lr$estimated * estimated_lr$dilution
head(estimated_lr)
```

## Summarise the data
 
 For summarising the "estimated_lr" based on "id" and "type". 

```{r plate-summary, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, out.width=50}

result<-dfsummary(estimated_lr,"estimated2",c("id","type"),
        c("STD","Blank"),"plate1", rm="FALSE",
        param=c(strict="FALSE",cutoff=40,n=12))

result
```

## Statistical test

For the t test (S1 as control)

```{r t test 1}
pval<-pvalue(result, control="S1", sigval=0.05)
head(pval)

```

          
\newpage

# Example 2: Analysing result from a 384 well plate.

This example data contain result of dose response of few drugs (drug1, drug2, drug3, drug4) at three concentrations (C1,C2,C3) from two different cell lines (hepg2 and huh7)

## Input data.

Reading the spectrophotometer readings in csv file

```{r 384 rawdata, eval=FALSE}
rawdata2<-read.csv("384.csv",stringsAsFactors = FALSE,strip.white = TRUE, 
                  na.strings = c("NA",""),header = TRUE,skip=1)
dim(rawdata2)
head(rawdata2)

```

```{r 384 rawdata2, echo= FALSE }
data(rawdata384)
rawdata2<-rawdata384
dim(rawdata2)
head(rawdata2)
```

Reading metadata

```{r 384 metadata, eval=FALSE}
metadata2<-read.csv("metafile_384_plate.csv",stringsAsFactors = FALSE,strip.white = TRUE, 
                   na.strings = c("NA",""),header = TRUE)
head(metadata2)

```

```{r 384 metadata2, echo=FALSE}
data(metafile384)
metadata2<-metafile384
head(metadata2)
```


## Rearranging the data.

For appropriate naming of rows and column of rawdata2

```{r 384 rearrange}

rawdata2<-data2plateformat(rawdata2,platetype = 384)
head(rawdata2)

```

## To convert data into a dataframe.

```{r 384 plate2df}
OD_df2 <- plate2df(rawdata2)
head(OD_df2)
```

## For an overview of the plate.

For heat map

```{r 384overview1, echo = TRUE, eval=TRUE, fig.width=4, fig.height=3.5}

data2<-matrix96(OD_df2,"value",rm="TRUE")
heatplate(data2,"Plate 384", size=1.5)
```


## For joining metadata2 and OD_df2.

```{r 384 innerjoin}
data_DF2<- dplyr::inner_join(OD_df2,metadata2,by=c("row","col","position"))
head(data_DF2)
```

For categorical view of cells

```{r 384overview2, echo = TRUE, eval=TRUE, fig.width=4, fig.height=3.5}
data3<-matrix96(data_DF2,"cell",rm="TRUE")
heatplate(data3,"Plate 384", size=2)
```

For categorical view of compound

```{r 384overview3, echo = TRUE, eval=TRUE, fig.width=4, fig.height=3.5}
data4<-matrix96(data_DF2,"compound",rm="TRUE")
heatplate(data4,"Plate 384", size=2)
```


## Sorting blank wells and reducing blanks.

The data contain separate blanks for drug1, drug2, drug3 and drug4 (blank1, blank2,blank3 and blank 4 respectively)

```{r 384 blank}
data_blk<-reduceblank(data_DF2, 
x_vector=c("drug1","drug2","drug3","drug4"),
blank_vector = c("blank1","blank2","blank3","blank4"), "value")
dim(data_blk)
```

```{r 384 blank2}
head(data_blk)
```


## To summarise the result from a plate.

For summarising the result in the order cell, compound,concentration,type and by omitting blanks.

```{r 384 summary 1}
result2<-dfsummary(data_blk,"blankminus",
        c("cell","compound","concentration","type"),
        c("blank1","blank2","blank3","blank4"),
        nickname="384well", 
        rm="FALSE",param=c(strict="FALSE",cutoff=40,n=12))
head (result2)
dim (result2)

```

For summarising the result in the order cell, compound and concentration and by omitting blanks (all blanks are marked as "B" in concentration), drug 2 and huh7. 

```{r 384 summary 2}
result3<-dfsummary(data_blk,"blankminus",
        c("cell","compound","concentration"),
        c("B","drug2","huh7"),
        nickname="", 
        rm="FALSE",param=c(strict="FALSE",cutoff=40,n=12))
head (result3)
dim (result3)

```

## Statistical test.

For t-test

```{r 384 ttest, echo=TRUE}
pvalue<-pvalue(result3,"C3",sigval=0.05)
pvalue

```
