---
title: "SSLR basics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This package includes a collection of methods to create models for semi-supervised learning (e.g. fitting the model, making predictions, etc), with a fairly intuitive interface that is easy to use.

In [`Model list section`](.././articles/models.html) you can see the list of different classification, regression and clustering models.

### Motivation

Current packages to do semi-supervised learning do not use an intuitive interface. In this package, trying to use semi-supervised learning in an easy and intuitive way.


`SSLR` tries to solve this by providing an interface to use different models, mainly using the parsnip model interface to make the use of this package easier.

`SSLR` connects with parsnip to create different models without using too many arguments in the fit functions.

In addition, it uses other packages such as [`RSSL`](https://CRAN.R-project.org/package=RSSL) to use the same interface in an easy way.

For example, to use different ones like [`RSSL`](https://CRAN.R-project.org/package=RSSL). It has a different interface. Thanks to SSLR you can use different options to use its fit functions.

### Process

To fit the model (for example SelfTraining), you must:

* Have a defined model using parsnip
* Use your _parameters_ or using by default
* Call `fit` with formula, `fit_xy` with x and y, or `fit_x_u` with x and unlabeled data. See [`Model fitting section`](.././articles/fit.html).


For example, with `fit` function:

```{r, eval = FALSE}
rf <-  rand_forest(trees = 100, mode = "classification") %>%
  set_engine("randomForest")

m <- selfTraining(learner = rf) %>% fit(Wine ~ ., data = train)
```

Or with `fit_xy` function:

```{r, eval = FALSE}
rf <-  rand_forest(trees = 100, mode = "classification") %>%
  set_engine("randomForest")

m <- selfTraining(learner = rf) %>% fit_xy(x = train[,-cls], y = train$Wine)
```

This uses the `parsnip` package that has an intuitive interface to create a Random Forest model and this can be used in the `SSLR` package in a simple way.
