---
title: "Table 2"
author: "Zhaohua Lu"
date: "2024-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load package

## Install BayesianHybridDesign

After install the package, load the functions in the package.
```{r}
#(1) Install devtools if not installed. Remove # below to install the package

#install.packages("devtools")

#(2) Install BayesianHybridDesign. Remove # below to install the package

#devtools::install_github("phe9480/BayesianHybridDesign")

library(BayesianHybridDesign)
```

## Load other packages

```{r}

library(knitr)
library(xtable)


library(foreach)
library(doParallel)
library(RBesT)

```


# Generate Data

```{r}
set.seed(123)
NCore = 7
#get_OC.betaMix

nc = c(45,45,45,45,45,45,45)
nt = nc
pc = c(0.15,0.2,0.25,0.30,0.35,0.4,0.45)
pt = pc + 0.2

nc.c = nc
nt.c = nc.c
pc.c = c(0.3,0.3,0.3,0.3,0.3,0.3,0.3)

nche = c(45,45,45,45,45,45,45)
pch=0.30
nch=180
ns = length(nc)

nsim = 100000#

Calibration.Idx = c(4,4,4,4,4,4,4)

theta = 0.5
eta = 1
a0c = 0.001
b0c = 0.001
a0t = 0.001
b0t = 0.001
delta_threshold = 0.1
alpha = 0.1

xt_list = list()
xt_typeI_list = list()
xc_list = list()

for(s in 1:ns){
  set.seed(2000)
  
  xt_typeI_list[[s]] = rbinom(nsim,size = nt[s],prob = pc[s])
  xc_list[[s]] = rbinom(nsim,size = nc[s], prob = pc[s])
  xt_list[[s]] = rbinom(nsim,size = nt[s],prob = pt[s])
  
}

```

# DPP simulation

## Demonstrate Dynamic Power Prior with 4 methods for $w_d$ 

```{r}

DPPMethodNames = c("Empirical Bayes",
                   "Bayesian p",
                   "Generalized BC",
                   "JSD")
nm = length(DPPMethodNames)

dn = list(paste("Setting",1:ns,sep=" "),DPPMethodNames)
dn3 = dn
dn3[[3]] = 1:nsim

TypeI_DPP = matrix(NA,ns,nm,dimnames=dn) 
Power_DPP = matrix(NA,ns,nm,dimnames=dn) 
DynamicBoundary_DPP = matrix(NA,ns,nm,dimnames=dn) 
pc.mi_DPP = matrix(NA,ns,nm,dimnames=dn) 
pc.sdi_DPP = matrix(NA,ns,nm,dimnames=dn) 
phat_pt_larger_pc_DPP = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)
tau_DPP = matrix(NA,ns,nm,dimnames=dn) 

TypeITime_DPP = matrix(NA,ns,nm,dimnames=dn) 
PowerTime_DPP = matrix(NA,ns,nm,dimnames=dn) 

for(m in 1:length(DPPMethodNames)){
  
  for(s in 1:ns){
    
    set.seed(2000)
    
    cs = Calibration.Idx[s]
    xt_typeI = xt_typeI_list[[cs]] 
    xc =  xc_list[[cs]]
    datamat_cal = cbind(xt_typeI,xc) 
    
    # Calibrate
    tau = calibration (nt=nt.c[s], pc.calib=pc.c[s], nc=nc.c[s], 
                       pch=pch, nche=nche[s], nch=nch,
                       alpha = alpha, 
                       a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                       delta_threshold=delta_threshold, 
                       method=DPPMethodNames[m], theta=theta, eta=eta,
                       datamat = datamat_cal, 
                       nsim = nsim, seed=2000)
    
    tau_DPP[s,m] = tau
    
    # Get data for type I and power
    
    xt = xt_list[[s]]
    xt_typeI = xt_typeI_list[[s]] 
    xc =  xc_list[[s]]
    # Type I error
    
    datamat_typeI = cbind(xt_typeI,xc) 
    datamat_Power = cbind(xt,xc) 
    
    
    # Calcualte Type I error rates
    start.time <- proc.time()
    TypeI = power.DPP(pt = pc[s], nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch, nche = nche[s],nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = delta_threshold,
                      method=DPPMethodNames[m], theta=theta, eta=eta,
                      datamat = datamat_typeI, 
                      nsim =nsim, seed = 2000)
    
    end.time <- proc.time()
    TypeITime_DPP[s,m] <- (end.time - start.time)[3]
    
    TypeI_DPP[s,m] = TypeI$power
    
    DynamicBoundary_DPP[s,m] = TypeI$delta.bound
    pc.mi_DPP[s,m] = TypeI$pc.PMD
    pc.sdi_DPP[s,m] = TypeI$pc.sd.PMD
    phat_pt_larger_pc_DPP[s,m,] = TypeI$phat_pt_larger_pc_all
    
    # Power
    set.seed(2000)
    
    start.time <- proc.time()
    Power = power.DPP(pt = pt[s], nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch, nche = nche[s],nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = delta_threshold,
                      method=DPPMethodNames[m], theta=theta, eta=eta,
                      datamat = datamat_Power, 
                      nsim =nsim, seed = 2000)
    
    end.time <- proc.time()
    TypeITime_DPP[s,m] <- (end.time - start.time)[3]
    
    Power_DPP[s,m] = Power$power
    
  }
}
```





## Table, Type I error, Comparing different methods
```{r}
cns = c("Study Treatment Sample Size", 
        "Study Control Sample Size", 
        "Historical Control Sample Size",
        "Study Treatment ORR",
        "Study Control ORR",
        "Historical Control ORR",
        DPPMethodNames)

Table_TypeI = matrix(NA,ns,length(cns))
colnames(Table_TypeI) = cns

Table_TypeI[,1] = nt
Table_TypeI[,2] = nc
Table_TypeI[,3] = nche
Table_TypeI[,4] = pt
Table_TypeI[,5] = pc
Table_TypeI[,6] = pch
Table_TypeI[,6+1:length(DPPMethodNames)] = TypeI_DPP


Table_Power = matrix(NA,ns,length(cns))
colnames(Table_Power) = cns

Table_Power[,1] = nt
Table_Power[,2] = nc
Table_Power[,3] = nche
Table_Power[,4] = pt
Table_Power[,5] = pc
Table_Power[,6] = pch
Table_Power[,6+1:length(DPPMethodNames)] = Power_DPP



Table_mi = matrix(NA,ns,length(cns))
colnames(Table_mi) = cns

Table_mi[,1] = nt
Table_mi[,2] = nc
Table_mi[,3] = nche
Table_mi[,4] = pt
Table_mi[,5] = pc
Table_mi[,6] = pch
Table_mi[,6+1:length(DPPMethodNames)] = pc.mi_DPP



Table_sdi = matrix(NA,ns,length(cns))
colnames(Table_sdi) = cns

Table_sdi[,1] = nt
Table_sdi[,2] = nc
Table_sdi[,3] = nche
Table_sdi[,4] = pt
Table_sdi[,5] = pc
Table_sdi[,6] = pch
Table_sdi[,6+1:length(DPPMethodNames)] = pc.sdi_DPP

Table1 = rbind(Table_TypeI,Table_Power,Table_mi,Table_sdi)

#write.table(Table1,"sdi.csv",sep=",",row=F,col=T)

kable(Table1)

```



## Simulation: Demonstrate each part of the Dynamic power prior


```{r Settings}
nt = c(45,45,45,45,45,45,45)
nc = nt
pc = c(0.15,0.2,0.25,0.3,0.35,0.4,0.45)
pt = pc + 0.2

setting_for_calibration = c(4,4,4,4,4,4,4)

nche = c(45,90,135,180)
pch=c(0.3,0.3,0.3,0.3,0.3,0.3,0.3)
nch=180
ns = length(nt)

nsim = 100000#

theta = 0.5
eta = 1
a0c = 0.001
b0c = 0.001
a0t = 0.001
b0t = 0.001

xt_list = list()
xt_typeI_list = list()
xc_list = list()

for(s in 1:ns){
  
  set.seed(2000)
  
  xt_typeI_list[[s]] = rbinom(nsim,size = nt[s],prob = pc[s])
  xc_list[[s]] = rbinom(nsim,size = nc[s], prob = pc[s])
  xt_list[[s]] = rbinom(nsim,size = nt[s],prob = pt[s])
}

```

## Dynamic Power Prior 


```{r}


nm = length(nche)

dn = list(paste("Setting",1:ns,sep=" "),nche)
dn3 = dn
dn3[[3]] = 1:nsim

TypeI_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
Power_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
DynamicBoundary_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
pc.mi_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
pc.sdi_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
phat_pt_larger_pc_DPP1 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)

wmean_DPP1 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)

TypeITime_DPP1 = matrix(NA,ns,nm,dimnames=dn) 
PowerTime_DPP1 = matrix(NA,ns,nm,dimnames=dn) 

for(s in 1:ns){
  
  for(m in 1:nm){
    
    set.seed(2000)
    
    # Calibrate
    cs = setting_for_calibration[s]
    
    datamat_typeI = cbind(xt_typeI_list[[cs]],xc_list[[cs]]) 
    
    tau = calibration (nt=nt[cs], pc.calib=pc[cs], nc=nc[cs], 
                       pch=pch[cs], nche=nche[m], nch=nch,
                       alpha = alpha, 
                       a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                       delta_threshold=delta_threshold, 
                       method="Empirical Bayes", theta=theta, eta=eta,
                       datamat = datamat_typeI, 
                       nsim = nsim, seed=2000)
    
    # Get data for type I and power
    
    
    xt = xt_list[[s]]
    xt_typeI = xt_typeI_list[[s]] 
    xc =  xc_list[[s]]
    
    # Type I error
    
    datamat_typeI = cbind(xt_typeI,xc) 
    datamat_Power = cbind(xt,xc) 
    
    start.time <- proc.time()
    TypeI = power.DPP(pt = pc[s], nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch[s], nche = nche[m], nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = delta_threshold, 
                      method="Empirical Bayes", theta=theta, eta=eta,
                      datamat = datamat_typeI, 
                      nsim =nsim, seed = 2000)
    
    
    end.time <- proc.time()
    TypeITime_DPP1[s,m] <- (end.time - start.time)[3]
    
    TypeI_DPP1[s,m] = TypeI$power
    
    DynamicBoundary_DPP1[s,m] = TypeI$delta.bound
    pc.mi_DPP1[s,m] = TypeI$pc.PMD
    pc.sdi_DPP1[s,m] = TypeI$pc.sd.PMD
    phat_pt_larger_pc_DPP1[s,m,] = TypeI$phat_pt_larger_pc_all
    
    wmean_DPP1[s,m,] = TypeI$w
    
    # Power
    set.seed(2000)
    
    start.time <- proc.time()
    Power = power.DPP(pt = pt[s],nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch[s], nche = nche[m],nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = delta_threshold, 
                      method="Empirical Bayes", theta=theta, eta=eta,
                      datamat = datamat_Power, 
                      nsim =nsim, seed = 2000)
    
    end.time <- proc.time()
    PowerTime_DPP1[s,m] <- (end.time - start.time)[3]
    
    Power_DPP1[s,m] = Power$power
    
    
  }
}
```

## Dynamic Power Prior w = a = nche / nch

```{r}

nm = length(nche)

dn = list(paste("Setting",1:ns,sep=" "),nche)
dn3 = dn
dn3[[3]] = 1:nsim

TypeI_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
Power_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
DynamicBoundary_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
pc.mi_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
pc.sdi_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
phat_pt_larger_pc_DPP2 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)


wmean_DPP2 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)

TypeITime_DPP2 = matrix(NA,ns,nm,dimnames=dn) 
PowerTime_DPP2 = matrix(NA,ns,nm,dimnames=dn) 

for(s in 1:ns){
  
  for(m in 1:nm){
    
    
    set.seed(2000)
    
    
    # Calibrate
    cs = setting_for_calibration[s]
    
    datamat_typeI = cbind(xt_typeI_list[[cs]],xc_list[[cs]]) 
    
    tau = calibration (nt=nt[cs], pc.calib=pc[cs], nc=nc[cs], 
                       pch=pch[cs], nche=nche[cs], nch=nch,
                       alpha = alpha, 
                       a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                       delta_threshold=100, 
                       method="Empirical Bayes", theta=theta, eta=eta,
                       datamat = datamat_typeI, w = nche[m]/nch, 
                       nsim = nsim, seed=2000)
    
    # Get data for type I and power
    
    
    
    xt = xt_list[[s]]
    xt_typeI = xt_typeI_list[[s]] 
    xc =  xc_list[[s]]
    
    # Type I error
    
    datamat_typeI = cbind(xt_typeI,xc) 
    datamat_Power = cbind(xt,xc) 
    
    
    start.time <- proc.time()
    TypeI = power.DPP(pt = pc[s],nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch[s], nche = nche[m], nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = 100,
                      method="Empirical Bayes", theta=theta, eta=eta,
                      datamat = datamat_typeI, w = nche[m]/nch, 
                      nsim =nsim, seed = 2000)
    
    
    end.time <- proc.time()
    TypeITime_DPP2[s,m] <- (end.time - start.time)[3]
    
    TypeI_DPP2[s,m] = TypeI$power
    
    DynamicBoundary_DPP2[s,m] = TypeI$delta.bound
    pc.mi_DPP2[s,m] = TypeI$pc.PMD
    pc.sdi_DPP2[s,m] = TypeI$pc.sd.PMD
    phat_pt_larger_pc_DPP2[s,m,] = TypeI$phat_pt_larger_pc_all

    
    wmean_DPP2[s,m,] = TypeI$w
    
    # Power
    set.seed(2000)
    
    start.time <- proc.time()
    Power =  power.DPP(pt = pt[s],nt = nc[s], 
                       pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                       pch = pch[s], nche = nche[m], nch = nch, 
                       tau = tau, 
                       a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                       delta_threshold = 100,
                       method="Empirical Bayes", theta=theta, eta=eta,
                       datamat = datamat_Power, w = nche[m]/nch, 
                       nsim =nsim, seed = 2000)
    
    end.time <- proc.time()
    TypeITime_DPP2[s,m] <- (end.time - start.time)[3]
    
    Power_DPP2[s,m] = Power$power
    
    
    
  }
}
```


## Dynamic Power Prior w = a * wd = nche / nch * wd

```{r}

nm = length(nche)

dn = list(paste("Setting",1:ns,sep=" "),nche)
dn3 = dn
dn3[[3]] = 1:nsim

TypeI_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
Power_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
DynamicBoundary_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
pc.mi_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
pc.sdi_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
phat_pt_larger_pc_DPP3 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)

wmean_DPP3 = array(NA,dim=c(ns,nm,nsim),dimnames = dn3)

TypeITime_DPP3 = matrix(NA,ns,nm,dimnames=dn) 
PowerTime_DPP3 = matrix(NA,ns,nm,dimnames=dn) 

for(s in 1:ns){
  
  for(m in 1:nm){
    
    set.seed(2000)
    
    # Calibrate
    cs = setting_for_calibration[s]
    
    datamat_typeI = cbind(xt_typeI_list[[cs]],xc_list[[cs]]) 
    
    tau = calibration (nt=nt[cs], pc.calib=pc[cs], nc=nc[cs], 
                       pch=pch[cs], nche=nche[cs], nch=nch,
                       alpha = alpha, 
                       a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                       delta_threshold=100, 
                       method="Empirical Bayes", theta=theta, eta=eta,
                       datamat = datamat_typeI,  
                       nsim = nsim, seed=2000)
    
    # Get data for type I and power
    
    xt = xt_list[[s]]
    xt_typeI = xt_typeI_list[[s]] 
    xc =  xc_list[[s]]
    
    # Type I error
    
    datamat_typeI = cbind(xt_typeI,xc) 
    datamat_Power = cbind(xt,xc) 
    
    start.time <- proc.time()
    TypeI = power.DPP(pt = pc[s], nt = nt[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch[s], nche = nche[m], nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = 100,
                      method="Empirical Bayes", theta=theta, eta=eta,
                      datamat = datamat_typeI, 
                      nsim =nsim, seed = 2000)
    
    
    end.time <- proc.time()
    TypeITime_DPP3[s,m] <- (end.time - start.time)[3]
    
    TypeI_DPP3[s,m] = TypeI$power
    
    DynamicBoundary_DPP3[s,m] = TypeI$delta.bound
    pc.mi_DPP3[s,m] = TypeI$pc.PMD
    pc.sdi_DPP3[s,m] = TypeI$pc.sd.PMD
    phat_pt_larger_pc_DPP3[s,m,] = TypeI$phat_pt_larger_pc_all
    
    wmean_DPP3[s,m,] = TypeI$w
    
    # Power
    set.seed(2000)
    
    start.time <- proc.time()
    Power = power.DPP(pt = pt[s], nt = nc[s], 
                      pc = pc[s], nc = nc[s], pc.calib=pc.c[s],
                      pch = pch[s], nche = nche[m], nch = nch, 
                      tau = tau, 
                      a0c=a0c, b0c=b0c, a0t=a0t, b0t=b0t,
                      delta_threshold = 100,
                      method="Empirical Bayes", theta=theta, eta=eta,
                      datamat = datamat_Power, 
                      nsim =nsim, seed = 2000)
    
    
    end.time <- proc.time()
    TypeITime_DPP3[s,m] <- (end.time - start.time)[3]
    
    Power_DPP3[s,m] = Power$power
    
    
    
  }
}

save.image("DPPDone.RData")
```


## Table , Type I error + Power+ mean PMD + SD PMD
```{r}
cns = c("$p_{c}$", 
        nche,
        nche,
        nche)

Table_TypeI = matrix(NA,ns,length(cns))
colnames(Table_TypeI) = cns

Table_TypeI[,1] = pc
Table_TypeI[,1+1:4] = TypeI_DPP1
Table_TypeI[,5+1:4] = TypeI_DPP3
Table_TypeI[,9+1:4] = TypeI_DPP2

Table_Power = matrix(NA,ns,length(cns))
colnames(Table_Power) = cns

Table_Power[,1] = pc
Table_Power[,1+1:4] = Power_DPP1
Table_Power[,5+1:4] = Power_DPP3
Table_Power[,9+1:4] = Power_DPP2


Table_mi = matrix(NA,ns,length(cns))
colnames(Table_mi) = cns

Table_mi[,1] = pc
Table_mi[,1+1:4] = pc.mi_DPP1
Table_mi[,5+1:4] = pc.mi_DPP3
Table_mi[,9+1:4] = pc.mi_DPP2

Table_sdi = matrix(NA,ns,length(cns))
colnames(Table_sdi) = cns

Table_sdi[,1] = pc
Table_sdi[,1+1:4] = (pc.sdi_DPP1)
Table_sdi[,5+1:4] = (pc.sdi_DPP3)
Table_sdi[,9+1:4] = (pc.sdi_DPP2)


Table2 = rbind(Table_TypeI,Table_Power,Table_mi,Table_sdi)

#write.table(Table2,"Table2.csv",sep=",",row=F,col=T)

kable(Table2)

print(xtable(Table2,digits=3), include.rownames = FALSE, include.colnames = TRUE, sanitize.text.function = I, hline.after = c(0,nrow(Table_mi),nrow(Table_mi)*2,nrow(Table_mi)*3,nrow(Table2)))

```



# SAMprior and rMAP


## Generate Data

```{r}
set.seed(123)
NCore = 7
#get_OC.betaMix

nc = c(45,45,45,45,45,45,45)
nt = nc
pc = c(0.15,0.2,0.25,0.30,0.35,0.4,0.45)
pt = pc + 0.2

# nc.c = nc
# nt.c = nc.c
# pc.c = c(0.3,0.3,0.3)+c(0,0,0)

nche = c(180,180,180,180,180,180,180)
pch=0.30
nch=180
ns = length(nc)

nsim = 100000#
Simulation.Calibration.Match = c(1,1,1,1,1,1,1)

# calibartion for SAM 
nc.sam.c = c(45)
nt.sam.c = nc.sam.c
pc.sam.c = c(0.3)
nche.c = c(180)

ncal = length(nc.sam.c)

theta = 0.5
eta = 1
a0c = 0.001
b0c = 0.001
a0t = 0.001
b0t = 0.001
delta_threshold = 0.1
alpha = 0.1

xt_list = list()
xt_typeI_list = list()
xc_list = list()



for(s in 1:ns){
  set.seed(2000)
  
  xt_typeI_list[[s]] = rbinom(nsim,size = nt[s],prob = pc[s])
  xc_list[[s]] = rbinom(nsim,size = nc[s], prob = pc[s]) 
  xt_list[[s]] = rbinom(nsim,size = nt[s],prob = pt[s])
  
}

xt_typeI_cal_list = list()
xc_cal_list = list()

# Data for calibration
for(s in 1:ncal){
  set.seed(2000)
  
  xt_typeI_cal_list[[s]] = rbinom(nsim,size = nt.sam.c[s], prob = pc.sam.c[s])
  xc_cal_list[[s]] = rbinom(nsim,size = nc.sam.c[s], prob = pc.sam.c[s]) 
  
  
}

```


### Calibaration

```{r}

## Calibaration
cl <- makeCluster(NCore)
registerDoParallel(cl)


cal.tau = foreach(s = 1:length(nc.sam.c),.packages = c("BayesianHybridDesign","RBesT"),.combine = rbind) %dopar%{ 
  
  set.seed(2000)
  HistData = data.frame(study = "TL01",r = round(nche.c[s]*pch),n = nche.c[s])
  
  set.seed(2000)
  map_HistData <- gMAP(cbind(r, n-r) ~ 1 | study,
                       family = binomial,
                       data = HistData, 
                       tau.dist = "HalfNormal", 
                       tau.prior = 1,
                       beta.prior = 2)
  
  map_automix <- automixfit(map_HistData)
  map_automix
  
  xt = xt_typeI_cal_list[[s]]
  xc = xc_cal_list[[s]]
  
  
  datamat_cal = cbind(xt,xc) 
  
  
  set.seed(2000)
  start.time <- proc.time()
  CalRes <- runSAM(if.prior = map_automix,       ## MAP prior from historical data
                   nf.prior = mixbeta(c(1,0.001,0.001)), #mixbeta(c(1,1,1)), ## Non-informative prior for treatment arm
                   delta    = delta_threshold,               ## CSD for SAM prior
                   ## Method to determine the mixture weight for the SAM prior
                   method.w = 'LRT',             
                   n   = nc.sam.c[s], ## Sample size for control and treatment arms
                   n.t = nt.sam.c[s], ## Sample size for treatment and treatment arms
                   ## Decisions
                   decision = decision2S(0.90, 0, lower.tail=FALSE), 
                   ntrial   = nsim,              ## Number of trials simulated
                   if.MAP   = TRUE,              ## Output robust MAP prior for comparison
                   weight   = 0.5,               ## Weight for robust MAP prior
                   ## Response rates for control and treatment arms
                   theta    = pc.sam.c[s],
                   theta.t  = pc.sam.c[s],
                   datamat = datamat_cal,
                   dist = TRUE
  )
  sam.tau = quantile(CalRes$res_SAM_dist,0.9,type = 3)
  rMAP.tau =  quantile(CalRes$res_rMAP_dist,0.9,type = 3)
  non.tau = quantile(CalRes$res_Non_dist,0.9,type = 3)
  sam.tau = matrix(c(sam.tau,rMAP.tau,non.tau),nrow=1)
}
stopCluster(cl)


sam.tau = cal.tau[Simulation.Calibration.Match,1]
rMAP.tau = cal.tau[Simulation.Calibration.Match,2]
non.tau = cal.tau[Simulation.Calibration.Match,3]
```

### Simulation after calibration

```{r}

SAMMethodNames = c("SAMprior",
                   "MAP",
                   "Noninfo")
nm = length(SAMMethodNames)

dn = list(paste("Setting",1:ns,sep=" "),SAMMethodNames)
dn3 = dn
dn3[[3]] = 1:nsim

TypeI_SAM = matrix(NA,ns,nm,dimnames=dn) 
Power_SAM = matrix(NA,ns,nm,dimnames=dn) 
pc.mi_SAM = matrix(NA,ns,nm,dimnames=dn) 
pc.sdi_SAM = matrix(NA,ns,nm,dimnames=dn) 
mean_SAM = array(NA,dim=c(ns,nm,nsim),dimnames = dn3) 


TypeITime_SAM = numeric(ns)
PowerTime_SAM = numeric(ns)
TypeITime_MAP = numeric(ns)


cl <- makeCluster(NCore)
registerDoParallel(cl)

ResultSAM = foreach(s = 1:ns,.packages = c("BayesianHybridDesign","RBesT")) %dopar%{ 
  
  TypeI_SAM1 = rep(NA,nm) 
  Power_SAM1 = rep(NA,nm) 
  pc.mi_SAM1 =rep(NA,nm)  
  pc.sdi_SAM1 = rep(NA,nm) 
  mean_SAM1 = array(NA,dim=c(nm,nsim)) 
  
  set.seed(2000)
  
  HistData = data.frame(study = "TL01",r = round(nche[s]*pch),n = nche[s])
  
  set.seed(2000)
  map_HistData <- gMAP(cbind(r, n-r) ~ 1 | study,
                       family = binomial,
                       data = HistData, 
                       tau.dist = "HalfNormal", 
                       tau.prior = 1,
                       beta.prior = 2)
  
  map_automix <- automixfit(map_HistData)
  map_automix
  
  xt = xt_list[[s]]
  xt_typeI = xt_typeI_list[[s]] 
  xc =  xc_list[[s]]
  
  
  datamat_typeI = cbind(xt_typeI,xc) 
  datamat_Power = cbind(xt,xc) 
  
  set.seed(2000)
  
  start.time <- proc.time()
  TypeI <- runSAM(if.prior = map_automix,       ## MAP prior from historical data
                  nf.prior = mixbeta(c(1,0.001,0.001)), #mixbeta(c(1,1,1)), ## Non-informative prior for treatment arm
                  delta    = delta_threshold,               ## CSD for SAM prior
                  ## Method to determine the mixture weight for the SAM prior
                  method.w = 'LRT',
                  n   = nc[s], ## Sample size for control and treatment arms
                  n.t = nt[s], ## Sample size for treatment and treatment arms
                  ## Decisions
                  decision = decision2S(0.90, 0, lower.tail=FALSE),
                  ntrial   = nsim,              ## Number of trials simulated
                  if.MAP   = FALSE,              ## Output robust MAP prior for comparison
                  weight   = 0.5,               ## Weight for robust MAP prior
                  ## Response rates for control and treatment arms
                  theta    = pc[s],
                  theta.t  = pc[s],
                  datamat = datamat_typeI
  )
  end.time <- proc.time()
  TypeITime_SAM1 = (end.time - start.time)[3]
  
  set.seed(2000)
  
  start.time <- proc.time()
  TypeI <- runSAM(if.prior = map_automix,       ## MAP prior from historical data
                  nf.prior = mixbeta(c(1,0.001,0.001)), #mixbeta(c(1,1,1)), ## Non-informative prior for treatment arm
                  delta    = delta_threshold,               ## CSD for SAM prior
                  ## Method to determine the mixture weight for the SAM prior
                  method.w = 'LRT',             
                  n   = nc[s], ## Sample size for control and treatment arms
                  n.t = nt[s], ## Sample size for treatment and treatment arms
                  ## Decisions
                  decision = decision2S(0.90, 0, lower.tail=FALSE), 
                  ntrial   = nsim,              ## Number of trials simulated
                  if.MAP   = TRUE,              ## Output robust MAP prior for comparison
                  weight   = 0.5,               ## Weight for robust MAP prior
                  ## Response rates for control and treatment arms
                  theta    = pc[s],
                  theta.t  = pc[s],
                  datamat = datamat_typeI
  )
  
  
  end.time <- proc.time()
  TypeITime_Both = (end.time - start.time)[3]
  
  TypeITime_MAP1 = TypeITime_Both - TypeITime_SAM1
  
  TypeI_SAM1[1] = mean(TypeI$res_SAM_dist > sam.tau[s]) 
  TypeI_SAM1[2] = mean(TypeI$res_rMAP_dist > rMAP.tau[s]) 
  TypeI_SAM1[3] = mean(TypeI$res_Non_dist > non.tau[s]) 
  
  for(i in 1:nsim){
    mean_SAM1[1,i] = summary(TypeI$post_theta_c_SAM_list[[i]])["mean"]
    mean_SAM1[2,i] = summary(TypeI$post_theta_c_MAP_list[[i]])["mean"]
    mean_SAM1[3,i] = summary(TypeI$post_theta_c_list[[i]])["mean"]
  }
  
  
  pc.mi_SAM1[1] = mean(mean_SAM1[1,] - mean_SAM1[3,])
  pc.mi_SAM1[2] = mean(mean_SAM1[2,] - mean_SAM1[3,])
  pc.mi_SAM1[3] = mean(mean_SAM1[3,] - mean_SAM1[3,])
  
  pc.sdi_SAM1[1] = sd(mean_SAM1[1,] - mean_SAM1[3,])
  pc.sdi_SAM1[2] = sd(mean_SAM1[2,] - mean_SAM1[3,])
  pc.sdi_SAM1[3] = sd(mean_SAM1[3,] - mean_SAM1[3,])
  
  
  
  start.time <- proc.time()
  Power <- runSAM(if.prior = map_automix,       ## MAP prior from historical data
                  nf.prior = mixbeta(c(1,0.001,0.001)), #mixbeta(c(1,1,1)), ## Non-informative prior for treatment arm
                  delta    = delta_threshold,               ## CSD for SAM prior
                  ## Method to determine the mixture weight for the SAM prior
                  method.w = 'LRT',             
                  n   = nc[s], ## Sample size for control and treatment arms
                  n.t = nt[s], ## Sample size for treatment and treatment arms
                  ## Decisions
                  decision = decision2S(0.90, 0, lower.tail=FALSE), 
                  ntrial   = nsim,              ## Number of trials simulated
                  if.MAP   = TRUE,              ## Output robust MAP prior for comparison
                  weight   = 0.5,               ## Weight for robust MAP prior
                  ## Response rates for control and treatment arms
                  theta    = pc[s],
                  theta.t  = pt[s],
                  datamat = datamat_Power
  )
  end.time <- proc.time()
  PowerTime_SAM1 = (end.time - start.time)[3]
  
  Power_SAM1[1] = mean(Power$res_SAM_dist > sam.tau[s]) 
  Power_SAM1[2] = mean(Power$res_rMAP_dist > rMAP.tau[s]) 
  Power_SAM1[3] = mean(Power$res_Non_dist > non.tau[s])
  
  list(TypeITime_SAM1 = TypeITime_SAM1,
       PowerTime_SAM1 = PowerTime_SAM1,
       TypeI_SAM1 = TypeI_SAM1, 
       Power_SAM1 = Power_SAM1, 
       pc.mi_SAM1 = pc.mi_SAM1, 
       pc.sdi_SAM1 = pc.sdi_SAM1, 
       mean_SAM1 = mean_SAM1,
       TypeITime_MAP = TypeITime_MAP1
  )
  
}

stopCluster(cl)
save.image("SAMDone.RData")
```





### Summarize the results for SAMprior and MAP prior
```{r}



for(s in 1:ns){
  
  TypeITime_SAM[s] = ResultSAM[[s]]$TypeITime_SAM
  TypeITime_MAP[s] = ResultSAM[[s]]$TypeITime_MAP
  PowerTime_SAM[s] = ResultSAM[[s]]$PowerTime_SAM
  TypeI_SAM[s,] = ResultSAM[[s]]$TypeI_SAM1 
  Power_SAM[s,] = ResultSAM[[s]]$Power_SAM1 
  pc.mi_SAM[s,] =ResultSAM[[s]]$pc.mi_SAM1 
  pc.sdi_SAM[s,] = ResultSAM[[s]]$pc.sdi_SAM1 
  mean_SAM[s,,] =ResultSAM[[s]]$mean_SAM1
  
}

```


```{r}
cns = c("Study Treatment Sample Size", 
        "Study Control Sample Size", 
        "Historical Control Sample Size",
        "Study Treatment ORR",
        "Study Control ORR",
        "Historical Control ORR",
        SAMMethodNames[1:2])

Table_TypeI = matrix(NA,ns,length(cns))
colnames(Table_TypeI) = cns

Table_TypeI[,1] = nt
Table_TypeI[,2] = nc
Table_TypeI[,3] = nche
Table_TypeI[,4] = pt
Table_TypeI[,5] = pc
Table_TypeI[,6] = pch
Table_TypeI[,6+1:2] = TypeI_SAM[,1:2]


Table_Power = matrix(NA,ns,length(cns))
colnames(Table_Power) = cns

Table_Power[,1] = nt
Table_Power[,2] = nc
Table_Power[,3] = nche
Table_Power[,4] = pt
Table_Power[,5] = pc
Table_Power[,6] = pch
Table_Power[,6+1:2] = Power_SAM[,1:2]


Table_mi = matrix(NA,ns,length(cns))
colnames(Table_mi) = cns

Table_mi[,1] = nt
Table_mi[,2] = nc
Table_mi[,3] = nche
Table_mi[,4] = pt
Table_mi[,5] = pc
Table_mi[,6] = pch
Table_mi[,6+1:2] = pc.mi_SAM[,1:2]


Table_sdi = matrix(NA,ns,length(cns))
colnames(Table_sdi) = cns

Table_sdi[,1] = nt
Table_sdi[,2] = nc
Table_sdi[,3] = nche
Table_sdi[,4] = pt
Table_sdi[,5] = pc
Table_sdi[,6] = pch
Table_sdi[,6+1:2] = pc.sdi_SAM[,1:2]

Table3 = rbind(Table_TypeI,Table_Power,Table_mi,Table_sdi)

#write.table(Table3,"SAM.csv",sep=",",row=F,col=T)

kable(Table3)

```

# Generate Table 2 in the manuscript

```{r}
Table = cbind(Table2,Table1[,c("Bayesian p", "Generalized BC","JSD")],Table3[,c("SAMprior", "MAP")])
```



## To latex tables

```{r}
require(xtable)

print(xtable(Table,digits=3), include.rownames = FALSE, include.colnames = TRUE, sanitize.text.function = I, hline.after = c(0,nrow(Table_TypeI),nrow(Table_TypeI)*2,nrow(Table_TypeI)*3,nrow(Table_TypeI)*4))


```




## Table, Computational Time 
```{r}
cns = c("Study Treatment Sample Size", 
        "Study Control Sample Size", 
        "Historical Control Sample Size",
        "Study Treatment ORR",
        "Study Control ORR",
        "Historical Control ORR",
        DPPMethodNames,
        "SAM","MAP")

Table_Time = matrix(NA,ns,length(cns))
colnames(Table_Time) = cns

Table_Time[,1] = nt
Table_Time[,2] = nc
Table_Time[,3] = nche
Table_Time[,4] = pt
Table_Time[,5] = pc
Table_Time[,6] = pch
Table_Time[,6+1:length(DPPMethodNames)] = TypeITime_DPP
Table_Time[,6+length(DPPMethodNames)+1] = TypeITime_SAM
Table_Time[,6+length(DPPMethodNames)+2] = TypeITime_MAP
#write.table(Table_Time,"Time.csv",sep=",",row=F,col=T)

kable(Table_Time)

```
```{r}
save.image(paste("Table2.RData",sep=""))
```

```{r}
sessionInfo()
```



