---
title: "ojsr-vignette"
author: "Gaston Becerra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ojsr-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(knitr)
```

# Overview

ojsr allows you to crawl OJS archives, issues, articles, galleys, and search results, and retrieve metadata from articles.

**Important Notes:**

- **ojsr functions rely on OJS routing conventions**, therefore, OJS installations using custom routing conventions will most probably fail (see <https://docs.pkp.sfu.ca/dev/documentation/en/architecture-routes>);
- **ojsr does not scrape the URL you provide**. ojsr composes the required URL for different scrapping scenarios (e.g., if you are scrapping articles from an issue, it will parse and browse the issue's ToC).

## About OJS

(from the OJS documentation, as of Jan.2020)

Open Journal Systems (OJS) is a journal management and publishing system that has been developed by the Public Knowledge Project through its federally funded efforts to expand and improve access to research.

OJS assists with every stage of the refereed publishing process, from submissions through to online publication and indexing. Through its management systems, its finely grained indexing of research, and the context it provides for research, OJS seeks to improve both the scholarly and public quality of refereed research.

OJS is open source software made freely available to journals worldwide for the purpose of making open access publishing a viable option for more journals, as open access can increase a journal’s readership as well as its contribution to the public good on a global scale (see PKP Publications).

## OJS API

Since OJS v3.1+ <https://docs.pkp.sfu.ca/dev/api/ojs/3.1> a Rest API is provided. We are positive a better R interface should use that API instead of web scraping. So, why ojsr? According to <https://pkp.sfu.ca/software/ojs/usage-data//>, as of 2019 (when v3.1+ was launched), at least 15,000 journals worldwide were using OJS. OJS is an excellent free publishing solution for institutions that could probably not publish otherwise, and, presumably, cannot afford to update constantly. ojsr aims to help crawling and retrieving info from OJS during this legacy period.

# Example

## Scraping complete journals

Let’s say we want to scrape metadata from a journal's issue collection of journals to compare them. 
We start with the journal's titles and URL, and can use ojsr to scrape their issues, articles, and metadata.

```{r fullexample , eval = FALSE }
# NOT RUN {

library(ojsr)

journal <- 'https://revistapsicologia.uchile.cl/index.php/RDP/'

issues <- ojsr::get_issues_from_archive(input_url = journal) 

articles <- ojsr::get_articles_from_issue(input_url = issues$output_url[1:2]) # only first 2 issues

metadata <- ojsr::get_html_meta_from_article(input_url = articles$output_url[1:5]) # only first 5 articles

# }

```

# Function reference

## get_issues_from_archive: Scraping issues’ URLs from the OJS issues archive

`get_issues_from_archive()` takes a vector of OJS URLs and scrapes the issues URLs from the *issue archive* 

You don't need to provide the actual URL to issue archives. `get_issues_from_archive()` parses the URL you provide to compose it. 
Then, it looks for links containing "/issue/view" in the href. Links are post-processed to comply to OJS routing conventions before returning.

```{r get_issues_from_archive , eval = FALSE }
journal <- 'https://revistapsicologia.uchile.cl/index.php/RDP/'

issues <- ojsr::get_issues_from_archive(input_url = journal)

```

The result is a long-format data frame (1 input_url may result in several rows, one for each output_url) containing:

1. *input_url* - the URL you provided
2. *output_url* - the issues' URL that has been scraped

## get_articles_from_issue: Scraping articles URLs from the ToC of OJS issues

`get_articles_from_issue()` takes a vector of OJS (issue) URLs and scrapes the links to articles from the issues table of content 

You don't need to provide the actual URL of the issues' ToC, but you must provide URLs that include issue ID (articles URLs do not include this info!). `get_articles_from_issue()` parses the URL you provide to compose the ToC URL. Then, it looks for links containing "/article/view" in the href. Links are post-processed to comply to OJS routing conventions before returning.

```{r get_articles_from_issue , eval = FALSE }
issue <- 'https://revistapsicologia.uchile.cl/index.php/RDP/issue/view/6031/'

articles <- ojsr::get_articles_from_issue(input_url = issue) 

```

The result is a long-format dataframe (1 input_url may result in several rows, one for each output_url), containing:

1. *input_url* - the URL you provided
2. *output_url* - the articles URL that has been scraped

## get_articles_from_search: Scraping OJS search results for a given criteria to retrieve articles’ URL

`get_articles_from_search()` takes a vector of OJS URLs and a string for search criteria to compose search result URLs, then it scrapes them to retrieve the articles’ URLs.

You don't need to provide the actual URL of the search result pages. `get_articles_from_search()` parses the URL you provide to compose the search result page(s) URL. If pagination is involved, necessary links are also included. Then, it looks for links containing "/article/view" in the href. Links are post-processed to comply to OJS routing conventions before returning.

```{r get_articles_from_search , eval = FALSE }
journal <- 'https://revistapsicologia.uchile.cl/index.php/RDP/'

criteria <- "psicologia+social"

articles_search <- ojsr::get_articles_from_search(input_url = journal, search_criteria = criteria)

```

The result is a long-format dataframe (1 input_url may result in several rows, one for each output_url), containing:

1. *input_url* - the URL  you provided
2. *output_url* - the article URL

## get_galleys_from_article: Scraping galleys URLs from OJS articles

*Galleys* are the final presentation version of the articles content. Most of the time, these include full content in PDF and other reading formats. Less often, they are supplementary files (tables, dataset) in different formats. 

`get_galleys_from_article()` takes a vector of OJS URLs and scrapes all the galleys URLs from the *article view* 

You may provide any article-level URL (article abstract view, inline view, PDF direct download, etc.). `get_galleys_from_article()` parses the URL you provide to compose the article view URL. Then, it looks for links containing "/article/view" in the href. Links are post-processed to comply to OJS routing conventions before returning (i.e., having a galley ID).

```{r get_galleys_from_article , eval = FALSE }
article <- 'https://dspace.palermo.edu/ojs/index.php/psicodebate/article/view/516/311' # inline reader

galleys <- ojsr::get_galleys_from_article(input_url = article) 

```

The result is a long-format dataframe (1 input_url may result in several rows, one for each output_url), containing:

1. *input_url* - the URL you provided
2. *output_url* - the galleys URL that has been scraped
3. *format* - the format of the galley (e.g., PDF, XML)
4. *download_url* - the conventional URL to force galley download. You may pass these to a download function of your own (e.g., <https://stackoverflow.com/questions/39246739/download-multiple-files-using-download-file-function>).

## get_html_meta_from_article: Scraping metadata from the OJS articles HTML

`get_html_meta_from_article()` takes a vector of OJS URLs and scrapes all metadata written in HTML from the *article view* (e.g., <https://publicaciones.sociales.uba.ar/index.php/psicologiasocial/article/view/593>).

You may provide any article-level URL (article abstract view, inline view, PDF direct download, etc.). `get_html_meta_from_article()` parses the URL you provide to compose the URL of the article view. Then, it looks for \<meta\> tags in the \<head\> section of the HTML. **Important!** This may not only retrieve bibliographic metadata; any other "meta" property detailed on the HTML will be obtained (e.g., descriptions for propagation on social network, etc.).

```{r get_html_meta_from_article , eval = FALSE }
article <- 'https://revistapsicologia.uchile.cl/index.php/RDP/article/view/75178'

metadata <- ojsr::get_html_meta_from_article(input_url = article) 

```

The result is a long-format dataframe (1 input_url may result in several rows, one for each output_url), containing:

1. *input_url* - the URL you provided
2. *meta_data_name* - name of the property/metadata (e.g., "DC.Date.created" for the Date of creation) 
3. *meta_data_content* - the actual metatag value 
4. *meta_data_scheme* - the standard in which the content is annotated
5. *meta_data_xmllang* - the language in which the metadata was entered

## get_oai_meta_from_article: Retrieving OAI records for OJS articles

An alternative to web scraping metadata from the article pages HTML is to retrieve their OAI-PMH (Open Archives Initiative Protocol for 'Metadata' Harvesting) records.

`get_oai_meta_from_article()` will try to access the OAI records within the OJS for any article's URL you have provided.

```{r get_oai_meta_from_article , eval = FALSE }
article <- 'https://dspace.palermo.edu/ojs/index.php/psicodebate/article/view/516/311' # xml galley

metadata_oai <- ojsr::get_oai_meta_from_article(input_url = article)

```

The result is a long-format dataframe (1 input_url may result in several rows, one for each output_url), containing:

1. *input_url* - the URL you provided
2. *meta_data_name* - name of the property/metadata (e.g., "DC.Date.created" for the Date of creation) 
3. *meta_data_content* - the actual metatag value
4. *meta_data_scheme* - **it always returns NA** (included just for easier binding with `get_html_meta_from_article()` results)
5. *meta_data_xmllang* - **it always returns NA** (included just for easier binding with `get_html_meta_from_article()` results)

**Note: This function is in a very preliminary stage**. If you are interested in working with OAI records, you may want to check Scott Chamberlain's **OAI package for R** <https://CRAN.R-project.org/package=oai>. If you only have the OJS home url, and would like to check all the article's OAI records at one shot, an interesting option is to parse it with `ojsr::parse_oai_url()` and passing the output_url to `oai::list_identifiers()`.

## parse_base_url: Parsing URLs against OJS routing conventions to retrieve the base URL

`parse_base_url()` takes a vector of OJS URLs and retrieves their base URL, according to OJS routing conventions.

```{r parse_base_url , eval = FALSE }
mix_links <- c(
   'https://dspace.palermo.edu/ojs/index.php/psicodebate/issue/archive',
   'https://revistapsicologia.uchile.cl/index.php/RDP/article/view/75178'
)

base_url <- ojsr::parse_base_url(input_url = mix_links)

```

The result is a vector of the same length of your input.

## parse_oai_url: Parsing URLs against OJS routing conventions to retrieve the OAI protocol URL

`parse_oai_url()` takes a vector of OJS URLs and retrieves their OAI entry URL, according to OJS routing conventions.

```{r parse_oai_url , eval = FALSE }
mix_links <- c(
   'https://dspace.palermo.edu/ojs/index.php/psicodebate/issue/archive',
   'https://revistapsicologia.uchile.cl/index.php/RDP/article/view/75178'
)

oai_url <- ojsr::parse_oai_url(input_url = mix_links)

```

The result is a vector of the same length of your input.
