---
title: "Classification as typical and deviant"
author: "Ingo Rohlfing"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Classification as typical and deviant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(MMRcaseselection)
library(ggplot2) # for illustration of plot customization
```

The classification (designation, assignment) of cases as typical and deviant needs to be distinguished from the actual choice of cases that are of a specific type. The package offers two ways of classifying cases. First, one can designate cases using the *prediction interval* or the *standard deviation of the residuals*. Both classification techniques require an `lm` object as input. 

The approach using the prediction interval additionally requires the specification of the width of the interval. The % prediction interval represents the range of outcome values within which 95% of future outcome values are expected to fall in repeated samples. Following Rohlfing and Starke ([2013](https://doi.org/10.1111/spsr.12052)), a case is designated as typical if the observed outcome value is inside the interval and as deviant otherwise. 

Here is a simple example using the `mtcars` dataset and a 90% prediction interval (for no specific reason, the default value is 0.95). The output of the function is a dataframe with the observed and fitted outcome values, the upper and lower bound of the prediction interval and the case status based on the fitted value and the prediction interval that is specific to each case.
```{r}
df <- lm(mpg ~ disp + wt, data = mtcars)
pi_df <- predint(df, piwidth = 0.9)
head(pi_df)
```

The designation of cases based on the prediction interval can be visualized with the `predint_plot()` command. The graph plots observed vs. fitted outcome values together with the % prediction interval. The plot is a `gg` object and can be edited with the usual commands of the `ggplot2` package. 

The input into the plotting function has to be a dataframe produced with `predint()`. The color scheme is the [viridis color palette](https://CRAN.R-project.org/package=viridis) that is accustomed for colorblind readers and also works when printed in grey scale. If the viridis palette is not preferred, one can substitute it with any other `scale_color_*` scheme by adding it to the command.
```{r, fig.height = 6, fig.width = 6}
predint_plot(pi_df)
# Using scale_color_brewer() instead of the viridis palette.
predint_plot(pi_df) + scale_color_brewer()
```

An alternative classification technique following Lieberman ([2005](https://doi.org/10.1017/S0003055405051762)) uses the *standard deviation of the residuals*. The command for assigning cases and typical and deviant is `residstd()`. The specification of `stdshare` determines how large the share of the residual standard deviation should be to distinguish typical from deviant cases. The default value is 1 (for no specific reason).
```{r}
resid_df <- residstd(df, stdshare = 1.5)
head(resid_df)
```

To my knowledge, it has not been proposed to plot the case classification resulting from this technique. However, it can be done and is available in this package with the command `residstd_plot()`. The standard deviation of the residual is necessarily the same for all cases and is represented by a symmetric corridor (parallel grey lines) around the bisecting line in a plot of observed vs. fitted outcome values. The plot is a `gg` object that can be accustomed with `ggplot2` commands.
```{r, fig.width = 6, fig.height = 6}
residstd_plot(resid_df)
```

## Packages used in this vignette
-   base (R Core Team 2020)
-   ggplot2 (Wickham 2016)
-   grateful (Rodriguez-Sanchez 2018)

R Core Team. 2020. *R: A Language and Environment for Statistical
Computing*. Vienna, Austria: R Foundation for Statistical Computing.
<https://www.R-project.org/>.

Rodriguez-Sanchez, Francisco. 2018. *grateful: Facilitate Citation of R Packages*.
<https://github.com/Pakillo/grateful>.

Wickham, Hadley. 2016. *ggplot2: Elegant Graphics for Data Analysis*.
Springer-Verlag New York. <https://ggplot2.tidyverse.org>.


