---
title: "Descriptive statistics by group"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive statistics by group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(qacBase)
```

## Overview

Getting summary statistics for a quantitative variable is a very common
task in data analysis. Unfortunately, **R** makes it surprisingly difficult.

The `qstats` function is an attempt to rectify the situation by making it simple to get any number of descriptive statistics for a numeric variable and to break these statistics down by the levels of one or more categorical variables (groups).

The general format is

**`qstats(data, variable, grouping variables, statistics, other options)`**

Note that variable names do not have to be quoted.

## Using default statistics
By default the *sample size*, *mean*, and *standard deviation* are provided.
Let's take a look at fuel efficiencies for 11,914 automobiles
in the `cardata` data frame.

```{r include=TRUE}
# simple summary statistics 
qstats(cardata, highway_mpg)

# summary statistics by vehicle_size
qstats(cardata, highway_mpg, vehicle_size)

# summary statistics by vehicle_size and drive type
qstats(cardata, highway_mpg, vehicle_size, driven_wheels)
```

## Specifying other statistics
You can supply a statistics argument with the "stats" parameter. You can pass a single statistic, or multiple statistics as a vector of names.

```{r include=TRUE}
# single statistic
qstats(cardata, highway_mpg, vehicle_size, stats = "median")

# multiple statistics
qstats(cardata, highway_mpg, vehicle_size, 
       stats = c("median", "min", "max"))
```


User-defined functions can also be used as a statistics. The only requirement
is that the function returns a single number.
```{r include=TRUE}
#custom statistics
p25 <- function(x) quantile(x, probs=.25)
p75 <- function(x) quantile(x, probs=.75)

#calling the built in and custom statistics
qstats(cardata, highway_mpg, vehicle_size, 
       stats = c("min", "p25", "p75", "max"))
```


## Other options

Other options include 

* **na.rm** When TRUE, NAs are removed. Default is TRUE.  
* **digits** The number of decimal points to print. Default = 2.  

```{r include=TRUE}
qstats(cardata, highway_mpg, vehicle_size,  
       stats=c("n", "mean","median","sd"),  
       na.rm=FALSE, digits=2)
```

























