---
title: "Recoding Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recoding Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(knitr)
library(kableExtra)
```

```{r setup, echo=FALSE}
library(qacBase)
df <- data.frame(sex=c(1,2,1,2,2,2),
                
                 race=c("b", "w", "a", "b", "w", "h"),
                 outcome=c("better", "worse", "same", "same", "better", "worse"),
                 Q1=c(20, 30, 44, 15, 50, 99),
                 Q2=c(15, 23, 18, 86, 99, 35),
                 age=c(12, 20, 33, 55, 30, 100),
                 rating =c(1,2,5,3,4,5))
```

The `recodes()` functions makes it very easy to recode one or more variables in the your data frame. The format is </br></br>

**`newdata <- recodes(olddata, variables, from values, to values)`**

### Original dataset

Consider the following data set (below). Lets make the following changes.

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```



#### sex

For `sex`, set 1 to "Male" and 2 to "Female".

```{r}
df <- recodes(data=df, vars="sex", 
               from=c(1,2), to=c("Male", "Female"))
```

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

#### race
Recode `race` to "White" vs. "Other".

```{r}
df <- recodes(data=df, vars="race", 
              from=c("w", "b", "a", "h"), 
              to=c("White", "Other", "Other", "Other"))
```

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

#### outcome
Recode `outcome` to 1 (better) vs. 0 (not better).

```{r}
df <- recodes(data=df, vars="outcome", 
              from=c("better", "same", "worse"), 
              to=c(1, 0, 0))
```

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```


#### Q1 and Q2

For `Q1` and `Q2` set values of 86 and 99 to missing.

```{r}
df <- recodes(data=df, vars=c("Q1", "Q2"), 
              from=c(86, 99), to=NA)
```

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

#### age

For `age`, set values 

* less than 20 or greater than 90 to missing, 
* 20 <= age <= 30 to "Younger",   
* 30 < age <= 50 to "Middle Aged", and   
* 50 < age <= 90 to "Older". 

You can use expressions in your `from` fields. When they are `TRUE`, the corresponding `to` values will be applied. We will use the dollar sign (**\$**) to represent the variable (age in this case). The symbols ( **\|**, **\&** ) mean **OR** and **AND** respectively.

```{r}
df <- recodes(data=df, vars="age", 
              from=c("$ <   20 | $ >  90", 
                     "$ >=  20 & $ <= 30",
                     "$ >   30 & $ <= 50",
                     "$ >   50 & $ <= 90"), 
              to=c(NA, "Younger", "Middle Aged", "Older"))
```

We can also write this as

```{r eval=FALSE}
df <- recodes(data=df, vars="age", 
              from=c("$ < 20", "$ <= 30", "$ <= 50", "$ <= 90", "$ > 90"), 
              to=  c(NA, "Younger", "Middle Aged", "Older", "NA"))
```

This works because once the age value for an observations meets a criteria that is `TRUE` (working left to right), it is recoded. It isn't changed again by later criteria in the same `recodes` statement.

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

#### rating

Finally, for the `rating` variable, reverse the scoring so that 1 to 5 becomes
5 to 1.

```{r}
df <- recodes(data=df, vars="rating", from=1:5, to=5:1)
```

```{r, echo=FALSE}
kbl(df) %>% kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

### Note

Remember that `recodes` returns a data frame, not a variable. 

* `df <- recodes(data=df, vars="rating", from=1:5, to=5:1)` **is correct**.  

* `df$rating <- recodes(data=df, vars="rating", from=1:5, to=5:1)` **is not**.

This allows you to apply the same recoding scheme to more than one variable at a time (e.g., Q1 and Q2 above).

**<span style="color: red;">And that's it (APPLAUSE, APPLAUSE)!</span>**
