---
title: "Spatial phylogentic data"
output: 
   rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{phylospatial-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
fig.dim = c(6, 4)
)
```

## Introduction

This vignette demonstrates the basic structure and creation of a spatial phylogenetic data set, which is the first step of any analysis using this R package. Spatial phylogenetic analyses require two essential ingredients: data on the geographic distributions of a set of organisms, and a phylogeny representing their evolutionary relationships. This package stores these data as objects of class `'phylospatial'`.

The core idea of spatial phylogenetics is that analyses account for every every single "lineage" on the phylogenetic tree, including terminals and larger clades. Each lineage has a geographic range comprising the collective ranges of all terminal(s) in the clade, and it has a single branch segment whose length represents the evolutionary history that is shared by those terminals and only those terminals. When calculating biodiversity metrics, every lineage's occurrence in a site gets weighted by its branch length.

In this vignette, we'll create a lightweight example of a `phylospatial` object, look through its components to understand how it is structured, and then demonstrate some more nuanced use cases with real data. Finally, we'll show how `phylospatial` objects can also be used for traditional non-phylogenetic biodiversity data analyses, in cases when incorporating a phylogeny is impossible or undesirable.


## A minimal example

Let's begin by creating a simple `phylospatial` object. To do this, we use the `phylospatial()` function, which has two key arguments: `tree`, a phylogeny of class `phylo`, and `comm`, a community data set representing the geographic distributions of the terminal taxa (usually species). In the code below, we simulate a random tree with five terminal taxa, and a raster data set with 100 grid cells containing occurrence probabilities for each terminal, with layer names corresponding to species on the tree. (A differentiating feature of the `phylospatial` library is that it supports quantitative data types like probabilities or abundances, in addition to binary community data.) Then we pass them to `phylospatial()`:

```{r setup, message=FALSE, warning=FALSE}
library(phylospatial); library(terra); library(ape); library(sf)

# simulate data
set.seed(1234)
n_taxa <- 5
x <- y <- 10
tree <- rtree(n_taxa)
comm <- rast(array((sin(seq(0, pi*12, length.out = n_taxa * x * y)) + 1)/2, 
                   dim = c(x, y, n_taxa)))
names(comm) <- tree$tip.label

# create phylospatial object
ps <- phylospatial(comm, tree)
ps
```


## Structure of a `phylospatial` object

#### Phylogeny

Our `phylospatial` object is a list with six elements. Let's look at each of these in turn, starting with the `tree`. This is the phylogeny we simulated, a tree of class `phylo` with `r n_taxa` tips and `r n_taxa - 2` larger clades. Note that the branch lengths of the input tree are scaled to sum to 1. We can use `plot()` function to view the tree.

```{r tree, message=FALSE, warning=FALSE}
names(ps)

ps$tree

plot(ps, "tree")
```

#### Community matrix

The other key component is `comm`, which is a `matrix` containing occurrence data. Although we supplied community data as a raster, it's stored here as a matrix, with a row for each grid cell and a column for each taxon. Let's take a look at the matrix. We can also `plot` the community data, which re-casts it as a spatial data set (a raster, in this case).

```{r comm, message=FALSE, warning=FALSE}
head(ps$comm)

plot(ps, "comm")
```

We can see that in addition to our `r n_taxa` terminal taxa, the data set also includes geographic ranges for the `r n_taxa - 2` larger clades. Internally, the `phylospatial()` function constructs ranges for every multi-tip clade on the tree, based on the topology of the tree and the community data for the tips. 

The specific way that these clade ranges are constructed depends on the type of community data being used. The package supports three data types: `"probability"`, `"abundance"`, and `"binary"`. Recall that our data were probabilities; we could have specified that explicitly by setting `data_type = "probability"` when we constructed our phylospatial object, but the function detected this based on the values in our data set, and we can confirm that it did so correctly by checking `ps$data_type`. For probabilities, the default function used to calculate clade occurrence values gives the probability that at least one member of the clade is present in a given site. Abundance and binary data have their own default functions. (You can also override the defaults by supplying your own `clade_fun`---for example, if you had occurrence probabilities that you knew were strongly non-independent among species, you could specify `clade_fun = max`.) The function that was used for a given data set can be accessed at `ps$clade_fun`.

Note that you can also specify your own clade ranges to `phylospatial()` rather than letting it build them for you, by setting `build = FALSE`. You might want to do this if, for example, you have modeled the distributions of every clade in addition to every terminal species in your data set.


#### Spatial data

The `spatial` component is the last key piece of our phylospatial object. (The only other element we haven't mentioned here is `dissim`, which is covered in the vignette on beta diversity.) The spatial component of the object contains spatial reference data on the geographic locations of the communities found in each row of the community matrix.

In this example, the spatial data is a raster layer inherited from the `SpatRaster` data we supplied as our `comm`. You can also supply vector data (points, lines, or polygons) as an `sf` object. If the spatial data is in raster, polygon, or line format, `phylospatial` will check that all features have equal area or length, which is an important assumption underlying various functions in the package.

Also note that spatial data isn't required; community data provided as a matrix works just fine.


## A realistic example

Now let's look at creating a phylospatial data set using real data. To do this, we'll use the example "moss" data set that ships with the package, representing a phylogeny and modeled occurrence probabilities for several hundred species of moss in California. The function `moss()` returns a pre-constructed `phylospatial` object based on these data, but here let's build one from scratch. In the code below we'll load a raster data set with a layer of occurrence probabilities for each species, and a phylogeny representing their evolutionary relationships. We'll then pass these to `phylospatial()`.

```{r moss, message=FALSE, warning=FALSE}
moss_comm <- rast(system.file("extdata", "moss_comm.tif", package = "phylospatial"))
moss_tree <- read.tree(system.file("extdata", "moss_tree.nex", package = "phylospatial"))
ps <- phylospatial(moss_comm, moss_tree)
plot(ps, "comm")
plot(ps, "tree", type = "fan", show.tip.label = FALSE)
```


#### Non-phylogenetic data

While the `phylospatial` library is obviously designed for phylogenetic analyses, it's worth noting that it also supports non-phylogenetic analyses. In cases where a phylogeny is unavailable or where a traditional species-based biodiversity analysis is desired, you can create a data set by calling `phylospatial()` without providing a tree. All major functions in the package will still work, and will assume that the taxa in `comm` are independent and equally weighted.

In fact, traditional species-based methods can actually be considered a specific case of more general phylogenetic methods, in which species are assumed to be connected on a "star" phylogeny with a single polytomy and equal branch lengths. In `phylospatial`, support for non-phylogenetic data is implemented by creating a star phylogeny if no phylogeny is provided by the user. Here's how this looks for the simple community data we created above:

```{r species, message=FALSE, warning=FALSE}
ps <- phylospatial(comm)
plot(ps, "tree", type = "fan")
```


