---
title: "Comparing Samples using hilbertSimilarity"
author: "Yann Abraham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Samples using hilbertSimilarity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
library(knitr)
opts_chunk$set(warning=FALSE,
               fig.width=8,
               fig.height=6,
               fig.retina=1,
               fig.keep='high',
               fig.align='center')
```

# Introduction

Comparing samples defined over a single dimension is a straightforward task that relies on standard, well established methods. Meanwhile distance between samples in high dimensional space remains a largely unexplored field. Available solutions rely on multivariate normal distributions, a condition that is both difficult to check and overlooking key behaviors in biological samples, where populations of interest often correspond to a small proportion (<1%) of all the points that have been measured.

We have developed `hilbertSimilarity` to address the problem of sample similarity in mass cytometry where samples are measured over up to 100 dimensions, and where each sample contains a slightly different number of points (or cells). Our method first transforms each sample into a probability vector, by dividing each dimension into a fixed number of bins and by associating each cell to a specific multidimensional cube. The proportion of cells in each hypercube is characteristic of a given sample. To characterize an compare samples we use measures from Information Theory, since their interpretation in terms of similarity and complexity is straightforward.

To demonstrate the power of `hilbertSimilarity` we applied the method to a subset of the bodenmiller *et al.* dataset, comparing the effect of different stimulations and identifying groups of cells that are significantly affected by different treatments.

Compared to other methods, `hilbertSimilarity` does not rely on expert-driven gating, or require any hypothesis about the number of populations that are present in the data. This makes it a powerful tool to quickly assess a dataset before using traditional methods, or when populations a not known *a priori*.

# Installation

`hilbertSimilarity` can be installed using the following command

```
devtools::install_github(yannabraham/hilbertSimilarity)
```

Once installed the package can be loaded using the standard `library` command.

```{r}
library(hilbertSimilarity)
```

# Loading some test data

We first need data from the `bodenmiller` package:

```{r}
library(bodenmiller)
data(refPhenoMat)
data(untreatedPhenoMat)
data(refFuncMat)
data(untreatedFuncMat)
data(refAnnots)
refAnnots$Treatment <- 'reference'
data(untreatedAnnots)
fullAnnots <- rbind(refAnnots[,names(untreatedAnnots)],
                    untreatedAnnots)
fullAnnots$Treatment <- factor(fullAnnots$Treatment)
fullAnnots$Treatment <- relevel(fullAnnots$Treatment,'reference')
refMat <- cbind(refPhenoMat,refFuncMat)
untreatedMat <- cbind(untreatedPhenoMat,
                      untreatedFuncMat)
fullMat <- rbind(refMat,untreatedMat)
fullMat <- apply(fullMat,2,function(x) {
  qx <- quantile(x,c(0.005,0.995))
  x[x<qx[1]] <- qx[1]
  x[x>qx[2]] <- qx[2]
  return(x)
})
```

In this dataset `r nrow(fullMat)` cells corresponding to `r nlevels(fullAnnots$Treatment)` have been measured over `r ncol(fullMat)` channels. Cells have been gated into `r nlevels(fullAnnots$Cells)` populations. The percentage of each population per treatment is shown below:

```{r,echo=FALSE}
library(dplyr)
library(tidyr)
fullAnnots %>% 
    count(Cells,Treatment) %>% 
    group_by(Treatment) %>% 
    mutate(n=n/sum(n),
           n=round(100*n,2)) %>% 
    spread(Treatment,n)
```

The smallest cell population, `r names(which.min(100*table(fullAnnots$Cells)/nrow(fullAnnots)))`, corresponds to `r round(min(100*table(fullAnnots$Cells)/nrow(fullAnnots)),3)`% of the total cells.

To demonstrate the use of `hilbertSimilarity` we will compare analyze the data at the treatment level.

# Defining a grid to compare samples

The `make.cut` function is used to prepare the grid that will be used to process the data matrix. It requires 2 arguments: the number of cuts (or bins), and the minimum number of cells to consider for a density. The latter depends on the technology and on the decision by the scientist running the analysis. For CyTOF we will require a population to correspond to at least 40 cells.

## Choosing a number of bins

The number of bins in each dimension is defined as follows:

$$c^{d} < N$$

Where $c$ is the number of bins, $d$ is the number of dimensions and $N$ is the total number of cells in the dataset. We chose the first integer $c$ that would generate at least 1 bin per cell. Because many combinations don't have any biological sense, the fraction of occupied bins will be lower than 1.

$c$ can be computed easily using the following formula

$$c = \max \left \{ \left \lfloor \sqrt[d]{N} \right \rfloor , 2 \right \}$$

The formula has been implemented in the `hilbert.order` function. For our example, where $N$ is `r nrow(fullMat)` and $d$ is `r ncol(fullMat)`, $c$ is `r hilbert.order(fullMat)`. The number of cuts to generate is the number of required bins plus 1.

After manual inspection, we used a $c$ of 3 to compute the cuts:

```{r}
nbins <- 2
cuts <- make.cut(fullMat,
                 n=nbins+1,
                 count.lim=40)
```

## Reviewing the grid

The `make.cut` function returns 2 cuts:

 - `fixed` corresponds to equally sized bins over the range of the channel
 - `combined` adjusts the bin size to the density of the channel

The cuts can be visualized using the `show.cut` function:

```{r,fig.height=8,fig.width=8}
show.cut(cuts)
```

The green lines correspond to the `fixed` limits, the red lines correspond to the adjusted limits (when applicable). For cases like CD3, pSlp76 and ZAP70, it allows for a better separation between the positive and negative populations.

# Applying the grid to the data

Given a dataset and a grid, the `do.cut` function will assign each cell to a particular bin in every dimension.

```{r}
cutFullMat <- do.cut(fullMat,cuts,type='combined')
```

Effectively, each cell is now associated to a voxel, and each voxel is enriched for a particular cell type that corresponds to the the unique combination of dimensions and ranges it corresponds to.

To uniquely identify each occupied voxel, one can compute a Hilbert index over the grid. The  after the dimensions have been ordered to better capture the 

## Ordering the dimensions

Intuitively, each voxel on the grid contains a specific cell type. In order for populations to be associated with **consecutive** specific bins, one must optimize the order of channels so that dimensions that are specific to a population are grouped together in the input matrix.

A simple way to achieve this is to use the normalized mutual information between the dimensions to cluster them into meaningful groups:

```{r}
library(entropy)
miFullMat <- matrix(0,nrow=ncol(fullMat),ncol = ncol(fullMat) )
for (i in seq(ncol(fullMat)-1)) {
  for (j in seq(i+1,ncol(fullMat))) {
    cur.tbl <- table(cutFullMat[,i],cutFullMat[,j])
    nent <- 1-mi.empirical(cur.tbl)/entropy.empirical(cur.tbl)
    miFullMat[i,j] <- miFullMat[j,i] <- nent
  }
}
dimnames(miFullMat) <- list(colnames(fullMat),colnames(fullMat))
hcFullMat <- hclust(as.dist(miFullMat))
plot(hcFullMat)
```

## Calculating the Hilbert index

After ordering the cut matrix using the normalized mutual information, it can be transformed into a Hilbert index using the `do.hilbert` function:

```{r}
col.order <- hcFullMat$labels[rev(hcFullMat$order)]
hc <- do.hilbert(cutFullMat[,col.order],nbins)
```

All cells are found in `r length(table(hc))` voxels out of the `r nbins^ncol(fullMat)` possible voxels defined over the initial `r ncol(fullMat)`-dimensional space (`r round(100*length(table(hc))/nbins^ncol(fullMat),2)`%).

# Visualizing the Hilbert curve
## Using the Andrews curve to visualize cell densities

Using a Hilbert index to describe the grid has the advantage that consecutive index correspond to consecutive voxels in the grid. Effectively it corresponds to a projection from *N* dimensions to a single one.

To visualize the Hilbert curve we use [Andrews plots](https://en.wikipedia.org/wiki/Andrews_plot) to standardize the display to a common range. First we compute the number of cells per Hilbert index, per treatment:

```{r}
treatment.table <- with(fullAnnots,
                        table(hc,Treatment))
treatment.pc <- apply(treatment.table,2,function(x) x/sum(x))
```

Next we prepare the Andrews vector; adjust the number of breaks to return a smoother curve :

```{r}
av <- andrewsProjection(t(treatment.pc),breaks=30)
```

Then we project the Hilbert curve of each treatment onto the Andrews vector:

```{r}
treatment.proj <- t(treatment.pc) %*% t(av$freq)
```

Now we can visualize the different treatment using line charts:

```{r}
library(ggplot2)
library(reshape2)
melt(treatment.proj) %>% 
    rename(AndrewsIndex=Var2) %>% 
    mutate(AndrewsIndex=av$i[AndrewsIndex]) %>% 
    ggplot(aes(x=AndrewsIndex,y=value))+
    geom_line(aes(group=Treatment,color=Treatment))+
    theme_light(base_size=16)+
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())
```

Specific treatments are associated with changes in specific bins. Please note that this method compresses the Hilbert curve by only considering indices that contain cells.

## Projecting the Hilbert curve in 2 Dimensions

The Hilbert curve can be re-folded back into any dimensionality using the `hilbertProjection` function. To visualize the Hilbert curve as a scatter plot, we use 2 as the target number of dimensions :

```{r}
proj <- hilbertProjection(hc,target = 2)
```

To visualize the Hilbert curve as a 2D projection, we first compute the number of cells per unique combination of annotation columns and Hilbert index:

```{r}
fullProj <- fullAnnots %>% 
    mutate(HilbertIndex=hc) %>% 
    group_by_at(vars(one_of(colnames(fullAnnots),'HilbertIndex'))) %>% 
    count() %>%
    bind_cols(as.data.frame(proj[.$HilbertIndex+1,]))
fullProjCount <- fullProj %>% 
    ungroup() %>% 
    count(HilbertIndex,Treatment) %>% 
    arrange(desc(n))
kable(head(fullProjCount))
```

The percentage of cells per treatment is visualized using `ggplot2` :

```{r,fig.height=8}
fullProj %>% 
    group_by(Treatment) %>% 
    mutate(PC=n/sum(n)) %>% 
    ggplot(aes(x=V1,y=V2))+
    geom_tile(aes(fill=PC),
              width=24,
              height=24)+
    facet_wrap(~Treatment)+
    scale_fill_gradient(low='grey80',high='blue')+
    theme_light(base_size=16)+
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank())
```

The quality of the projection will depend on the density of the Hilbert curve. As this curve is sparse (only `r round(100*length(table(hc))/nbins^ncol(fullMat),2)`% of the Hilbert index contain at least 1 cell) this projection is only provided as an example.

# Comparing cells using Hilbert index

With each cell now associated to a specific hilbert index, each sample can be described by the percentage of cells from a given sample that corresponds to a particular index. The resulting table can be visualized as a heat map:

```{r,fig.width=6,fig.height=6}
heatmap(log10(treatment.pc),
        scale = 'none',
        Colv = NA,
        Rowv = NA,
        labRow = NA,
        col = colorRampPalette(c('white',blues9))(256),
        margin = c(12,1))
```

In this experiment, each sample corresponds to a specific cell type and treatment: to compute the distance between samples we use a distance derived from information theory, the [Jensen-Shannon distance](https://www.cise.ufl.edu/~anand/sp06/jensen-shannon.pdf). This is done through the `js.dist` function. The resulting distance matrix can be used to compute a hierarchical cluster:

```{r}
treatment.dist <- js.dist(t(treatment.table))
treatment.hc <- hclust(treatment.dist)
```

When ordering samples using `treatment.hc` we see patterns emerging:

```{r,fig.width=6,fig.height=8}
heatmap(log10(treatment.pc),
        scale = 'none',
        Colv = as.dendrogram(treatment.hc),
        Rowv = NA,
        labRow = NA,
        col = colorRampPalette(c('white',blues9))(256),
        margin = c(12,2))
```

Samples corresponding to **reference** and **BCR/FcR-XL** cluster together, while **PMA/Ionomycin** and **vanadate** samples show strong differences with every other sample.
