---
title: "Minimal Working Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Minimal Working Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Feel free to report any issue

An issue can be an **error** but also an **improvement**. And improvement can be
a new function as well as a better documentation (like a simple typo).
Everything is welcome. There is no mistake when helping.

(Click here to report an issue)[https://gitlab.com/qonfluens/model/rbioacc/-/issues]

To help returning issue, we provide here a set of *Minimal Working Example* (MWE)
that you can modify to obtain yours.

Below are MWE where **each chunk** (piece of R code in grey area) work on its own.


```{r setup, eval=FALSE}
library(rbioacc)
```

### Prior Posterior

```{r plotPP, eval=FALSE}
data("Male_Gammarus_Single")
modelData_MGS <- modelData(Male_Gammarus_Single, time_accumulation = 4)
fit_MGS <- fitTK(modelData_MGS, iter = 10, chains = 2)

# Compute a dataframe with priors and posteriors
df_MGS <- df_PriorPost(fit_MGS)

# Compute a plot with priors and posteriors
plt_MGS <- plot_PriorPost(fit_MGS)
```

