---
title: "Introduction to DataSetsVerse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to DataSetsVerse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(DataSetsVerse)
```

# Introduction to DataSetsVerse

The `DataSetsVerse` is a metapackage that brings together a curated collection of R packages containing domain-specific datasets. It includes *time series data, educational metrics, crime records, medical datasets, and oncology research data*.

Designed to provide researchers, analysts, educators, and data scientists with centralized access to structured and well-documented datasets, this metapackage facilitates:

- Reproducible research

- Data exploration

- Teaching applications

across a wide range of domains.

## Installation

To install the `DataSetsVerse` package, use the following:

```{r install_package, eval=FALSE}

# Install from CRAN 
install.packages("DataSetsVerse")

# Then load the package:
library(DataSetsVerse)


```


## Using the DataSetsVerse() Function

Once the package is loaded, you can call the DataSetsVerse() function to display the list of included dataset packages and their versions:

```{r datasetsverse_function, eval=FALSE}

DataSetsVerse()

```


## Important Note on Detaching Packages

`DataSetsVerse` imports and depends on several subpackages. Therefore, you cannot detach an individual subpackage (like OncoDataSets) while DataSetsVerse is still loaded.

Example of an Error


```{r detach, eval=FALSE}

# This will raise an error
detach("package:OncoDataSets", unload = TRUE)

```


## Correct Way to Detach

To properly unload a subpackage, you must first detach DataSetsVerse:

```{r detach_way, eval=FALSE}

# First detach the metapackage
detach("package:DataSetsVerse", unload = TRUE)

# Now you can safely detach the subpackage
detach("package:OncoDataSets", unload = TRUE)

```


By installing the `DataSetsVerse` package this will attach the following packages to your R session:

- `timeSeriesDataSets`

- `educationR`

- `crimedatasets`

- `MedDataSets`

- `OncoDataSets`


## Included Packages

### timeSeriesDataSets

A comprehensive collection of time series datasets from multiple domains including:

- Economics

- Finance

- Energy

- Healthcare

Each dataset includes a suffix to denote its structure. Examples:

`AirPassengers_ts`: Monthly airline passengers (1949–1960)

`taylor_30_min_df_ts`: Half-hourly electricity demand

### educationR

- Contains datasets related to:

- Student performance

- Learning methods

- Test scores

- Absenteeism

Each dataset includes a suffix to denote its structure. Examples:

`Develop_tbl_df`: Dev Students: 2-Year & 4-Year College Demographics

`Devmath_tbl_df`: Fall '95 Developmental Math: Failed Student Scores

### crimedatasets

- Focuses exclusively on:

- Crimes and criminal activities

- Criminology

- Socio-economic analysis related to crime

Each dataset includes a suffix to denote its structure. Examples:

`TerrorismGlobal_table`: Global Terrorism Database (GTD) Yearly Summaries

`USATerror_data_df`: Terrorism Incidents in the USA (1968-1974)

### MedDataSets

Medical datasets covering:

- Drug effectiveness

- Vaccine trials

- Survival rates

- Public health and treatments

Each dataset includes a suffix to denote its structure. Examples:

`Aids2_df`: Australian AIDS Survival Data

`Cushings_df`: Diagnostic Tests on Patients with Cushing's Syndrome

### OncoDataSets

Provides rich datasets focused on cancer research, including:

- Survival rates

- Genetic studies

- Biomarkers

Cancer types (melanoma, leukemia, breast, ovarian, lung, etc.)

Each dataset includes a suffix to denote its structure. Examples:

`UKLungCancerDeaths_df`: Lung Cancer Deaths among UK Physicians

`USCancerStats_df`: US Cancer Incidence, Mortality, and Survival Changes


## Summary of Included Packages

| Package           | Domain            | Example Datasets                  |
|------------------|-------------------|-----------------------------------|
| timeSeriesDataSets | Time Series       | `AirPassengers_ts`, `taylor_30_min_df_ts` |
| educationR        | Education          | `Develop_tbl_df`, `Devmath_tbl_df` |
| crimedatasets     | Criminology        | `TerrorismGlobal_table`, `USATerror_data_df` |
| MedDataSets       | Medical Research   | `Aids2_df`, `Cushings_df`         |
| OncoDataSets      | Oncology           | `UKLungCancerDeaths_df`, `USCancerStats_df` |


## Conclusion

The `DataSetsVerse` provides a unified framework for accessing diverse, domain-specific datasets in R. Whether you're working in education, healthcare, criminology, or cancer research, this metapackage helps streamline your data workflow.
For more details and full documentation of each package, please refer to the reference manual and help files included with the package.
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
