---
title: "cycleTrendR Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cycleTrendR-overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(cycleTrendR)
set.seed(1)
```
# Introduction
cycleTrendR provides a unified framework for analyzing time‑series that contain
both trend and cyclic components. It supports:

LOESS, GAM, and GAMM trend estimation

Automatic Fourier harmonic selection (AICc/BIC)

Bootstrap confidence intervals (IID or MBB)

Change‑point detection

Lomb–Scargle periodogram for irregular sampling

Rolling‑origin forecasting

Publication‑quality ggplot2 visualizations

This vignette demonstrates the main workflows using simulated data.

# Simulated Example Data
We simulate a noisy cyclic signal with irregular sampling:
```{r} 
dates <- as.Date("2020-01-01") + cumsum(sample(1:3, 300, replace = TRUE))
signal <- sin(2*pi*as.numeric(dates)/20) + rnorm(300, 0, 0.3)
```
# LOESS Trend + Automatic Fourier Selection
```{r}
res_loess <- adaptive_cycle_trend_analysis(
  signal = signal,
  dates = dates,
  trendmethod = "loess",
  usefourier = TRUE,
  auto_fourier_select = TRUE,
  nboot = 50
)

res_loess$Plot$Trend
```

# GAM Trend
```{r} 
res_gam <- adaptive_cycle_trend_analysis(
  signal = signal,
  dates = dates,
  trendmethod = "gam",
  usefourier = TRUE,
  nboot = 50
)

res_gam$Plot$Trend
```

# GAMM Trend with Random Effects
```{r} 
group <- rep(letters[1:4], length.out = length(signal))

res_gamm <- adaptive_cycle_trend_analysis(
  signal = signal,
  dates = dates,
  trendmethod = "gam",
  use_gamm = TRUE,
  group_var = "subject",
  group_values = group,
  usefourier = FALSE,
  nboot = 20
)

res_gamm$Plot$Trend
```

# Irregular Sampling + Fourier + Lomb–Scargle
Irregular sampling is detected automatically and the Lomb–Scargle periodogram is used:
```{r} 
  res_irreg <- adaptive_cycle_trend_analysis(
  signal = signal,
  dates = dates,
  trendmethod = "loess",
  usefourier = TRUE,
  auto_fourier_select = TRUE,
  nboot = 50
)

res_irreg$Plot$Spectrum
```

# Change‑Point Detection
```{r} 
res_loess$ChangePoints
```

# Bootstrap Confidence Intervals
```{r} 
head(res_loess$CI$lower)
head(res_loess$CI$upper)
```

# Conclusion
cycleTrendR provides a flexible and robust toolkit for analyzing complex
time‑series with trend and cyclic components, especially when sampling is
irregular or noise levels are high. It is suitable for biomedical assay
monitoring, environmental signals, and forecasting tasks.
