
---
title: "cycleTrendR in practice: Universal Trend and Cycle Analysis"
author: "Pietro Piu"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cycleTrendR in practice}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(cycleTrendR)
library(ggplot2)
```

# Introduction

cycleTrendR provides a unified framework for trend estimation, cycle detection, spectral analysis, bootstrap confidence intervals, and change-point detection across **any type of time series**.

Version **0.3.0** introduces a major enhancement:

- full support for **multiple time formats** through the argument dates_type;

- a universal internal time index (timenum);

- automatic switching between STL, Lomb–Scargle, Fourier, LOESS, GAM, and GAMM;

- compatibility with **daily**, **sub-daily**, and **numeric** time series.

This vignette demonstrates how to use cycleTrendR in three real-world scenarios:

1. **Daily data** (Date)

2. **Wearable / PhysioNet data** (POSIXct)

3. **Neuroscience / spike trains** (numeric time)


# Supported time formats

adaptive_cycle_trend_analysis() accepts three types of temporal indexing:


|dates_type|	Input class |	Typical use case               |
|----------|--------------|--------------------------------|
| "date"	 |Date	        |Epidemiology, climate, economics|
|"posix"	 |POSIXct	      |Wearable sensors, HR, EDA, EEG  |
|"numeric" |numeric	      |Spike trains, firing rate, ...  |


The function automatically adapts:

- STL only for "date"

- Lomb–Scargle for "posix" and "numeric"

- Fourier harmonics in time units of timenum

- Trend estimation via LOESS, GAM, or GAMM

- Change-points in the same scale as the input time


# Example 1 — Daily data (Date)

```{r example_1}
set.seed(1)

dates <- as.Date("2020-01-01") + 1:200
signal <- sin(2*pi*(1:200)/30) + rnorm(200, 0, 0.2)

res_date <- adaptive_cycle_trend_analysis(
  signal = signal,
  dates = dates,
  dates_type = "date",
  trendmethod = "loess",
  usefourier = TRUE
)

res_date$Plot$Trend
res_date$Plot$Spectrum
```

This example illustrates:

- STL decomposition

- automatic seasonality detection

- Fourier harmonics

- LOESS trend

- model-based or bootstrap CI


# Example 2 — PhysioNet in practice (POSIXct)

```{r example_2}
# Example: synthetic POSIXct HR-like signal
t <- as.POSIXct("2020-01-01 00:00:00") + seq(0, by = 60, length.out = 1000)
hr <- 70 + 5*sin(2*pi*(1:1000)/200) + rnorm(1000, 0, 1)

res_posix <- adaptive_cycle_trend_analysis(
  signal = hr,
  dates = t,
  dates_type = "posix",
  trendmethod = "gam",
  usefourier = TRUE
)

res_posix$Plot$Trend
res_posix$Plot$Spectrum
```

Key features:

- Lomb–Scargle periodogram

- GAM or GAMM trend

- Fourier in seconds

- change-points in POSIXct

- robust CI

This is the recommended workflow for:

- PhysioNet HR

- PhysioNet EDA

- accelerometry

- EEG timestamps


# Example 3 — Neuroscience in practice (numeric time)

```{r example_3}
time <- seq(0, 10, length.out = 2000)
spike_rate <- sin(2*pi*time*5) + rnorm(2000, 0, 0.2)

res_num <- adaptive_cycle_trend_analysis(
  signal = spike_rate,
  dates = time,
  dates_type = "numeric",
  trendmethod = "loess",
  usefourier = TRUE
)

res_num$Plot$Trend
res_num$Plot$Spectrum
```


This workflow is ideal for:

- spike trains

- firing rate

- LFP power

- simulated time series

- any non-calendar time axis


# Best practices
- Use "date" when the time axis is calendar-based

- Use "posix" for sub-daily physiological signals

- Use "numeric" for experimental or simulated time

- Prefer GAM/GAMM for long or noisy signals

- Use Fourier only when a dominant period is meaningful

- Use bootstrap CI when residuals show autocorrelation


# Session info

```r
```r{r info}
sessionInfo()
```
