---
title: 'The flifo package'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The flifo package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(flifo)
```

The flifo package provides a few functions to create and manipulate FIFO (First In First Out), LIFO (Last In First Out), and NINO (Not In or Never Out) stacks in R.

Functions `fifo`, `lifo`, and `nino` are made to 
create empty stacks. For instance: 

```{r}
# Create an empty LIFO
s <- lifo()
print(s)

is.empty(s)
is.fifo(s)
is.lifo(s)
```

Then `push` and `pop` enable one to add elements to and retrieve 
elements from the stack, respectively. 

```{r}
# Add values to 's'
push(s, 0.3)
push(s, data.frame(x=1:2, y=2:3))
print(s)
size(s)# in bytes

# Retrive the last element inserted
pop(s)
size(s)
```

A maximum number of elements can be specified at the creation of the stack (no limit in the number of elements is the default). 

```{r, eval=FALSE}
s <- fifo(max_length = 3)
max_length(s)
#> [1] 3

# max_length can be changed
max_length(s) <- 2
push(s, 1)
push(s, 2)
push(s, 3) # generates an error
#> Error in push(s, 3) : '.stack' is full
```

If an object exists in the current environment `e` and is pushed into the stack, it disappears from `e`: 

```{r}
s <- lifo()
x <- 3.14
exists("x")
push(s, x)
exists("x")
```

The `nino` function creates a stack from which we cannot retrieve anything: 

```{r, eval=FALSE}
s <- nino()
push(s, "foo")
print(s)
#> NINO stack: no element can be reached
pop(s) # generates an error
#> Error in pop(s) : cannont retrieve elements from a 'nino' stack
```
