
# Greta

We finally run the `greta` version of our model with `runMCMC_btadjust()`. `greta` is rather different from `JAGS` and `NIMBLE` in that the model defines objects in R and thus does not require a model code to be passed to `runMCMC_btadjust()`, nor Data or Constants. The coding with `greta` is as follows:

```{r greta model preparation}
#in my setting I need to load not only greta but R6 & tensorflow packages
library(greta)
library (R6)
library(tensorflow)

#first requirement of greta: declaring the data that will be analyzed with the function as_data
Y<-as_data(y1000)

#we then proceed by writing the model directly in R, starting with the priors of the parameters using greta functions for probability distributions - here uniform()
population.mean<-uniform(0,5000)
population.sd<-uniform(0,100)
    
#we then define the distribution of the data - here with the normal distribution - by default parametrized with a standard deviation in greta:
try({distribution(Y)<-normal(population.mean,population.sd) })

#we finally declare the greta model, which will be the object passed to runMCMC_btadjust 
m<-model(population.mean, population.sd)

### we finally have to prepare initial values with a specific greta function - initials:
ModelInits.Greta <- function()
    {initials(population.mean = rnorm(1,600,90), population.sd = runif(1, 1, 30))}

set.seed(1)
  Inits.Greta<-lapply(1:Nchains,function(x){ModelInits.Greta()})

```

We are now ready to fit the model with `runMCMC_btadjust()`, specifying `MCMC_language="Greta"` and giving the argument `model` instead of `code` and `data`: 

```{r greta model running with runMCMC_btadjust, cache=FALSE}
out.mcmc.greta<-runMCMC_btadjust(model=m, MCMC_language="Greta",
    Nchains=Nchains,params=params,inits=Inits.Greta,
		niter.min=1000,niter.max=300000,
    nburnin.min=100,nburnin.max=200000,
		thin.min=1,thin.max=1000,
		conv.max=1.05, neff.min=1000)


```

```{r greta model running with runMCMC_btadjust table, echo=FALSE}

knitr::kable(format(summary(out.mcmc.greta)[[1]],digits=2,scientific=FALSE),align="r",caption="Summary of the statistical parameters of the greta model:")

```


We first check that estimations are similar to those with `NIMBLE` and `JAGS` with paired Kolmogorov-Smirnov tests:

```{r Nimble-Jags-Greta comparison parameters, echo=FALSE, warning=FALSE }
compar_MCMC_params<-format(cbind(c(ks.test(unlist(out.mcmc[,"population.mean"]),unlist(out.mcmc.greta[,"population.mean"]))$p.value, ks.test(unlist(out.mcmc.Jags[,"population.mean"]),unlist(out.mcmc.greta[,"population.mean"]))$p.value),c(ks.test(unlist(out.mcmc[,"population.sd"]),unlist(out.mcmc.greta[,"population.sd"]))$p.value
,ks.test(unlist(out.mcmc.Jags[,"population.sd"]),unlist(out.mcmc.greta[,"population.sd"]))$p.value
)),digits=4,scientific=FALSE)
rownames(compar_MCMC_params)<-c("Nimble vs. greta","Jags vs. greta")
knitr::kable(compar_MCMC_params,col.names=c("mean","sd"),align="r",caption="P-values of paired Kolmogorov-Smirnov tests of output parameters of the greta model with the default NIMBLE model and the JAGS model:")

```

We then report the efficiency of the MCMCs.

```{r Nimble-Jags-Greta comparison efficiency, echo=FALSE}
compar_MCMC_times<-format(cbind(unlist(attributes(out.mcmc)$final.params),unlist(attributes(out.mcmc.Jags)$final.params),unlist(attributes(out.mcmc.greta)$final.params)),digits=4,scientific=FALSE)
knitr::kable(compar_MCMC_times[1:(dim(compar_MCMC_times)[1]-1),],col.names=c("Nimble.default","Jags","Greta"),align="r",caption="Comparison of the efficiency of the default NIMBLE, the JAGS and the greta models:")

```

MCMC time (rows `duration.MCMC.transient` & `duration.MCMC.asymptotic`) was far greater with `greta` than with `JAGS` and `NIMBLE`, for a minimum number of effective values with `greta` of `r attributes(out.mcmc.greta)$final.diags$neff_synth[,"min"]`. Total duration is rather close with `greta` compared with `NIMBLE`, due to the great time required by `NIMBLE` for MCMC preparation - while this preparation is done outside `runMCMC_btadjust()` with `greta`. Yet, when we compare CPU total durations (`CPUduration`), `greta` gets worse than `NIMBLE` while it was the reverse for total duration (`duration`), simply because `greta` parallelized its process and therefore required more CPU time per time unit. 

We tried to give a second chance to `greta`, based on the following post: <https://forum.greta-stats.org/t/size-and-number-of-leapfrog-steps-in-hmc/332>. The idea was to let `greta` have more information to adapt its hmc parameters during the warm-up phase by just having more chains to run - hereafter, 15.

```{r second greta model running with runMCMC_btadjust, cache=FALSE}
Nchains.Greta<-15
ModelInits.Greta <- function()
    {initials(population.mean = rnorm(1,600,90), population.sd = runif(1, 1, 30))}

set.seed(1)
Inits.Greta<-lapply(1:Nchains.Greta,function(x){ModelInits.Greta()})
  
  out.mcmc.greta.morechains<-runMCMC_btadjust(model=m, MCMC_language="Greta",
    Nchains=Nchains.Greta,params=params,inits=Inits.Greta,
		niter.min=1000,niter.max=300000,
    nburnin.min=100,nburnin.max=200000,
		thin.min=1,thin.max=1000,
		conv.max=1.05, neff.min=1000)
  

```

```{r second greta model running with runMCMC_btadjust - inv, echo=FALSE}

knitr::kable(format(summary(out.mcmc.greta.morechains)[[1]],digits=2,scientific=FALSE),align="r",caption=paste0("Summary of the statistical parameters of the greta model with ",Nchains.Greta," chains:"))


```

This run was indeed much faster. Parameter estimates were still not significantly different from those with `NIMBLE` and `JAGS` based on paired Kolmogorov-Smirnov tests:

```{r Nimble-Jags-second Greta comparison parameters, echo=FALSE, warning=FALSE}
compar_MCMC_params<-format(cbind(c(ks.test(unlist(out.mcmc[,"population.mean"]),unlist(out.mcmc.greta.morechains[,"population.mean"]))$p.value, ks.test(unlist(out.mcmc.Jags[,"population.mean"]),unlist(out.mcmc.greta.morechains[,"population.mean"]))$p.value),c(ks.test(unlist(out.mcmc[,"population.sd"]),unlist(out.mcmc.greta.morechains[,"population.sd"]))$p.value
,ks.test(unlist(out.mcmc.Jags[,"population.sd"]),unlist(out.mcmc.greta.morechains[,"population.sd"]))$p.value
)),digits=4,scientific=FALSE)
rownames(compar_MCMC_params)<-c("Nimble vs. greta.morechains","Jags vs. greta.morechains")
knitr::kable(compar_MCMC_params,col.names=c("mean","sd"),align="r",caption="P-values of paired Kolmogorov-Smirnov tests of output parameters of the greta model with 15 chains with the default NIMBLE model and the JAGS model:")

```

We now report the efficiency of the MCMCs:

```{r Nimble-Jags-second Greta comparison efficiency, echo=FALSE}
compar_MCMC_times<-format(cbind(unlist(attributes(out.mcmc)$final.params),unlist(attributes(out.mcmc.Jags)$final.params),unlist(attributes(out.mcmc.greta.morechains)$final.params)),digits=4,scientific=FALSE)
knitr::kable(compar_MCMC_times[1:(dim(compar_MCMC_times)[1]-1),],col.names=c("Nimble.default","Jags","Greta.morechains"),align="r",caption="Comparison of the efficiency of the default NIMBLE, the JAGS and the greta.morechains models:")

```

We still observed more CPU duration with `greta`, although the associated number of effective values for `greta` was now `r attributes(out.mcmc.greta.morechains)$final.diags$neff_synth[,"min"]`, which rendered MCMC CPU efficiency with `greta` closer to `NIMBLE`.

