
# JAGS
We now turn to analyzing the same data with the same statistical model using `JAGS` with `runMCMC_btadjust()`. We rely on the data simulated above. In `JAGS`, we now put all the data in the same list:

```{r Jags Data}
ModelData.Jags <-list(mass = y1000, nobs = length(y1000))

```

We then propose the use of `JAGS` with a specification of the model from within `R` - which we find more convenient. We therefore write the `JAGS` model within `R` as a character chain:

```{r Jags ModelCode}

modeltotransfer<-"model {

		# Priors
			population.mean ~ dunif(0,5000)
			population.sd ~ dunif(0,100)

			# Normal distribution parameterized by precision = 1/variance in Jags
    	population.variance <- population.sd * population.sd
      precision <- 1 / population.variance

			# Likelihood
			for(i in 1:nobs){
			  mass[i] ~ dnorm(population.mean, precision)
			}
		}"


```

The other objects useful or required for running `runMCMC_btadjust` with `JAGS` are similar to those required with `NIMBLE` (`Inits`, `Nchains`, `params`) and are not repeated here.

We then launch `runMCMC_btadjust()` with `MCMC_language="Jags"`, specifying arguments `code` and `data` which are required in this case. Note that if we had written the `JAGS` code in a text file named `"ModelJags.txt"`, we would just have replaced in the command above `code=modeltotransfer` by `code="ModelJags.txt"`.


```{r Jags First model, cache=FALSE}

set.seed(1)
out.mcmc.Jags<-runMCMC_btadjust(code=modeltotransfer,  data = ModelData.Jags, MCMC_language="Jags", 
    Nchains=Nchains, params=params, inits=Inits,
    niter.min=1000,niter.max=300000,
    nburnin.min=100,nburnin.max=200000,
    thin.min=1,thin.max=1000,
		conv.max=1.05,neff.min=1000)

```

Here is a summary of the parameter estimates:

```{r Jags First model inv, echo=FALSE}

knitr::kable(format(summary(out.mcmc.Jags)[[1]],digits=2,scientific=FALSE),align="r",caption="Summary of the statistical parameters of the Jags model:")

```


Results seem in line with those of `NIMBLE`. We check this using a paired Kolmogorov-Smirnov tests with `NIMBLE` models:

```{r Nimble-Jags comparison parameters, echo=FALSE, warning=FALSE}
compar_MCMC_params<-format(cbind(c(ks.test(unlist(out.mcmc.Geweke[,"population.mean"]),unlist(out.mcmc.Jags[,"population.mean"]))$p.value
,ks.test(unlist(out.mcmc.Coda.Geweke[,"population.mean"]),unlist(out.mcmc.Jags[,"population.mean"]))$p.value
,ks.test(unlist(out.mcmc[,"population.mean"]),unlist(out.mcmc.Jags[,"population.mean"]))$p.value
),c(ks.test(unlist(out.mcmc.Geweke[,"population.sd"]),unlist(out.mcmc.Jags[,"population.sd"]))$p.value
,ks.test(unlist(out.mcmc.Coda.Geweke[,"population.sd"]),unlist(out.mcmc.Jags[,"population.sd"]))$p.value
,ks.test(unlist(out.mcmc[,"population.sd"]),unlist(out.mcmc.Jags[,"population.sd"]))$p.value
)),digits=4,scientific=FALSE)
rownames(compar_MCMC_params)<-c("Nimble.Geweke vs. Jags","Nimble.Coda.Geweke vs. Jags","Nimble.Default vs. Jags")
knitr::kable(compar_MCMC_params,col.names=c("mean","sd"),align="r", caption="P-values of paired Kolmogorov-Smirnov tests of output parameters (columns) of the Jags model with the three NIMBLE models (rows):")

```

Our results do confirm that the `JAGS` result cannot be considered as stemming from a different probability distribution than `NIMBLE` results.

We finally compare the efficiency of the `JAGS` and default `NIMBLE` MCMCs:

```{r Nimble-Jags comparison efficiency, echo=FALSE}
compar_MCMC_times<-format(cbind(unlist(attributes(out.mcmc)$final.params),unlist(attributes(out.mcmc.Jags)$final.params)),digits=4,scientific=FALSE)
knitr::kable(compar_MCMC_times[1:(dim(compar_MCMC_times)[1]-1),],col.names=c("Nimble.default","Jags"),align="r", caption="Comparison of the efficiency of the default NIMBLE model and the Jags model:")

```

The conclusion is that `JAGS` is much faster than `NIMBLE` on this example (row named `duration` in the previous table), due to much less time devoted to MCMC preparation - as well as to burn-in/thinning adjustment (rows named `duration.MCMC.preparation` and `duration.btadjust` in the previous table). Actually there is no adjustment with `JAGS` (`niter.tot` is equal to the initial number of iterations). Yet, `NIMBLE` is quicker regarding MCMC updating by iteration since it took `NIMBLE` less time than `JAGS` for the transient phase (respectively less than twice time for the asymptotic phase) although using more than `r as.double(compar_MCMC_times["burnin",1])/as.double(compar_MCMC_times["burnin",2])` (resp. `r (as.double(compar_MCMC_times["niter.tot",1])-as.double(compar_MCMC_times["burnin",1]))/(as.double(compar_MCMC_times["niter.tot",2])-as.double(compar_MCMC_times["burnin",2]))` for the asymptotic phase) times more iterations than `JAGS`.

At first sight, we would also conclude that MCMC efficiency per effective value is also better with `NIMBLE` since both languages had the same target for the minimum number of effective value - 1,000 - and the total MCMC time was lower with `NIMBLE`. Yet, the number of effective values are different:

```{r Nimble-Jags second comparison efficiency, echo=FALSE}
compar_MCMC_times<-format(cbind(c(attributes(out.mcmc)$final.diags$neff_synth[,"min"
], (attributes(out.mcmc)$final.params$CPUduration.MCMC.transient+attributes(out.mcmc)$final.params$CPUduration.MCMC.asymptotic)/attributes(out.mcmc)$final.diags$neff_synth[,"min"
]),c(attributes(out.mcmc.Jags)$final.diags$neff_synth[,"min"
], (attributes(out.mcmc.Jags)$final.params$CPUduration.MCMC.transient+attributes(out.mcmc.Jags)$final.params$CPUduration.MCMC.asymptotic)/attributes(out.mcmc.Jags)$final.diags$neff_synth[,"min"
])),digits=4,scientific=FALSE)
row.names(compar_MCMC_times)<-c("Min. Number Eff. values","MCMC CPU time per Effective Value")

```

```{r Nimble-Jags second comparison efficiency bis, echo=FALSE}
knitr::kable(compar_MCMC_times,col.names=c("Nimble.default","Jags"),align="r",caption="Comparison of the number of effective values between the default NIMBLE model and the JAGS model:")

```

`r if (as.double(compar_MCMC_times[1,2])>=as.double(compar_MCMC_times[1,1])) {paste0("Indeed, \"JAGS\" with just the first iterations produced a higher number of effective values - actually bigger than the targeted \"neff.min\"  - than \"NIMBLE\".")}`
`r if (as.double(compar_MCMC_times[2,1])<as.double(compar_MCMC_times[2,2])) { "Yet, the MCMC time per effective value remained lower with \"NIMBLE\" than with \"JAGS\" with this model (cf. table above)."}`
`r if (as.double(compar_MCMC_times[2,1])>as.double(compar_MCMC_times[2,2])) { "In line with this, the MCMC time per effective value was actually lower with \"JAGS\" than with \"NIMBLE\" with this model (cf. table above)."}`

