
# Automatic removal of chains

We finally present an attractive feature of `runMCMC_btadjust()` which is to remove chains whose parameters are all invariable during the firs cycle of the MCMC. This is an issue that can appear for some chains, especially when initial values of parameters are far away from likely values and the MCMC sampler cannot move from the corresponding value. We here illustrate this on a simplified version of the above model in which the standard deviation is given (i.e. is not estimated) - simply because in the case the sampler can move for this parameter. We introduce a low standard deviation value in the normal distribution. Here is the code:

```{r Nimble new code}

ModelCode<-nimbleCode(
  {
    # Priors
    population.mean ~ dunif(0,5000)
    
    # Likelihood
    for(i in 1:nobs){
      mass[i] ~ dnorm(population.mean, 1)
    }
  })


```

We then update the initial values, introducing a very unlikely value in the second chain:

```{r Nimble new initial values...}
Nchains <- 3

ModelInits <- function()
{list (population.mean = rnorm(1,600,2))}
  

set.seed(1)
Inits<-lapply(1:Nchains,function(x){ModelInits()})

Inits[[2]]<-list(population.mean=-600)


#specifying the names of parameters to analyse and save:
params <- c("population.mean") 


```


We then launch `runMCMC_btadjust()` in this setting:

```{r Nimble pbIV first run, cache=FALSE}

out.mcmc.pbIV<-runMCMC_btadjust(code=ModelCode, constants = ModelConsts, data = ModelData, MCMC_language="Nimble",
    Nchains=Nchains, params=params, inits=Inits,
    niter.min=1000, niter.max=300000,
    nburnin.min=100, nburnin.max=200000, 
    thin.min=1, thin.max=1000,
    conv.max=1.05, neff.min=1000)


```

Despite the very unlikely value for chain 2, the model converges and reached the number of effective values. We also notice in the above that only two chains were updated - chains number 1 and 3. Actually, only two chains are present in the results and the output specifies that chain 2 was removed.

```{r output of the above, cache=FALSE}
length(out.mcmc.pbIV)

attributes(out.mcmc.pbIV)$final.params$final.Nchains

attributes(out.mcmc.pbIV)$final.params$removed.chains


```

Note that if the removal of chains makes the number of chains pass from more than 1 to 1, the MCMC is stopped because the convergence diagnostic type and its targets would not adapt to a situation where there is only one chain.


```{r Nimble Nimble pbIV with 2 chains, cache=FALSE}

out.mcmc.pbIV.2chains<-runMCMC_btadjust(code=ModelCode, constants = ModelConsts, data = ModelData, MCMC_language="Nimble",
    Nchains=2, params=params, inits=Inits[1:2],
    niter.min=1000, niter.max=300000,
    nburnin.min=100, nburnin.max=200000, 
    thin.min=1, thin.max=1000,
    conv.max=1.05, neff.min=1000)


```

Removed chains are removed from the convergence and effective number diagnostics as well as from the outputs of `runMCMC_btadjust()`. If used with `Nimble`, it is also removed from the updated chains - thus saving computer resources. Such was not found possible with `Jags` and `Greta`. This 
