
# Hamiltonian Monte Carlo sampler

We will in the sequel try two other possibilities provided by Nimble. The first one is to use an Hamiltonian sampler - similar in principle to the one provided by `STAN` - which is a priori well suited to such badly behaving cases. This will require rather numerous steps to do so. First, this will require the loading of the new `nimbleHMC` library as well as modifications of `nimbleOptions` in the component `parallelizeInitExpr` of parameter `control.MCMC`; this is done in what follows in the variable `newParallelizeInitExpr`. Second, we need to refer to the HMC sampler in the `confModel.expression.toadd` component. Third, we need to turn to TRUE the `buildDerivs` component of parameter `control.MCMC`. This actually means that the log posterior density should be derivable relative to all the parameters to use this sampler - or at least relative to the parameters on which this sampler is applied. This is what we now perform:

```{r Nimble second run, with HMC, cache=FALSE}

newParallelizeInitExpr<-expression({
      library(nimble)
      library(nimbleHMC)
      nimbleOptions(MCMCusePredictiveDependenciesInCalculations = TRUE)
      nimbleOptions(MCMCorderPosteriorPredictiveSamplersLast = FALSE)
      nimbleOptions(enableDerivs = TRUE)
      })

sampler.expression.toadd<-expression(
  {ConfModel[[i]]$removeSamplers(c("Intercept","Slope"))
   nimbleHMC::addHMC(ConfModel[[i]], target=c("Intercept","Slope"))
  }  )

out.mcmc.HMC<-runMCMC_btadjust(code=ModelCode, constants = ModelConsts, data = ModelData, MCMC_language="Nimble",
    Nchains=Nchains, params=params, inits=Inits,
    niter.min=1000, niter.max=Inf,
    nburnin.min=100, nburnin.max=Inf, 
    thin.min=1, thin.max=Inf,
    conv.max=1.05, neff.min=1000,
    control=list(neff.method="Coda", time.max=120, print.diagnostics=FALSE),
    control.MCMC=list(confModel.expression.toadd=sampler.expression.toadd, parallelizeInitExpr= newParallelizeInitExpr,buildDerivs=TRUE, parallelize=TRUE, n.adapt=1000))
```

`r if(as.logical(attributes(out.mcmc.HMC)$final.params$converged)&!as.logical(attributes(out.mcmc.HMC)$final.params$neffs.reached)){ paste0("With these adaptations, within two minutes, the Hamiltonian Monte Carlo sampler does not go very far; it converges but does not reach the required number of effective values. This is due to a longer time per iteration and a longer time of MCMC preparation than with other samplers. The associated summary is given below:")}`

`r if(!as.logical(attributes(out.mcmc.HMC)$final.params$converged)&!as.logical(attributes(out.mcmc.HMC)$final.params$neffs.reached)){ paste0("With these adaptations, within two minutes, the Hamiltonian Monte Carlo sampler does not go very far; it does not converge and does not reach the required number of effective values. This is due to a longer time per iteration and a longer time of MCMC preparation than with other samplers. The associated summary is given below:")}`


```{r summary of Nimble second run, with HMC, cache=FALSE}
summary(out.mcmc.HMC)

effectiveSize(out.mcmc.HMC)
```

