
## NIMBLE

We start with fitting the example with `NIMBLE` (cf. <https://r-nimble.org/>).

```{r Nimble setup}

library(runMCMCbtadjust)
library(nimble)

```

As `NIMBLE` distinguishes data that have random distributions from other data, we specify two distinct lists to contain these:

```{r Nimble Data}

ModelData <-list(mass = y1000)
ModelConsts <- list(nobs = length(y1000))

```

We then write our Bayesian code within R with the `nimbleCode` function in the `nimble` package:

```{r Nimble Code}
 ModelCode<-nimbleCode(
  {
    # Priors
    population.mean ~ dunif(0,5000)
    population.sd ~ dunif(0,100)
    
    # Normal distribution parameterized by precision = 1/variance in Nimble
    population.variance <- population.sd * population.sd
    precision <- 1 / population.variance
  
    # Likelihood
    for(i in 1:nobs){
      mass[i] ~ dnorm(population.mean, precision)
    }
  })

```

The model is a simple linear - and Gaussian - model with only an intercept -, actually the same model - for the likelihood section -  as the probabilistic model used to generate the data.

Our - optional - next step is to specify starting values for model's parameters. This is done by first writing a function that is repetitively called for each chain. We - also optionally - indicate the names of parameters to be saved and diagnosed in a vector called `params`:

```{r Nimble Inits}
ModelInits <- function()
{list (population.mean = rnorm(1,600,90), population.sd = runif(1, 1, 30))}
  
Nchains <- 3

set.seed(1)
Inits<-lapply(1:Nchains,function(x){ModelInits()})

#specifying the names of parameters to analyse and save:
params <- c("population.mean", "population.sd") 

#devising the maximum level allowed for the Geweke diagnostic of convergence (cf. following)
npars<-length(params)
Gew.Max<-as.double(format(quantile(sapply(1:100000,function(x,N){max(abs(rnorm(N)))},npars),0.95),digits=3,scientific=FALSE))

```
