---
title: "Using Cluster.OBeu with OpenCPU"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Cluster.OBeu with OpenCPU}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

This document describes the use of the functions implemented in *Cluster.OBeu* package in *OpenCPU* environment, after *[installing OpenCPU](https://www.opencpu.org/download.html)* and *Cluster.OBeu* package on your OpenCPU server address(/ocpu/test).

# How to use functions

- Go to: yourserver/ocpu/test

- Type to the endpoint:

```{r, eval=FALSE, include=TRUE}
 ../library/ {name of the library} /R/ {function}
```

- If you want to see the function parameters you should: 

    - Select Method: **`Get`**
    
- To run a function you should:

    - Select Method: **`Post`**

- Then you can click on **Ajax Request**.

# OpenCPU and Cluster.OBeu

## cl.analysis

In this example we will use `cl.analysis` function that returns in a single call a json string or a list with the following components:

| Component      | Description                        |
| :------------: | :--------------------------------: |
| cluster.method | Label of the clustering algorithm  |
| raw.data       | Input data                         |
| data.pca       | Principal components               |
| modelparam     | Clustering model specifications    |
| compare        | Clustering measures                |


Table: *`cl.analysis`* components

## Select library and function

1. Go to: yourserver/ocpu/test

2. Copy and paste the following function to the endpoint 
```{r, eval=FALSE, include=TRUE}
../library/Cluster.OBeu/R/cl.analysis
# library/ {name of the library} /R/ {function}
```

3. **Select Method**: **`Post`**

## Adding parameters parameters

Click **add parameters** every time you want to add a new parameters and values.

4. Define the input data:

    - **Param Name**: **`cl.data`**
    - **Param Value**: e.g. **`sample_city_data`** 

5. Define the *cluster algorithm* parameter:

    - **Param Name**: **`cl.meth`**
    - **Param Value**: **`pam`**
    
6. Define the *cluster number* parameter:

    - **Param Name**: **`clust.numb`**
    - **Param Value**: **`3`**

For larger datasets you can specify likewise a specific clustering feature (*cl_feature*), the amounts and the aggregation if needed by defining *cl_feature*, *amount*, *cl.aggregate* parameters, see Cluster.OBeu *reference manual* for further details.

7. Ready! Click on **Ajax request**!

## Results

8. copy the **/ocpu/tmp/{this_id_number}/R/.val** (second on the right panel)

9. finally, paste **`yourserver/ocpu/tmp/{this_id_number}/R/.val`** on a new tab.


# Further Details

+ [HTTP in OpenCPU](https://www.opencpu.org/api.html)
+ [OpenCPU Help](https://www.opencpu.org/help.html)
+ [OpenCPU JavaScript Client](https://www.opencpu.org/jslib.html)
+ [OpenCPU on CRAN](https://cran.r-project.org/package=opencpu)

# Github
+ [OpenCPU package *development version*](https://github.com/opencpu/opencpu)
+ [Cluster.OBeu package *development version*](https://github.com/okgreece/Cluster.OBeu)
