---
title: "CrownScorchTLS Workflow"
author: "Jeffery Cannon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
Suggests: rmarkdown(>= 2.23)
vignette: >
  %\VignetteIndexEntry{CrownScorchTLS Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

#remotes::install.packages('tiagodc/TreeLS')
```

# Introduction

This vignette demonstrates the main workflows in the `CrownScorchTLS` package for estimating crown scorch from terrestrial lidar scans collected with a RIEGL vz400i, following [Cannon et al. 2025](https://doi.org/10.1186/s42408-025-00420-0).

# Workflow Part 1: Predict Scorch with Default Model

## Load Individual Tree

```{r load-read, fig.width=4, fig.height=4}
library(CrownScorchTLS)
library(lidR)
# Download external file to a temporary .las/.laz file from data repo
url = "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/D-03-10867_post.laz"
las_file <- tempfile(fileext = paste0(".", tools::file_ext(url)))
download.file(url, las_file, mode = "wb", quiet = TRUE)
las <- readLAS(las_file)
#plot(las, color = 'Intensity')
```

## Generate Reflectance Histogram

```{r histogram, fig.width=4, fig.height=4, eval=FALSE}
crown <- remove_stem(las)
crown <- add_reflectance(crown)
histogram <- get_histogram(crown)
plot(density ~ intensity, data = histogram, xlab='Reflectance (dB)', type='l')
```

## Predict Scorch

```{r predict, eval=FALSE}
scorch <- predict_scorch(las)
scorch
```

# Workflow Part 2: Multiple Trees

```{r multiple-trees, eval=FALSE, fig.width=4, fig.height=6}
directory <- system.file('extdata', package = 'CrownScorchTLS')
filenames <- list.files(directory, pattern='.laz', full.names=TRUE)

par(mfrow = c(3,2), mar = c(4,4,1,1))
for(f in filenames) {
  las <- readLAS(f)
  scorch <- suppressMessages(predict_scorch(las, plot=TRUE))
  cat('file:\t', basename(f), '\t', 'scorch:\t', round(scorch,3),'\n')
}
```

# Workflow Part 3: Custom Prediction Model

The steps for creating a customized model are similar: segment trees, generate histograms, train a custom random forest model, and predict scorch on new objects.

You can find a detailed example at <https://github.com/jbcannon/CrownScorchTLS>

```{r custom-model, fig.width=4, fig.height=4, eval=FALSE}
## ================================
## 1. Load segmented trees
## ================================

# --- Option A: Use example data from GitHub (default for vignette) ---
urls <- c(
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/M-04-15549_post.laz",
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/L-05-14669_post.laz",
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/E-08-9269_post.laz",
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/B-04-4286_post.laz",
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/D-03-10867_post.laz",
  "https://raw.githubusercontent.com/jbcannon/CrownScorchTLS-data/main/data/manual-clip-trees/C-04-11029_post.laz"
)

# Download each file to a temporary .las/.laz file
filenames <- sapply(urls, function(u) {
  tf <- tempfile(fileext = paste0(".", tools::file_ext(u)))
  download.file(u, tf, mode = "wb", quiet = TRUE)
  tf
})


# --- Option B: User supplies their own local files ---
# directory <- "C:/mydata/segmented_trees/"
# filenames <- list.files(directory, pattern = "\\.(las|laz)$", full.names = TRUE)


## ================================
## 2. Extract histogram features
## ================================

library(tidyr)
library(dplyr)

prediction_data <- lapply(filenames, function(f) {
  las  <- readLAS(f)
  crown <- tryCatch(remove_stem(las), error = function(e) las)
  crown <- add_reflectance(crown)

  hist <- get_histogram(crown)
  hist$intensity <- round(hist$intensity, 1)

  pivot_wider(hist,
    names_from = intensity,
    values_from = density,
    names_prefix = "intensity_"
  )
})

prediction_data <- bind_rows(prediction_data)


## ================================
## 3. Train prediction model
## ================================

scorch_training <- c(0.05, 0.20, 0.50, 0.75, 0.95, 0.95)

library(randomForest)
library(Boruta)

boruta.sel <- Boruta(x = prediction_data, y = scorch_training)
important <- names(boruta.sel$finalDecision[boruta.sel$finalDecision != "Rejected"])

model.RF <- randomForest(
  x = prediction_data[, important],
  y = scorch_training
)

print(model.RF)
varImpPlot(model.RF)


## ================================
## 4. Predict scorch on new trees
## ================================

# --- User-supplied new lidar object ---
# new_las <- readLAS("C:/mydata/newtree1.las")

# Example using the first file:
new_las <- readLAS(filenames[1])

predicted_scorch <- predict_scorch(new_las, model = model.RF)
predicted_scorch

```
