---
title: "Get started with cancerscreening"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with cancerscreening}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
auth_success <- tryCatch(
  khisr:::khis_cred_docs(),
  khis_cred_internal_error = function(e) e
)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE,
  purl = khisr::khis_has_cred(),
  eval = khisr::khis_has_cred()
)
```

```{r eval = !khisr::khis_has_cred(), echo = FALSE, comment = NA}
cancerscreening:::cancerscreening_bullets(c(
  "Code chunks will not be evaluated, because:",
  strsplit(auth_success$message, split = "\n")[[1]]
))
khisr::khis_cred_clear()
```

```{r, include = FALSE}
library(cancerscreening)
```

This article takes a quick tour of the main features of cancerscreening.

Remember to see [the articles](https://cancerscreening.damurka.com/articles/index.html) for more detailed treatment of all these topics and more.


## Key Functions for Data Download:

These functions directly call the KHIS server to download the data require the setting of KHIS credentials. See [setting the credentials](https://cancerscreening.damurka.com/articles/set-your-credentials.html) for more information. The cancer screening data that are being tracked breast cancer, cervical cancer,
colorectal cancer and laboratory diagnostic data.

### Cervical Cancer:

* `get_cervical_screened()`: Downloads data on women screened for cervical cancer.
* `get_cervical_positive()`: Retrieves data on women who tested positive for cervical pre-cancer or cancer.
* `get_cervical_treated()`: Accesses data on women who received treatment for cervical pre-cancer or cancer.

### Breast Cancer:

* `get_breast_cbe()`: Downloads data on women who received CBE for breast cancer screening.
* `get_breast_ultrasound()`: Downloads data on women who received ultrasound for breast cancer screening.
* `get_breast_mammogram()`: Downloads data on women who received mammograms for breast cancer screening.

### Colorectal Cancer:

* `get_colorectal_fobt()`: Retrieves data on individuals who receive FOBT for colorectal cancer screening tests.
* `get_colorectal_colonoscopy()`: Retrieves data on individuals who receive colonoscopy for colorectal cancer screening tests.

### Laboratory Data
* `get_lab_fluid_cytology()`: Retrieve data on fluid cytology done in the lab. They include Ascitic fluid, cerebral spinal fluid, pleural fluid and urine.
* `get_lab_tissue_histology()`: Retrieve data on tissue histology done in the lab. They include Breast, colorectal, oesphageal, head & neck, hepatobiliary, lymph nodes, oral, ovary, prostate, skin and uterine tissues;
* `get_lab_bone_marrow()`: Retrieve data on bone marrow done in the lab.
* `get_lab_fna()`: Retrieve data on fine-needle aspiration done in the lab. They include breast, liver, lymph node and thyroid.
* `get_lab_smears()`Retrieve data on smears done in the lab. They include pap smear, tissue impressions, and touch preparation.

To get the data the following calls can be made

```{r eval = khisr::khis_has_cred()}
# Get data for those screening for cervical cancer
cervical_screened <- get_cervical_screened('2022-01-01', end_date = '2022-06-30')
cervical_screened

# Get data for those screening for colorectal cancer using FOBT
colorectal_screened <- get_colorectal_fobt('2022-01-01', end_date = '2022-06-30')
colorectal_screened

# Get data for those screening for breast cancer using mammogram
breast_screened <- get_breast_mammogram('2022-01-01', end_date = '2022-06-30')
breast_screened
```


## Target Population Functions:

These functions do *not* require to access the KHIS server the project and calculate the target population guided the 
[Kenya housing and population census 2019](https://www.knbs.or.ke) and the 
[Kenya National Cancer Screening guidelines 2019](https://www.iccp-portal.org/system/files/plans/KENYA%20NATIONAL%20CANCER%20CONTROL%20STRATEGY%202017-2022_1.pdf).

The functions include: `get_cervical_target_population()`, `get_colorectal_target_population()`, `get_breast_cbe_target_population()`
and `get_breast_mammogram_target_population()`. 

If these function do not meet your criteria you can make your target population using the `get_filtered_population()`.

```{r eval = khisr::khis_has_cred()}
# Get the cervical screening target population for 2022
cervical_target_population <- get_cervical_target_population(2022)
cervical_target_population

# Get the colorectal cancer screening target population for 20223 by county
colorectal_target_population <- get_colorectal_target_population(2023, level = 'county')
colorectal_target_population

# Get the population of women 15-49 year for the year 2024
wra_pop <- get_filtered_population(year = 2024, min_age = 15, max_age = 49, pop_sex = 'female')
wra_pop
```



