---
title: "Generating STICS text files from XML files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating STICS text files from XML files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  javastics_path: /tmp/JavaSTICS-1.5.2-STICS-10.2.0
  workspace_path: /tmp/JavaSTICS-1.5.2-STICS-10.2.0/example
  output_path: /tmp/example
  eval: FALSE
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r eval=FALSE}
library(SticsRFiles)
```

```{r include=FALSE}
suppressWarnings(library(SticsRFiles))
```

```{r, echo = FALSE, eval = TRUE}
javastics_path <- params$javastics_path
workspace_path <- params$workspace_path
output_path <- params$output_path
chunk_eval <- params$eval
#java_cmd <- params$java_cmd
gen_usms_xml2txt <- SticsRFiles::gen_usms_xml2txt
```

## Introduction

In this part information are focusing on functionalities related to generation of text input files for the STICS model.

Generated files will be directly usable for running the model with **SticsOnR** package functions. See the documentation [here](https://sticsrpacks.github.io/SticsOnR/).

## Converting JavaSTICS workspace files
For using the model directly neither using the JavaSTICS graphical interface nor the `run_javastics` function interface, we provide a function, `gen_usms_xml2txt`, for converting JavaSTICS XML files to STICS text files from a JavaStics workspace. 

Then, the model can be run using the [`run_stics()`](https://sticsrpacks.github.io/SticsOnR/reference/run_stics.html) function from [SticsOnR](https://sticsrpacks.github.io/SticsOnR/)

Observation files may also be copied if they have a standard name corresponding to a usm name and a `.obs` extension. If not, they must be renamed to do so.

```{r paths_def, eval = FALSE}
# Specifying the JavaSTICS folder
javastics_path <- "/path/to/JavaSTICS-1.5.2-STICS-10.2.0"

# Specifying a workspace as a subfolder of JavaSTICS
workspace_path <- "example"
# or an absolute path to an external folder
# workspace_path <- "/path/to/javastics/workspace"

# Specifying an output folder path
output_path <- "/path/to/output/folder"

```

## Converting files into separated folders (one per usm)
```{r convert_to_separate_1, eval = chunk_eval}
## Generating files for all the usms contained in the workspace

# Into the workspace directory
gen_usms_xml2txt(javastics_path,
                 workspace_path)
#> ℹ USM 'SugarCane' successfully created
#> ℹ USM 'potato' successfully created
#> ℹ USM 'banana' successfully created
#> ℹ USM 'sorghum' successfully created
#> ℹ USM 'barley' successfully created
#> ℹ USM 'sugarbeet' successfully created
#> ℹ USM 'wheat' successfully created
#> ℹ USM 'maize' successfully created
#> ℹ USM 'soybean' successfully created
#> ℹ USM 'lettuce' successfully created
#> ℹ USM 'tomato' successfully created
#> ℹ USM 'DurumWheat' successfully created
#> ℹ USM 'rapeseed' successfully created
#> ℹ USM 'sunflower' successfully created
#> ℹ USM 'grass' successfully created
#> ℹ USM 'BareSoil' successfully created
#> ! Obs file not found for USM
#> "demo_Wheat1": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/demo_Wheat1.obs'
#> ℹ USM 'demo_Wheat1' successfully created
#> ! Obs file not found for USM
#> "demo_BareSoil2": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/demo_BareSoil2.obs'
#> ℹ USM 'demo_BareSoil2' successfully created
#> ! Obs file not found for USM
#> "demo_maize3": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/demo_maize3.obs'
#> ℹ USM 'demo_maize3' successfully created
#> ! Obs file not found for USM
#> "DurumWheat_4years": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/DurumWheat_4years.obs'
#> ℹ USM 'DurumWheat_4years' successfully created
#> ! Obs file not found for USM
#> "maize_4years": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/maize_4years.obs'
#> ℹ USM 'maize_4years' successfully created
#> ℹ USM 'strawberry' successfully created
#> ℹ USM 'pea' successfully created
#> ℹ USM 'vine' successfully created
#> ℹ USM 'fescue' successfully created
#> ℹ USM 'flax' successfully created
#> ! Obs file not found for USM
#> "intercrop_pea_barley": '/tmp/JavaSTICS-1.5.2-STICS-10.2.0/example/intercrop_pea_barley.obs'
#> ℹ USM 'intercrop_pea_barley' successfully created
#> ℹ USM 'timothy' successfully created
#> ℹ USM 'DurumWheat_snow' successfully created
#> ℹ USM 'Turmeric' successfully created
#> ℹ USM 'cc_BristleOat' successfully created
#> ℹ USM 'cc_mustard' successfully created
#> ℹ USM 'cc_ItalianRyegrass' successfully created
#> ℹ USM 'cc_vetch' successfully created
#> ℹ USM 'cc_CrimsonClover' successfully created
#> ℹ USM 'proto_rice' successfully created

```

```{r convert_to_separate_2, eval = chunk_eval}

# Into a specific output folder, with verbose mode turned off
gen_usms_xml2txt(javastics_path,
                 workspace_path,
                 out_dir = output_path,
                 verbose = FALSE)


## Generating files for a subset of usms
# Into the workspace directory
gen_usms_xml2txt(javastics_path,
                 workspace_path,
                 usm = c("banana", "wheat"))
#> ℹ USM 'banana' successfully created
#> ℹ USM 'wheat' successfully created

# Into a specific folder
gen_usms_xml2txt(javastics_path,
                 workspace_path,
                 usm = c("banana", "wheat"),
                 out_dir = output_path)
#> ℹ USM 'banana' successfully created
#> ℹ USM 'wheat' successfully created

```

```{r convert_to_separate_3, eval = chunk_eval}
## Getting returned information about files generation
gen_info <- gen_usms_xml2txt(javastics_path,
                             workspace_path,
                             usm = c("banana", "wheat"),
                             out_dir = output_path)
#> ℹ USM 'banana' successfully created
#> ℹ USM 'wheat' successfully created


gen_info
#> $usms_path
#> [1] "/path/to/output/folder/banana"
#> [2] "/path/to/output/folder/wheat"
#>
#> $files
#>  [1] "climat.txt"      "param.sol"       "ficini.txt"      "ficplt1.txt"
#>  [5] "fictec1.txt"     "station.txt"     "new_travail.usm" "tempopar.sti"
#>  [9] "tempoparv6.sti"  "ficplt2.txt"     "fictec2.txt"
#>
#> $copy_status
#> [1] TRUE TRUE
#>
#> $obs_copy_status
#> [1] TRUE TRUE
#>
#> $lai_copy_status
#> [1] FALSE FALSE
```

## Converting files into one folder (overwriting case) 
```{r convert_to_one, eval = chunk_eval}
# Generating files directly in the workspace or a specific folder
# (no usm sub-folder)
# In this case the model files are overwritten at each gen_usms_xml2txt call !
# In the workspace
gen_usms_xml2txt(javastics_path,
                 workspace_path,
                 usm = "banana",
                 dir_per_usm_flag = FALSE,
                 verbose = FALSE)

# In a specific folder
gen_usms_xml2txt(javastics_path,
                 workspace_path,
                 usm = "banana",
                 out_dir = output_path,
                 dir_per_usm_flag = FALSE,
                 verbose = FALSE)
```
