---
title: "Introduction to iClusterVB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to iClusterVB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# iClusterVB


iClusterVB allows for fast integrative clustering and feature selection for high dimensional data.
    
Using a variational Bayes approach, its key features - clustering of mixed-type data, automated determination of the number of clusters, and feature selection in high-dimensional settings - address the limitations of traditional clustering methods while offering an alternative and potentially faster approach than MCMC algorithms, making __iClusterVB__ a valuable tool for contemporary data analysis challenges.

There is a simulated dataset included as a list in the package that we can use to illustrate iClusterVB.

### Data pre-processing

```{r sim_data_example}
library(iClusterVB)

# sim_data comes with the iClusterVB package.
dat1 <- list(
  gauss_1 = sim_data$continuous1_data[c(1:20, 61:80, 121:140, 181:200), 1:75],
  gauss_2 = sim_data$continuous2_data[c(1:20, 61:80, 121:140, 181:200), 1:75],
  poisson_1 = sim_data$count_data[c(1:20, 61:80, 121:140, 181:200), 1:75],
  multinomial_1 = sim_data$binary_data[c(1:20, 61:80, 121:140, 181:200), 1:75]
)

# We re-code `0`s to `2`s

dat1$multinomial_1[dat1$multinomial_1 == 0] <- 2

dist <- c(
  "gaussian", "gaussian",
  "poisson", "multinomial"
)
```

### Running the model

```{r model}
fit_iClusterVB <- iClusterVB(
  mydata = dat1,
  dist = dist,
  K = 4,
  initial_method = "VarSelLCM",
  VS_method = 1,
  max_iter = 50
)
```


### Summary of the Model

```{r summary}
# We can obtain a summary using summary()
summary(fit_iClusterVB)
```


### Generic Plots

```{r plots, fig.width=6, fig.height=6}
plot(fit_iClusterVB)
```


### Probability of Inclusion Plots

```{r piplot, fig.width=6, fig.height=6}
# The `piplot` function can be used to visualize the probability of inclusion

piplot(fit_iClusterVB)
```


### Heat maps to visualize the clusters


```{r chmap, echo = TRUE, fig.width=6, fig.height=6}
# The `chmap` function can be used to display heat maps for each data view

chmap(fit_iClusterVB, rho = 0,
      cols = c("green", "blue",
               "purple", "red"),
      scale = "none")
```

