---
title: "Frequency report"

date: "`r Sys.Date()`"
output:
  html_document:
  theme: flatly
highlight: tango

params:
  robj: NULL
  freq.out_species: NULL

---
```{r setup_freq, include=FALSE}
# This chunk sets global options to hide all R code from appearing in the output

knitr::opts_chunk$set(echo = FALSE)
library(growthTrendR)
robj <- params$robj
# print(robj$freq.parms["label_data"])
```

<br>
<br>

This report presents an analysis of the distribution of samples across geographical locations, stratified by species, providing insights into sampling coverage and spatial patterns.
<br>

  source data: <b><u>`r  robj$freq.parms["freq.label_data"] `</u> </b>

  uid level: <b><u>`r robj$freq.parms["freq.uid_level"] `</u> </b>
  
  cuttoff year: <b><u>`r robj$freq.parms["freq.cutoff_year"] `</u> </b>
  
  Spatial resolution: <b><u>`r robj$freq.parms$freq.geo_resolution[1] `</u> </b> degree (longitude); <b><u>
  `r robj$freq.parms$freq.geo_resolution[2] `</u> </b> (latitude)


<br>
  <br>

```{r  results='asis', warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.align='left'}


plt.lst <- plot_freq(dt.freq = robj, out.species = params$freq.out_species)


    # Loop through the plot lists and arrange them on each page
   for (i in 1:length(plt.lst)) {

  print(plt.lst[[i]])
}

```

