---
title: "Data summary report"

date: "`r Sys.Date()`"
output:
  html_document:
  theme: flatly
highlight: tango

params:
  robj: NULL
  data_report.reports_sel: NULL
  qa.min_nseries: !r 50
  scale.N_nbs: !r 10
  scale.max_dist_km: !r 20
  qa.label_data: !r NULL
  qa.label_trt: !r NULL
  

---
```{r setup_data, include=FALSE}
# This chunk sets global options to hide all R code from appearing in the output
if (!requireNamespace("htmltools", quietly = TRUE)) {
  # stop("Package 'htmltools' is required to run this function. Please install it.")
  install.packages("htmltools")
}
knitr::opts_chunk$set(echo = FALSE)
library(data.table)
library(growthTrendR)
library(gt)
# library(htmltools)
library(stringr)
# 
 robj <- params$robj
 reports.sel <- params$data_report.reports_sel
 label_data <- params$qa.label_data
 label_trt <- params$qa.label_trt
 qa.min_nseries <- params$qa.min_nseries

 # max.dist_km <- params$parms.qa["max.dist_km"]
 N.nbs.qa <- params$scale.N_nbs
scale.max_dist_km <- params$scale.max_dist_km
 
 # for test
 
 
  if (!inherits(robj, "cfs_format")) stop("please check the input , make sure it's the result of CFS_format() function")

  tr_long <- robj$tr_all_long
  
 tr_w6 <- robj$tr_all_wide[, c("uid_site", "site_id", "uid_tree", "uid_sample",  "uid_radius", "radius_id", "rw_yend", "rw_ystart", "latitude", "longitude", "species", "dbh_cm", "ht_tot_m")]
tr_7 <- merge(tr_w6, tr_long$tr_7_ring_widths, by ="uid_radius")

spc.lst <- unique(tr_long$tr_3_trees$species)

dt.freq <- CFS_freq(tr_w6) 
dist.uids <- melt(dt.freq$dist_uids, id.vars = c("uid_label", "species" , "ord", "N", "pct.species" , "lat"),
                  variable.name = "lon",
                  value.name = "nuids")[!is.na(nuids)]
dist.uids[, lon:= as.numeric(as.character(lon))]
setorder(dist.uids, ord, lon, lat)

# label_trt <- "differentiated"

# min.nseries.qa <- 20; N.nbs.qa <- 6;
# by species
# qa<- vector("list", length(spc.lst))
# for (i.spc in seq_along(spc.lst)){
# print(i.spc)

rep_proj <- 1; rep_species <- 2; rep_site <- 3; rep_radii <- 4
```


<br>
<br>

This report provides an overview of the tree ring data's quality and characteristics at four levels:  

    1. project: Data Completeness: Assessment of missing or incomplete data of the whole data;
    2. project-species: Data Summary: Summary statistics and descriptions;
    3. project-species-site: data summary tables and series graphing; 
    4. project-species-site-radii: Correlation Analysis and quality assessment. 
  

<br>

project name: <b><u>`r tr_long$tr_1_projects$project_name`</u> </b>

selected reports: <b><u>`r sort(reports.sel)`</u> </b>

<br>


```{r qa, eval = any(reports.sel > 1), results = "hide", echo=FALSE, warning=FALSE, message=FALSE}



# qa <- lapply(1:length(spc.lst), function(i.spc, min.nseries = min.nseries.qa, N.nbs = N.nbs.qa){
if (!("RW_trt" %in% names(tr_7))) {
  setorder(tr_7, uid_radius, year)
  # tr_7[, RW_trt:= rw_mm - shift(rw_mm), by = uid_radius]
  tr_7$RW_trt <-
  ave(
    as.numeric(tr_7$rw_mm),
    tr_7$uid_radius,
    FUN = function(x)
      if (length(x) > 1L) c(NA_real_, diff(x)) else NA_real_
  )

 label_trt <- "differentiated"
  }
 qa <- lapply(1:length(spc.lst), function(i.spc){
 
  # plot-level summary
tr_6i <- tr_w6[species == spc.lst[i.spc]]
tr_7i <- tr_7[species == spc.lst[i.spc]]
rw_ref <- tr_7i


if (any(c(2,4) %in% reports.sel)) {
  rep_radii.fin <- ( TRUE & nrow(tr_6i) >= qa.min_nseries)
  # dt.radii.spc <- table_spc_site_radii(tr_6i, tr_7i, rep_radii = rep_radii.fin)
  } else rep_radii.fin <- FALSE
  # dt.radii.spc <- table_spc_site_radii(tr_6i, tr_7i, rep_radii = rep_radii.fin)
  
  
  if (rep_radii.fin == TRUE) {
  tr_7ispc<- copy(tr_7i)
  # setnames(tr_7ispc, c("uid_radius", "year", "rw_mm"), c("SampleID", "Year" ,"RawRing"))
  tr_7ispc[, c("SampleID", "Year" ,"RawRing"):= .(uid_radius, year, rw_mm)]
  # check duplication
 # if (nrow( tr_7ispc[, .N, by = .(SampleID, Year)][N>1]) > 0) stop("SampleID-Year is not unique key, please check...")
  # include RW_trt, this is the series that qa process works on,
  # tr_7ispc[, RW_trt:= RawRing - shift(RawRing), by = SampleID]; label_trt <- "differentiated"
  # if (!("RW_trt" %in% names(tr_7ispc))) stop("please assign the column RW_trt for running the qa process...")
  # acf.trt <- tr_7ispc[!is.na(RW_trt), .( ar1_trt = round(acf(RW_trt, plot = FALSE)$acf[2],2) ), by = .(SampleID)]

  # run quality assessment procedure
  qa.trt.ccf <-CFS_qa(dt.input = tr_7ispc, qa.label_data = params$qa.label_data, qa.label_trt = label_trt, qa.min_nseries = qa.min_nseries)
  dt.radii.spc <- copy(qa.trt.ccf$dt.stats)
  qa_code <- qa.trt.ccf$qa.parms$qa.code_desc
  dt.radii.spc[, uid_radius:= as.integer(SampleID)]
  dt.radii.spc <- merge(tr_6i[, c("uid_radius", "site_id", "radius_id")], dt.radii.spc, by = "uid_radius")

}else{
  # qa.trt.ccf <- list()
  qa_code <- NULL
  dt.radii.spc <- tr_7i[, .(N = .N, rw.mean = mean(rw_mm), rw.sd = sd(rw_mm), rw.min = min(rw_mm), rw.max = max(rw_mm), ymin = min(year), ymax = max(year), ar1_rw = acf(rw_mm, plot = FALSE)$acf[2] ), by = .(site_id, radius_id, uid_radius)]
  # if more than 10 series, running the correlation with mean
  if (nrow(tr_6i) > 10){
    # mean.rw <- tr_7i[, mean_rw:= mean(rw_mm), by = uid_radius]
    tmp <- merge(tr_7i, tr_7i[, .(mean_rw= mean(rw_mm)), by = year], by = "year")
    chk <- tmp[, .(ncorr_mean_rw = .N, result = cor.test(mean_rw, rw_mm)), by = .(uid_radius)]

    chk.corr <- chk[, .SD[4], by = uid_radius][, corr_mean_rw:= as.numeric(unlist(result))]

    chk.pvalue <- chk[, .SD[3], by = uid_radius][, pcorr_mean_rw:= as.numeric(unlist(result))]

    dt.radii.spc <- dt.radii.spc[chk.corr[, c("uid_radius", "corr_mean_rw", "ncorr_mean_rw")], on = "uid_radius"][chk.pvalue[, c("uid_radius", "pcorr_mean_rw")], on = "uid_radius"]
  }
}
  
  
  
  

if (rep_site %in% reports.sel) dt.site.out <- growthTrendR:::table_spc_site(dt.radii.spc, rw_ref, scale.max_dist_km, N.nbs.qa) else dt.site.out <- data.table()
if (rep_species %in% reports.sel)summary_spc <- growthTrendR:::table_spc(tr_6i,dt.radii.spc) else summary_spc <- data.table()

return(list(dt.spc_radii=dt.radii.spc, dt.spc_site = dt.site.out, dt.spc = summary_spc, qa_code = qa_code) )
})
```

```{r 1_project_tab,eval=rep_proj %in% reports.sel, results='asis'}



cat("

###  data completeness

<br>

This table presents the completeness of each variable of the whole dataset as a percentage. A value of 0 indicates no effective data. Please carefully verify that all required data has been included in the submission.

")
# library(gt)
# library(htmltools)
# if (usage != 1) return()
# Split the data into three blocks
split_dfs <- split(robj$complete_vars, cut(seq_len(nrow(robj$complete_vars)), breaks = 3, labels = FALSE))

# Generate gt tables
gt1_html <- gt(split_dfs[[1]]) %>% as_raw_html()
# %>%
# tab_header(title = "Block 1")
gt2_html <- gt(split_dfs[[2]]) %>% as_raw_html() 
# %>%
# tab_header(title = "Block 2")
gt3_html <- gt(split_dfs[[3]]) %>% as_raw_html() 


if (requireNamespace("htmltools", quietly = TRUE)) {
  # Use htmltools functions here
  grid_layout <- htmltools::tagList(
  htmltools::tags$div(style = "display: grid; grid-template-columns: repeat(3, 1fr); gap: 20px;",
           htmltools::tags$div(htmltools::HTML(gt1_html)),
           htmltools::tags$div(htmltools::HTML(gt2_html)),
           htmltools::tags$div(htmltools::HTML(gt3_html))
  )
)
} else {
  # Provide fallback or error message
  stop("Package 'htmltools' is required but not available")
}



# Output the grid layout
grid_layout



```
  
```{r 2_spc_plot, eval = rep_species %in% reports.sel, results='asis', warning=FALSE, message=FALSE, fig.width=10, fig.height=5, fig.align='left'}
cat("
### Data Summary (Species)

<br>

This section presents key summary statistics, including spatial and temporal ranges, summary of ring width measurements, series length, etc., categorized by species.

<br>

In this dataset, ", if (length(spc.lst) == 1) "there's" else "there're", 
length(spc.lst), "species: ", paste(spc.lst, collapse = ', '),  "

<br>
")

library(knitr)
library(kableExtra)
library(gt)
library(patchwork)
library(ggplot2)

for (i.spc in seq_along(spc.lst)){
# print(i.spc)
tr_6i <- tr_w6[species == spc.lst[i.spc]]
tr_7i <- tr_7[species == spc.lst[i.spc]]

# # summary table of spc
summary_spc <- qa[[i.spc]]$dt.spc

  
table_out <- summary_spc %>%
  gt() %>%
  # tab_spanner(label = "Series", columns = everything()) %>%
  cols_align(
    align = "left",  # Align all columns to the left
    columns = everything()  # Apply to all columns
  ) %>%

# %>%
  tab_options(
    column_labels.hidden = TRUE,  # Hide the header
    table.border.top.style = "none",  # Remove top border
    table.border.bottom.style = "none",  # Remove bottom border
    # table_body.border.style = "none",  # Remove all horizontal lines
    table_body.hlines.style = "none",  # Remove horizontal lines between rows
    table_body.vlines.style = "none"  # Remove vertical lines between columns
  )
table_grob <- table_out %>% gt::as_gtable()

# Convert table_grob to a ggplot object
table_plot <- ggplot() + 
  annotation_custom(table_grob, xmin = 0, xmax = 1, ymin = 0, ymax = 1) + 
  theme_void() +
  theme(plot.margin = margin(2,20, 0,0))  # Add margin around the table


p.ispc<- growthTrendR:::plot_ds(list(tr_6i, tr_7i))



p.ispc <- p.ispc + 
  theme(plot.margin = margin(50, 50, 50, 50))
# Combine the table and plot side by side using patchwork with layout adjustments
final_output <- table_plot + p.ispc + 
 plot_layout(widths = c(1,1.5)) +
  plot_annotation() &
  theme(plot.margin = margin(5, 10, 5, 10))

# Display the combined output
# final_output

# print(final_output)
print(table_plot | p.ispc  + plot_layout(widths = c(1,20),heights = c(7,5)))

}

cat(paste(" 

 *Number of series that passed the test of CFS_qa() on ", label_trt, " series 
<br> 
 
**The values refer to mean ± sd (min, max)  
 <br>
 
 "))

```


<br>

```{r 3_site_plot, eval = rep_site %in% reports.sel, results='asis', echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=5, fig.align='left'}  
cat("

###  site-level data summary

<br>

This section presents site-level data summaries, including a figure showing ring width measurements over time and a table with key statistics.

<br>

")
# plot rw by site_id
library(ggplot2)
library(ggforce)

for (i.spc in seq_along(spc.lst)){
  
  
dt.radii.spc <- qa[[i.spc]]$dt.spc_radii
# dt.radii[, year:= as.numeric(year)]
# plotting 
tr_7i <- tr_7[species == spc.lst[i.spc]]

plot_list.rw <- growthTrendR:::plot_facet(data = tr_7i, varcols = c("year", "rw_mm", "site_id"), xylabels = c("Year", "Ring Width (mm)"), nrow = 2, ncol = 2)
# plot_list.rw[[4]] <- NULL
library(patchwork)

cat(paste("\n\n"))
cat(paste("#####", spc.lst[i.spc], "on ring width measurement", "\n\n"))

 n_plots <- length(plot_list.rw)
plots_per_page <- 2

for (page in seq(1, n_plots, by = plots_per_page)) {
  # Select plots for this page (1 or 2)
  plots_this_page <- plot_list.rw[page:min(page + 1, n_plots)]
  
  # Combine and print
  print(
    wrap_plots(plots_this_page, ncol = length(plots_this_page)) +
      plot_annotation() &
      theme(plot.margin = margin(5, 5, 5, 5))
  )
}



}



```

<br>

```{r 3_site_tab, eval = rep_site %in% reports.sel, results='asis', echo=FALSE, message=FALSE, warning=FALSE, warning=FALSE}  
cat("


 This table provides a site-level summary, including the series length, raw ring width measurements, the median ring width, the median ring width of its", N.nbs.qa, " closest neighbors, and the ratio between them. This offers insights into potential outliers caused by scaling issues. 
 <br><br> 
")

for (i.spc in seq_along(spc.lst)){
  
  dt.site.out <- qa[[i.spc]]$dt.spc_site
  # if ("Nradius" %in% names(dt.site.out)) setnames(dt.site.out, "Nradius", "Nb. radiuses")
  # if ("Nsamples" %in% names(dt.site.out)) setnames(dt.site.out, "Nsamples", "Nb. samples")
if ("Nsamples" %in% names(dt.site.out)) dt.site.out$Nsamples <- NULL  
if ("Nradius" %in% names(dt.site.out)) dt.site.out$Nradius <- NULL  
  setnames(dt.site.out, c("latitude", "longitude", "Ntrees",   "summ_rw", "summ_len"),
           c("lat", "lon", "Nb. trees",   "rw(mm)**", "len.series**"))
  
  if (all(c( "rw.median", "rw.median.nbs", "ratio_median") %in% names(dt.site.out)))
  setnames(dt.site.out, c( "rw.median", "rw.median.nbs", "ratio_median"),
           c( "median rw(mm)", "neigh median rw(mm)", "ratio_rw*"))
  
  
  
  gt_table <- gt(dt.site.out[, -c("uid_site", "species")]) %>%
    tab_header(title = spc.lst[i.spc]) %>%
    tab_style(
      style = cell_text(weight = "bold"),  # Set header text to bold
      locations = cells_column_labels(everything())  # Apply style to all headers
    ) %>%
    cols_align(
      align = "center", # Center align content
      columns = everything() # Apply to all columns
    )%>%
    #  cols_width(
    #   everything()[7] ~ pct(8),         # First column: 20% width
    #   everything()[8] ~ pct(12),        # Second column: 30% width
    #   everything()[9] ~ pct(12)         # Third column: 50% width
    # )%>%
    tab_options(
      table.width = "100%",
      heading.title.font.weight = "bold",  # Make the header bold
      table.border.top.width = px(0),
      table_body.border.top.width = px(0),
      table_body.border.bottom.width = px(0)
    )
  
  # Add scrolling through CSS
  if (requireNamespace("htmltools", quietly = TRUE)) {
  scrollable_table <- htmltools::div(
    gt_table %>% as_raw_html(),
    style = "overflow-y: auto; height: 400px; border: 1px solid #ddd; padding: 10px;"
  )
  }else {
  # Provide fallback or error message
  stop("Package 'htmltools' is required but not available")
}

  # scrollable_table
  # Print the table
  print(scrollable_table, browse = FALSE)  # Render the scrollable table
  rm(gt_table, scrollable_table)
  # cat("<br><br>")
}

cat(paste("
<br>
<br>
    *ratio between median of rw of the site and median of rw of its ", N.nbs.qa, " nearest neighbors. 
<br>    
    **The values refer to mean ± sd (min, max)
 
    "))

```

<br>


```{r 4_radii_tab, eval = rep_radii %in% reports.sel, results='asis', warning=FALSE, fig.width=10, fig.height=5, fig.align='left'}  


cat(paste("

<br>
<br>

###  Correlation and quality assessment code

<br>

This table provides a summary for each series, including raw ring width measurements, autocorrelation, correlation with chronologies for both raw and ", label_trt, " data, and quality assessment code (qa_code) which was derived from the CFS_qa() function. The chronologies includes all the series with qa_code 'pass'.

<br>
    "))


# qa_code <- qa[[1]]$qa_code
if (!is.null(qa[[1]]$qa_code))
  gt(qa[[1]]$qa_code) %>%
  tab_header(
    title = "Description of qa_code"
  )%>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = vars(qa_code))
  ) %>%
  opt_css(css = "td:nth-child(1) { padding-right: 20px; }")

cat(paste("

<br>
<br>"))

for (i.spc in seq_along(spc.lst)){

dt.radii.spc <- qa[[i.spc]]$dt.spc_radii

# dt.radii.spc[, summ_rw := paste0(round(rw.mean,2), " ± ", round(rw.sd,2), " ( ", round(rw.min,2), ", ", round(rw.max,2), " )")]

if ("corr_mean_rw" %in% names(dt.radii.spc)) dt.radii.spc[!is.na(corr_mean_rw),summ_corr_rw:= paste0(round(corr_mean_rw,2), " ( ", round(pcorr_mean_rw,2), " )") ]
if ("corr_mean_trt" %in% names(dt.radii.spc)) dt.radii.spc[!is.na(corr_mean_trt),summ_corr_trt:= paste0(round(corr_mean_trt,2), " ( ", round(pcorr_mean_trt,2), " )") ]
if ("ar1_rw" %in% names(dt.radii.spc)) dt.radii.spc[, ar1_rw:=round(ar1_rw, 2)]
setorder(dt.radii.spc, uid_radius) 
if ("ar1_trt" %in% names(dt.radii.spc)) {
  dt.radii.spc.stats <- dt.radii.spc[, c("site_id", "radius_id", "ymin", "ymax", "N", "ar1_rw", "summ_corr_rw","ar1_trt",  "summ_corr_trt", "qa_code")]
setnames(dt.radii.spc.stats, c("ar1_trt", "summ_corr_trt", "qa_code"), c("trt AR1**", "trt corr_mean**&", "qa_code**%"))
} else {
  if (!("summ_corr_rw" %in% names(dt.radii.spc)))dt.radii.spc[, summ_corr_rw:=""]
  dt.radii.spc.stats <- dt.radii.spc[, c("site_id", "radius_id", "ymin", "ymax", "N", "ar1_rw", "summ_corr_rw")]
  
  }
 setnames(dt.radii.spc.stats, c("ymin", "ymax", "N", "ar1_rw","summ_corr_rw"), 
           c("year from", "year to", "len series", "raw AR1*", "raw corr_mean*&"))   

  

   
gt_table <- gt(dt.radii.spc.stats) %>%
  tab_header(title = spc.lst[i.spc]) %>%
 tab_style(
    style = cell_text(weight = "bold"),  # Set header text to bold
    locations = cells_column_labels(everything())  # Apply style to all headers
  ) %>%
  cols_align(
    align = "center", # Center align content
    columns = everything() # Apply to all columns
  )%>%
  # cols_width(
  #     everything()[1] ~ pct(7),
  #     everything()[2] ~ pct(10),
  #   everything()[3] ~ pct(5),         # First column: 20% width
  #   everything()[4] ~ pct(5),
  #   everything()[5] ~ pct(6),
  #   everything()[7] ~ pct(7),
  #   everything()[9] ~ pct(7),
  #   everything()[11] ~ pct(10)# Third column: 50% width
  # )%>%
  tab_options(
    table.width = "100%",
    heading.title.font.weight = "bold",  # Make the header bold
    table.border.top.width = px(0),
    table_body.border.top.width = px(0),
    table_body.border.bottom.width = px(0)
  )

# Add scrolling through CSS
if (requireNamespace("htmltools", quietly = TRUE)) {
scrollable_table <- htmltools::div(
  gt_table %>% as_raw_html(),
  style = "overflow-y: auto; height: 400px; border: 1px solid #ddd; padding: 10px;"
)
}else {
  # Provide fallback or error message
  stop("Package 'htmltools' is required but not available")
}


# Print the table
print(scrollable_table, browse = FALSE)  # Render the scrollable table
rm(gt_table, scrollable_table)
# cat("<br><br>")
}

cat(paste("

<br><br>
     *developed  from raw series
<br><br>
    **developed  from ",  label_trt, " series
<br><br>
     &correlation with chronologies, the value represents correlation (p-value)
<br><br>
     %qa_code is identified using the current data as reference dataset

"))

```





 

