---
title: "Manipulating STICS XML files"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Manipulating STICS XML files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
#params:
#  javastics: "/home/plecharpent/tmp/TEST_UPDATE_XML_V9_V10/JavaSTICS-1.41-stics-9.2"
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE
)
```

```{r eval=FALSE}
library(SticsRFiles)
```

```{r include=FALSE}
suppressWarnings(library(SticsRFiles))
suppressWarnings(library(tibble))

load_param_names <- function() {
  tibble::as_tibble(
    read.csv2(file = "param_names.csv", stringsAsFactors = FALSE)
    )
}
```

```{r echo=FALSE}
stics_version <- get_stics_versions_compat()$latest_version
examples_path <- get_examples_path("xml", stics_version = stics_version)
```

```{r setup, results="hide", echo=FALSE}
xml_root <- normalizePath(tempdir(), winslash = "/", mustWork = FALSE)
xml_loc_dir <- file.path(xml_root, "XML")
if (!dir.exists(xml_loc_dir)) dir.create(xml_loc_dir)

```


```{r rm_files, results='hide', echo=FALSE}
files_list <- list.files(path = xml_loc_dir, pattern = "\\.xml$",
                         full.names = TRUE)
files_list
if (length(files_list))  {
  print("Removing files")
  file.remove(files_list)
}
```

# Introduction

The purpose of this document is to provide starting information about:

* how parameters are defined/structured in STICS XML files
* how to find information about parameters among an XML files set
* how to use functions for extracting data from XML STICS input files
* how to use functions for replacing data into them

Simple reproducible use cases based on 3 essential functions **get_param_info**, **get_param_xml** and **set_param_xml** are described in this document.

For the moment, queries for selecting information among `usms`, `soils` are limited.
For example, getting soil parameters for a list of soils is possible, as getting usms informations using a usms list.

Queries inspecting elements downwards are not available yet; as for example, getting soil names/parameters using a criterion like `argi` values > 20.

Exploring files content for files sets are also possible, but not replacing parameters values.

This document has been generated using the latest STICS version **`r stics_version`**.


# Understanding what are parameters in STICS XML files

## XML files structures describing parameters

XML files are self documented files and hierarchically structured. Syntaxic rules are based on what we call nodes (i.e. nodes names, similar to html tags) and attributes (nodes attributes values).

STICS XML files are filled with different kinds of parameters description
structures according to their type/role in the model formalisms.

As we can see in the above XML file extract from a general parameter file:

* Parameters are grouped according to formalisms (in a `xml <formalisme>...</formalisme>` block)

* Parameters may be defined by single values
This kind of definition is common for all files types and the parameter name is stored in the `nom` attribute of a `param` XML node. So the first parameter name in the example is `beta` and its value is `1.40000` (inside `param` tags).

* Option parameters (with or without attached parameters, eventually with sub-option)
The corresponding parameter name for an **option** node is stored in an attribute named `nomParam` and the corresponding value is given in the attribute `choix` of the option node. For example, the value of the parameter `codesymbiose` is `2` in the example (*choix="2"*).

```{r , eval = TRUE, echo = FALSE, class.output="xml"}
par_xml <- file.path(examples_path, "param_gen.xml")
par <- readLines(par_xml)
idx_water <- grep(pattern = "formalisme", par)[9:10]
```

````{r results='asis', echo = FALSE}
cat("```xml\n")
cat(paste(par[idx_water[1]:idx_water[2]], collapse = "\n"))
cat("\n```\n")
````

* Parameters with several values (for soil layers, technical operations,...)
Several values may be defined for example relatively to soil parameters (different layers) or for crop management operations (with time repetitions).

* Example from soil files
In the above file extract for soil parameters, each set of parameters (included in `tableau` nodes) is repeated for each soil layer (layers names are set in the `nom` attributes from *layer 1* to *layer 5*). The parameters names are defined in a `nom` attribute in each `colonne` XML tag.

```{r , eval = TRUE, echo = FALSE, class.output="xml"}
sol_xml <- file.path(examples_path, "sols.xml")
sol <- readLines(sol_xml)
lines_sol <- c(sol[3], "...", sol[70:79], "...", sol[110:119], "...")
```

````{r results='asis', echo = FALSE}
cat("```xml\n")
cat(paste(lines_sol, collapse = "\n"))
cat("\n```\n")
````

* Example from tec files
In the above file extract for crop management parameters, each set of parameters is repeated for each operation (undefined number) corresponding to an `intervention` node . Inside each `intervention` node, parameters names are defined in a `nom` attribute for each `colonne` XML tag.

```{r , eval = TRUE, echo = FALSE, class.output="xml"}
tec_xml <- file.path(examples_path, "file_tec.xml")
tec <- readLines(tec_xml)[3:22]
tec <- c(tec, "...", tec[12:20], "...")
tec[23] <- gsub(pattern = "112", replacement = "220", tec[23])
```

````{r results='asis', echo = FALSE}
cat("```xml\n")
cat(paste(tec, collapse = "\n"))
cat("\n```\n")
````


## Functions for XML files manipulations

Rules have been defined to easily search and extract information from XML files in order to simplify the way of using functions dedicated to XML files manipulations.

In some cases, information relative to upward dependence are needed for extracting parameters values, but in most cases only the parameters names are mandatory in functions arguments.

# Getting XML files examples from the `SticsRFiles` library

Several XML examples files have bee included  in the package in order to use them in `reproducible`
**manipulations** and **results** described in this document. These examples files are extracted from the JavaSTICS
standard distribution (some of them have been renamed).

## Getting the XML files source directory

In xml_dir, we store the directory path of the XML files available in the SticsRFiles installation directory.

```{r set_xml_dir, echo = FALSE}
xml_dir <- get_examples_path(file_type = "xml", stics_version = stics_version)

```

```{r, eval=FALSE}
xml_dir <- get_examples_path(file_type = "xml", stics_version = stics_version)

# For linux
#> "/path/to/user/R/x86_64-pc-linux-gnu-library/3.6/SticsRFiles/
#> extdata/xml/examples/V10.2.0"

# For windows
#> "C:/Users/username/Documents/R/win-lib/3.6/SticsRFiles/
#> extdata/xml/examples/V10.2.0"

```

## Examples files

```{r}

xml_files <- list.files(path = xml_dir, pattern = ".xml$", full.names = TRUE)

# Listing only the first three files of the entire list

# For linux
#> [1] "/path/to/user/R/x86_64-pc-linux-gnu-library/3.6/SticsRFiles/extdata/xml
#> /examples/V10.2.0/file_ini.xml"
#> [2] "/path/to/user/R/x86_64-pc-linux-gnu-library/3.6/SticsRFiles/extdata/xml
#> /examples/V10.2.0/file_plt.xml"
#> [3] "/path/to/user/R/x86_64-pc-linux-gnu-library/3.6/SticsRFiles/extdata/xml
#> /examples/V10.2.0/file_sta.xml"

# For windows
#> [1] "C:/Users/username/Documents/R/win-lib/3.6/SticsRFiles/extdata/xml/
#> examples/V10.2.0/file_ini.xml"
#> [2] "C:/Users/username/Documents/R/win-lib/3.6/SticsRFiles/extdata/xml/
#> examples/V10.2.0/file_plt.xml"
#> [3] "C:/Users/username/Documents/R/win-lib/3.6/SticsRFiles/extdata/xml/
#> examples/V10.2.0/file_sta.xml"

```

``` {r, echo=FALSE, results="hide"}
files_list <- list.files(path = xml_dir, pattern = ".xml$")
legend <- c("initializations", "plant", "station", "crop management",
            "general", "general (new formalisms)", "soils", "usms")
dt <- data.frame(files = files_list, groups = legend)

```

The corresponding files types to file names is given in the above table:

``` {r, echo = FALSE}
knitr::kable(dt)
#, caption = "Correspondence between XML example files names
# and parameters groups")

```

## Copying files in a local directory

### A specific file

```{r}
# Setting a local directory path
# xml_loc_dir <- "/path/to/local/directory"

file.copy(from = file.path(xml_dir, "sols.xml"),
          to = file.path(xml_loc_dir, "sols.xml"), overwrite = TRUE)
```

### All files

```{r}

file.copy(from = xml_files, to = xml_loc_dir)
```

# Getting information about parameters

## Searching parameters names from files

The `get_param_info` function search is performed in a parameter table stored internally in the SticsRFiles library with respect to the model version (`r stics_version` as default one).

This function uses 2 optional arguments i.e.:
* `param` a name or a part of a parameter name (or a vector of) for searching in parameters names
* and/or `keyword` for searching in the parameters description information.

The information columns in the returned tibble are:

* `name`: parameter name
* `file`: file type of the parameter (keyword), see above table for correspondence
with files


```{r echo=FALSE}
knitr::kable(read.csv(file = "param_files_keywords.csv", sep = ";"), col.names = c("keyword", "xml file", "parameter kind"))
```

* `min`: minimal bound
* `max`: maximal bound
* `definition`: parameter description


Whatever the arguments list is when calling the function, the returned tibble always contains all of these columns.

Information about `bounds` may be missing according to parameters kind.

> For example, for initialization or usm parameters and/or character type parameters (files names, residues type,...), bounds dont make any sense and `NA` is used as missing values.

### Getting all STICS parameters information

Using the `get_param_info` function without any argument, or with specifying the desired `STICS` version, allows to get a table containing all input parameters.

In the above code block, an extract of the returned information table is shown:

```{r, eval = FALSE}

param_names <- get_param_info()

head(param_names)

```
```{r echo = FALSE}
param_names <- get_param_info()

head(param_names)
```


The full parameters data.frame content is presented in the above paged table.

```{r echo=FALSE}
# Displaying the returned data as a paged table
rmarkdown::paged_table(param_names)
```

### Getting information about one or more parameters

If parameters names are known and are given with the right syntax, information can be retrieved as follows

```{r, eval = FALSE}

get_param_info(param = "albedo")

get_param_info(param = c("albedo", "latitude", "humcapil"))

```
```{r echo = FALSE}
param_names[grep("albedo", param_names$name), ]

idx <- grepl("albedo", param_names$name) |
  grepl("latitude", param_names$name) |
  grepl("humcapil", param_names$name)

param_names[idx, ]

```

### Getting parameters information using partial names matching

A search with incomplete names may be done as follows

```{r}

get_param_info(param = "hum")


param_names <- get_param_info(param = c("alb", "hum"))

```

The found parameters data.frame content is presented in the above paged table.

```{r echo=FALSE}
# Displaying the returned data as a paged table
rmarkdown::paged_table(param_names)
```


## Searching parameters using keywords

The `keyword` argument (one or several possible strings) may be used to search in all textual columns as `name` or `description`.

```{r, warning = FALSE, eval = FALSE}

get_param_info(keyword = "hum")


param_names <- get_param_info(keyword = c("alb", "hum"))


```
```{r echo = FALSE}
param_names <- get_param_info(keyword = c("alb", "hum"))

```

The found parameters data.frame content is presented in the above paged table.

```{r, warning=FALSE, echo = FALSE}
# Displaying the returned data as a paged table
rmarkdown::paged_table(param_names)

```

# Getting parameters values from XML files

The `get_param_xml` function is used for extracting parameters values in XML files (for one file or a list of), optionally providing it a parameter name or a names list.

The result of the function call is a named list (with files names), and each element of the list contains a named list of the parameters values (see examples below).

For usms or soils parameter, a conditional selection may be used to filter parameter values respectively through a list of usms or soils.

Among STICS parameters files types, XML structure may be different, and they can contain either one/several occurrence(s) of parameters with scalar values (soils or usms description), or one/several occurrence(s) of parameters with vector values (along soil depth, crop management operations,...).

So parameters values extraction with the `get_param_xml` requires different input arguments combination to do so.

## Scalar parameters extraction

* For one parameter and one occurrence

```{r}
# Fixing files paths
sols <- file.path(xml_loc_dir, "sols.xml")
par_gen <- file.path(xml_loc_dir, "param_gen.xml")

# A option parameter
get_param_xml(par_gen, param = "codeactimulch")

# A simple parameter
get_param_xml(par_gen, param = "tnitopt")

# Using a conditional selection
get_param_xml(sols, param = "argi", select = "sol",
              select_value = "solcanne")

```

* For one parameter and several occurences

```{r}

# For all soils

get_param_xml(sols, param = "argi")
```

* For several parameters and several occurrences

```{r}

# For all soils
get_param_xml(sols, param = c("argi", "pH"))
```

* For several parameters and one occurrence (conditional selection)

```{r}

# For one soil
get_param_xml(sols, param = c("argi", "pH"), select = "sol",
              select_value = "solcanne")
```

## Vector parameters extraction

Vectors parameters are typically present in soils file or crop management files.

* Getting all values for given parameters and all soils

```{r}
# For all soil layers
get_param_xml(sols, param = c("epc", "infil"))
```

* Getting all values for given parameters (for a given soil, or crop management operation)

```{r}
# For all soil layers
get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              value = "solcanne")
```

```{r}
# For all irrigation supplies
tec <- file.path(xml_loc_dir, "file_tec.xml")
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"))

```

* Selecting with values index (among 5 soil layers, several irrigations supply)

```{r}
# For soil layers 1 to 3
get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              select_value = "solcanne", ids = 1:3)
```

```{r}

# For irrigation operations 1 to 5
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"), ids = 1:5)

```

## Getting all parameters values from files

Using the function without any argument produces a list of all parameters values of a file.

```{r}
tec_param_values <- get_param_xml(tec)[[1]]
#
# Displaying only a subset of the list
head(tec_param_values, n = 10)

```

A files list may be used also to get all parameters as follows in the same list

```{r}
param_values <- get_param_xml(c(tec, sols))

# Files list names
names(param_values)

# param_values extract of 5 elements from each file sub-list
head(param_values$file_tec.xml, n = 5)

head(param_values$sols.xml, n = 5)
```

# Replacing parameters values in XML files

## Defining the output file using function arguments

The **set_param_xml** function allows to set the output file with the following optional arguments as follows:

* For overwriting the existing file: use ```overwrite = TRUE```

* For specifying a new file name and directory: use  `save_as` for giving the new xml file path (including the .xml extension)

In the following examples, the original file is always overwritten. Before and after each `set_param_xml` call, the original and new written values are checked with a `get_param_xml` call.

## Scalar parameters

* For one parameter and one occurrence

```{r}


## An option parameter

# Initial value
get_param_xml(par_gen, param = "codeactimulch")

# Setting a new one
set_param_xml(par_gen, param = "codeactimulch", values = 2, overwrite = TRUE)

# Controlling the written value
get_param_xml(par_gen, param = "codeactimulch")



## A simple parameter

# Initial value
get_param_xml(par_gen, param = "tnitopt")

# Setting a new one
set_param_xml(par_gen, param = "tnitopt", values = 29.5, overwrite = TRUE)

# Controlling written value
get_param_xml(par_gen, param = "tnitopt")


## Using a conditional selection

# Initial value
get_param_xml(sols, param = "argi", select = "sol",
              select_value = "solcanne")

# Setting a new one
set_param_xml(sols, param = "argi", values = 33, select = "sol",
              select_value = "solcanne", overwrite = TRUE)

# Controlling written value
get_param_xml(sols, param = "argi", select = "sol",
              select_value = "solcanne")

```

* For one parameter and several occurences

For passing several parameters values (for one or more parameters) or single values for several parameters, the `param_value` argument must contain a list of vectors of values with the same length as `param` vector.

```{r}

## For all soils
soils_number <- length(get_soils_list(sols))

# Initial values
get_param_xml(sols, param = "argi")

# One value per occurence
set_param_xml(sols, param = "argi", values = list(1:soils_number),
              overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = "argi")

# Setting the same value for all occurences
set_param_xml(sols, param = "argi", values = 40, overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = "argi")


```

* For several parameters and several occurrences

```{r}

## For all soils
soils_number <- length(get_soils_list(sols))

# Initial values

# Setting one value per parameters occurence
set_param_xml(sols, param = list("argi", "pH"),
              values = list(1:soils_number, soils_number:1),
              overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = c("argi", "pH"))

# Setting the same value for all occurences
set_param_xml(sols, param = c("argi", "pH"), values = list(50, 8),
              overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = c("argi", "pH"))


```

* For several parameters and one occurrence (conditional selection)

```{r}

## For one soil

# Initial values
get_param_xml(sols, param = c("argi", "pH"), select = "sol",
              select_value = "solcanne")

# Setting new values
set_param_xml(sols, param = c("argi", "pH"), values = list(50, 8),
              select = "sol", select_value = "solcanne", overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = c("argi", "pH"), select = "sol",
              select_value = "solcanne")
```

## Vector parameters

* All values (for all soil layers, or crop management operations)

```{r}
## For all soil layers

# Initial values
get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              select_value = "solcanne")

# Setting new values
set_param_xml(sols, param = c("epc", "infil"), values = list(18:22, 48:52),
              select = "sol", select_value = "solcanne", overwrite = TRUE)

# Controlling written values
get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              select_value = "solcanne")
```

```{r}
## For all irrigation operations
tec <- file.path(xml_loc_dir, "file_tec.xml")

# Initial values
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"))

# Setting new values
set_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"),
              values = list(200:215, 20:35), overwrite = TRUE)

# Controlling written values
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"))

```

* Some values

In the examples below, `ids` contains the position of the parameters values in the vector, for a given soil (5 layers) name or for irrigation supplies operations amounts. In that case, used values in `ids` are contiguous, but they do not need to be.

A vector of logical values may be used instead of position values for queries or replacement.

```{r}
## For soil layers 1 to 3

# Initial values
get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              select_value = "solcanne", ids = 1:3)

# Setting new values
set_param_xml(sols, param = c("epc", "infil"), values = list(20:18, 50:48),
              select = "sol", select_value = "solcanne", overwrite = TRUE,
              ids = 1:3)

# Controlling written values

get_param_xml(sols, param = c("epc", "infil"), select = "sol",
              select_value = "solcanne", ids = 1:3)
```

```{r}

## For irrigation operations 1 to 5 (same indices for all parameters)

# Initial values
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"))

# Setting new values
set_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"),
              values = list(204:200, 24:20), overwrite = TRUE, ids = 1:5)

# Controlling written values
get_param_xml(tec, param = c("julapI_or_sum_upvt", "amount"))
```
