---
title: "Compare MPT Analysis Approaches"
author: "Marius Barth"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Compare MPT Analysis Approaches}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
  , comment = "#>"
  , warning = FALSE
  , message = FALSE
)
```


## Example 2: Smith et al. (2011)

```{r model-and-data}
# load packages:
library("MPTmultiverse")


# Make sure that working directory is correct via either:
# - setwd() / getwd()
# - Rstudio->Sessions->Set Working Directory->'To Source File Location'

# If you're running the analysis from an .rmd file, you only need to ensure that
# the .rmd, .eqn, and .csv files are all in the same directory.


# ------------------------------------------------------------------------------
# MPT model definition & Data

EQN_FILE <- "prospective_memory.eqn"
DATA_FILE <- "smith_et_al_2011.csv"


### if .csv format uses semicolons ";" (German format):
# data <- read.csv2(DATA_FILE, fileEncoding = "UTF-8-BOM")
### if .csv format uses commata "," (international format):
data <- read.csv(DATA_FILE, fileEncoding = "UTF-8-BOM")

head(data)
TreeBUGS::plotFreq(data, boxplot = FALSE, eqn = EQN_FILE)

COL_CONDITION <- "WM_EX"  # name of the variable encoding group membership

# experimental condition should be labeled meaningfully ----
unique(data[[COL_CONDITION]])

data[[COL_CONDITION]] <- factor(
  data[[COL_CONDITION]]
  , levels = 1:2
  , labels = c("low_WM", "high_WM")
)


# check input data.frame ----
head(data)
```

### Options

Every time the package `MPTmultiverse` is loaded, it automatically sets some more or less useful defaults for model estimation,
usage of multiple processor cores, number of posterior predictive samples, etc.
By calling `mpt_options()` without any parameters, you can inspect these default values.
If you want to change them, call `mpt_options` with the respective parameter specified, i.e. `mpt_options(n.iter = 1000)`.
For testing purposes, you can also specify `mpt_options("test")`, which is a shorthand for setting fast, but highly unreliable settings.
You can set options to defaults, again, by typing the shorthand `mpt_options("default")`.

```{r options, results = 'hide'}
# How to change a single option:
mpt_options(n.iter = 1e3)

# For testing purposes, you can use this shorthand to set fast, but unreliable options:
mpt_options("test")

# List all options that were set for the different analysis approaches:
mpt_options()
```

### Estimation

In the next chunk, the main computations are done.
Type `?fit_mpt` in the R console if you want to find out more about the parameters of the function.

```{r analysis, results = 'hide'}
all_supported_methods <- c(
  "asymptotic_complete"
  , "asymptotic_no"
  , "pb_no"
  , "simple"
  , "simple_pooling"
  , "trait"
  , "trait_uncorrelated"
  , "beta"
)

results <- fit_mpt(
  method = all_supported_methods
  , dataset = DATA_FILE
  , data = data
  , model = "prospective_memory.eqn"
  , condition = COL_CONDITION
)
```


### Post-processing of Results

```{r results, eval = FALSE}
# print convergence results
checks <- (check_results(results))

used_options <- getOption("MPTmultiverse")
# store results
save(
  results
  , data
  , EQN_FILE
  , DATA_FILE
  , used_options
  , checks
  , file = "results.RData"
)

# Write model checks to a file:
# write_check_results(paste0(EQN_FILE, "-", DATA_FILE), results)
```

The analysis output `results` is an object of class `multiverseMPT`, that has its own `plot()` method.
Type `?plot.multiverseMPT` to see the documentation of possible arguments to this method.

To plot group-level parameter estimates, type:

```{r}
plot(results, save = FALSE, "est")
```

To plot between-subjects comparisons, type:

```{r}
plot(results, save = FALSE, "test_between")
```

To plot overall goodness-of-fit, type:

```{r}
plot(results, save = FALSE, "gof1")
```

To plot group-wise goodness-of-fit, type:

```{r}
plot(results, save = FALSE, "gof2")
```



