---
title: "Refining Finna Metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Refining Finna Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The `refine_metadata` function is designed to clean and standardize metadata retrieved from the Finna API. This function enhances metadata usability by:

- **Validating Required Fields:** Ensures the presence of specified metadata fields, returning `NULL` if any are missing.
- **Selecting Relevant Fields:** Allows users to specify which metadata fields to retain for streamlined analysis.
- **Handling Missing Values (Optional):** If `fill_na = TRUE`, missing values (`NA`) are replaced with placeholders.
- **Logging Missing Data (Optional):** If `verbose = TRUE`, prints a summary of missing values to assist in data quality assessment.

This preprocessing step ensures metadata consistency, improving its reliability for subsequent analysis and visualization.

## Example Usage

```{r message = FALSE, warning = FALSE}
library(finna)
library(ggplot2)

# Retrieve metadata from Finna API
sibelius_data <- search_finna("sibelius")

# Refine metadata and display missing data summary
refined_data <- refine_metadata(sibelius_data)
print(refined_data)
```

## Visualizing Metadata Distribution

The `top_plot()` function can be used to visualize key metadata distributions, such as author frequency and yearly publication distribution.

### **Author Distribution Analysis**

```{r message = FALSE, warning = FALSE, fig.alt = "author_distribution"}
# Retrieve and refine metadata
sibelius_data <- search_finna("sibelius")
refined_data <- refine_metadata(sibelius_data)

# Plot top 10 authors with frequency percentages
top_plot(refined_data$Author, field = "Author", ntop = 10, show.percentage = TRUE) +
    xlab("Author") +  
    ylab("Percentage")
```

### **Yearly Publication Distribution**

```{r message = FALSE, warning = FALSE, fig.alt = "year_distribution"}
library(finna)

# Retrieve and refine metadata
sibelius_data <- search_finna("sibelius")
refined_data <- refine_metadata(sibelius_data)

# Plot publication year distribution
top_plot(refined_data$Year, field = "Year", ntop = 10, show.percentage = TRUE) +
  xlab("Publication Year") +  
  ylab("percentage distribution of Publications")
```

This vignette demonstrates how `refine_metadata()` improves metadata usability, facilitating effective analysis and visualization using the `top_plot()` function.
