---
title: "Overview of Finna Collection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of Finna Collection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


## Introduction

This section helps us to have a general overview of the general collection in Finna. For this purpose we use a separate function `fetch_finna`.
This document demonstrates the use of the `fetch_finna` function to retrieve data from the Finna API with various queries and parameters.


## Example 1: Fetching EAD Records

**EAD records** refer to archival descriptions encoded using the **Encoded Archival Description (EAD)** standard, an XML-based framework designed to describe archival materials, collections, and finding aids in a structured, machine-readable format.

### **Purpose of EAD Records**
EAD records provide detailed information about archival collections, enabling better organization, discovery, and access to historical and cultural resources stored in archives, libraries, and museums.

### **What EAD Records Contain**

An EAD record typically includes metadata such as:

- **Collection Title**: Name of the archival collection.  
- **Creator(s)**: Person(s) or organization(s) responsible for the collection.  
- **Dates**: Time period covered by the materials.  
- **Extent**: Size or physical description (e.g., number of boxes or folders).  
- **Scope and Content**: Overview of what the collection contains.  
- **Biographical/Historical Note**: Background on the creator or related historical context.  
- **Arrangement**: How the materials are organized.  
- **Access and Use Conditions**: Restrictions or guidelines for accessing the materials.  
- **Container List**: Detailed inventory of items or series in the collection.  


```{r message = FALSE, warning = FALSE}
library(finna)
result <- fetch_finna(
  query = "record_format:ead",
  limit = 0,
  facets = "building",
  lng = "fi",
  prettyPrint = TRUE
)
print(result)
```


## Example 2: Fetching EAD3 Records

```{r message = FALSE, warning = FALSE}
library(finna)
result <- fetch_finna(
  query = "record_format:ead3",
  limit = 0,
  facets = "building",
  lng = "fi",
  prettyPrint = TRUE
)
print(result)
```

More Metadata Formats in Finna can be found in the [link](https://www.kiwi.fi/display/Finna/OAI-PMH+Harvesting+Interface+for+Finna%27s+Index) here.


## Conclusion

This document provides examples of how to use the `fetch_finna` function for retrieving various types of records from the Finna API. The tibbles contain structured data for further analysis.


