---
title: "Phylogenetic beta diversity"
output: 
   rmarkdown::html_vignette:
      toc: true
vignette: >
  %\VignetteIndexEntry{beta-diversity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
fig.dim = c(6, 4)
)
```

This vignette covers topics related to phylogenetic beta diversity, including calculation of pairwise dissimilarity between sites, and use of these dissimilarity values in ordination and regionalization analyses.

To get started, let's load the `phylospatial` library, as well as `tmap` for visualization. Note that the functions covered here all require a `phylospatial` object as input; see `vignette("phylospatial-data")` for details on constructing data sets. We'll use the `moss()` example data here.

```{r setup, message=FALSE, warning=FALSE}
library(phylospatial); library(tmap); library(magrittr)
ps <- moss()
```

## Dissimilarity

This package provides a range of methods for calculating pairwise community phylogenetic distances among locations. It can calculate phylogenetic versions of any quantitative community dissimilarity metric available trough the `vegan` package, including the various predefined indices provided through `vegan::vegdist` as well as custom indices specified through `vegan::designdist`. The default metric is Bray-Curtis distance, also known as quantitative Sorensen's dissimilarity. Additional choices allow for partitioning dissimilarity in to turnover and nestedness components. 

Dissimilarity is computed using the function `ps_add_dissim()`, which adds a distance matrix to the `dissim` slot of your `phylospatial` data set. Or if you just want the matrix itself, you can use `ps_dissim()`. 

In addition to specifying the dissimilarity index to use, these functions include options for different ways to scale the phylogenetic community matrix before calculating dissimilarity. Setting `endemism = TRUE` will scale every lineage's occurrence values to sum to 1 across all sites, giving greater weight to narrowly distributed taxa. Setting `normalize = TRUE` scales every site's total occurrence value to sum to 1 across taxa, which results in a distance matrix that emphasizes proportional differences in composition rather than alpha diversity gradients.

Let's run an example using quantitative Sorensen's index, weighted by endemism. Printing the result, we can see it now contains dissimilarity data:

```{r beta, message=FALSE, warning=FALSE}
ps <- ps_add_dissim(ps, method = "sorensen", endemism = TRUE, normalize = TRUE)
ps
```



## Ordination

Having done this, we can then assess spatial turnover patterns in a couple ways: ordination or clustering. Ordination, which is implemented in the function `ps_ordinate()`, reduces the dimensionality of the phylogenetic community matrix, making it possible to visualize or analyze the dominant axes of variation. Various ordination algorithms can be selected. Let's perform a PCA, and make maps of the first four dimensions:

```{r ordinate, message=FALSE, warning=FALSE}
ps %>%
      ps_ordinate(method = "pca", k = 4) %>%
      tm_shape() +
      tm_raster(col.scale = tm_scale_continuous(values = "brewer.pi_yg"),
                col.free = FALSE) +
      tm_title("Phylogenetic community ordination")
```

We can also qualitatively visualize compositional patterns by converting ordination axes to a set of colors representing how similar two sites are to each other, using the `ps_rgb()` function. Let's do that here, using the `"cmds"` (classical multidimensional scaling) ordination algorithm, and then plot the result using `tmap::tm_rgb()`:

```{r rgb, message=FALSE, warning=FALSE}
ps %>%
      ps_rgb(method = "cmds") %>%
      tm_shape() +
      tm_rgb(col.scale = tm_scale_rgb(max_color_value = 1), 
             interpolate = FALSE) +
      tm_title("Phylogenetic community ordination")
```



## Regionalization

We can also perform a more formal cluster analysis that splits the landscape into a set of evolutionary bioregions, using the `ps_regions()` function. To do this, we need to specify the clustering `method` and the number of clusters (`k`). Choices of method include k-means and various hierarchical clustering methods; note that results are sometimes highly sensitive to which method is selected. The hierarchical methods require a dissimilarity matrix calculated by first running `ps_add_dissim()`, while k-means does not.

Choosing `k` is usually subjective. Many alternative methods have been proposed in the literature to identify the "optimal" number of clusters in a data set, but ecological data are often inherently characterized by continuous gradients rather than discrete provinces, in which case no value of `k` may clearly fit best. You can use the function `ps_regions_eval()` to help evaluate how well different choices for `k` fit the data by comparing the variance explained by different numbers of clusters. Let's do that here, using the `"average"` hierarchical clustering method:

```{r k, message=FALSE, warning=FALSE}
ps_regions_eval(ps, k = 1:15, plot = TRUE, method = "average")
```
From the evaluation plot, it looks like a value of `k = 4` stands out as the most distinct "elbow" in explained variance ("SSE"), with near-maximum distance to the 1:1 line ("dist11"), and with more negative "curvature" than its neighbors, though other choices could also be reasonable. Let's generate results for four regions, and then make a map of these zones:

```{r regions, message=FALSE, warning=FALSE}
ps %>%
      ps_regions(k = 4, method = "average") %>%
      tm_shape() +
      tm_raster(col.scale = tm_scale_categorical(values = "brewer.dark2"),
                col.legend = tm_legend(show = FALSE)) +
      tm_title("phylogenetic region")
```
