---
title: "Dealing with Timezones"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dealing with Timezones}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
#| message: false
library(ARUtools)
library(lubridate) # For working with date/times
library(dplyr) # For manipulating data
```

Timezones can cause a lot of confusion, but, unfortunately are important descriptors of a moment in time.

> **To deal with timezones in ARUtools you only need two things:**
> 
> - To make sure the timezones match between the recordings (`meta`) and your site index (`site`)
> - And that you *know* what the timezones are (or at least that they can be derived from their coordinates).

To explain this in more detail, let's talk about how ARUtools treats timezones.

In R, a date/time column can only have *one* timezone specified.
However, when working with sites around a timezone divide, it's possible that you 
may occasionally have recordings made in different timezones which you would like to process together.

To facilitate this we use the convention of 'local' times marked with "UTC". 
Here we mean 'local' to reflect the timezone that the ARU is recording in.
This means that the `date_time` column may contain a time recorded in Eastern Daylight savings, but the
'official' timezone according to R is still UTC.

If we were to try to use non-UTC times, we'll be warned off.

To illustrate this, let's create a cleaned sites index data frame with the 
`date_time`s in the America/Toronto timezone.
```{r}
s <- example_sites_clean

# Force to a non-UTC timezones
s$date_time_start <- force_tz(s$date_time_start, "America/Toronto")
s$date_time_end <- force_tz(s$date_time_end, "America/Toronto")
```

If we try to add these sites to our cleaned metadata we can see that the timezones are removed
```{r}
m <- clean_metadata(project_files = example_files)
m <- add_sites(m, s)
```

## What timezone are my data in?

If we ask R what timezone these data are in, R will say "UTC"

```{r}
tz(m$date_time)
```

But that's probably not *really* the case.

There are three possible options for what the timezones might look like:

1) All times are in the same timezone that was programmed into the ARUs before
they were set out. This would likely be either the timezone of the region to
which they were being deployed, or the timezone of the lab or home base (it
doesn't really matter which, as long as they're all the same and you know which
one it is).

2) There are several different timezones among these recordings, which correspond
to where they were deployed. This would likely happen if ARUs were set to use
GPS to get the timezone and if a study area straddled a timezone boundary. 

3) There are several different timezones among these recordings, but they do **not**
correspond to where they were deployed. This might happen if the timezones were
set on the ARUs for different projects and were not corrected before deployment. 

In ARUtools, we have options to deal withe the first two scenarios. 
However, if you find yourself in the third scenario, the best thing would be to 
split your files by timezone and run through the workflow individually with each batch.


## Calculating time to sunrise/sunset
For simplicity, we don't need to worry about the 'real' timezone except for when we calculate 
the time to sunrise/sunset. 

This is where it's important to know what timezone patterns you have in your data.

In our first scenario, we know our our recordings all have the same timezone and we know what that timezone is.

Here we can specify that timezone specifically:
```{r}
m_est <- calc_sun(m, aru_tz = "America/Toronto")
```

Alternatively, in our second scenario, we know that the timezones may be different, 
but importantly, that they correspond to the location where the unit was deployed.
Here we can use `aru_tz = "local"` and `calc_sun()` will use the recording 
coordinates to figure out what the timezone was.
```{r}
m_local <- calc_sun(m, aru_tz = "local")
```

Finally, in our final scenario, we know what the timezones are, 
but they are not all the same and they do not
correspond to the location where the unit was deployed.

In this case we'll split the data and use the specific timezones.
Let's assume that we know the timezones and that sites P06_1 and P09_1 are in 
Central, and the rest in Eastern. 

```{r}
# Split by timezone
m1 <- filter(m, site_id %in% c("P06_1", "P09_1")) # Get P06_1 and P09_1
m2 <- filter(m, !site_id %in% c("P06_1", "P09_1")) # Get all except the above

# Calculate time to sunrise/sunset individually
m1_cst <- calc_sun(m1, aru_tz = "America/Winnipeg")
m2_est <- calc_sun(m2, aru_tz = "America/Toronto")

# Join them back in
m_joint <- bind_rows(m1_cst, m2_est)
```

Because we actually use the same timezone that the sites were located in,
if you compare `m_joint` to `m_local` you'll see that with the exception of what
the timezone is called they have the same results 
("America/Detroit" is the same timezone as "America/Toronto").

You'll also note that those with the Eastern timezone (America/Toronto or America/Detroit),
all match those in `m_est`. 


## Important things to note
- The timezone of your recordings, must be the same as in your site index, or they won't
  join properly with `add_sites()`
  
- Unless you know the timezones match the deployment location, you *must* know 
  what timezone your recordings were made in. 
  
- Because `calc_sun()` returns *relative* time to sunrise and sunset, it's 
  important that you know what the timezone of the recording is in, but it's 
  **not** important that the timezone matches that of it's deployment area.


## An example

Let's assume we have two sites, one in the Eastern timezone, one in Western.
However, they are both programmed to record at 4am, 5am and 6am *Eastern*.

We'll first use some of our example data to create this mini meta data set.
```{r}
m_mini <- filter(m, site_id %in% c("P01_1", "P06_1")) |>
  select(aru_id, site_id, longitude, latitude) |>
  distinct() |>
  cross_join(data.frame(date_time = c(
    "2020-05-02 05:00:00",
    "2020-05-02 06:00:00",
    "2020-05-02 07:00:00"
  ))) |>
  mutate(
    date = as_date(date_time),
    path = paste0(aru_id, "_", site_id, "_", hour(date_time), ".csv")
  )
```

If we now calculate the time to sunrise/sunset (`t2sr` and `t2ss`) we find that
the difference between these sites is about 15min,
accounting for the fact that site P06_1 is farther west than P01_1 and 
so the recording at 6am occurs 28.8min before sunrise, 
whereas P01_1's 6am recording occurs only 14.9 min before sunrise.

```{r}
calc_sun(m_mini, aru_tz = "America/Toronto") |>
  arrange(date_time)
```

However, if we were to incorrectly assume that the ARU unit located in the central timezone
was recording in that timezone, we would get very different results.

```{r}
calc_sun(m_mini, aru_tz = "local") |>
  arrange(date_time)
```

Here all the times to sunrise/sunset for site P06_1 are offset by an hour, because we're assuming the
wrong timezone (which is an hour different from the correct one). 


**Therefore the take home is that you only need two things:**

- To make sure the timezones match between the recordings and your site index
- And that you *know* what the timezones are (or at least that they can be derived from their coordinates).
