---
title: "theme_duke()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{theme_duke()}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.asp = 0.618
)
```

This vignette is intended to demonstrate the breadth of `theme_duke()` and its various applications. 
Similar to the other functions within the `theme()`, this function can be applied to all visualizations made with ggplot2. 

# Plot Examples

For these visualizations, we will use the `penguins` dataset from the **palmerpenguins** package.

```{r penguins, warning = FALSE}
library(palmerpenguins)
library(ggplot2)
library(duke)
```

## Scatter Plot

```{r scatter, warning = FALSE}
p <- ggplot(penguins, aes(bill_length_mm, flipper_length_mm)) +
  geom_point(aes(colour = species)) +
  labs(
    title = "Bill Length vs. Flipper Length", 
    caption = "There are three different species of penguins.", 
    x = "Bill Length (mm)", 
    y = "Flipper Length (mm)"
  )

p +
  theme_duke()
```


## Bar Plot

```{r bar, warning = FALSE}
p <- ggplot(penguins, aes(species)) +
  geom_bar(aes(fill = species)) +
  labs(
    title = "Species Count", 
    caption = "There are three different species of penguins.", 
    x = "Species", 
    y = "Count"
  )

p +
  theme_duke()
```

## Histogram

```{r hist, warning = FALSE}
p <- ggplot(penguins, aes(body_mass_g)) +
  geom_histogram(aes(fill = species)) +
  labs(
    title = "Distribution of Penguin Body Mass", 
    caption = "There are three different species of penguins.", 
    x = "Body Mass (g)", 
    y = "Count"
  )

p +
  theme_duke()
```

## Box Plot

```{r box}
p <- ggplot(penguins, aes(sex, body_mass_g)) +
  geom_boxplot() +
  labs(
    title = "Comparison of Body Mass By Sex", 
    x = "Sex", 
    y = "Body Mass (g)"
  )

p +
  theme_duke()
```

## Density Plot

```{r density, warning = FALSE}
p <- ggplot(penguins, aes(bill_depth_mm)) +
  geom_density() +
  labs(
    title = "Density of Penguin Bill Depth", 
    x = "Bill Depth (mm)", 
    y = "Densiy"
  )

p +
  theme_duke()
```


## Jitter Plot

```{r jitter, warning = FALSE}
p <- ggplot(penguins, aes(year, body_mass_g)) +
  geom_jitter() +
  labs(
    title = "Comparison of Body Mass By Year", 
    x = "Year", 
    y = "Body Mass (g)"
  )

p +
  theme_duke()
```
