---
title: "User Stories Report"
output: html_document
params:
  storiesdf: stories_related[[2]]
---

```{r setup, include=FALSE}
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)


knitr::opts_chunk$set(echo = TRUE)

storiesdf <- params$stories
row.names(storiesdf) <- NULL

keywords <- readKeywords()

```


The following report analises structural data from the user stories, and determines basic weaknesses.


<br />


## Objectives Stories
The following stories have been identified as objectives.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Obtain the data
sub_obj <- subset(storiesdf, type == "objective")

# Calculate values for vague objectives
warning_vagueObj <- sapply(sub_obj$goal, function(x) any(is.na(x) | x == '' )) %>% as.logical()


# Get data to calculate conflicts
directions <- unique(sub_obj$goal)
directions <- directions[!is.na(directions)]
#objDirs <- unique(readRDS("objKeywords.rds")$direction)
objDirs <- objKeywords$direction
objDirs <- objDirs[!is.na(objDirs)]
# Get the conflict
warning_confObj <- all( objDirs %in% directions )
    
```


`r if(!warning_confObj && !warning_vagueObj) {"<!--"}`

<div style="color: red;">
Warnings:

* Conflicting optimisation directions: `r warning_confObj`
* Vague objectives: `r warning_vagueObj`
</div>

`r if(!warning_confObj && !warning_vagueObj) {"-->"}`



```{r, echo=FALSE, message=FALSE, warning=FALSE}

# Keep a subset
sub_obj_show <- sub_obj[,c("storyId", "who", "what", "why", "goal")]
rownames(sub_obj_show) <- NULL
color_rows <- which(is.na(sub_obj_show$goal))


# Show it
sub_obj_show %>%
  kable(row.names = NA) %>% kable_styling("striped") %>%
  row_spec(color_rows, bold = T, color = "white", background = "#D7261E")

```





<br />


## Input Data Stories

The following stories have been identified as input data.

```{r echo=FALSE}
sub_input <- subset(storiesdf, type == "input")
warning_inputLocation <- sapply(sub_input$location, function(x) any(is.na(x) | x == '' )) %>% any() %>% as.logical()
```



`r if(!warning_inputLocation) {"<!--"}`
<div style="color: red;">
Warnings:

* Missing input data locations: `r warning_inputLocation`
</div>
`r if(!warning_inputLocation) {"-->"}`

<br />


```{r echo=FALSE}

# Keep a subset
sub_input_show <- sub_input[,c("storyId", "who", "what", "why", "location")]
rownames(sub_input_show) <- NULL
color_rows <- which(is.na(sub_input_show$location))


# Show it
sub_input_show %>%
  kable(row.names = NA) %>% kable_styling("striped") %>%
  row_spec(color_rows, bold = T, color = "white", background = "#D7261E")
```




<br />




## Output Data Stories

The following stories have been identified as output data.

```{r echo=FALSE}
# Get the data
sub_output <- subset(storiesdf, type == "output")

# Calculate warnings
warning_outputLocation <- subset(sub_output, goal == "write")$location %>%
  sapply(function(x) any(is.na(x) | x == '' )) %>% any() %>% as.logical()
warning_outputVisualization <- subset(sub_output, goal == "visualise")$location %>%
  sapply(function(x) any(is.na(x) | x == '' )) %>% any() %>% as.logical()
```

`r if(!warning_outputLocation && !warning_outputVisualization) {"<!--"}`
<div style="color: red;">
Warnings:

* Missing location for writtings: `r warning_outputLocation`
* Missing visualisation types: `r warning_outputVisualization`
</div>
`r if(!warning_outputLocation && !warning_outputVisualization) {"-->"}`


```{r echo=FALSE}
sub_output_show <- sub_output[,c("storyId", "who", "what", "why", "goal", "location")]
rownames(sub_output_show) <- NULL
color_rows <- which(is.na(sub_output_show$location))


# Show it
sub_output_show %>%
  kable(row.names = NA) %>% kable_styling("striped") %>%
  row_spec(color_rows, bold = T, color = "white", background = "#D7261E")
```





<br />


## Constraints

The following stories have been identified as constraints:

```{r echo=FALSE}
# Get the data
sub_const <- subset(storiesdf, type == "constraint")
row.names(sub_const) <- NULL

# Show it
kable(sub_const[,c("storyId", "who", "what", "why")]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```



<br />


## Scenarios

The following stories have been identified as scenarios:

```{r echo=FALSE}
# Get the data
sub_scenario <- subset(storiesdf, type == "scenario")
row.names(sub_scenario) <- NULL

# Show it
kable(sub_scenario[,c("storyId", "who", "what", "why")]) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```



<br />
 
## Semantic Groups
 
The stories (excluding objectives) have been grouped semantically, according to the similarity of words used. The following groups have been detected:
 
 
```{r echo=FALSE}

grouped_stories <- storiesdf[,c("group", "storyId", "who", "what", "why", "type", "belonging")]
grouped_stories <- grouped_stories[order(grouped_stories$group), ]
row.names(grouped_stories) <- NULL



kable(grouped_stories, booktabs = T) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1, latex_hline = "major", valign = "middle") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```




## Proportion of Stories

The following is the proportion of types of stories by type of user, found in this set:

```{r echo=FALSE}
ggplot(storiesdf, aes(x=type, fill = who)) + geom_bar(stat = "count")


```
