---
title: "LMoFit | Advanced L-Moment Fitting of Distributions"
author: "Mohanad Zaghloul, Simon Michael Papalexiou, Amin Elshorbagy"
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    number_sections: yes
  rmarkdown::pdf_document:
    toc: yes
    toc_depth: 2
    number_sections: yes
    extra_dependencies: ["float"]
vignette: >
  %\VignetteIndexEntry{LMoFit | Advanced L-Moment Fitting of Distributions} 
  %\VignetteEncoding{UTF-8} 
  %\VignetteEngine{knitr::rmarkdown}
header-includes: \usepackage{amsmath}
editor_options:
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  fig.height = 3*0.65,
  fig.width = 4,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
library(LMoFit)
```

***
Citation to this package:

"*Zaghloul M, Papalexiou S, Elshorbagy A (2020). _LMoFit: Advanced L-Moment Fitting of
Distributions_. R package version 0.1.6.*"

***
# Introduction {#section_1}

In the practice of frequency analysis, some probability distributions are abandoned due to the absence of a convenient way of handling them. Burr Type-III (`BrIII`), Burr Type-XII (`BrXII`), and Generalized Gamma (`GG`) distributions are examples of such abandoned distributions. Most past studies involving frequency analysis ignored the unconventional distributions regardless of their remarkable advantages. For example, [Zaghloul et al. (2020)](https://doi.org/10.1016/j.advwatres.2020.103720) reveals the importance of using `BrIII` and `BrXII` distributions instead of the commonly used Generalized Extreme Value (`gev`) distribution for flood frequency analysis across Canada. Also, [Papalexiou and Koutsoyiannis (2016)](https://doi.org/10.1016/j.advwatres.2016.05.005) recommended the use of `BrXII` and `GG` distributions for describing daily precipitation across the globe.

Various conventional and unconventional distribution are covered by `LMoFit`, which is the first package that facilitated the use of some unconventional distributions that are:

(1) Burr Type-III (`BrIII`)
(2) Burr Type-XII (`BrXII`)
(3) Generalized Gamma (`GG`)

While the commonly used conventional distributions are masked from the '`lmom`' package by [Hosking, J.R.M. (2019)](https://CRAN.R-project.org/package=lmom). These conventional distributions are:

(4) Normal (`nor`)
(5) Log-Normal (`ln3`)
(6) Pearson Type-3 (`pe3`)
(7) Generalized Normal (`gno`)
(8) Generalized Pareto (`gpa`) 
(9) Generalized Logistic (`glo`)
(10) Gamma (`gam`)
(11) Generalized Extreme Value (`gev`) with different parameterization

For each of the above-mentioned distributions, `LMoFit` provides functions that:

* Fit the distributions, i.e. parameter estimation.
* Estimate cumulative probabilities of non-exceedance of specific quantiles.
* Estimate probability densities of specific quantiles.
* Estimate quantiles of specific return periods or specific cumulative probabilities of non-exceedance.
* Estimate return periods of specific quantiles.

***
# Overview of functions {#section_2}

## Estimation of sample L-moments and L-moment ratios

`LMoFit` follows the method of [Hosking, J.R.M. (1990)](https://doi.org/10.1111/j.2517-6161.1990.tb01775.x) in estimating sample L-moments. The function is referred to as `get_sample_lmom()` and can be implemented as follows. For an embedded sample `FLOW_AMAX` of annual maximum streamflow in cfs observed at a gauge in BC, Vancouver, Canada @ Lat: 51°14'36.8¨N, Long:116°54'46.6¨W.

```{r}
FLOW_AMAX
```

the sample L-moments and L-moment ratios are estimated as

```{r}
sample_lmoments <- get_sample_lmom(x = FLOW_AMAX)
knitr::kable(sample_lmoments, digits = 2, caption = "Sample L-moments and L-moment ratios")
```

where

- sl1 is the 1st sample L-moment,
- sl2 is the 2nd sample L-moment,
- sl3 is the 3rd sample L-moment,
- sl4 is the 4th sample L-moment,
- st2 is the 2nd sample L-moment ratio = sl2/sl1,
- st3 is the 3rd sample L-moment ratio = sl3/sl2,
- st4 is the 4th sample L-moment ratio = sl4/sl2.

## Fitting (parameter estimation) functions

Fitting functions are called by adding a prefix `fit_` before a distribution name. e.x. `fit_BrIII`, `fit_BrXII`, `fit_GG`, `fit_gev`, etc. An example of estimating the fitted parameters of some randomly generated sample can be implemented as

Assume that the sample L-moments are:

* 1st L-moment (sl1) = 436
* 2nd L-moment ratio (st2) = 0.144
* 3rd L-moment ratio (st3) = 0.103

```{r}
# Fitting of BrIII distribution
parameters <- fit_BrIII(sl1 = 436, st2 = 0.144, st3 = 0.103)
scale <- parameters$scale
shape1 <- parameters$shape1
shape2 <- parameters$shape2
```

This function results in the fitted parameters that are scale, shape1, and shape2 parameters for BrIII distribution.

```{r}
knitr::kable(parameters[1:3], digits = 2, caption = "Estimated parameters of BrIII distribution")
```

## Cumulative probability distribution functions

A cumulative probability distribution function (CDF) estimates the probability of non-exceedance. CDF functions are called in `LMoFit` by adding a prefix 'p' before a distribution name. e.x. `pBrIII`, `pBrXII`, `pGG`, `pgev`, etc. An example of implementing such functions is provided for the `BrXII` distribution as follows.

Assume that the fitting of `BrXII` to a sample results in the following parameters:

* scale parameter = 322
* shape1 parameter = 6.22
* shape2 parameter = 0.12

The probability of non-exceedance of 500 is estimated as

```{r}
scale <- 322; shape1 <- 6.22; shape2 <- 0.12
probability <- pBrXII(x = 500, para = c(scale, shape1, shape2))
probability
```

This function can also be implemented to a vector of quantities of interest such as

```{r}
probability <- pBrXII(x = c(400, 450, 500, 550, 600, 1000), para = c(scale, shape1, shape2))
probability
```

## Probability density functions 

The probability density functions (PDF) are called in `LMoFit` by adding a prefix 'd' before a distribution name. e.x. `dBrIII`, `dBrXII`, `dGG`, `dgev`, etc. An example of implementing such functions is provided for the `gev` distribution as follows.

Assume that the fitting of `gev` to a sample results in the following parameters:

* location parameter = 388
* scale parameter = 99
* shape parameter = -0.11

The probability density of a value of 200 is estimated as

```{r}
location <- 388; scale <- 99; shape <- -0.11
density <- dgev(x = 200, para = c(location, scale, shape))
density
```

This function can also be implemented to a vector of quantities of interest such as

```{r}
density <- dgev(x = c(100, 150, 200, 250, 300), para = c(location, scale, shape))
density
```

## Quantile distribution functions

The quantile functions in `LMoFit` are called by adding a prefix 'q' before a distribution name. e.x. `qBrIII`, `qBrXII`, `qGG`, `qgev`, etc. This is one of the most useful and handy functions in `LMoFit` because it enables the user to estimate quantiles corresponding to specific probabilities or directly corresponding to specific return periods.

Assume the estimated BrIII parameters are

```{r}
scale <- 565.29; shape1 <- 5.75; shape2 <- 0.13
```

then the quantile that corresponds to a non-exceedance probability of 0.99 is

```{r}
qua <- qBrIII(u = 0.99, para = c(scale, shape1, shape2))
qua
```

and the quantile that corresponds to a return period of 100 years is

```{r}
qua <- qBrIII(RP = 100, para = c(scale, shape1, shape2))
qua
```

The quantile functions can also be implemented for multiple probabilities or return periods as

```{r}
qua <- qBrIII(RP = c(5, 10, 25, 50, 100, 200), para = c(scale, shape1, shape2))
qua
```

## Return period functions

A return period function is the same as the CDF but after transforming probabilities to their corresponding return periods. This transformation is a simple process that was previously neglected and left for the user. However, `LMoFit` is developed to be handy and user friendly. So, the return period functions are included in `LMoFit` and are called by adding a prefix 't' before a distribution name. e.x. `tBrIII`, `tBrXII`, `tGG`, `tgev`, etc. An example of implementing such functions is provided for the `BrXII` distribution as follows.

Assume that the fitting of `BrXII` to a sample results in the following parameters:

* scale parameter = 322
* shape1 parameter = 6.22
* shape2 parameter = 0.12

The return period of a quantity equal to 800 is estimated as

```{r}
scale <- 322; shape1 <- 6.22; shape2 <- 0.12
return_period <- tBrXII(x = 800, para = c(scale, shape1, shape2))
return_period
```

This function can also be implemented to a vector of quantities of interest such as

```{r}
return_period <- tBrXII(x = c(500, 600, 700, 800), para = c(scale, shape1, shape2))
return_period
```

## Theoretical L-space of BrIII distribution

Distributions with two shape parameters such as the BrIII distribution are presented on the L-moment ratio diagram as a theoretical L-space. `LMoFit` developed the theoretical L-space of the BrIII distribution and embedded its results as a ggplot in the package data. Users can call the L-space plot of the BrIII as `lspace_BrIII` and apply any desired adjustments to it as for regular ggplots.

```{r, fig.cap = "Theoretical L-space of `BrIII` Distribution"}
lspace_BrIII
```

## Theoretical L-space of BrXII distribution

The BrXII Distributions also has two shape parameters and can be presented on the L-moment ratio diagram as a theoretical L-space. `LMoFit` developed the theoretical L-space of the BrXII distribution and embedded its results as a ggplot in the package data. Users can call the L-space plot of the BrXII as `lspace_BrXII` and apply any desired adjustments to it as for regular ggplots.

```{r, fig.cap = "Theoretical L-space of `BrXII` Distribution"}
lspace_BrXII
```

## Theoretical L-space of GG distribution

The GG Distributions is another case of distributions having two shape parameters so it can also be presented on the L-moment ratio diagram with a theoretical L-space. `LMoFit` developed the theoretical L-space of the GG distribution and embedded its results as a ggplot in the package data. Users can call the L-space plot of the GG as `lspace_GG` and apply any desired adjustments to it as for regular ggplots.

```{r, fig.cap = "Theoretical L-space of `GG` Distribution"}
lspace_GG 
```

## Goodness of fitting on L-moment ratio diagrams

There are numerous criteria for comparing the goodness of fitting of probability distributions. A very important preliminary test should be a visual inspection of the L-moment ratio diagrams. The sample, that the distributions are fitted to, is presented on the L-moment ratio diagram as an L-point. If this L-point lies outside the L-space of any two-shape parametric distribution, then the distribution is not valid to describe the sample.

For a single sample such as the sample provided as `FLOW_AMAX`, the test can be implemented by calling the function `com_sam_lspace()` accounting for 'compare a sample with an L-space'. This function is valid for the two-shape parametric distributions that are `BrIII`, `BrXII`, and `GG`. The L-point of the sample is presented by a red point mark on the diagrams.

```{r, fig.cap = "Example of single L-points on the L-space of `BrIII` Distribution"}
com_sam_lspace(sample = FLOW_AMAX, type = "s", Dist = "BrIII")
```

```{r, fig.cap = "Example of single L-points on the L-space of `BrXII` Distribution"}
com_sam_lspace(sample = FLOW_AMAX, type = "s", Dist = "BrXII")
```

```{r, fig.cap = "Example of single L-points on the L-space of `GG` Distribution"}
com_sam_lspace(sample = FLOW_AMAX, type = "s", Dist = "GG")
```

For multiple samples, such as streamflow observed at various sites, the sample of each site is presented by one L-point on the L-moment ratio diagram. Therefore, by implementing the visual test to multiple samples we can identify distributions that have their L-spaces overlaying the greatest portion of L-points and therefore we can select the best regionally valid distribution. `com_sam_lspace` is developed to be flexible and user friendly. The user can use the same function with multiple sites by changing the `type` condition from `type = "s"` to `type = "m"`. Ten hypothetical samples are developed at 10 hypothetical sites and embedded in the package data as `FLOW_AMAX_MULT`. `FLOW_AMAX_MULT` is a dataframe with 10 columns where each column describes the sample at one site. 

```{r}
colnames(FLOW_AMAX_MULT) <- paste0("site.", 1:10)
knitr::kable(head(FLOW_AMAX_MULT), caption = "The first few observations of streamflow at 10 sites")
```

An example of a regional visual test is implemented as

```{r, fig.cap = "Example of multiple L-points on the L-space of `BrIII` Distribution"}
com_sam_lspace(sample = FLOW_AMAX_MULT, type = "m", Dist = "BrIII", shape = 20)
```

```{r , fig.cap = "Example of multiple L-points on the L-space of `BrXII` Distribution"}
com_sam_lspace(sample = FLOW_AMAX_MULT, type = "m", Dist = "BrXII", shape = 20)
```

```{r, fig.cap = "Example of multiple L-points on the L-space of `GG` Distribution"}
com_sam_lspace(sample = FLOW_AMAX_MULT, type = "m", Dist = "GG", shape = 20)
```

These diagrams are still in ggplot format and can be further adjusted by the user. 

In the last example, the visual test with multiple samples `FLOW_AMAX_MULT` reveals that the L-space of `BrIII` combines the greatest amount of L-points and therefore `BrIII` should be a better candidate compared to `BrXII` and `GG` distributions.

## Condition of sample L-points as inside/outside of L-spaces

As the number of multiple points increases, the corresponding number of L-points increases, and it gets harder to make a visual judgment. For that reason, `LMoFit` provides an extra function to test the condition of each L-point of the multiple samples and identify it as inside or outside the L-space of interest. All L-points that are overlaid by the L-space are assigned a flag '`lpoint_inside_lspace`', or '`lpoint_outside_lspace`' otherwise.

```{r}
flags_BrIII <- con_sam_lspace(sample = FLOW_AMAX_MULT, type = "m", Dist = "BrIII")
knitr::kable(head(flags_BrIII), caption = "Flags obtained for BrIII's L-space")
flags_GG <- con_sam_lspace(sample = FLOW_AMAX_MULT, type = "m", Dist = "GG")
knitr::kable(head(flags_GG), caption = "Flags obtained for GG's L-space")
```

By counting the number of times the flag was `lpoint_inside_lspace` we can exactly identify the number of L-points inside each L-space as

```{r}
counter_BrIII <- nrow(flags_BrIII[flags_BrIII$condition == "lpoint_inside_lspace",])
paste0("the number of L-points inside the L-space of BrIII = ", counter_BrIII)
counter_GG <- nrow(flags_GG[flags_GG$condition == "lpoint_inside_lspace",])
paste0("the number of L-points inside the L-space of GG = ", counter_GG)
```

Accordingly, we should use `BrIII` rather than `GG` as a regional distribution in the latter example.

Note: `con_samlmom_lspace()` is similar to `con_sam_lspace()`, but it needs the sample L-moments rather than the sample itself and is only applicable to single samples. The sample L-moments are estimated by using `get_sample_lmom()`.

***
# Step by step guide to frequency analysis {#section_3}

Frequency analysis can be implemented by `LMoFit` in three steps:

 1. Estimate sample L-moments using '`get_sample_lmom()`'.
 2. Fit a specific distribution, for example, `BrIII`, to the estimated sample L-moments, using '`fit_BrIII()`'.
 3. Estimate probabilities "`pBrIII()`", quantiles "`qBrIII()`", densities "`dBrIII()`", or even return periods "`tBrIII()`".

The three steps are further explained by an example of fitting `BrIII` to the embedded sample `FLOW_AMAX`. Since `BrIII` distribution is to be used, we will check its validity to describe the sample.

```{r}
con_sam_lspace(sample = FLOW_AMAX, type = "s", Dist = "BrIII")
```

`BrIII` passes the visual test since the L-point of the sample is located inside the L-space of `BrIII` distribution. `com_sam_lspace()` can also be used for the same purpose.

Next, the sample L-moments can be estimated as follows.

```{r}
# Step 1
sample <- FLOW_AMAX
samlmoms <- get_sample_lmom(x = sample)
```

After that, the desired distribution can be fitted to the sample L-moments to determine its fitted parameters.

```{r}
# Step 2
parameters <- fit_BrIII(sl1 = samlmoms$sl1, st2 = samlmoms$st2, st3 = samlmoms$st3)
```

Once the fitted parameters are obtained, frequency analysis can be conducted in different forms to estimate probabilities, quantiles, densities, and return periods.

```{r}
# Step 3
quantile <- qBrIII(RP = c(5, 10, 25, 50, 100),
                    para = c(parameters$scale, parameters$shape1, parameters$shape2))
prob <- pBrIII(x = quantile, 
               para = c(parameters$scale, parameters$shape1, parameters$shape2))
dens <- dBrIII(x = quantile, 
               para = c(parameters$scale, parameters$shape1, parameters$shape2))
T_yrs <- tBrIII(x = quantile, 
                para = c(parameters$scale, parameters$shape1, parameters$shape2))
```

The results of this example are concluded in the table below.

```{r}
output <- cbind(Q = round(quantile, digits = 0), 
                CDF = round(prob, digits = 4),
                PDF = round(dens, digits = 5),
                T_yrs)
knitr::kable(output, caption = "Example of fitting BrIII distribution to FLOW_AMAX")
```

***
# Acknowledgement {#section_4}

This research is financially supported by the Integrated Modeling Program for Canada (IMPC) project under the umbrella of the Global Water Futures (GWF) program at the University of Saskatchewan, Canada. S.M.P. acknowledges the support of the Natural Sciences and Engineering Research Council of Canada (NSERC Discovery Grant: RGPIN ‐2019 ‐06894 ). In addition, M.Z. acknowledges the Department of Civil, Geological, and Environmental Engineering for the financial support of research through the Departmental Devolved Scholarship.

***
# References {#section_5}

* Hosking, J.R.M. (2019). L-Moments. R package, version 2.8. [link](https://CRAN.R-project.org/package=lmom)
* Hosking, J.R.M. (1990). L-Moments: Analysis and Estimation of Distributions using Linear Combinations of Order Statistics. Journal of the Royal Statistical Society. Series B 52, 105–124. [link](https://doi.org/10.1111/j.2517-6161.1990.tb01775.x)
*	Papalexiou, S.M., Koutsoyiannis, D. (2016). A global survey on the seasonal variation of the marginal distribution of daily precipitation. Advances in Water Resources 94, 131–145. [link](https://doi.org/10.1016/j.advwatres.2016.05.005)
*	Zaghloul, M., Papalexiou, S.M., Elshorbagy, A., Coulibaly, P. (2020). Revisiting flood peak distributions: A pan-Canadian investigation. Advances in Water Resources 145, 103720. [link](https://doi.org/10.1016/j.advwatres.2020.103720)
