---
title: "Optimising the LOESS method used for automatic identification of the outlier thresholds."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimising the LOESS method used for automatic identification of the outlier thresholds.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
cran <- identical(tolower(Sys.getenv("NOT_CRAN")), "false")
if (cran || !curl::has_internet()) {
  knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
} else {
  knitr::opts_chunk$set(eval = TRUE, collapse = TRUE, comment = "#>")
}

```

```{r librarydata}
library(specleanr)
```

## Identifying the optimal threshold for identifying absolute outliers using the local regression method (LOESS).

In the *speclceanr* package, besides using the naive methods and data classification, we incorporated the local regression method (LOESS) in setting the optimal threshold to identify the absolute outliers. LOESS is a non-parametric scatterplot smoothing method that uses point-wise linear regression to smooth scatter plots *(Cleveland 1979; Cleveland and Devlin 1988)*. Because it is a non-parametric method, the relationship between the independent and dependent variables should not be set beforehand (Jacoby 2000). It can reveal complex data patterns compared to traditional statistical methods *(Jacoby 2000)*.

LOESS is less computationally intensive for small datasets, easy to compute, and highly resistant to outliers *(Cleveland 1979)*. Therefore, we use it to predict the optimal threshold after modeling the relationship between the data points retained at each threshold. The thresholds for outlier detection range from 0 to 1 (0 means not an outlier or the record has not been flagged in any of the outlier detection methods used, and 1 represents a perfect outlier where the record has been flagged in all the methods). Therefore, at each threshold, the data points are retained or flagged out. Then, at the optimal threshold (global or local maximum), the data points retained are asymptotic, meaning no more records are flagged out. The optimal threshold is then used to retain the quality-controlled dataset used in further analysis.

**NOTE**

* If errors in the workflow is associated with FishBase, rerun after 2 to 5 minutes.

## Examples using fish species 

### Data processing: species and environmental variables

```{r dataprocessing, fig.width = 6, fig.height= 4, fig.align='center' }

data(efidata) 

data(jdsdata) 

danube <- sf::st_read(system.file('extdata', "danube.shp.zip",
                                  package = 'specleanr'), quiet=TRUE)


df_online <- getdata(data = c("Squalius cephalus", 'Salmo trutta',"Thymallus thymallus"),
                     extent = danube,
                     gbiflim = 50,
                     inatlim = 50,
                     vertlim = 50,
                     verbose = FALSE)


mergealldfs <- match_datasets(datasets = list(efi= efidata, jds = jdsdata,
                                              onlinedata = df_online),
                              country = c('JDS4_sampling_ID'),
                              lats = 'lat', lons = 'lon',
                              species = c('speciesname', 'scientificName'))
#Cleaning data

cleannames_df <- check_names(data = mergealldfs, colsp = 'species', pct = 90,
                             merge = TRUE, verbose = FALSE)

spfilter <- cleannames_df[cleannames_df$speciescheck %in%
                                   c("Squalius cephalus", 'Salmo trutta',
                                     "Thymallus thymallus","Anguilla anguilla", 
                                     'Barbatula barbatula'),]

worldclim <- terra::rast(system.file('extdata/worldclim.tiff', package = 'specleanr'))

#Get basin shapefile to delineate the study region: optional

danube <- sf::st_read(system.file('extdata', 'danube.shp.zip',
                                  package = 'specleanr'), quiet=TRUE)
```

### Oultier detection, threshold optimisation using LOESS and plotting

```{r outlierdetoptplot, fig.width = 6, fig.height= 4.5, fig.align='center', dpi=120 }
parm <- par(mfrow = c(2, 2),
    mar = c(3,3, 1.5, 0.5),
    oma = c(0, 0, 0, 0),
    mgp = c(1.7, 0.8, 0)
)

spp <- unique(spfilter$speciescheck)

pltout <- lapply(spp, function(s){
  
  spout <- spfilter[spfilter[,'speciescheck'] %in%s,]
  
  refdata <-  pred_extract(data= spout, raster= worldclim,
                           lat = 'decimalLatitude',
                           lon = 'decimalLongitude',
                           colsp = 'speciescheck',
                           bbox  = danube,
                           list= TRUE,
                           minpts = 10)
  
  outdet <- multidetect(data = refdata, multiple = FALSE,
                        var = 'bio6', output = 'outlier',
                        exclude = c('x','y'),
                        methods = c('zscore', 'adjbox',
                                    'logboxplot', 'distboxplot',
                                    'iqr', 'semiqr','seqfences',
                                    'hampel','kmeans',
                                    'jknife', 'onesvm',
                                    'iforest'), 
                        warn = FALSE)
  print(nrow(refdata))
  
  opt <- optimal_threshold(refdata = refdata, outliers = outdet, 
                             plotsetting = list(plot = TRUE, group = s))
  opt
  
})
par(parm)
```

### Summary explanation

To ensure the **LOESS** model was only fitted if a group or species had absolute outliers, we set a cutoff of **0.6**. So, LOESS will not be fitted if outliers are not detected in less than half of the outlier detection methods. For example, in the figure above, the threshold was only optimized for *Salmo trutta*, *Anguilla anguilla*,  and *Thymallus thymallus*. In *Barbatula barbatula* and *Squalius cephalus* the outliers with the highest weight was 0.41 and 0.5 respectively.

### Simulating for small datasets less than 25 records

```{r simulated data }
set.seed(113554333)
a <- rnorm(30, 32, 1)
b <- rnorm(30, 4, 1)
c <- rnorm(30, 0, 1)
d <- rnorm(30, 6, 1)
#add outlier rows
out <- c(409, 43, 76, 23)
out1 <- c(-0.2409, 10, 43, 22)
out2 <- c(1509, 0.43, 76, 23)

df <- data.frame(a, b, c, d)

df2 <- rbind(df, out, out1, out2)

```

**Outlier detection for small datasets**

```{r outlier detection }
outdet2 <- multidetect(data = df2, multiple = FALSE,
                      var = 'a', output = 'outlier',
                      methods = c('zscore', 'adjbox',
                                  'logboxplot', 'distboxplot',
                                  'iqr', 'semiqr','seqfences',
                                  'hampel','kmeans',
                                  'jknife', 'onesvm',
                                  'iforest'), 
                      warn = FALSE)
```

**Visualize the threshold**

```{r visualise data, fig.width = 6, fig.height= 4, fig.align='center'}
par(mar = c(3, 3, 1.5, 1.5))

opt1 <- optimal_threshold(refdata = df2, 
                          outliers = outdet2, 
                         plotsetting = list(plot = TRUE))
opt1


```

* In the above scenario, the outlier detection exponentially increased and therefore absolute outliers were recorded denoted when the predictions exponentially increased. The optimal threshold of 0.7 will be automatically determined which filters out the three absolute outliers introduced in the data.

**Record weights used quality controlled data extraction.**

```{r check for the outlier weights and data cleaning}
#get the weights for the flagged records

weights <- ocindex(x = outdet2, absolute = TRUE, props = TRUE, threshold = 0.1, warn = FALSE)

print(weights)

dfclean <- extract_clean_data(refdata = df2, outliers = outdet2, loess = TRUE)

print(dfclean)

```

**NOTE**

* The outliers are removed from the dataset from the dataset automatically using the LOESS method.
* The legend is sometimes hidden.

### References

1. Cleveland, W. S. 1979. Robust Locally Weighted Regression and Smoothing Scatterplots. - J Am Stat Assoc 74
2. Cleveland, W. S. and Devlin, S. J. 1988. Locally Weighted Regression: An Approach to Regression Analysis by Local Fitting. - J Am Stat Assoc 83
3. Jacoby, W. G. 2000. Electoral inquiry section Loess: a nonparametric, graphical tool for depicting relationships between variables. - Elect Stud 19












