---
title: "Quick Start"
author: "Gilles Colling"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(couplr)
library(dplyr)
```

## What couplr Does

couplr creates matched samples from two groups of observations. Given a "left" group (e.g., treatment) and a "right" group (e.g., control), it finds optimal pairings based on similarity across variables you specify. couplr supports both **one-to-one matching** (each treatment unit paired with one control) and **full matching** (variable-ratio groups where every unit is assigned).

**Common use cases:**

- Matching treated patients to similar controls in observational studies

- Pairing survey respondents for comparison

- Creating balanced samples for causal inference

- Full matching when discarding unmatched units is undesirable

### Documentation Roadmap

| Vignette | Focus | Audience |
|----------|-------|----------|
| **Quick Start** (this) | Basic matching with `match_couples()` | Everyone |
| [Matching Workflows](matching-workflows.html) | Full pipeline: preprocessing, blocking, diagnostics | Researchers |
| [Algorithms](algorithms.html) | Mathematical foundations, solver selection | Technical users |
| [Comparison](comparison.html) | vs MatchIt, optmatch, designmatch | Package evaluators |

**Start here**, then proceed to whichever vignette matches your use case.

---

## Your First Match

The simplest workflow uses `match_couples()`:
```{r first-match}
library(couplr)
library(dplyr)

# Create example data: treatment and control groups
set.seed(123)
treatment <- tibble(
  id = 1:50,
  age = rnorm(50, mean = 45, sd = 10),
  income = rnorm(50, mean = 55000, sd = 12000)
)

control <- tibble(
  id = 1:80,
  age = rnorm(80, mean = 50, sd = 12),
  income = rnorm(80, mean = 48000, sd = 15000)
)

# Match on age and income
result <- match_couples(
  left = treatment,
  right = control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# View matched pairs
head(result$pairs)
```

**What happened:**

1. couplr calculated how similar each treatment unit is to each control unit

2. It found the optimal one-to-one pairing that minimizes total distance

3. Each treatment unit gets matched to exactly one control unit

### Understanding the Output

```{r output-explained}
# Quick overview with summary()
summary(result)

# Or access specific info
result$info$n_matched
```

The `result$pairs` table contains:

- `left_id`: Row number from the treatment group

- `right_id`: Row number from the control group

- `distance`: How different the matched units are (lower = more similar)

---

## Why Scaling Matters

Without scaling, variables with larger values dominate the matching. Income (measured in thousands) would overwhelm age (measured in decades):

```{r scaling-demo}
# BAD: Without scaling, income dominates
result_unscaled <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = FALSE
)

# GOOD: With scaling, both variables contribute equally
result_scaled <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# Compare mean distances
cat("Unscaled mean distance:", round(mean(result_unscaled$pairs$distance), 1), "\n")
cat("Scaled mean distance:", round(mean(result_scaled$pairs$distance), 3), "\n")
```

**Rule of thumb:** Always use `auto_scale = TRUE` unless you have a specific reason not to.

---

## Checking Match Quality

After matching, verify that treatment and control groups are now balanced:

```{r balance-check}
# Get the matched observations
matched_treatment <- treatment[result$pairs$left_id, ]
matched_control <- control[result$pairs$right_id, ]

# Compare means before and after matching
cat("BEFORE matching:\n")
cat("  Age difference:", round(mean(treatment$age) - mean(control$age), 1), "years\n")
cat("  Income difference: $", round(mean(treatment$income) - mean(control$income), 0), "\n\n")

cat("AFTER matching:\n")
cat("  Age difference:", round(mean(matched_treatment$age) - mean(matched_control$age), 1), "years\n")
cat("  Income difference: $", round(mean(matched_treatment$income) - mean(matched_control$income), 0), "\n")
```

For formal balance assessment, use `balance_diagnostics()` (covered in [Matching Workflows](matching-workflows.html)).

### Visualizing Match Quality

Use `plot()` to see the distribution of match distances:

```{r plot-result, fig.width=6, fig.height=4, fig.alt="Histogram showing distribution of match distances, with most matches having low distances near zero"}
plot(result)
```

The histogram shows how similar matched pairs are. A distribution concentrated near zero indicates good matches.

---

## Large Datasets: Use Greedy Matching

For datasets larger than a few thousand observations, optimal matching becomes slow. Use `match_couples(method = "greedy")` instead; it's 10-100x faster with nearly identical results:

```{r greedy-example}
# Create larger datasets
set.seed(456)
large_treatment <- tibble(
  id = 1:2000,
  age = rnorm(2000, 45, 10),
  income = rnorm(2000, 55000, 12000)
)

large_control <- tibble(
  id = 1:3000,
  age = rnorm(3000, 50, 12),
  income = rnorm(3000, 48000, 15000)
)

# Fast greedy matching
result_greedy <- match_couples(
  large_treatment, large_control,
  vars = c("age", "income"),
  auto_scale = TRUE,
  strategy = "row_best"  # fastest strategy
, method = "greedy")

cat("Matched", result_greedy$info$n_matched, "pairs\n")
cat("Mean distance:", round(mean(result_greedy$pairs$distance), 3), "\n")
```

**When to use which:**

| Dataset size | Recommended function |
|--------------|---------------------|
| < 1,000 per group | `match_couples()` |
| 1,000 - 5,000 | Either works; greedy is faster |
| > 5,000 | `match_couples(method = "greedy")` |

---

## Setting a Maximum Distance (Caliper)

Sometimes you want to reject poor matches rather than force bad pairings. Use `max_distance` to set a caliper:

```{r caliper-example}
# Allow any match
result_loose <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE
)

# Only allow close matches
result_strict <- match_couples(
  treatment, control,
  vars = c("age", "income"),
  auto_scale = TRUE,
  max_distance = 0.5  # reject pairs more different than this
)

cat("Without caliper:", result_loose$info$n_matched, "pairs\n")
cat("With caliper:", result_strict$info$n_matched, "pairs\n")
```

Stricter calipers mean fewer but better matches.

---

## Matching Within Groups (Blocking)

When you have natural groups in your data (e.g., hospitals, regions, study sites), you can match within each group separately. This ensures exact balance on the grouping variable.

First, create blocks with `matchmaker()`, then pass the result to `match_couples()`:

```{r blocking-example}
# Data from multiple hospital sites
set.seed(321)
treated <- tibble(
  id = 1:60,
  site = rep(c("Hospital A", "Hospital B", "Hospital C"), each = 20),
  age = rnorm(60, 55, 10),
  severity = rnorm(60, 5, 2)
)

controls <- tibble(
  id = 1:90,
  site = rep(c("Hospital A", "Hospital B", "Hospital C"), each = 30),
  age = rnorm(90, 52, 12),
  severity = rnorm(90, 4.5, 2.5)
)

# Step 1: Create blocks by hospital site
blocks <- matchmaker(
  left = treated,
  right = controls,
  block_type = "group",
  block_by = "site"
)

# Step 2: Match within each block
result_blocked <- match_couples(
  left = blocks$left,
  right = blocks$right,
  vars = c("age", "severity"),
  block_id = "block_id",
  auto_scale = TRUE
)

# Verify: matches stay within their block
result_blocked$pairs |> count(block_id)
```

Blocking guarantees that Hospital A patients are only matched to Hospital A controls, etc.

---

## Full Matching: Keep Every Unit

One-to-one matching discards unmatched controls. If you want every unit in a group, use `full_match()`. It creates variable-ratio groups (e.g., 1 treatment + 3 controls, or 2 treatments + 1 control) that minimize total distance:

```{r full-match-example}
result_full <- full_match(
  left = treatment,
  right = control,
  vars = c("age", "income")
)

result_full

# Each group has one or more left and right units with matching weights
head(result_full$groups)
```

Full matching is useful when your control pool is much larger than treatment and you don't want to waste data. See `vignette("matching-workflows")` for details on constraints (`min_controls`, `max_controls`, `caliper`) and the choice between `method = "optimal"` (default, globally optimal) and `method = "greedy"` (faster).

---

## Other Matching Methods

couplr also supports several alternative matching strategies. Each is covered in detail in `vignette("matching-workflows")`:
- **`cem_match()`** — Coarsened exact matching: bins continuous variables and matches exactly within strata, avoiding model dependence

- **`subclass_match()`** — Propensity score subclassification: divides units into PS strata with target estimand weighting (ATT, ATE, ATC)

- **`ps_match()`** — Propensity score matching with a logit caliper

- **`cardinality_match()`** — Maximizes sample size subject to strict balance constraints

All result types work with `balance_diagnostics()`, `match_data()`, and `as_matchit()` for ecosystem interoperability with cobalt and marginaleffects.

---

## Complete Example

Here's a realistic workflow from start to finish:

```{r complete-example}
# 1. Prepare your data
set.seed(789)
patients_treated <- tibble(
  patient_id = paste0("T", 1:100),
  age = rnorm(100, 62, 8),
  bmi = rnorm(100, 28, 4),
  smoker = sample(0:1, 100, replace = TRUE, prob = c(0.6, 0.4))
)

patients_control <- tibble(
  patient_id = paste0("C", 1:200),
  age = rnorm(200, 58, 10),
  bmi = rnorm(200, 26, 5),
  smoker = sample(0:1, 200, replace = TRUE, prob = c(0.7, 0.3))
)

# 2. Match on clinical variables
matched <- match_couples(
  left = patients_treated,
  right = patients_control,
  vars = c("age", "bmi", "smoker"),
  auto_scale = TRUE
)

# 3. Check how many matched
cat("Treated patients:", nrow(patients_treated), "\n")
cat("Successfully matched:", matched$info$n_matched, "\n")
cat("Match rate:", round(100 * matched$info$n_matched / nrow(patients_treated), 1), "%\n")

# 4. Extract matched samples for analysis
treated_matched <- patients_treated[matched$pairs$left_id, ]
control_matched <- patients_control[matched$pairs$right_id, ]

# 5. Verify balance
cat("\nBalance check (difference in means):\n")
cat("  Age:", round(mean(treated_matched$age) - mean(control_matched$age), 2), "\n")
cat("  BMI:", round(mean(treated_matched$bmi) - mean(control_matched$bmi), 2), "\n")
cat("  Smoker %:", round(100*(mean(treated_matched$smoker) - mean(control_matched$smoker)), 1), "\n")
```

---

## Next Steps

You now know the basics of matching with couplr. Here's where to go next:

**For production research workflows:**

- [Matching Workflows](matching-workflows.html) covers preprocessing, blocking, formal balance diagnostics, and publication-ready output

**For understanding algorithm choices:**

- [Algorithms](algorithms.html) explains when different solvers are faster or more appropriate

**For comparing with other packages:**

- [Comparison](comparison.html) shows how couplr differs from MatchIt, optmatch, and designmatch

---

## Additional: Direct Assignment Problem Solving

If you need to solve assignment problems directly (not matching workflows), couplr also provides lower-level functions.

### lap_solve(): Matrix-Based Assignment

Given a cost matrix where entry (i,j) is the cost of assigning row i to column j:

```{r lap-solve-basic}
# Cost matrix: 3 workers x 3 tasks
cost <- matrix(c(
  4, 2, 5,
  3, 3, 6,
  7, 5, 4
), nrow = 3, byrow = TRUE)

result <- lap_solve(cost)
print(result)
```

Row 1 is assigned to column 2 (cost 2), row 2 to column 1 (cost 3), row 3 to column 3 (cost 4). Total cost: 9.

### Forbidden Assignments

Use `NA` or `Inf` for impossible assignments:

```{r forbidden}
cost_forbidden <- matrix(c(
  4, 2, NA,   # Row 1 cannot go to column 3
  Inf, 3, 6,  # Row 2 cannot go to column 1
  7, 5, 4
), nrow = 3, byrow = TRUE)

lap_solve(cost_forbidden)
```

### Maximization

For preference or profit maximization:

```{r maximize}
preferences <- matrix(c(
  8, 5, 3,
  4, 7, 6,
  2, 4, 9
), nrow = 3, byrow = TRUE)

lap_solve(preferences, maximize = TRUE)
```

### Grouped Data

Solve multiple assignment problems at once using grouped data frames:

```{r grouped-lap}
# Weekly nurse-shift scheduling: solve each day separately
schedule <- tibble(
  day = rep(c("Mon", "Tue", "Wed"), each = 9),
  nurse = rep(rep(1:3, each = 3), 3),
  shift = rep(1:3, 9),
  cost = c(4,2,5, 3,3,6, 7,5,4,   # Monday costs
           5,3,4, 2,4,5, 6,4,3,   # Tuesday costs
           3,4,5, 4,2,6, 5,5,4)   # Wednesday costs
)

# Solve all three days at once
schedule |>
  group_by(day) |>
  lap_solve(nurse, shift, cost)
```

This solves each day's assignment problem independently and returns all results in one tidy table.

### K-Best Solutions

Find multiple near-optimal solutions:

```{r kbest}
cost <- matrix(c(1, 2, 3, 4, 3, 2, 5, 4, 1), nrow = 3, byrow = TRUE)

kbest <- lap_solve_kbest(cost, k = 3)
print(kbest)
```

---

## See Also

- `?match_couples` - Optimal one-to-one matching

- `?full_match` - Full matching (variable-ratio groups)

- `?match_couples` with `method = "greedy"` - Fast approximate matching

- `?balance_diagnostics` - Formal balance assessment

- `?lap_solve` - Direct assignment problem solving
