---
title: "VPR_processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VPR_processing}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::knitr}
author: " Emily O'Grady, Kevin Sorochan, Catherine Johnson"
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(vprr)
```

# Section 1: Background

This document was produced at Bedford Institute of Oceanography (BIO) to
accompany the vprr package, a processing and visualization package for
data obtained from the Digital Auto Video Plankton Recorder (VPR)
produced by SeaScan Inc. The VPR consists of a CPU, CTD, and camera
system with different optical settings (i.e., magnifications). It
captures underwater images and records their corresponding salinity,
temperature, and depth. The vprr package functions to join environmental
and plankton data derived from the CTD and camera, respectively, and
calculate plankton concentration and averaged environmental variables
along the path of the VPR. The package does not include automated image
classification; however, there is an optional manual classification
module, which can be used to review and correct outputs from automated
image classification while providing a record of any
(re)classifications.

The VPR outputs two raw files (.dat and .idx) for a given time period in
a deployment. These files are processed together in a software provided
with the VPR (i.e., AutoDeck), which decompresses the images, extracts
"regions of interest" (ROIs), and outputs ROI image files and a
corresponding CTD data file (.dat). The ROI file names are numeric
consisting of 10 digits. The first 8 digits correspond to the number of
milliseconds elapsed in the day at the time the image was captured. The
last two digits correspond to the ROI identifier (01-99). The ROIs and
corresponding CTD data are linked by their 8 digit time stamp. After the
ROIs have been extracted from the raw files they may be sorted into
categories manually or by an automated classification procedure. In
vprr, file naming conventions and directory structures are inherited
from a VPR image classification and analysis software, Visual Plankton.
However, the functionality of vprr is not dependent on the use of VP.

The data inputs for processing in vprr consist of the following file
types: aid (.txt), aidmeas (.txt), and CTD (.dat). The aid and aidmeas
files are derived from separate image classification and measurement
steps outside of vprr. Each "aid" (i.e., autoid) file contains file
paths to individual ROIs that have been classified to the category of
interest. The corresponding "aidmeas" file contains morphological data
for the ROIs (e.g., long axis length, perimeter, etc.).

![Figure 1. VPR data processing flow chart. Blue boxes represent
software, green and yellow boxes represent data sets, where yellow is
visual data and green is text format data. This document outlines the
processes within 'Processing and Visualization (R).](vp_flowchart.PNG)

# Section 2: Summary of vprr data processing steps

The processing steps in R from the 'Post Processing and Visualization
box in Figure 1 are detailed in Figure 2.

![Figure 2. Processing steps in R using vprr. Blue boxes represent each
section of processing, green ovals represent sub steps or routines
within each step.](r_flowchart.png)

## Section 2.1: Processing Environment

Before beginning data processing with vprr, it is recommended that a
processing environment be created containing commonly used variables and
file paths. The simplest and most reproducible way to achieve this is to
write an R script where all the mission and system specific variables
are contained, then save the environment as a RData file to be loaded at
the start of any processing scripts. This processing environment
contains reference to a station identifier csv file which should be
created for each mission. This file links station names from deck sheets
to the day and hour values assigned by AutoDeck. Day and hour values
represent the Julian day (3 digit) and two digit hour (24 hour clock)
when sampling was done. Note that the day and hour values will be in the
time zone of the computer used to run AutoDeck. Ensure that this matches
the time zone of the VPR CPU at the time of data collection to avoid a
time offset between data sources.

Another important part of setting up the processing environment is
ensuring the proper directory structure is in place, see Appendix 1 for
details on the required directory structure.

    # set VPR processing environment

    # WORKING DIRECTORY
    wd <- "C:/VPR_PROJECT/"
    setwd(wd)

    # MISSION
    cruise <- 'COR2019002'

    year <- 2019

    # CSV FILE WITH STATION NAMES AND CORRESPONDING DAY/HOUR INFO
    station_names_file <- paste0("station_names_", cruise, ".csv")
    # note columns should be labeled : station, day, hour

    # DIRECTORY FOR CTD DATA (output from AutoDeck)
    castdir <- paste0('D:/', cruise, "/", cruise, "_autodeck/")


    # AUTOID FOLDER FOR MEASUREMENT DATA (aidmeas & aid files)
    drive <- 'C:/'
    auto_id_folder <- paste0(drive, "cruise_", cruise, "/", autoid)
     #!!NO BACKSLASH AT END OF STRING

    # PATH TO AUTOID CATEGORY FOLDERS
    auto_id_path <- list.files(paste0(auto_id_folder, "/"), full.names = T) 

    # CREATE STANDARD DIRECTORY FOR SAVED DATA FILES PER MISSION
    savedir <- paste0(cruise, '_data_files')
    dir.create(savedir, showWarnings = FALSE)

    # CREATE STANDARD DIRECTORY FOR SAVED DATA PRODUCTS PER MISSION AND STATION
    stdir <- paste('data_product/', cruise, sep = "")
    dir.create(stdir, showWarnings = FALSE, recursive = TRUE)

    # BIN SIZE FOR COMPUTING DEPTH AVERAGED PLANKTON CONCENTRATION ALONG PATH OF THE VPR
    binSize <- 3

    # CATEGORY OF INTEREST (PLANKTON CATEGORIES TO BE PROCESSED AND VISUALIZED)
    category_of_interest <-
      c(
        'krill',
        'Calanus',
        'chaetognaths',
      )

    #### SAVE ####
    # SAVE ALL FILE PATHS AND SETTINGS AS PROJECT ENVIRONMENT

    save.image(file = paste0(cruise,'_env.RData'))

An example of the station names csv file looks like this:

```{r, eval = TRUE}
csv <- read.csv('station_names_COR2019002.csv')

head(csv)

```

Once this environment is set, it can be loaded into any processing
session by using

    load('COR2019002_env.RData') # where COR2019002 is mission name

If sharing processing code with colleagues on version control, keeping
the environment variables separate (outside of the git project) will
allow collaboration while avoiding inconsistencies in file paths or
folder names.

## Section 2.2: Image Copying (optional):

ROIs are organized into folders corresponding to their assigned
classification categories from automated image classification. The
information in each aid file is used to create a folder of images that
have been classified to that category. This step is only required if
manual re-classification (see Section 2.2) is intended. Further details
on image copying are provided in Section 3.

## Section 2.3: Manual re-classification (optional):


Automated classifications from are manually checked, which allows for
manual correction and addition of categories not previously used for
automated classification. ROIs that have been copied in Section 2.1 are 
manually sorted to correct for misclassifications. Updated aid and aidmeas files are produced. 
Further details on manual re-classification are provided in Section 4.

## Section 2.4: Processing:

Data outputs from Autodeck (ctd .dat files), automatic classifications (aid files)
and measurements (aidmeas files) are joined together. The aid and aidmeas files, which may have been updated
(see Section 2.3) are joined with CTD text files by the 8 digit time
stamp. The data are then averaged in user-defined vertical bins to
produce a time series of plankton concentrations and environmental
variables. Quality controlled data products (before and after binning)
are then exported in simple formats (csv, RData, oce) for plotting and
analysis. Further details on data processing are provided in Section 5.


# Section 3: Image Copying

In this step, ROIs are copied to folders that are organized based on the
day and hour of data collection and classification category assigned
from automatic classification (see Appendix 1: 'Image Folders'). The
images are organized by AutoDeck into day and hour; however,
reorganizing them based on classification allows easier human
interaction with the data and visual inspection of classifications.
Moreover, this directory structure is used by the next step of
processing (i.e., manual re-classification). To implement this step use
the function `vprr::vpr_autoid_copy()` For more information on input
variables, please see documentation for `vpr_autoid_copy()`

    # create variables
    # ---------------------
    basepath <- "C:\\data\\cruise_COR2019002\\autoid\\" 
    # note this is the same as the auto_id_folder environment variable except the file separator is different, because this script will run source code in command line which does not recognize '/' as a file separator

    day <- "123"
    hour <- "01" # note leading zero 
    classifier_type <- "svm" 
    classifier_name <- "myclassifier"

    # run file organizer
    # ---------------------
    vpr_autoid_copy(basepath, day, hour, classifer_type, classifier_name)

# Section 4: Manual Re-classification

Manual re-classification of some categories after automated
classification may be required to achieve identification accuracy
standards. In this step, ROIs are displayed on the screen one at a time
for manual verification. If an image has been misclassified or if it
falls into a new user-defined category (described below), the image can
be re-classified. This is especially useful for classification of rare
categories that were not defined prior to automatic classification.
After completing manual re-classification for a day-hour set, new aid
and aidmeas files are created for new categories, which are identical in
format to original aid and aidmeas files.

## Section 4.1: Preparing the environment by setting some variables

-   Load the processing environment, which includes the `auto_id_folder`
    variable.
-   Set day and hour of interest.
-   Set category of interest. These categories are the existing automated
    classification categories which require manual re-classification, as
    well as any new categories. The `vprr::vpr_category_create()`
    function sets up the folder structure for any new categories which
    have been added to the list of interest.
-   Run manual re-classification with
    `vprr::vpr_manual_classification()`. This function has a few
    optional arguments to customize the manual re-classification
    experience, notably `gr` which is a logical value determining
    whether or not manual re-classification options appear as pop ups or
    in the console, as well as `img_bright`, a logical which determines
    whether or not the original image is appended with an extra bright
    version of the image. Having a bright version of the image allows
    the user to see the outline of the organism better, any thin
    appendages become more clear and gelatinous organisms like
    chaetognaths or ctenophores are easier to distinguish.

```{=html}
<!-- -->
```
        #### MANUAL RE-CLASSIFICATION
        # -------------------------------------

        # Once classified images are sorted by taxa
        # verify classification accuracy by manually 
        # looking through classified images

        #### USER INPUT REQUIRED ####     

        load('COR2019002_env.RData')

        day <- '235' 
        hr <- '19' # keep leading zeros, must be two characters


        category_of_interest <-
          c(
            'krill',
            'Calanus',
            'chaetognaths',
            'ctenophores',
            'Other',
            'larval_fish',
            'marine_snow',
            'small_copepod',
            'other_copepods',
            'larval_crab',
            'amphipod',
            'Metridia',
            'Paraeuchaeta',
            'cnidarians',
            'speciesA', # new category
          )


        # add new category (optional)

        vpr_category_create(taxa = category_of_interest, auto_id_folder)
        # ensures there is proper folder structure for all categories of interest

        # reclassify images
        vpr_manual_classification(day = day, hour= hr, basepath = auto_id_folder,gr = FALSE, 
                  taxa_of_interest = category_of_interest, scale = 'x300',
                  opticalSetting = 'S3')

## Section 4.2: Generate new aid and aidmeas files

The function`vprr::vpr_manual_classification()` produces two files
('misclassified' and 're-classified' text files) as a record of manual
re-classification, which are found in the R project working directory in
folders named by the day and hour that the data were collected. The
function`vprr::vpr_autoid_create()`takes these files and outputs new aid
and aidmeas files in the R working directory in folders named by
classification category. This step should be run after each hour of data
is manually re-classified.

    #### REORGANIZE ROI AND ROIMEAS DATA
    # -----------------------------------------

    day_hour_files <-  paste0('d', day, '.h', hr)

    misclassified <- list.files(day_hour_files, pattern = 'misclassified_', full.names = TRUE)

    reclassify <- list.files(day_hour_files, pattern = 'reclassify_', full.names = TRUE)

    # MOVE ROIS THAT WERE MISCLASSIFIED INTO CORRECT FILES & REMOVE MISCLASSIFIED ROIS
    vpr_autoid_create(reclassify, misclassified, auto_id_folder)


The aid and aidmeas files are both text files which are
specifically formatted to record classification outputs for further
processing. The format and naming conventions of these files has been inherited from a VPR image classification and data processing tool called Visual Plankton (written in Matlab); however, the functionality of vprr is independent from that of Visual Plankton. The aid files are text records of image paths,
where each individual text file represents a classification category. Each line of
the aid file is the full path to an image which was classified into the
designated category. Note that the naming scheme of aid files does not
include the category name in the file title and the category is only
identifiable by the folder in which it is located. For example the 'krill'
classification aid file might be named 'oct10_1svmaid.d224.h01' but be located within the
'krill' autoid folder. The aidmeas files are also text files which represent a
variety of different measurements taken of the object(s) within a ROI image. The columns of the aidmeas files are
c('Perimeter','Area','width1','width2','width3','short_axis_length','long_axis_length').Examples
of each of these files can be found below.


```{r, eval =  TRUE}

aid <- read.table(file = system.file("extdata/COR2019002/autoid/bad_image_blurry/aid/sep20_2svmaid.d222.h04", package = 'vprr', mustWork = TRUE))

head(aid)

aidmeas <- readLines(system.file("extdata/COR2019002/autoid/bad_image_blurry/aidmea/sep20_2svmaid.mea.d222.h04", package = 'vprr', mustWork = TRUE))

head(aidmeas)
```

## Section 4.3: File check


The last step of manual re-classification includes some manual file 
organization and final checks. These files should be manually reorganized
in a new directory which will become the new auto_id_folder (see Appendix 1:
Directory Structure). Any aid and aidmeas files from categories which were 
not manually checked and re-classified should also be added to this new auto_id_folder
if they are to be included in further processing (e.g., computation of concentration
in user-specified depth bins). After the updated aid and aidmeas files have been
manually reorganized they can be quality controlled using vprr::vpr_autoid_check().
The user could also manually check the files. The automated check function removes
any empty aid files created.  Empty files can cause errors in processing down the line.
This function also checks that (1) aid and aidmeas files are aligned within an hour of data;
(2) aid and aidmeas files include the same number of ROIs; and (3) the
VPR tow number for all files is the same.


    #### FILE CHECK
    # --------------------------------


    # aid check step
    # removes empty aid files, and checks for errors in writing

    vpr_autoid_check(basepath, cruise) #OUTPUT: text log 'CRUISE_aid_file_check.txt’ in working directory

# Section 5: Data Processing

This is the main chunk of coding required to generate data products.
This step does not require image copying (Section 3) or manual
re-classification (Section 4) steps; however, if these steps were taken
the aid and aidmeas files generated from manual re-classification and
integrated into the directory structure (as specified in Section 4) are
used as an input. The following is a walk-through of processing data
from a DFO field mission (i.e. mission COR2019002) in the southern Gulf
of St. Lawrence in 2019. First, all libraries should be loaded and the
processing environment, described in Section 2.4 should be loaded.

    ##### PROCESSING  --------------------------------------------------------------------------------------------------------------------
    library(vprr)

    #### FILE PATHS & SETTINGS --------------------------------------------------------------------------------------------------------------------
    # loads processing environment specific to user

    load('COR2019002_env.RData')

This section allows a user to process all stations of a particular
mission in a loop. This can be modified or removed based on personal
preference

    ##### STATION LOOP ----------------------------------------------------------------------------------------------------------------------------

    all_stations <- read.csv(station_names_file, stringsAsFactors = FALSE) 
    all_stations_of_interest <- unique(all_stations$station) 


    for (j in 1:length(all_stations_of_interest)){
      
      station_of_interest <- all_stations_of_interest[j] 
      
      cat('Station', station_of_interest, 'processing... \n')
      cat('\n')

Optical settings and image volume variables should be set. If they are
consistent throughout the mission, they could also be added to the
processing environment (Section 2.4).

      #==========================================#
      #   Set optical settings  & Image Volume   #
      #   !Should be updated with each mission!   #
      #==========================================#
      if(cruise == "COR2019002") {
        
        #VPR OPTICAL SETTING (S0, S1, S2 OR S3)
        opticalSetting <- "S2"
        imageVolume <- 108155 #mm^3
      }

CTD data are loaded in using `vprr::vpr_ctd_files` to find files and
`vprr::vpr_ctd_read` to read in files. During CTD data read in, a
seawater density variable `sigmaT` is derived using the function
`oce::swSigmaT`, and `depth` (in meters) is derived from pressure using
the function `oce::swDepth`. For more information on the oce package,
see dankelley/oce on GitHub.

      #get day and hour info from station names list
      dayhour <- vpr_dayhour(station_of_interest, file = station_names_file)
      
      
      ##### PULL CTD CASTS ----------------------------------------------------------------------------------------------------------------------------
      # get file path for ctd data
      
      # list ctd files for desired day.hours
      ctd_files <- vpr_ctd_files(castdir, cruise, dayhour) 
      
      ##### READ CTD DATA ----------------------------------------------------------------------------------------------------------------------------
      
      ctd_dat_combine <- vpr_ctd_read(ctd_files, station_of_interest)
      
      cat('CTD data read complete! \n')
      cat('\n')
      

The aid and aidmea files, which reflect manual classification (if used,
see Section 4), are then found

      ##### FIND VPR DATA FILES ----------------------------------------------------------------------------------------------------------------------
      
      # Path to aid for each taxa                               
      aid_path <- paste0(auto_id_path, '/aid/')              
      # Path to mea for each taxa                               
      aidmea_path <- paste0(auto_id_path, '/aidmea/')        
      
      # AUTO ID FILES
      aid_file_list <- list()
      aidmea_file_list <- list()
      for (i in 1:length(dayhour)) {
        aid_file_list[[i]] <-
          list.files(aid_path, pattern = dayhour[[i]], full.names = TRUE)
        # SIZE DATA FILES
        aidmea_file_list[[i]] <-
          list.files(aidmea_path, pattern = dayhour[[i]], full.names = TRUE)
      }
      
      aid_file_list_all <- unlist(aid_file_list)
      aidmea_file_list_all <- unlist(aidmea_file_list)
      
      remove(aid_file_list, aidmea_file_list, aid_path, aidmea_path)
      

aid and aidmeas files are read in using `vprr::vpr_autoid_read()`.

      ##### READ ROI AND MEASUREMENT DATA ------------------------------------------------------------------------------------------------------------
      
      
      # ROIs
      roi_dat_combine <-
        vpr_autoid_read(
          file_list_aid = aid_file_list_all,
          file_list_aidmeas = aidmea_file_list_all,
          export = 'aid',
          station_of_interest = station_of_interest,
          opticalSetting = opticalSetting
        )
      
      # MEASUREMENTS
      roimeas_dat_combine <-
        vpr_autoid_read(
          file_list_aid = aid_file_list_all,
          file_list_aidmeas = aidmea_file_list_all,
          export = 'aidmeas',
          station_of_interest = station_of_interest,
          opticalSetting = opticalSetting
        )
      
      cat('ROI and measurement data read in complete! \n')
      cat('\n')

Next, CTD and aid data are merged to create a data frame describing both
environmental variables (eg. temperature, salinity) and classified
images. The function used is `vprr::vpr_ctdroi_merge()`.

      ##### MERGE CTD AND ROI DATA ---------------------------------------------------------------------------------------------------------------------
      ctd_roi_merge <- vpr_ctdroi_merge(ctd_dat_combine, roi_dat_combine)
      
      cat('CTD and ROI data combined! \n')
      cat('\n')

Before final export of data products, the following variables are added
to the data frame: time in hours (time_hr) is calculated, and a time
stamp (ymdhms) with POSIXct signature in Y-M-D h:m:s format is added
using the function `vpr_ctd_ymd`.

      ##### CALCULATED VARS ----------------------------------------------------------------------------------------------------------------------------
      
      # add time_hr and sigma T data and depth
        data <- ctd_roi_merge %>%
        dplyr::mutate(., time_hr = time_ms / 3.6e+06) 
     
        data <- vpr_ctd_ymd(data, year)
        
      # ensure that data is sorted by time to avoid processing errors
        data <- data %>%
        dplyr::arrange(., time_ms)
      
      cat('Initial processing complete! \n')
      cat('\n')
      
      # clean environment
      remove(ctd_roi_merge)


Average plankton concentration and environmental variables (e.g.,
temperature, salinity, density, etc.) are then computed within a user
defined depth bin. The computation of plankton concentration is dependent
on the assumption that the same animals are not re-sampled by the
instrument. The bin-averaging step standardizes plankton concentrations
when the VPR does not sample the water column evenly. This can occur 
due to characteristics of the deployment or variability in the sampling
rate, which is not necessarily constant in older versions of the VPR. 
Binning also reduces noise in the data. First, an oce CTD object is 
created using `vprr::vpr_oce_create()`. Then, bin-averaging is done 
using `vprr::bin_cast()`. Concentrations are calculated for each 
category of interest.


      ##### BIN DATA AND DERIVE CONCENTRATION ----------------------------------------------------------------------------------------------------------
      
      ctd_roi_oce <- vpr_oce_create(data)
      
    # bin and calculate concentration for all taxa (combined)
      vpr_depth_bin <- bin_cast(ctd_roi_oce = ctd_roi_oce, binSize =  binSize, imageVolume = imageVolume)
      
      # get list of valid taxa
      taxas_list <- unique(roimeas_dat_combine$taxa)
      
      # bin and calculate concentrations for each category
      taxa_conc_n <- vpr_roi_concentration(data, taxas_list, station_of_interest, binSize, imageVolume)  

      cat('Station', station_of_interest, 'processing complete! \n')
      cat('\n')
      
      # bin size data
      
      size_df_f <- vpr_ctdroisize_merge(data, data_mea = roimeas_dat_combine, taxa_of_interest = category_of_interest)
      
      size_df_b <- vpr_size_bin(size_df_f, bin_mea = 3)
      

Finally, data are saved as RData and csv files for export and plotting.
Data are also saved as an `oce` object in order to preserve both data
and metadata in an efficient format.

      ##### SAVE DATA ---------------------------------------------------------------------------------------------------------------------------------
      # Save oce object
      oce_dat <- vpr_save(taxa_conc_n)
      save(file = paste0(savedir, '/oceData_', station_of_interest,'.RData'), oce_dat) # oce data and metadata object
      
      
      # Save RData files
      save(file = paste0(savedir, '/ctdData_', station_of_interest,'.RData'), ctd_dat_combine) #CTD data
      save(file = paste0(savedir, '/stationData_', station_of_interest,'.RData'), data) # VPR and CTD data
      save(file = paste0(savedir, '/meas_dat_', station_of_interest,'.RData'), roimeas_dat_combine) #measurement data
      save(file = paste0(savedir, '/bin_dat_', station_of_interest,'.RData'), vpr_depth_bin) # binned data with cumulative concentrations
      save(file = paste0(savedir, '/bin_size_dat_', station_of_interest,'.RData'), size_df_b) # binned data including measurements
      
      cat('CTD, ROI-VPR merge, ROI measurement saved as RData! \n')
      cat('\n')
      
      # Write csv files
      # write.csv(file = paste0(stdir, '/vpr_data_unbinned', station_of_interest, '.csv'), data, row.names = F) # VPR and CTD data (not binned)
      # write.csv(file = paste0(stdir, '/vpr_meas', station_of_interest, '.csv'), roimeas_dat_combine) # measurement data
      write.csv(file = paste0(stdir, '/vpr_data_binned', station_of_interest, '.csv'), taxa_conc_n) # VPR and CTD data with concentrations by taxa
      
      cat('ROI measurments, ROI-CTD merge-unbinned, and ROI-CTD merge-binned written to csv! \n')
      cat('\n')
      
    } #end of station loop

# Section 6: Plotting

Although not primarily a plotting package, vprr can produce contour
plots, profile plots and temperature-salinity (TS) plots from VPR data
sets. A few example plots are provided in the following code. The first
step to plotting is properly loading in the processed VPR data objects
created in processing. The environment, described in Section 2.4 should
also be loaded. The individual data files are found by distinct names
(e.g., "stationData"). The directory structure may be different
depending on the `savedir` where data files were saved during
processing. Note that the following plotting examples are tailored for
tow-yo pattern VPR deployments.

    ##### FILE PATH & SETTINGS -----------------------------------------------------------------------------------------------------------------------

    library(vprr)

    # loads all file paths and environment vars specific to User
    load('COR2019002_env.RData')

    #find all data files
    fn_all_st <- list.files(paste0(cruise, "_data_files/"), pattern = "stationData", full.names = T)
    fn_all_meas <- list.files(paste0(cruise, "_data_files/"), pattern = "meas", full.names = T)
    fn_all_conc <- list.files(paste0("data_product/", cruise, "/"), pattern = "data_binned", full.names = T)
    fn_all_bin <- list.files(paste0(cruise,"_data_files/"), pattern = 'bin_dat', full.names = T)

Once files are loaded, plots for all stations in a mission can be
generated using a loop, in order to efficiently generate comparable
plots. The example below uses a loop to run through a list of stations
described by a csv file. This loop also isolates two specific
classification categories to plot (eg. "Calanus" and "krill").

    ####START STATION LOOP ---------------------------------------------------------------------------------------------------------------------------

    setwd(wd)

    all_stations <- read.csv(station_names_file, stringsAsFactors = FALSE)
    all_stations_of_interest <- unique(all_stations$station)

    taxa_to_plot <- c("Calanus", "krill")


    for (j in 1:length(all_stations_of_interest)){
      
      setwd(wd)
      station <- all_stations_of_interest[j]
      
      cat('station', station ,'starting to plot.... \n')
      cat('\n')

Data files are loaded for the specific station of interest. This loads
in all relevant RData files as well as the concentration data saved as a
csv file.

      #load station roi and ctd data
      fn_st <- grep(fn_all_st, pattern = station, value = TRUE, ignore.case = TRUE)
      fn_meas <- grep(fn_all_meas, pattern = station, value = TRUE, ignore.case = TRUE)
      fn_conc <- grep(fn_all_conc, pattern = station, value = TRUE, ignore.case = TRUE)
      fn_bin <- grep(fn_all_bin, pattern = station, value = TRUE, ignore.case = TRUE)
      
      load(fn_st)
      load(fn_meas)
      load(fn_conc)
      load(fn_bin)
      
      
     # load concentration data
      taxa_conc_n <- read.csv(fn_conc, stringsAsFactors = F)
      
      station_name <- paste('Station ', station)

The final section of set up indicates the directory in which plots will
be saved and provides generic plot size arguments which will control how
large the saved .png files are.

      # directory for plots
      stdir <- paste0('figures/', cruise, '/station', station)
      dir.create(stdir, showWarnings = FALSE, recursive = TRUE)
      setwd(stdir)
      
      width = 1200
      height = 1000

The following example presents a plot of the concentrations of a taxon
as scaled bubbles along the tow path, overlain on contours of an
environmental variable from the CTD. The main function used is
`vprr::vpr_plot_contour()` which uses a standard VPR data frame
(`taxa_conc_n` - produced from processing (Section 5)) and plots the
background contours. Interpolation methods can be adjusted based on data
or preference. The VPR tow path can be added on top of contours. This
method can be repeated with various environmental variables (e.g.,
temperature, salinity etc.) used to calculate the contours, by changing
the `var` argument in `vprr::vpr_plot_contour()`.

      # Density (sigmaT)
      png('conPlot_taxa_dens.png', width = width, height = height)
      p <- vpr_plot_contour(taxa_conc_n[taxa_conc_n$taxa %in% c(taxa_to_plot),], var = 'density', dup = 'strip', method = 'oce', bw = 0.5)
      p <- p + geom_line(data = data, aes(x = time_hr - min(time_hr), y = pressure), col = 'snow4', inherit.aes = FALSE) +
        geom_point(data = taxa_conc_n[taxa_conc_n$taxa %in% c(taxa_to_plot),], aes(x = time_hr, y = min_pressure, size = conc_m3), alpha = 0.5)+
        ggtitle(station_name ) +
        labs(size = expression("Concentration /m" ^3), fill = 'Density')+
        scale_size_continuous(range = c(0, 10)) +
        facet_wrap(~taxa, ncol = 1, scales = 'free') +
        theme(legend.key.size = unit(0.8, 'cm'),
              axis.title = element_text(size = 20),
              strip.text = element_text(size = 20),
              plot.title = element_text(size = 32), 
              axis.ticks = element_line(size = 1, lineend = 'square'),
              axis.text = element_text(size = 30),
              legend.text = element_text(size = 20),
              legend.title = element_text(size = 25)
        )
      
      print(p)
      dev.off()

Vertical profiles of plankton concentration and environmental variables
compressed over the sampling duration can be generated using
`vprr::vpr_plot_profile()`. This type of plot indicates the overall
pattern in vertical distribution over the VPR deployment.

        png('profilePlots.png', width = 1000, height = 500)
        p <- vpr_plot_profile(taxa_conc_n, taxa_to_plot)
        print(p)
        dev.off()

Temperature-salinity (TS) plots can be generated to visualize how
plankton concentration varies across different water masses. In the
example below, a TS plot is produced in ggplot (with labeled
isopycnals), and concentration bubbles for each selected classification
group are overlaid on the plot. The basic TS bubble plot can be easily
manipulated using ggplot2 grammar, for example the plots can be faceted
by classification group or axis labels and sizing can be adjusted (see
ggplot2 package for more information).

    ####TS BUBBLE PLOT ----------------------------------------------------------------------------------------------------------------------------
      
      # plot by taxa
      taxa_conc <- taxa_conc_n[taxa_conc_n$conc_m3 > 0,]
      png('TS_conc_taxa.png', width = 1000, height = 500)
      p <- vpr_plot_TS(taxa_conc[taxa_conc$taxa %in% c(taxa_to_plot),], var = 'conc_m3') +
        facet_wrap(~taxa, nrow = 1) +
        theme(strip.text = element_text(size = 18),
              axis.title = element_text(size = 20),
              panel.spacing = unit(2, 'lines'))
      print(p)
      dev.off()
      
      
      cat('station', station, 'complete! \n')
      cat('\n')
      
    } # end station loop

# Section 7: Disclaimer

The functions in vprr were created for a specific project and have not
been tested on a broad range of field mission data. It is possible that
deviations in data format and directory structure from that described
herein may result in errors when using vprr. The vprr package was
developed for the purpose of processing data collected during tow-yo VPR
deployments and image classification using VP. The purpose of this
document is to provide a template for processing and visualizing VPR
data that can be adapted by other users for their own objectives.

# Appendix 1: Directory Structure

Visual Plankton (Matlab image classification software) requires a very
specific directory structure in order to function. Since this processing
is meant to directly follow this image classification, the VP directory
structure is used for consistency. This allows a smooth transition
between the Matlab classifications and the completion of processing in
R. The directory structure required is described below

-   C:/
    -   data

        -   cruise_name

            -   autoid

                -   taxa
                    -   aid
                    -   aidmea
                    -   image folders

            -   rois

                -   vprtow#
                    -   day
                        -   hour

            -   trrois

                -   vprtow#
                    -   day
                        -   hour

This is your project directory, where your R scripts and work products
will be stored:

-   ...
    -   VPR_PROJECT

    -   R

        -   R scripts/ workflows

    -   new_autoid

        -   taxa

        -   aid

        -   aidmea

    -   manual_reclassification_record

        -   day/hour

        -   misclassified

        -   reclassified

    -   figures

    -   station names (csv)

# Appendix 2: Glossary

**Aid files** - Visual Plankton file output text file, listing file path
information for ROI's of a specific classification group

**AidMeas files (AutoID measurements)** - Visual Plankton output text
file, listing measurement data for ROI's of a specific classification
group. Unit is pixels and columns are 'Perimeter', 'Area', 'width1',
'width2', 'width3', 'short_axis_length', 'long_axis_length'

**Auto Deck** - software which pulls plankton images from Video Plankton
Recorder frames based on specific settings

**Auto ID** - The automatic classification given to an image from Visual
Plankton machine learning algorithm

**AutoID files** - Includes both Aid and AidMeas files as part of Visual
Plankton's automatic classifications

**BIO** - Bedford Institute of Oceanography, a research institute in
Halifax NS, Canada

**Classification category (Taxa)** - A defined group under which VPR
images can be classified, often represents a taxonomic group (e.g.
Krill), but can also be defined by image type (e.g. 'bad_image_blurry'),
or other (e.g. 'marine_snow'), should be one continuous string (no
spaces)

**CPU** - Central processing unit (computer processor)

**CTD** - Conductivity, Temperature and depth sensor instrument

**Day** - Julian calendar day on which VPR data was collected (three
digits)

**Hour** - Two digit hour (24 hour clock) describing time at which VPR
data was collected

**Image volume** - The measured volume of water captured within a VPR
image. Calculated based on optical setting and VPR standards. This is
based on AutoDeck settings, it is calculated from the VPR calibration
file (unique to each instrument). It will change based on AutoDeck
settings and should be updated with each cruise/ processing batch. It is
measured in cubic mm

**Optical Setting** - A VPR setting controlling image magnification and
field of view, which can be S0, S1, S2 or S3, where S0 has the greatest
magnification and smallest image volume, and S3 has the least
magnification and largest image volume

**ROI** - Region of interest, images identified by autodeck within VPR
frames based on settings defined in autoDeck program

**SeaScan** - Oceanographic instrument manufacturing company

**station** - A named geographic location, where the VPR was deployed

**Tow-yo** - A VPR deployment method where the VPR is towed behind a
vessel while being raised and lowered through the water column in order
to sample over both depth and distance

**TRROIS** - Training set of images used to train machine learning
algorithm in Visual Plankton

**VP** - Visual Plankton program run in Matlab

**VPR** - Video Plankton Recorder, oceanographic instrument used to
image small volumes of water for the purpose of capturing images of
plankton

**vprtow#** - A numeric code which is unique to each VPR deployment

**Working Directory** - File path on your computer that defines the
default location of any files you read into R, or save out of R
