---
title: "Introduction to lifelogr"
date: "2017-03-21"
output: 
  pdf_document:
    fig_width: 3.5
    fig_height: 1.9
urlcolor: blue

vignette: >
  %\VignetteIndexEntry{Introduction to lifelogr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


```{r, eval = FALSE}
vignette("Report", package = "lifelogr")
```


```{r}
library(lifelogr)
```

[Lifelogging](http://lifestreamblog.com/lifelogging/) is the "process of 
tracking personal data generated by our own behavioral activities."  Examples of
lifelogging include tracking one's exercise and sleep. While devices like 
iPhones and Fitbits make collecting data on oneself easy, analyzing and 
interpreting that data is more difficult.

The lifelogr package makes life logging analysis and interpretation easy by:

- Providing a framework for working with multiple sources of personal data.
- Allowing users to easily create three sets of meaningful visualizations.
- Creating an ecosystem for experimentation - a feedback loop between self-data,
and the process generating it.

## Working with Multiple Sources of Data: the "Person" class

This package is built around the Person class, an R6 class. An object of Person 
class encapsulates a complete picture of a user, as seen through her data.
The goal of this class is to make working with multiple sources of data as easy 
as possible by joining and standardizing multiple datasets. Users can then 
access the data from and across sources for visualization and experimentation.

Upon instantiation, a user can provide information about himself, the filepath 
to a csv file with Apple data, Fitbit account information, and/or dataframes of
data from other sources.

### Loading Multi-Source Data into Person Objects

Fitbit users can provide their account username and password upon instantiation.
We use functions from the fitbitScraper package to collect all of a user's 
Fitbit data between the start and end date of interest.
```{r, eval = FALSE}
EX <- Person$new(fitbit_user_email = "example@domain.com", 
                 fitbit_user_pw = "password",
                 apple_data_file = "apple.csv",
                 addl_data = NA,
                 user_info = list("name" = "EX", "age" = 29, "gender" = "male"),
                 target_steps = 10000,
                 group_assignments = list(data.frame(NA), data.frame(NA)),
                 start_date = "2017-01-19", end_date = "2017-02-17")
```

```{r, echo = FALSE}
load("../data/EX.rda")
```

iPhone users can provide a filepath to a csv with their Apple health data. Users
can download their data using an iOS app called QS Access (quantified self),
which allows for selection of variables and time scales (by day or by hour), to 
include in a csv file. For greater compatibility with the lifelogr package, 
users are highly recommended to select the "by hour" time interval.

Users can provide data from other sources and self-tracking apps as pre-cleaned
dataframes in the `addl_data` or `addl_data2` arguments. To make maximal use of
lifelogr's functions, the column names of the same variables in the dataframe 
should match those referenced in the Person class, and class documentation 
(when the variables in the dataframe are the same as those referred to in the documentation).

### Example of Person Class

EX is an example of a Person class for a user who provided their Fitbit account 
information upon instantiation. Functions within the Person class gathered and 
manipulated the data into data from three main "sources":

* `fitbit_daily` has Fitbit data recorded on a daily basis
* `fitbit_intraday` has Fitbit data recorded on an intraday basis, generally 
  every 15 minutes, although for certain variables, such as heart rate, data are 
  recorded every 5 minutes.
*  `fitbit_util` has information about the dates in the range specified 
  by the user upon Person instantiation, such as which day of the week a 
  specific date is.
  
  
### Dates, Times, and Date-Times

Dates and times come in numerous forms. lifelogr recognizes three types of
dates/times as acceptable time-related variables to join datasets on:

* date: "2017-03-20"
* time: "08:00:00"
* datetime: "2017-03-20 08:00:00", a combination of a date and a time

Upon instantiation, the Person class automatically converts and renames Fitbit 
and Apple health data to adhere to this variable naming and format convention. 
If users provide their own addl_data as dataframes, they will need to have one 
or more of these time-related variables as columns in their dataframes in order
to be able to join across their data source, and Fitbit/Apple health data.

## Visualization

There are 3 functions which allow for a series of plots with just one function
call: `plot_sleep_all`, `plot_daily_all`, and `plot_intraday_all`.  Each acts 
like the `plot.lm` function, where users must click "enter" to see the next plot.

Each plot within each of the three series can also be generated individually 
with a call to its appropriate plot function.

For example, using `plot_sleep_all` to generate all the sleep plots for EX:
```{r, fig.show = "hold"}
plot_sleep_all(EX)
```

Calling `plot_daily_all` and `plot_intraday_all` will result in similar series
of plots.

Users can also call each function individually using 
`plot_sleep(person, plot_type)` (or `plot_daily` or `plot_intraday`), and also 
pass in other arguments for plots with other arguments.

For example:
```{r}
plot_sleep(EX, "by_start_end_time", "day_of_week")
```

Using `plot_daily`:
```{r}
plot_daily(EX, "steps")
```

Using `plot_intraday`:
The default is for `plot_intraday` to group the data by time intervals 
within each day so that data for a "typical day" is displayed.
```{r}
plot_intraday(EX, "distance", unit = "km")
```

However, it is also possible to specify that the plots use the raw data and plot
over all datetimes.

```{r}
plot_intraday(EX, "bpm", FALSE)
```

## Experimentation Framework

While almost all the visualizations shown above had date, date-time, or time on 
the x axis, the experimentation framework allows for greater flexibility.  Users
can select any variables of interest and examine their relationship.

Users can specify

* x variables of interest
* y measures of interest
* type of analysis ("plot", "correlation", "anova", "compare_groups", or 
"regression")
* time variable (date, time, or datetime)

The `experiment` function studies the effect of each variable, or all variables
together (depending on the analysis), of the variable (X) and measure variables 
(Y) given.

Users can also perform each analysis using the functions
`l_plot`, `correlation`, `l_anova`, `compare_groups`, or `l_regression` 
instead of the higher-level function `experiment`.

The analysis functions below perform specific analyses, but using the 
standardized datasets in the fields of the Person class, the `create_dataset` 
function,
which joins datasets across source types, and the templates provided by these
analysis functions, users can easily perform their own unique analyses.

1.  Plots

    To plot the relationship between sleep duration and resting heart 
    rate and the relationship between day of week and resting heart rate on a 
    date basis:
    ```{r, fig.show = "hold"}
    experiment(person = EX, 
               variables = list("fitbit_daily" = c("sleepDuration"),
                                "util" = c("day_of_week")),
               measures = list("fitbit_daily" = c("restingHeartRate")), 
               analysis = c("plot"),
               time_var = c("date"))
    ```

2.  Correlation

    To calculate the correlation between sleep duration and number of steps on
    a date basis:
    ```{r}
    experiment(person = EX,
               variables = list("fitbit_daily" = c("sleepDuration")),
               measures = list("fitbit_intraday" = c("distance")),
               analysis = c("correlation"),
               time_var = c("date"))
    ```
  `l_plot` will generate the same result:
    ```{r}
        dataset <- create_dataset(person = EX,
                                  all_variables = list("fitbit_daily" = c("sleepDuration"),
                                                       "fitbit_intraday" = c("distance")), 
                                  time_var = c("date"))
        
        correlation_df <- correlation(dataset, person = EX, 
                                      variables = list("fitbit_daily" = c("sleepDuration")),
                                      measures = list("fitbit_intraday" = c("distance")),
                                      time_var = "date")
    ```


3.  ANOVA

    To create ANOVAs for the effect of sleep duration and steps on resting
    heart rate on a date basis:
    ```{r}
    experiment(person = EX,
               variables = list("fitbit_daily" = c("sleepDuration", "steps")),
               measures = list("fitbit_daily" = c("restingHeartRate")),
               analysis = c("anova"),
               time_var = c("date"))
    ```

4.  Compare Groups

    To compare across time variables not already defined, such as months, users
    can set up groups. Groups must be part of the Person instance, so it is 
    oftentimes easier to just use `compare_groups`, which instead allows users to 
    pass in new dataframes defining groupings.
    
    In this case, only data from January and February is
    in the sample instance of Person, so only two groups are compared:
    ```{r}
     dataset <- create_dataset(person = EX,
                          all_variables = list("util" = c("month"), 
                                               "fitbit_daily" = 
                                                 c("sleepDuration", 
                                                   "steps",
                                                   "restingHeartRate")), 
                          time_var = c("date"))

    indiv_months <- data.frame("month" = c("Jan", "Feb", "Mar", "Apr", "May",
                                          "Jun", "Jul", "Aug",
                                          "Sep", "Oct", "Nov", "Dec"),
                               "group" = c(1:12))
    
    compare_groups(dataset, person = EX, 
                         addl_grouping_assignments = list("indiv_months" = 
                                                            indiv_months), 
                         names_of_groupings = c("indiv_months"),
                         variables_to_compare = c("steps", "restingHeartRate"))
    ```

5.  Regression

    To run a regression of resting heart rate on steps ^ 2
    ```{r}
    experiment(person = EX,
                   variables = list("fitbit_daily" = c("steps")),
                   measures = list("fitbit_daily" = c("restingHeartRate")),
                   analysis = c("regression"),
                   time_var = c("date"))
    ```

## Shiny Application

We provide a sample Shiny application for visualization of the EX object's data.