---
title: 'postGGIR part 6: NonWear matrix in GGIR outputs '
author: "Wei Guo"
date_xxx
output:
  html_document:
    df_print: paged
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# {r, results="hide"} - The chunks is run but all results are hidden. The code shows in the doc,  
# {r, include=FALSE} - the code is run but neither the code or the results are shown
# {r, echo=FLASE} - The code is not shown, but results are

```


```{r ,include=FALSE  }
currentdir_xxx 
studyname_xxx
bindir_xxx
outputdir_xxx 
epochIn_xxx
epochOut_xxx 
flag.epochOut_xxx 
log.multiplier_xxx 

```

 
## Non-wear time Matrix for `r studyname` data
      
The non-wear data was saved in M-metalong-timestamp and M-metalong-nonwearscore under R data of /meta/basic/meta----.bin.RData. GGIR output the matrix to clarify when data was imputed for each long epoch time window and the reason for imputation. NonWear time was summarized in 15 minutes window.
Value = 1 indicates imputation. 

   +  column 1 = monitor non wear
   +  column 2 = clipping  
   +  column 3 = additional nonwear  
   +  column 4 = protocol based exclusion  
   +  column 5 = sum of column 1,2,3 and 4. 

This is a program to read nonwear matrix and reformat it for JIVE.

   + 1) Change 15 minutes data to 1 minute level by carry 15 minutes forword    
   + 2) Carry forward for replace NAs between two nonwear=1,2,3,4,5 values
   + 3) Redefine nonwear =1 for those 2-5 values

Complications:

   +  For 9:12/97:100th  columns, we might have duplicate columns due to time change such as tc_02:00:00, tc_02:15:00, tc_02:30:00,tc_02:45:00. We just take the better value with each pair.   
   
  
 

```{r chunk1,include=FALSE,results="hide"} 
 
setwd(currentdir) 
datadir<-paste(currentdir,"/data",sep="")

  
listFN<-list.files(datadir,recursive=TRUE)
idMFN <-listFN[ grep("IDMatrix.",listFN )]  
if (length(idMFN)>1) stop("please choose which idMFN should be used?") 
idM<-read.csv(paste("./data/",idMFN,sep=""),head=1,stringsAsFactors=F)
 
 
# find such as nonwearscore_Sydney_01_44_900s.csv files 
csvFN <-listFN[ grep(".csv",listFN )] 
nwFN <-intersect(csvFN[ grep("nonwearscore",csvFN)],
                 csvFN[ grep("900s",csvFN)])  
print(nwFN) 
if (length(nwFN)>1) stop("please choose which nonwear.data should be used?")  
nwFN.key<-gsub(900,epochOut,nwFN)
outFN<-paste("./data/JIVEraw_",nwFN.key,sep="") 
outFN2<-paste("./data/JIVEimpu_",nwFN.key,sep="") 
# Change 900s nonwearscore into 60s format

myser<-c(paste(0,0:9,sep=""),10:99)
timeSer24h_1s<-NULL 
for (t1 in 1:24)
for (t2 in 1:60)  
for (t3 in 1:60) 
timeSer24h_1s<-c(timeSer24h_1s,paste(myser[t1],myser[t2],myser[t3],sep=":"))
timeSer24h<-timeSer24h_1s[1:(24*60*60/epochOut)*epochOut-epochOut+1]
length(timeSer24h)  #2880 for 30s,1440 for 60s

pickmin<-function(x,y){x<-as.numeric(x)
                       y<-as.numeric(y)
                       Z<-NULL
                       for (i in 1:length(x)){
                       if (is.na(x[i]) & is.na(y[i])) Z[i]<-NA else Z[i]<-min(c(x[i],y[i]),na.rm=TRUE)} 
                       return(Z)}
timestamp.switch<-function(x="X00.00.00") {x<-gsub("X","",x) 
                      return(paste(unlist(strsplit(x,"\\.")),collapse=":")) }  

n1440=24*60*60/epochOut
n15=900/epochOut 
```

# Read `r nwFN` file and merge with ID matrix

```{r chunk2}

data<-read.csv(paste("./data/",nwFN,sep=""),head=1,stringsAsFactors=F) 
temp<-gsub("meta_","",data[,"filename"])
data["filename"]<-gsub(".RData","",temp)  
nonwear.mer<-merge(idM,data[,-3],by=c("filename","Date"),all.x=FALSE,all.y=TRUE)
dim(nonwear.mer)
Ctop<-which(colnames(nonwear.mer)=="X00.00.00")-1
  
# nonwear matrix, NW, 96 columns for normal data, 100 columns for time change day; 00:00:00 00:15:00 ... 23:45:00 
NW<-nonwear.mer[,(Ctop+1):ncol(nonwear.mer)] 
dim(NW)
if (ncol(NW)==96+4) {
for (j in 9:12) { temp<-pickmin(NW[,j],NW[,j+88]) 
                  NW[,j]<-temp} 
NW<-NW[,1:96]
} else {
print(paste(c("NO time change columns, dim=",dim(NW)),collapse=" "))
if (ncol(NW)!=96) stop ("check nonwear matrix: not 96 columns")
}

NW[NW>=1]<-1 
colnames(NW)<- unlist(lapply(colnames(NW), timestamp.switch)) 
if (sum(order(colnames(NW)) != 1:96)!=0) stop ("nonwear columns are not increasing..") 
nw.NAs.b<-rowSums(is.na( NW)) *n15  #number of NAs in NW matrix,900s->60s
```

# Plot nonwear status for those with nonwear input 

Among `r nrow(NW)` lines, there are `r length(which(nonwear.mer[, "NonWearMin"]>=1))` lines with Nonwear input. The profit of nonwear status is plotted as below.

```{r c3plot,echo=FALSE, fig.width=10, fig.height=8}
  
 
for (f in 0:1){ 
  S0<-which(nonwear.mer[, "NonWearMin"]>=1 & nonwear.mer[, "remove16h7day"]==f)   
  Data3<-NW[S0,1:96]
  #print(dim(Data3))
  #par(mai=c(1.02,1.52,0.82,0 )) # bottom,left,up,right
  par(las=1)
  col.light = rep(rainbow(7),100000)
  x.range = 1:ncol(Data3)
  plot(x.range,Data3[1,],type="n",ylim=c(0,nrow(Data3)),xaxt="n",  ylab =paste("remove16h7day=",f,sep=""),xlab="Hour",col=col.light,main=paste("Nonwear for ",length(S0),"/",nrow(NW),sep="") )
 
  
  for(i in  1:nrow(Data3)) {
  yi<-nrow(Data3)-i+1 
  S<- which(Data3[i,]==1)  
  points(x=S,y=rep(yi,length(S)),pch=20,cex=0.17,col=col.light[i]) 
  } 
  mylabels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "23:00")
  mylabels.at<-which(colnames(Data3) %in% paste(mylabels,":00",sep=""))
  mylabels = substr(colnames(Data3)[mylabels.at],1,5) 
  axis(1, at = mylabels.at,labels = mylabels)
} 
 
```
 
  

# Expand to `r epochOut` seconds level and Carry forward for nonwear!=NA

```{r c5expand}
jive<-array(NA,dim=c(nrow(NW),n1440)) 
print(dim(NW)) 
for (j in 1:ncol(NW)){
 for (k in 1:n15)  jive[,(j-1)*n15+k]<-NW[,j]      
} 
colnames(jive)<-timeSer24h
jive<-cbind(nonwear.mer[,1:Ctop],nw.NAs.b,jive) 
write.csv(jive,file=outFN,row.names=F)
Ctop<-which(colnames(jive)=="00:00:00")-1
  

# 2) How to handle NAs? non wear score observation carried forward (refer to haochang's old code)  
#  As a result, 0---NA---0 or 0--NA was change to 0--------0, i.e NA was replaced by earlier score.
jive2<-jive 
for (i in 1:nrow(jive)){ 
  S2<-setdiff(which(!is.na(jive[i,])),1:Ctop) # NonMissingIndex   
  if (length(S2)>=1) {gap<-S2[-1]-S2[-length(S2)] 
  if (  length(setdiff(gap,1))>=1   ){ #found NA in middle/end
  #print(table(gap))
  for(j in 1:length(S2)){
    if (j != length(S2)){
      jive2[i,(S2[j]+1):(S2[j+1])-1] <- jive[i,S2[j]]
    }else{
      if (S2[j]!= ncol(jive))  jive2[i,(S2[j]+1):ncol(jive)] <- jive[i,S2[j]]  #if last nonNA= last column then do nothing
    } 
  }  
  print(paste(c(i,length(S2),jive2[i,1:3],j,S2[j],":", S2[which(gap>1)+(-1):1]) ,collapse=" "))
}}}

```
Note. We "fix" NAs for nonwear by replacing NA with previous score (earlier score carry forward)

 
# Plot nonwear status after carry forward NonWear scores

```{r c6nas,echo=FALSE, fig.width=10, fig.height=5}
 
nw.NAs<-rowSums(is.na( jive2[,-(1:Ctop)])) 
 
for (f in 0:1){ 
  if (f==0) S0<-which(jive2[, "NonWearMin"]>=1 & jive2[, "remove16h7day"]==0)   
  if (f==1) S0<-which(nw.NAs>=1 & jive2[, "remove16h7day"]==0)  
  Data3<-jive2[S0,Ctop+1:n1440]
  #print(dim(Data3))
  #par(mai=c(1.02,1.52,0.82,0.42)) # bottom,left,up,right
  par(las=1)
  col.light = rep(rainbow(10),100000)
  x.range = 1:ncol(Data3)
  plot(x.range,Data3[1,],type="n",ylim=c(0,nrow(Data3)),xaxt="n",  ylab =paste("remove16h7day=",0,sep=""),xlab="Hour",col=col.light,main=ifelse(f==0,paste("Nonwear for ",length(S0),"/",nrow(NW),sep=""),"Check NAs") ) 
  
  for(i in  1:nrow(Data3)) {
  yi<-nrow(Data3)-i+1 
  if (f==0) S<- which(Data3[i,]==1)  
  if (f==1) S<- which(is.na(Data3[i,]))  

  points(x=S,y=rep(yi,length(S)),pch=20,cex=0.15,col=col.light[i]) 
  } 
  mylabels = c("01:00", "04:00", "08:00", "12:00", "16:00", "20:00", "23:00")
  mylabels.at<-which(colnames(Data3) %in% paste(mylabels,":00",sep=""))
  mylabels = substr(colnames(Data3)[mylabels.at],1,5) 
  axis(1, at = mylabels.at,labels = mylabels)
} 
 
```


# Handle NAs in the head and tail for those incomplete missing lines  
    + replace head NAs with 1 in the head, treat them as nonwear in JIVE  
    + replace tail NAs with the last nonNA score in the tail     
    + wearMatrix = 1- nonwearMarix in JIVE

```{r c7nas}

jive3<-jive2  
for (i in 1:nrow(jive3)){ 
  S2<-setdiff(which(!is.na(jive3[i,])),1:Ctop) # NonMissingIndex 
  if (  length(S2)<1440 ){  
  #gap<-S2[-1]-S2[-length(S2)]
  gap<-diff(S2)
  if (  length(S2)>=1 & (length(setdiff(gap,1))>=1  ) )  stop("Check nonconsective Nonwear scores")
  #print(c(i,length(S2)))
  
  if (S2[1]>Ctop+1) jive3[i,(Ctop+1):(S2[1]-1)]<-1
  if (S2[length(S2)]<ncol(jive3) ) jive3[i, (S2[length(S2)]+1):ncol(jive3)]<-jive3[i, S2[length(S2)]]
}
} 
nw.NAs<-rowSums(is.na( jive3[,-(1:Ctop)]))
table(nw.NAs)
NonWearMin.a<-rowSums(jive3[,-(1:Ctop)]==1) 
jive3<-cbind(jive3[,1:Ctop],NonWearMin.a,jive3[,-(1:Ctop)])
write.csv(jive3,file=outFN2,row.names=F)
```

# NonWear plot after "imputation" for NAs

```{r c8plot,echo=FALSE, fig.width=10, fig.height=8.6}
par(mfrow=c(2,2))
boxplot(nw.NAs.b~remove16h7day,jive3,main="Number of NAs before imputation",col=c("blue","red"))
plot(x=1:nrow(jive3),jive3[,"nw.NAs.b"],col="red",ylab="Number of NAs",xlab="ids x days",main="Number of NAs before imputation")
S2=which(jive3[,"remove16h7day"]==0)
points(x=S2,y=jive3[S2,"nw.NAs.b"],col="blue")


# .b = before impu and .a=after impu 
boxplot(NonWearMin.a~remove16h7day,jive3,main="Nonwear minutes after imputation",ylab="NonWearMinutes",col=c("blue","red"))
plot(x=1:nrow(jive3),jive3[,"NonWearMin.a"],col="red",ylab="NonWearMinutes",xlab="ids x days",main="Nonwear minutes after imputation")
S1=which(jive3[,"remove16h7day"]==0)
points(x=S1,y=jive3[S1,"NonWearMin.a"],col="blue")  
```

Note: 

 +  red points are for "removed" samples with   remove16h7day=1        
 +  blue points are for "kept" samples with   remove16h7day=0
   

The nonwear matrix was written into `r outFN2`.  
 
 
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

How to run this? 



 