---
title: "spatialfusion: short demo"
author: "Craig Wang (craig.wang@math.uzh.ch)"
date: "June 21, 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{spatialfusion: short demo}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, fig.align = "center")
Sys.setenv("OMP_THREAD_LIMIT" = 2)
```
This brief demo provides code and output for fitting spatial fusion models using R package **spatialfusion**. The first section analyze the built-in synthetic dataset with INLA implementation while the second section analyze a simulated dataset with Stan implementation. The `method` argument in `fusionData()` function decides on which implementation to use.

# 1. Spatial fusion modelling with INLA on built-in synthetic data

## Load libraries
```{r}
library(spatialfusion)
library(tmap, quietly = TRUE)
library(sp, quietly = TRUE)
library(sf, quietly = TRUE)
```

## Load and view built-in synthetic data

```{r}
summary(dataGeo)
summary(dataLattice)
summary(dataPP)
```

## Plot data

```{r, message = FALSE, out.width = '50%'}
tm_shape(dataLattice) + tm_polygons(col = "white") +
  tm_shape(dataGeo) + tm_dots(size = 0.1) +
  tm_add_legend(type = "symbol", shape = 16, size = 0.3, col = "black", label = "geostatistical") +
  tm_shape(dataPP) + tm_symbols(col = "red", shape = 4, size = 0.02) +  
  tm_add_legend(type = "symbol", shape = 4, size = 0.2, col = "red", label = "point pattern") +
  tm_layout(main.title = "dataGeo, dataPP", main.title.size = 1, 
            frame = F, fontface = 2, legend.outside = T)
```

```{r, message = FALSE, out.width = '50%'}
tm_shape(dataLattice) +
  tm_fill(col = "mr", style = "fixed", breaks = c(0, 0.5, 2, 10, 30),
          title = "Mortality rate \n(per thousand)", legend.reverse = T) + tm_borders() +
  tm_layout(main.title = "dataLattice", main.title.size = 1, frame = F, fontface = 2,
            legend.position = c(0.77,0.8), legend.text.size = 0.5, legend.title.size = 0.5)
```

## Data preparation
```{r, message = FALSE}
dat <- fusionData(geo.data = dataGeo, geo.formula = lungfunction ~ covariate,
                  lattice.data = dataLattice,
                  lattice.formula = mortality ~ covariate + log(pop),
                  pp.data = dataPP, distributions = c("normal", "poisson"),
                  method = "INLA")

dat
```

## Fit a spatial fusion model
```{r}
if (require(INLA)){
mod <- fusion(data = dat, n.latent = 1, bans = matrix(c(0,0,0), ncol = 1),
              pp.offset = 400, prior.range = c(0.1, 0.5),
              prior.sigma = c(1, 0.5),  mesh.locs = dat$locs_point,
              mesh.max.edge = c(0.05, 0.5))
}
```

## Inspect the fit
```{r, fig.width = 8, fig.height = 5}
if (require(INLA)){
mod_fit <- fitted(mod, type = "link")
oldpar <- par(mfrow = c(1,2))
plot(dataGeo$lungfunction, mod_fit$point1,
     xlab = "Observed", ylab = "Fitted")
abline(0,1)
plot(log(dataLattice$mortality), mod_fit$area1,
     xlab = "Observed", ylab = "Fitted")
abline(0,1)
par(oldpar)
}

```

## Check parameter estimates
```{r}
if (require(INLA)){
summary(mod, digits = 3)
}
```

## Diagnostic plots
```{r, message = FALSE}
if (require(INLA)){
plot(mod, interactive = FALSE)
}
```

```{r, message = FALSE, out.width = '50%'}
if (require(INLA)){
plot(mod, posterior = FALSE)
}
```

## Predict latent surface
```{r, out.width = '50%'}
if (require(INLA)){
pred.locs <- st_as_sf(spsample(as_Spatial(dataDomain), 20000, type = "regular"))
mod.pred <- predict(mod, pred.locs)
mod.pred <- st_set_crs(mod.pred, "+proj=longlat +ellps=WGS84")
tm_shape(mod.pred) +
  tm_symbols(col = "latent.s11", shape = 15, size = 0.05, style = "cont",
             midpoint = NA, legend.col.reverse = TRUE, palette = "Oranges",
             title.col = "Latent process") +
  tm_shape(dataLattice) + tm_borders() +
  tm_layout(frame = FALSE, legend.outside = TRUE)
}
```

# 2. Spatial fusion modelling with Stan on simulated data

## Simulate data

```{r, message = FALSE}
dat <- fusionSimulate(n.point = 200, n.area = 30, n.grid = 5, n.pred = 100,
                      psill = 1.5, phi = 1, nugget = 0, tau.sq = 0.2,
                      dimension = 10, domain = NULL, point.beta = list(rbind(1,5)),
                      area.beta = list(rbind(1.5, 1.5)), nvar.pp = 1,
                      distributions = c("normal","poisson"),
                      design.mat = matrix(c(2, 0.5, 1), ncol = 1), 
                      pp.offset = 0.5, seed = 1)

geo.data <- st_as_sf(data.frame(y = dat$mrf[dat$sample.ind,"x"],
                       x = dat$mrf[dat$sample.ind,"y"],
                       cov.point = dat$data$X_point[,2],
                       outcome =dat$dat$Y_point[[1]]), 
                     coords = c("x","y"),
                     crs = st_crs("+proj=longlat +ellps=WGS84"))
            
                               
lattice.data <- cbind(dat$poly,
                     (data.frame(outcome = dat$data$Y_area[[1]],
                                 cov.area = dat$data$X_area[,2])))
lattice.data <- st_set_crs(lattice.data, "+proj=longlat +ellps=WGS84")

pp.data <- dat$data$lgcp.coords[[1]]
pp.data <- st_set_crs(pp.data, "+proj=longlat +ellps=WGS84")
```

## Plot data

```{r, out.width = '45%'}
tm_shape(lattice.data) + tm_polygons(col = "white") + 
  tm_shape(geo.data) + tm_dots(size = 0.1) +
  tm_add_legend(type = "symbol", shape = 16, size = 0.3, col = "black", label = "geostatistical") +
  tm_shape(pp.data) + tm_symbols(col = "red", shape = 4, size = 0.02) +
  tm_add_legend(type = "symbol", shape = 4, size = 0.2, col = "red", label = "point pattern") +
  tm_layout(main.title = "geo.data, pp.data", main.title.size = 1, 
            frame = FALSE, fontface = 2, legend.outside = TRUE)
```


```{r, out.width = '35%'}
tm_shape(lattice.data) + 
  tm_fill(col="outcome", style="fixed", breaks=c(0, 10, 50, 200, 1200),
          title = "Outcome", legend.reverse = T) + tm_borders() +
  tm_layout(main.title="lattice.data", main.title.size = 1, frame = FALSE, fontface = 2,
            legend.position = c(0.77,0.8), legend.text.size = 0.5, legend.title.size = 0.5)
```


## Data preparation
```{r}
dat2 <- fusionData(geo.data = geo.data, geo.formula = outcome ~ cov.point,
                   lattice.data = lattice.data, lattice.formula = outcome ~ cov.area,
                   pp.data = pp.data, distributions = c("normal","poisson"), 
                   method = "Stan")
dat2
```

## Fit a spatial fusion model
```{r, results = "hide"}
mod <- fusion(data = dat2, n.latent = 1, bans = matrix(c(0,0,0), ncol = 1),
              pp.offset = 0.5, prior.phi = list(distr = "normal", pars = c(1, 1)),
              nsamples = 1000, nburnin = 500, nchain = 1, ncore = 1)
```

## Inspect the fit
```{r, fig.width = 8, fig.height = 5}
mod_fit <- fitted(mod, type = "link")
oldpar <- par(mfrow = c(1,2))
plot(dat$data$Y_point[[1]], mod_fit$point1, xlab = "Observed", ylab = "Fitted")
abline(0,1)
plot(log(dat$data$Y_area[[1]]), mod_fit$area1, xlab = "Observed", ylab = "Fitted")
abline(0,1)
par(oldpar)
```

## Check parameter estimates
```{r}
summary(mod, digits = 2) 
```

## Diagnostic plots
```{r, message = FALSE}
plot(mod, interactive = FALSE)
```

```{r, message = FALSE}
plot(mod, posterior = FALSE) 
```

## Predict latent surface and compare with simulated truth
```{r, fig.width = 5, fig.height = 5}
mod.pred <- predict(mod, new.locs = dat$pred.loc, type = "summary")
oldpar <- par(mfrow = c(1,1))
plot(dat$mrf[dat$pred.ind, c("sim1")], mod.pred$latent1, 
     xlab = "Truth", ylab = "Predicted")
abline(0,1)
par(oldpar)
```

