---
title: "Asiaverse: A Metapackage for Asian Countries RESTful APIs and Curated Datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Asiaverse: A Metapackage for Asian Countries RESTful APIs and Curated Datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Asiaverse)
```

# Introduction 

The **`Asiaverse`** is a metapackage that brings together a comprehensive collection of R packages providing access to APIs functions and curated datasets from **China**, **Japan**, **South Korea**, **India**, and **Indonesia**. It integrates both real-time and historical data retrieved through public RESTful APIs, such as *Nager.Date*, *World Bank API*, and *REST Countries API*, while also offering extensive curated collections of open datasets covering economics, demographics, public health, environmental data, natural disasters, political indicators, and social metrics.  

The `Asiaverse` package installs and attaches five core subpackages. The first is **ChinAPIs**, which offers APIs functions and curated datasets for **China** and **Hong Kong**, including resources on air quality, demographics, input–output tables, epidemiology, political structure, and social indicators. The second is **JapanAPIs**, which provides APIs functions and curated datasets for Japan, with emphasis on natural disasters, economic production, the vehicle industry, air quality, demographics, and administrative divisions. The third is **SouthKoreAPIs**, a package focused on South Korea, covering public health outbreaks, social surveys, elections, economic indicators, natural disasters, climate data, energy consumption, cultural information, and financial markets. The fourth is **IndiAPIs**, which delivers APIs functions and curated datasets for India, offering comprehensive collections and real-time access to economic, demographic, and geopolitical indicators. Finally, **IndonesiAPIs** provides APIs functions and curated datasets for Indonesia, featuring data on holidays, economic indicators, consumer prices, poverty probability, food prices by region, tourism destinations, and minimum wage statistics.

## Installation

To install the `Asiaverse` package, use the following:

```{r install_package, eval=FALSE}

# Install from CRAN 
install.packages("Asiaverse")

# Then load the package:
library(Asiaverse)


```

## Using the Asiaverse() Function

Once the package is loaded, you can call the Asiaverse() function to display the list of included packages and their versions:

```{r asiaverse_function, eval=FALSE}

Asiaverse()

```


## Important Note on Detaching Packages

`Asiaverse` imports and depends on several subpackages. Therefore, you cannot detach an individual subpackage (like SouthKoreAPIs) while Asiaverse is still loaded.

### Example of an Error


```{r detach, eval=FALSE}

# This will raise an error
detach("package:SouthKoreAPIs", unload = TRUE)

```


## Correct Way to Detach

To properly unload a subpackage, you must first detach Asiaverse:

```{r detach_way, eval=FALSE}

# First detach the metapackage
detach("package:Asiaverse", unload = TRUE)

# Now you can safely detach the subpackage
detach("package:SouthKoreAPIs", unload = TRUE)

```

By installing the `Asiaverse` package this will attach the following packages to your R session:

- `ChinAPIs`

- `JapanAPIs`

- `SouthKoreAPIs`

- `IndiAPIs`

- `IndonesiAPIs`


## Included Packages in the Asiaverse

### ChinAPIs

The `ChinAPIs` package provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **China**. It allows users to retrieve up-to-date information on topics such as economic indicators, population statistics, unemployment rates, holidays, and basic geopolitical details.

In addition to API-access functions, the package includes one of the largest curated collections of open datasets related to China and Hong Kong. These datasets cover areas such as air quality, demographic indicators, input-output economic tables, epidemiology, administrative divisions, name distributions, political structure, and various social indicators.

#### Some ChinAPIs Functions

 - `get_country_info_cn()`: Get key geopolitical and administrative information about China from the REST Countries API.

- `get_china_gdp()`: Get China’s Gross Domestic Product (current US$) from the World Bank

- `get_china_cpi()`: Get China’s Consumer Price Index from World Bank

### Some ChinAPIs Datasets

- `bj_air_quality_tbl_df`: Tibble of hourly air pollution and weather measurements from Dongsi, Beijing.

- `china_admin_divisions_df`: Data frame containing the codes and names of China’s administrative divisions.

- `shanghai_factories_df`: Data frame containing data on per capita output of workers in 17 factories located in Shanghai.


### JapanAPIs

The `JapanAPIs` package provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **Japan**. It allows users to retrieve up-to-date or historical information on topics such as economic indicators, population statistics, national holidays, and basic geopolitical details.

In addition to API-access functions, the package includes one of the largest curated collections of open datasets related to Japan. These datasets cover a wide range of topics including natural disasters, economic production, the vehicle industry, air quality, demographic trends, and administrative divisions.

#### Some JapanAPIs Functions

- `get_japan_hospital_beds()`: Get Hospital Beds per 1,000 People in Japan from World Bank

- `get_japan_life_expectancy()`: Get Japan’s Life Expectancy at Birth from World Bank

- `get_japan_literacy_rate()`: Get Japan’s Literacy Rate (Age 15+) from World Bank

#### Some JapanAPIs Datasets

- `jpn_usd_exchange_df`: A Data frame containing weekly observations of the yen-dollar exchange rate from 1975 to 1989.

- `jpn_prefectures_tbl_df`: A tibble representing Japan’s 47 prefectures arranged in a 7x7 grid layout.

- `kobe_quake_1995_ts`: A time series containing data related to the 1995 Kobe earthquake.


### SouthKoreAPIs

The `SouthKoreAPIs` package provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **South Korea**. It allows users to retrieve up-to-date or historical information on topics such as economic indicators, population statistics, national holidays, and key geopolitical details.

In addition to API-access functions, the package includes one of the largest curated collections of open datasets related to South Korea. These datasets cover a wide range of topics including public health outbreaks, demographics, social surveys, elections, economic indicators, natural disasters, administrative divisions, air quality, climate data, energy consumption, cultural information, and financial markets.

#### Some SouthKoreAPIs Functions

- `get_southkorea_hospital_beds()`: Get South Korea’s Hospital Beds (per 1,000 people) data from the World Bank.

- `get_southkorea_life_expectancy()`: Get South Korea’s Life Expectancy at Birth data from the World Bank.

- `get_southkorea_unemployment()`: Get South Korea’s Total Unemployment Rate data from the World Bank.

#### Some SouthKoreAPIs Datasets

- `MERSKorea2015_list`: A list containing two data frames with information collected during the first weeks of the 2015 MERS-CoV outbreak in South Korea.

- `KoreanSocialSurvey_tbl_df`: A tibble containing a sample of data from the Korean General Social Survey (KGSS) conducted in 2023.

- `SeoulMosquito_tbl_df`: A tibble containing daily mosquito indicator data and weather measurements for Seoul, South Korea, from 2016 to 2019.

### IndiAPIs

The `IndiAPIs` package provides a unified interface to access open data from the World Bank API and the REST Countries API, with a focus on **India**. It allows users to retrieve up-to-date or historical information on topics such as economic indicators, international demographic statistics, and key geopolitical details related to India.

In addition to API-access functions, the package includes one of the largest curated collections of open datasets focused on India. These datasets cover a wide range of topics including population, economy, weather, politics, health, biodiversity, sports, agriculture, cybercrime, infrastructure, and more.

#### Some IndiAPIs Functions

- `get_india_child_mortality()`: Get India’s Under-5 Mortality Rate from World Bank.

- `get_india_cpi()`: Get India’s Consumer Price Index (2010 = 100) from World Bank.

- `get_india_energy_use()`: Get India’s Energy Use (kg of oil equivalent per capita) from World Bank.

#### Some IndiAPIs Datasets

- `BurdwanWeather_df`: A data frame containing weekly weather data for the rice growing season in the Burdwan district of West Bengal, India.

- `birds_watching_tbl_df`: A tibble containing detailed information on bird species observed in India, including species names, scientific names, the date of last observation.

- `DelhiPotatoPrices_ts`: A time series containing the monthly average potato prices of the Delhi market from January 2010 to July 2020.

### IndonesiAPIs

The `IndonesiAPIs` package provides a unified interface to access open data from the **World Bank API**, **Nager.Date API**, and the **REST Countries API**, with a focus on **Indonesia**. It allows users to retrieve up-to-date or historical information on topics such as economic indicators, population statistics, national holidays, and basic geopolitical details.

In addition to API-access functions, the package includes a curated collection of open datasets related to **Indonesia**. These datasets cover a wide range of topics including consumer prices, poverty probability, food prices by region, tourism destinations, and minimum wage statistics.

#### Some IndonesiAPIs Functions

- `get_indonesia_child_mortality()`: Get Under-5 Mortality Rate in Indonesia from World Bank

- `get_indonesia_cpi()`: Get Indonesia's Consumer Price Index from World Bank

- `get_indonesia_energy_use()`: Get Indonesia's Energy Use (kg of oil equivalent per capita) from World Bank

#### Some IndonesiAPIs Datasets

- `Bali_tbl_df`: A tibble containing food price data in Bali, Indonesia.

- `indonesia_minwage_tbl_df`: A tibble containing monthly minimum wage or Upah Minimum Regional (UMR) data in Indonesian Rupiah (IDR) across various regions of Indonesia from 1997 to 2025.

- `DKIJakarta_tbl_df`: A tibble containing food price data in DKI Jakarta, Indonesia.


### Summary of Included Packages

| Package        | Domain Focus |
|----------------|--------------|
| ChinAPIs       | Provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **China**. |
| JapanAPIs      | Provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **Japan**. |
| SouthKoreAPIs  | Provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **South Korea**. |
| IndiAPIs       | Provides a unified interface to access open data from the World Bank API and the REST Countries API, with a focus on **India**. |
| IndonesiAPIs   | Provides a unified interface to access open data from the World Bank API, Nager.Date API, and the REST Countries API, with a focus on **Indonesia**. |

## Conclusion

The **`Asiaverse`** metapackage represents a significant contribution to the R ecosystem by consolidating structured, reliable, and diverse data resources from five of Asia's most influential countries: **China, Japan, South Korea, India, and Indonesia**. By integrating access to real-time or historical information through RESTful APIs such as the *World Bank API*, *Nager.Date API*, and *REST Countries API*, alongside curated collections of open datasets, the package provides researchers, analysts, educators, and policymakers with a centralized platform for data-driven inquiry and comparative analysis.  

Each of the included subpackages — **ChinAPIs**, **JapanAPIs**, **SouthKoreAPIs**, **IndiAPIs**, and **IndonesiAPIs** — offers a thematic and geographic focus, making the data both country-specific and internationally comparable. Users can easily retrieve key indicators related to economics, demographics, health, politics, natural disasters, climate, and social metrics. Moreover, the curated datasets embedded within each subpackage extend beyond API access, ensuring that both historical and domain-specific data are readily available for reproducible research and educational use.


Ultimately, the **`Asiaverse`** metapackage advances the accessibility, transparency, and reproducibility of Asian data in R, serving as a bridge between open data initiatives and the global data science community.  

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>
