---
title: "Getting Information From Leanpub"
author: "John Muschelli"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Information From Leanpub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# leanpubr Package: 
The goal of `leanpubr` is to provide provides access to the 'Leanpub' 'API' <https://leanpub.com/help/api> for gathering information about publications and submissions to the 'Leanpub' platform.

## Setting the API key
The main thing you need to do is get an API key from Leanpub.  You can do this by visiting https://leanpub.com/user_dashboard/api_key if you are an author.  If you are not an author, I don't believe access to the API is possible.  You can pass this `api_key` to functions in Leanpub (almost all prefixed by `lp_`), or you can set the environmental variable `LEANPUB_API_KEY`.  You can set this in an interactive `R` session such as:

```r
Sys.setenv("LEANPUB_API_KEY" = "YOUR_KEY")
```

or put this in your `~/.Renviron`:

```r
LEANPUB_API_KEY = "YOUR_KEY"
```

## Getting Data about a book

You can use `lp_summary` or `lp_book_info`, these are identical, to get information about any book in Leanpub:

```{r, echo = FALSE}
Sys.setenv("LEANPUB_API_KEY" = "asdf")
```

```{r}
library(leanpubr)
slug = "biostatmethods"
res = lp_summary(slug, error = FALSE, verbose = TRUE)
res$content
```
