---
title: "Quick Tree Building"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Tree Building}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PhyInsight)
```

## Introduction

The `PhyInsight` package is used to efficiently conduct DNA sequence analyses. In this vignette, use of the package functions in order to create a phylogenetic tree will be demonstrated. The functions in this package are used to manipulate DNA strings, remove mismatched strings, and remove strings that cause distance outliers where necessary. These processes are performed automatically when the appropriate functions are used and can save the user much time and effort.

&nbsp;

Any taxon can be used provided the DNA strings are in the correct format. The Ursidae family will be used as an example of how one can use `PhyInsight` to manipulate DNA strings and build a phylogenetic tree in just four functions once the specimen dataframe is ready. 

&nbsp;

## Subset Data

```{r}

# query the data using the taxon name
specdata <- querySpecData("Ursidae")

# check the number of DNA marker codes
sort(table(specdata$markercode))

```

&nbsp;

Choosing which gene region(s) to use is an important step and will vary depending on the research question and the taxa. In this example, only sequences from the `COI-5P` region will be used for the analysis. Also, `getSpeciesRepr()` is used to automatically select representative observations for each species.

```{r, results = "hide"}

# subset results that only have nucleotides from the COI-5P region
specdata <- subset(specdata, markercode == "COI-5P")

# get one observation per species
specdata <- getSpeciesRepr(specdata)

```

&nbsp;

## Manipulate DNA Strings

Here, functions from the package are used to automatically create a DNA bin, create a DNA string set object, then automatically manipulate the strings.

```{r, results = "hide"}

# generate a DNA bin
DNABin <- genDNABin(specdata)

# use the DNA bin to create a DNA string set
DNAStringset <- genDNAStringSet(DNABin)

# automatically manipulate the DNA strings 
DNAStringSet <- ManipStringSet(DNAStringset)

```

&nbsp;

## Create Phylogenetic Tree

Here, the `genPhytree()` function is used to automatically create a phylo tree object. The labels names, offset, and size can also be edited.

```{r}

# automatically generate a phylo tree
PhyloTree <- genPhytree(DNAStringSet)


# change the label names
PhyloTree$tip.label <- paste(specdata$genus_name, "|", specdata$species_name)

```

&nbsp;

## Plot Phylogenetic Tree

```{r, fig.show='hold', fig.width=16, fig.height=14}
# plot the phylo tree
plot(
  PhyloTree,
  label.offset = 0.001,
  cex = 1.5
)

```


