---
title: "Automated String Clean Up"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated String Clean Up}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PhyInsight)
```

## Introduction

The `rmBadStrings` functions (`rmBadStrings_1()`, `rmBadStrings_2()`, and  `rmBadStrings_3()`) can be used to automatically clean a DNA string set that would otherwise be unsuitable for analysis. These functions use slightly different methods to automatically remove strings that are mismatched with others and those whose distances return `NaN` values or are considered outliers.

&nbsp;

## Subset Data

```{r, results = "hide"}
# query the data using the taxon name
specdata <- querySpecData("Nepenthes")

# subset results that only have nucleotides from the matK region
specdata <- subset(specdata, markercode == "matK")

# get one observation per species
specdata <- getSpeciesRepr(specdata)

```

&nbsp;

## Manipulate DNA Strings

```{r, results = "hide"}
# generate a DNA bin
DNABin <- genDNABin(specdata)

# use the DNA bin to create a DNA string set
DNAStringset <- genDNAStringSet(DNABin)

# automatically manipulate the DNA strings 
DNAStringSet_manip <- ManipStringSet(DNAStringset)

```

&nbsp;

## String Clean Up

At this point attempting to create a phylo tree will result in an error. Using `DECIPHER::BrowseSeqs(DNAStringSet_manip)` to view the string set will reveal a series of mismatched and fragmented strings. Using `rmBadStrings_1()` will automatically remove all mismatched strings. After which, a tree can be created and and plotted.

```{r}
# use the function to remove unsuitable strings and store into an object
BadStringsRemoved <- rmBadStrings_1(
  DNAStringSet = DNAStringSet_manip,
  specimen_dataframe = specdata
)

```

## Store New Objects

```{r}
DNAStringSet_new <- BadStringsRemoved[[1]]
specdata_new <- BadStringsRemoved[[2]]

```

## Create Phylogenetic Tree

```{r, fig.show='hold', fig.width=16, fig.height=14}
# automatically generate a phylo tree
PhyloTree <- genPhytree(DNAStringSet_new)

# change the label names
PhyloTree$tip.label <- specdata_new$species_name

# plot the phylo tree
plot(
  PhyloTree,
  label.offset = 0.0001,
  cex = 1
)

```
  
&nbsp;
&nbsp;
&nbsp;

# Remove Outliers

The `rmBadStrings` functions also have optional arguments to remove strings whose DNA distances are considered outliers. `rmOutliers` is a logical argument that when set to `TRUE` will automatically remove outliers as well 
as performing its regular processes. `max_Z_score` is a numerical value that allows the user to change the maximum Z score for each string's DNA distance. The default value for this argument is 3 as a score higher than this is generally considered an outlier. 

## Remove Outliers (Below 3 Z-Score)

```{r}
# use the function to remove unsuitable strings and remove outliers
BadStringsRemoved <- rmBadStrings_1(
  DNAStringSet = DNAStringSet_manip,
  specimen_dataframe = specdata,
  
  rmOutliers = TRUE
)

```

## Store New Objects

```{r}
DNAStringSet_new <- BadStringsRemoved[[1]]
specdata_new <- BadStringsRemoved[[2]]

```

## Create Phylogenetic Tree

```{r, fig.show='hold', fig.width=16, fig.height=14}
# automatically generate a phylo tree
PhyloTree <- genPhytree(DNAStringSet_new)

# change the label names
PhyloTree$tip.label <- specdata_new$species_name

# plot the phylo tree
plot(
  PhyloTree,
  label.offset = 0.0001,
  cex = 1
)

```

## Remove Outliers (Below 2 Z-Score)

```{r}
# use the function to remove unsuitable strings and remove outliers
BadStringsRemoved <- rmBadStrings_1(
  DNAStringSet = DNAStringSet_manip,
  specimen_dataframe = specdata,
  
  rmOutliers = TRUE,
  
  max_Z_score = 2
)

```

## Store New Objects

```{r}
DNAStringSet_new <- BadStringsRemoved[[1]]
specdata_new <- BadStringsRemoved[[2]]

```

## Create Phylogenetic Tree

```{r, fig.show='hold', fig.width=16, fig.height=14}
# automatically generate a phylo tree
PhyloTree <- genPhytree(DNAStringSet_new)

# change the label names
PhyloTree$tip.label <- specdata_new$species_name

# plot the phylo tree
plot(
  PhyloTree,
  label.offset = 0.0001,
  cex = 1
)

```

