---
title: "scale_duke_fill_discrete()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{scale_duke_fill_discrete()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette is intended to demonstrate the functionality of `scale_duke_fill_discrete()`. The function allows users to utilize a discrete color scale of Duke official branding colors on ggplot2 visualizations. 
It is fully integrated with ggplot2 visualizations.

# Plot Examples

For these visualizations, we will use the `penguins` dataset from the **palmerpenguins** package.

```{r penguins, warning = FALSE}
library(palmerpenguins)
library(duke)
library(ggplot2)
library(ggmosaic)
```

## Bar Plot

```{r bar plot, warning = FALSE}
plot <- ggplot(palmerpenguins::penguins, aes(x = species, fill = species)) +
  geom_bar() +
  labs(title = "Distribution of Species", caption = "(Colors used) \n Duke Royal Blue, Duke Navy Blue, Copper", x = "Species", y = "Count")

plot +
  scale_duke_fill_discrete()
```

## Histogram Plot

```{r histogram plot, warning = FALSE}
plot2 <- ggplot(palmerpenguins::penguins, aes(x = bill_length_mm)) +
  geom_histogram(aes(fill = species)) +
  labs(title = "Distribution of Bill Length", caption = "(Colors used) \n Duke Royal Blue, Duke Navy Blue, Copper", x = "Bill Length (mm)", y = "Count")

plot2 +
  scale_duke_fill_discrete()
```
