---
title: "Comparing Distributions with the `distfreereg` Package"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Comparing Distributions with the `distfreereg` Package}
  %\VignetteEngine{knitr::rmarkdown_notangle}
  %\VignetteEncoding{UTF-8}
linkcolor: blue
link-citations: true
csl: american-statistical-association.csl
bibliography: bibliography.bib
---



\def\eqdef{\mathrel{\raise0.4pt\hbox{$:$}\hskip-2pt=}}

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, out.width = "75%", fig.dim = c(6,5),
                      fig.align = "center")
is_check <- ("CheckExEnv" %in% search()) ||
  any(c("_R_CHECK_TIMINGS_", "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(eval = !is_check)

library(distfreereg)
```

<!-- https://kbroman.org/pkg_primer/pages/vignettes.html -->






# Comparing Observed and Simulated Statistics

The procedure implemented by `distfreereg()` relies on asymptotic behavior for
accurate results. In particular, when the assumptions of the procedure are met,
the asymptotic distribution of the observed statistic is equal to the
distribution of the simulated statistics. How large the sample size must be for
this asymptotic behavior to be sufficiently achieved is dependent on the details
of each application. To facilitate the exploration of sample size requirements,
`distfreereg` provides the function `compare()`.

In the following example, we explore the required sample size when the true mean
is $$f(X;\theta) = \theta_1 + \theta_2X_1 + \theta_3X_2,$$ $\theta=(2,5,-1)$,
and the errors are independent standard normal errors. The covariate matrix is
generated randomly. We start with a sample size of 10.

```{r comp_1, message = FALSE}
set.seed(20240913)
n <- 10
func <- function(X, theta) theta[1] + theta[2]*X[,1] + theta[3]*X[,2]
theta <- c(2,5,-1)
X <- matrix(rexp(2*n, rate = 1), nrow = n)
comp_dfr <- compare(theta = theta, true_mean = func, test_mean = func,
                    true_X = X, true_covariance = list(Sigma = 3), X = X,
                    covariance = list(Sigma = 3),
                    theta_init = rep(1, length(theta)))
```

This function generates a response vector `Y` from `true_mean` using the given
values of `true_X`, `true_covariance` (by default, the errors are multivariate
normal), and `theta`. It then passes `Y` and most of the remaining arguments to
`distfreereg()`. The observed statistics and p-values corresponding to each
simulated data set are saved, and the process is repeated. The results are
returned in an object of class `compare`.




# Plotting the Results

These results can be explored graphically. Several options are available. The
default behavior of the `compare` method for `plot()` produces a comparison of
estimated cumulative distribution functions (CDFs) for the observed and
simulated statistics. If the sample size is sufficiently large, these two curves
should be nearly identical.

```{r cdf}
plot(comp_dfr)
```

This plot gives some cause for concern.

Another way of exploring this uses quantile--quantile plots. Below is a Q--Q
plot that compares observed and simulated statistics. If the sample size is
sufficiently large, these points should lie on the line $y=x$.

```{r qq}
plot(comp_dfr, which = "qq")
```

The curvature indicates an insufficient sample size. Similar to this is the next
plot, which is a Q--Q plot comparing p-values to uniform quantiles. Once again,
if the sample size is sufficiently large, these points should lie on the line
$y=x$.

```{r qqp}
plot(comp_dfr, which = "qqp")
```

These plots all indicate that this sample size is insufficient for the
asymptotic behavior for both statistics.

More details on this plot method are discussed
[here](../doc/v3_plotting.html).


# Formal Comparison Testing

The `compare` method for `ks.test()`, which compares the observed and simulated
distributions, confirms that they are still different:

```{r ks.test}
ks.test(comp_dfr)
```

The default statistic is whatever is the first statistic appearing in the
supplied object:

```{r}
names(comp_dfr$obs)
ks.test(comp_dfr, stat = "CvM")
```




# Increasing the Sample Size

Let us repeat the simulation with a sample size of 100.
```{r comp_larger_n, message = FALSE}
n_2 <- 100
X_2 <- matrix(rexp(2*n_2, rate = 1), nrow = n_2)
comp_dfr_2 <- update(comp_dfr, true_X = X_2, X = X_2)
```

The following plots indicate that this sample size is sufficient.

```{r qqplots_2}
plot(comp_dfr_2)
plot(comp_dfr_2, which = "qq")
plot(comp_dfr_2, which = "qqp")
```

A formal test confirms this assessment.

```{r ks.test_2}
ks.test(comp_dfr_2)
```



# Power Considerations

The examples using `compare()` above all illustrate behavior of the function
when the null hypothesis is true. It is also important to investigate the
behavior when it is false to learn about the test's power against particular
alternative hypotheses. Using the example in [Comparing Observed and Simulated
Statistics], consider the case in which the true mean function is quadratic in
$X_2$: $f(x;\theta) = \theta_0 + \theta_1x_1 + \theta_2x_2 + x_2^2$. Having
verified above that a sample size of 100 is sufficient for the asymptotic
behavior to be sufficiently approximated, this can be investigated as follows.

```{r power, message = FALSE}
alt_func <- function(X, theta) theta[1] + theta[2]*X[,1] + theta[3]*X[,2] + 0.5*X[,2]^2
comp_dfr_3 <- update(comp_dfr_2, true_mean = alt_func, true_covariance = list(Sigma = 3))
```

```{r}
plot(comp_dfr_3)
```

As hoped, these curves are clearly different. Power estimates can be obtained
with `rejection()`.

```{r}
rejection(comp_dfr_3)
```

Power for different significance levels can be found by changing the `alpha`
argument.

```{r}
rejection(comp_dfr_3, alpha = 0.01)
```





