---
title: "IN-vivo reSPonsE Classification of Tumours ('INSPECTumours')"
author: ""
date: "`r Sys.time()`"
output: html_document
params:
  r: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
		echo = FALSE, 
		message = FALSE,
		warning = FALSE,
		results = "asis"
)
```
##### Application Version: `r packageVersion("INSPECTumours")`

# 1. Load Data

## 1.1 Data information 

#### Loaded Data File(s): 
```{r file names}
shiny::tagList(lapply(params$r$file_names, p))
```

```{r summary new_data}
knitr::kable(params$r$summary_new_data)
```

```{r summary control}
if (is.null(params$r$summary_control)) {
  cat("No historical control data")
} else {
  knitr::kable(params$r$summary_control)
}
```


## 1.2	Quality Control 

```{asis, eval = !is.null(params$r$min_points)}
##### End day of modelling: `r params$r$min_points`
```


##### Excluded data:
```{r excluded data}
if (!is.null(params$r$excluded_data)) {
 df_grouped <- dplyr::group_by(params$r$excluded_data, study, treatment, animal_id, reason)
 df_sum <- dplyr::summarise(df_grouped, days = paste(unique(day), collapse = ", ")) 
 knitr::kable(df_sum)
} else {
cat("Nothing is excluded")
}
```

# 2. Analysis 
## 2.1 Analysis settings 

##### End day of modelling: `r params$r$mod_end_day`

##### Consecutive measurements per week:  `r params$r$con_meas`

### Model selection: `r params$r$model_type`

This is a model for all the control data, including historical control data, to explain the normal tumour growth profile without treatment.  


```{asis, echo = params$r$model_type == "Linear model"}
$$log_{10}(tumour.volume) \sim a + b \times time + (1|animal.id) + (1|study),$$
where $a$ and $b$ are parameters that correspond to the log initial volume and growth rate. The random effect terms $(1|animal.id)$ and $(1|study)$ are used to capture the animal-to-animal and study-to-study variability, respectively.
```

```{asis, echo = params$r$model_type == "Two-stage non-linear model"}
$$log_{10}(tumour.volume) \sim f(time) + (1|animal.id) + (1|study),$$
where a continuous hinged function is used to allow different growth rates for the early stage (“unstable” phase) and the late stage (“stable” stage) of the tumour growth:
$$f(time) = a + b_0 \times (time - t_{change}) + (b_1 - b_0) \times \delta \times log(1 + \frac{exp(time - t_{change})}{\delta}),$$
with parameters: 
	$a$: log initial tumour volume 
	$b_0$: growth rate of the unstable stage
	$b_1$: growth rate of the stable stage 
	$t_{change}$: time point that the growth rates change 
	$\delta$: smoothness of the curve 
```



## 2.2	Model output 

```{r control_growth plot}
params$r$p_control_growth
```


## 2.3	Tumour classification 

The 95% predication interval is derived from the model for the control data. The tumours that are higher than the lower bound of 95% prediction interval are considered as not significantly different from the “normal” tumour growth. For an individual tumour, if there are m=consecutive measurements that are smaller than the lower bound of 95% prediction interval, the tumour is considered as “Responder”. Otherwise, the tumour is classified as “Non-responder”. 

All responders are classified into sub-categories based on the tumour growth rates. If the growth rate is not significantly different from 0, the tumour is classified as “Stable responder”. We consider a tumour with growth rate significantly larger than 0 as a “Modest responder”, which shows a response to the treatment although the tumour still growth slowly. A strong drug effect is concluded if the tumour growth rate is significantly smaller than 0 and the tumour is classified as “Regressing responder”. 


### Classification over tumour volume
```{r plot of the tumour classification}
invisible(lapply(params$r$classification_tv_plot, print))
```

### Classification over growth rate
```{r classification_gr}
params$r$classification_gr_plot
```

### Waterfall plot 
```{r waterfall}
invisible(lapply(params$r$classification_waterfall_plot, print))
```

```{asis, echo = params$r$stat_analysis_done}
## 2.4	Statistical test for tumour classification 
### 2.4.1 Animal numbers for tumour classification 
```


```{r df_animal_number, eval = params$r$stat_analysis_done}
knitr::kable(params$r$df_animal_number)
```


```{r plot_animal_number, eval = params$r$stat_analysis_done}
params$r$plot_animal_number
```



```{asis, echo = params$r$stat_analysis_done}
### 2.4.2 Ordered logistic model
By converting the tumour classification outcomes into ordinal data, we are able to assess the drug efficacy by using an ordered logistic regression: 
$$y^* \sim treatment + (1|study),$$
where $y^*$ is a latent continuous variable defined by the following: 

$$
Tumour.classification =  
\begin{cases}
\textit{0 (non-responder), if } \; y^* \leq \mu_1 \\
\textit{1 (modest responder), if} \; \mu_1 < y^* \leq \mu_2 \\
\textit{2 (stable responder), if} \; \mu_2 < y^* \leq \mu_3 \\
\textit{3 (regressing responder) other}\\
\end{cases}
$$
```

```{asis, echo = params$r$stat_analysis_done}
##### Drug efficacy
```

```{r df_drug_efficacy, eval = params$r$stat_analysis_done}
knitr::kable(params$r$df_efficacy)
cat("Asterisks indicate that the drug was considered significantly effective by comparing to the control.")
```

```{asis, echo = params$r$stat_analysis_done}
##### Classification categories
```

```{r df_categories, eval = params$r$stat_analysis_done}
knitr::kable(params$r$df_categories)
```

```{r plot_categories, eval = params$r$stat_analysis_done}
params$r$plot_categories
```


### Session Info
```{r session info}
pander::pander(sessionInfo())
```

