---
title: "GPCERF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPCERF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installation

```{r, eval=FALSE}
library("devtools")
install_github("NSAPH-Software/GPCERF", ref = "main")
```
```{r}
library("GPCERF")
```


## Usage

### Estimating GPS values

Input parameters:

**`cov_mt`** Covariate matrix containing all covariates. Each row represents a 
sample, and each column is a covariate.   
**`w_all`** A vector of observed exposure level.    

```{r}
mydata <- generate_synthetic_data()
gps_m <- estimate_gps(cov_mt = mydata[, c("cf1", "cf2", "cf3", "cf4",
                                         "cf5", "cf6")],
                  w_all = mydata$treat,
                  sl_lib = c("SL.xgboost"),
                  dnorm_log = FALSE)
```

### Estimating exposure response using Gaussian Processes

Input parameters:

**`data`** A data.frame of observation data.   
  - Column 1: Outcome (Y)   
  - Column 2: Exposure or treatment (w)   
  - Column 3~m: Confounders (C)   
**`w`** A vector of exposure level to compute CERF.   
**`gps_m`** A data.frame of GPS vectors.   
  - Column 1: GPS   
  - Column 2: Prediction of exposure for covariate of each data sample (e_gps_pred)   
  - Column 3: Standard deviation of e_gps (e_gps_std)   
**`params`** A list of parameters that is required to run the process.     
  - alpha: A scaling factor  for the GPS value.   
  - beta: A scaling factor for the exposure value.   
  - g_sigma: A scaling factor for kernel function (gamma/sigma).   
  - tune_app: A tuning approach. Available approaches:   
    - all: try all combinations of hyperparameters.   

alpha, beta, and g_sigma can be a vector of parameters. 

**`nthread`** An integer value that represents the number of threads to be used in a shared memory system.    
**`kernel_fn`** A kernel function. A default value is a Gaussian kernel.    

```{r}
set.seed(129)
sim_data <- generate_synthetic_data(sample_size = 400, gps_spec = 3)

# Estimate GPS function
gps_m <- estimate_gps(cov_mt = sim_data[, -(1:2)],
                      w_all = sim_data$treat,
                      sl_lib = c("SL.xgboost"),
                      dnorm_log = TRUE)

# exposure values
q1 <- stats::quantile(sim_data$treat, 0.05)
q2 <- stats::quantile(sim_data$treat, 0.95)

w_all <- seq(q1, q2, 1)

params_lst <- list(alpha = c(0.05),
                   beta = c(12),
                   g_sigma = c(0.1),
                   tune_app = "all")
cerf_gp_obj <- estimate_cerf_gp(sim_data,
                               w_all,
                               gps_m,
                               params = params_lst,
                               outcome_col = "Y",
                               treatment_col = "treat",
                               covariates_col = paste0("cf", seq(1,6)),
                               nthread = 1)
summary(cerf_gp_obj)
plot(cerf_gp_obj)
```

### Estimating exposure response using nearest-neighbor Gaussian Processes

Input parameters:

**`data`** A data.frame of observation data.   
**`w`** A vector of exposure level to compute CERF.   
**`gps_m`** A data.frame of GPS vectors.   
  - Column 1: GPS   
  - Column 2: Prediction of exposure for covariate of each data sample (e_gps_pred)   
  - Column 3: Standard deviation of e_gps (e_gps_std)   
**`params`** A list of parameters that is required to run the process.    
  - alpha: A scaling factor  for the GPS value.   
  - beta: A scaling factor for the exposure value.   
  - g_sigma: A scaling factor for kernel function (gamma/sigma).   
  - tune_app: A tuning approach. Available approaches:   
    - all: try all combinations of hyperparameters.   
  - n_neighbors: Number of nearest neighbors on one side.    
  - block_size: Number of samples included in a computation block. Mainly used to balance speed and memory requirement. Larger \code{block_size} is faster, but requires more memory.    

alpha, beta, and g_sigma can be a vector of parameters. 

**`outcome_col`** An outcome column name in `data`.
**`treatment_col`** A treatment column name in `data`.
**`covariates_col`** Covariates columns name in `data`.

**`nthread`** An integer value that represents the number of threads to be used in a shared memory system.    
**`kernel_fn`** A kernel function. A default value is a Gaussian kernel.    

```{r}
set.seed(19)
sim_data <- generate_synthetic_data(sample_size = 1000, gps_spec = 3)

# Estimate GPS function
gps_m <- estimate_gps(cov_mt = sim_data[, -(1:2)],
                      w_all = sim_data$treat,
                      sl_lib = c("SL.xgboost"),
                      dnorm_log = TRUE)
# exposure values
q1 <- stats::quantile(sim_data$treat, 0.05)
q2 <- stats::quantile(sim_data$treat, 0.95)

w_all <- seq(q1, q2, 1)

params_lst <- list(alpha = c(0.05),
                   beta = c(12),
                   g_sigma = c(0.1),
                   tune_app = "all",
                   n_neighbor = 20,
                   block_size = 1e4)
cerf_nngp_obj <- estimate_cerf_nngp(sim_data,
                                    w_all,
                                    gps_m,
                                    params = params_lst,
                                    outcome_col = "Y",
                                    treatment_col = "treat",
                                    covariates_col = paste0("cf", seq(1,6)),
                                    nthread = 1)

summary(cerf_nngp_obj)
plot(cerf_nngp_obj)
```

### Logging

The GPCERF package logs internal activities into the screen or `GPCERF.log` file. Users can change the logging file name (and path) and logging threshold. The logging mechanism has different thresholds (see [logger](https://cran.r-project.org/package=logger) package). The two most important thresholds are INFO and DEBUG levels. The former, which is the default level, logs more general information about the process. The latter, if activated, logs more detailed information that can be used for debugging purposes. The log file is located in the source file location and will be appended. 
