---
title: "IAT-example"
author: Ottavia M. Epifania
date: "`r Sys.Date()`"
bibliography: vignette.bib
output:  rmarkdown::html_vignette
#  pdf_document: default
vignette: >
  %\VignetteIndexEntry{IAT-example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  #fig.path = "",
  warning = FALSE, 
  message = FALSE
)
```

```{r setup}
library(implicitMeasures)
```

This vignette illustrates how to use the `implicitMeasures` package for computing the IAT *D* score. The illustration is based on the data set `raw_data` that comes with the package.

### First thing first: Import and explore data

Labels containing specification `.iat` in variable `blockcode` identify IAT blocks.

```{r}
data("raw_data")
# explore the dataframe
str(raw_data)
# explore the levels of the blockcode variable to identify the IAT blocks
levels(raw_data$blockcode)
```

Once the IAT blocks have been identified, it is possible to clean the IAT data by using the `clean_iat()` function. Since the data set also includes respondents' demographic information (`demo` in the `blockcode` variable), it is possible to extract and store these information in a separate data frame:  

```{r eval = T}
iat_cleandata <- clean_iat(raw_data, sbj_id = "Participant",
                          block_id = "blockcode",
                          mapA_practice = "practice.iat.Milkbad",
                          mapA_test = "test.iat.Milkbad",
                          mapB_practice = "practice.iat.Milkgood",
                          mapB_test = "test.iat.Milkgood",
                          latency_id = "latency",
                          accuracy_id = "correct",
                          trial_id = "trialcode",
                          trial_eliminate = c("reminder", "reminder1"),
                          demo_id = "blockcode",
                          trial_demo = "demo")
```

Since also the demographic data has been specified, `clean_iat()` results in a list of 3 elements: 

```{r}
str(iat_cleandata)
```

`data_keep` is a `data.frame` with class `iat_clean`. It contains the data set for the `compute_iat()` function. 

`data_eliminate` is a `data.frame` that contains all the discarded blocks and trials. 

`demo` is a `data.frame` that contains all the trials identified as `demo` in the `blockcode` variable.

Store the first `data_keep` element in a data frame for the `compute_iat()` function. 

```{r}
iat_data <- iat_cleandata[[1]]
head(iat_data)
```


### Compute IAT D score

Once that IAT data have been cleaned with the `clean_iat()` function, it is possible to compute the *D* score by using the `compute_iat()` function.

This function only takes two arguments. The first argument is the data frame with class `iat_clean`, the second argument is a character specifying the *D* score algorithm for the computation.  To compute multiple *D* score algorithms at the same time, use the`multi_dscore()` function. 

```{r}
dscore <- compute_iat(iat_data, Dscore = "d3")
str(dscore)
```

The `compute_iat()` function results in a `data.frame` with class `dscore` containing a number of rows equal to the number of participants. The columns contain the *D* score and otehr useful information on the performance  of each respondent (see the documentation of the `compute_iat()` function for further details). The `IAT_rel()`, `descript_d()`, `d_point()`, and `d_density()` functions  require the object resulting from function `compute_iat()`. 

The IAT D is computed as such that positive scores indicate slower response times in condition identified by Mapping B. 

### IAT descriptive statistics and reliability

The descriptive statistics of the *D* scores computed on the practice and test blocks, and the actual *D* scores can be easily obtained with the `descript_d()` function: 

```{r}
descript_d(dscore) # Data frame containing IAT Ds
```

By specifying `latex = TRUE`, function `descript_d()` print the results in LaTeX code: 

```{r}
descript_d(dscore, # Data frame containing IAT Ds
           latex = TRUE) # obtain the code for latex tables
```

The `IAT_rel()` function computes the reliability of the IAT by correlating the *D* score obtained from practice blocks with the *D* score obtained from test blocks [see @gaw2017 for further details]: 

```{r}
IAT_rel(dscore)
```

### Plotting the results

The `implicitMeasures` package comes with several functions for obtaining clear representations of the results at both individual respondent and sample levels. 
Additionally, it includes a function for computing and plotting multiple IAT *D* score algorithms at the same time. 

#### Individual respondent plot

The `d_point()` function plots the IAT *D* score for each respondent. 

```{r, fig.align='center', fig.width=6, fig.cap="d_point() function with default settings"}
d_point(dscore) # Data frame containing IAT D scores
```

In case of large sample size, the label identifying each respondent is difficult to read. It can be eliminated by setting `x_values = FALSE`. 

Respondents can be arranged by increasing or decreasing *D* scores by setting argument `order_sbj` equal to `"D-increasing"`  or `"D-decreasing"`, respectively. 
Descriptive statistics ($M_{\text{D}}\pm 2sd$) can be added by setting `include_stats = TRUE`. Finally, the color of the points can be changed by using argument `col_point`. 

```{r, fig.align='center', fig.width=6, fig.cap="d_point() function with   settings change"}
d_point(dscore, # dataframe containing IAT Ds
       order_sbj = "D-decreasing", # change respondents order
       x_values = FALSE, # remove respondents' labels
       include_stats = TRUE, # include descriptive statistics
       col_point = "lightskyblue") # change points color
```



#### Sample level plot

The `d_density()` function plots the distribution of the IAT *D* scores. It allows for choosing the most appropriate representation. 


```{r, fig.align='center', fig.width=6, fig.cap="d_density() function with default settings"}
d_density(dscore) # dataframe containing IAT Dscores
```

The number of bins can be changed with argument `n_bin`. Argument `graph` can be used for changing the graphical representation of the data. 
It is possible to choose an histogram representation (`graph = "histogram"`, default), a representation of the density distribution (`graph = "density"`), or a boxplot (`graph = "boxplot"`). 
Argument `col_fill` can be used to change the color of the points representing each respondent's score in the violin plot. 
Descriptive statistics (i.e., $M_{\text{D}}  \pm 2sd$) can be added to the graph by setting argument `include_stats = TRUE`. 

```{r sampleSettings, fig.align='center', fig.width=6, fig.cap="\\label{fig:sampleSettings}d_density() function with settings change"}
d_density(dscore, # dataframe containing IAT Dscores
        graph = "boxplot", # change graphical representation
        include_stats = TRUE) # include descriptive statistics
```


#### Multiple D scores

The `multi_dscore()` function computes and plot multiple *D* score algorithms. The *D* score algorithms that can be computed depend on the IAT administration. If the IAT administration included a feedback strategy (i.e., built-in correction), only *D1* and *D2* algorithms should be computed. If the IAT administration did not include a feedback strategy, then algorithms *D3*, *D4*, *D5*, and *D6* should be computed. An exhaustive and detailed illustration of the *D* score algorithms is provided in the "Implicit Measures" vignette. To specify the algorithms that can be computed, argument `ds` must be set equal to either `"built-in"` (for the computation of *D1* and *D2*) or `error-inflation` (for the computation of all other algorithms). 

```{r}
multi_scores <- multi_dscore(iat_data, # object with class "iat_clean"
                             ds = "error-inflation") # string specifying the 
                                            # algorithms to compute
```

The `multi_dscore()` function results in a `list` containing two objects. The first object is a `data.frame` that contains all the computed algorithms and the respondent IDs. 

```{r}
multi_d <- multi_scores[[1]]
head(multi_d)
str(multi_d)
```

The second object is a `ggplot` graph displaying the distribution of the *D* scores computed with each algorithm in violin plots.

```{r, fig.align='center', fig.width=6, fig.cap="Multiple IAT D scores representation"}
multi_scores[[2]]
```

# References
