---
title: "Lottery Brasil"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lotterybr)
```
This will open the shiny app for better visualisation of the datasets

## get_data

The function can be use in the following way

```{r}
quina_ganhadores = get_data("quina","winners","ptbr")
lotofacil_numbers = get_data("lotofacil","numbers","eng")
```

## Format

Each game has two datasets, numbers and winners, with the following columns:

### numbers
#### date

Date type, the date that the game occurred in.

#### course

Integer type, which course the game took place 

#### accumulated

Char type, if the game prize is accumulated "yes" or not "no"

#### numbers

Factor type with 60 levels, the drawn numbers.

### winners

#### date

Date type, the date that the game occurred in.

#### course

Integer type, which course the game took place .

#### accumulated

Char type, if the game prize is accumulated "yes" or not "no"

#### match

Char type, the possible match.

#### winners

Integer type, the number of winners of each possible match.

#### prize

Integer type, the prize for each type.

Some games, like "duplasena" or "diadesorte", may contain additional columns since each game has a different way of drawing numbers. Therefore, a brief explanation of each game is as follows.

### maismilionaria

Maismilionaria involves selecting five numbers from one pool and one additional number from another pool, with the potential to win multimillion-dollar jackpots.

### megasena

Six numbers are drawn from a pool of 60, with players aiming to match as many as possible for substantial prizes.

### lotofacil

Lotofacil involves picking 15 numbers from a pool of 25, with prizes awarded for varying levels of matches.

### quina

Players select five numbers from a pool of 80, aiming to match as many as possible for prizes.

### lotomania

Lotomania involves selecting 50 numbers from a pool of 100, with prizes awarded for matching a certain number of drawn numbers.

### duplasena

Dupla-Sena involves two separate draws, each with six numbers drawn from a pool of 50, offering double the chances to win.

### diadesorte

In Dia de Sorte, players choose seven numbers and a month, with prizes awarded for matching numbers and/or the month drawn.


