---
title: "A Simple Example With Migration"
author: "Eric C. Anderson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{simple-example-with-migration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette we show how to use CKMRpop/spip in a
multi-population version with asymmetrical migration rates.  It would
be good to read the "about_spip" vignette beforehand to understand how
migration is implemented in spip.

Our goal here is to model three "populations" of a fish species in which 
migration occurs during the larval stage, and it is  influenced by
currents that flow down the coast and make migration occur primarily in 
one direction.  Each of the populations will have the same life tables,
etc.  We will make it simple.  

First load some packages:
```{r, message=FALSE}
library(tidyverse)
library(CKMRpop)
```
## Basic Demography 

The basic demography for each, is given below.  We assume no
mortality from newborn to age one.  In effect we are only modeling the
the newborns that survive to age 1. Only 3, 4, and 5 year-olds mate,
and older fish have somewhat more reproductive success.
```{r}
pars <- list()
pars$`max-age` <- 5
pars$`fem-surv-probs` <- c(1, 0.7, 0.8, 0.8, 0.8)
pars$`male-surv-probs` <- c(1, 0.7, 0.8, 0.8, 0.8)
pars$`fem-prob-repro` <- c(0, 0, 1, 1, 1)
pars$`male-prob-repro` <- c(0, 0, 1, 1, 1)
pars$`fem-asrf` <- c(0, 0, .5, .7, 1)
pars$`male-asrp` <- c(0, 0, .5, .7, 1)
pars$`fem-rep-disp-par` <- 0.25
pars$`male-rep-disp-par` <- 0.25
pars$`sex-ratio` <- 0.5
```

We will run these simulations for 25 years and just assume
an annual cohort size of 200 fish in each population, and we will
go for a constant population size and start off with the stable
age distribution.
```{r}
pars$`number-of-years` <- 25

# given cohort sizes of 250 the stable age distribution
# can be found and used for the initial number of indivs
L <- leslie_from_spip(pars, 250)


# then we add those to the spip parameters
pars$`initial-males` <- floor(L$stable_age_distro_fem)
pars$`initial-females` <- floor(L$stable_age_distro_male)

pars$`cohort-size` <- "const 250"
```

## Sampling

We will simulate a sampling program in which juveniles are sampled
when they recruit after their larval phase.  We assume that the larval
phase starts at birth, but is over by age 1.  So we can sample new
recruits via prekill sampling at age 1.  We will pretend the sampling
is non-lethal. We will also non-lethally sample a fraction of the older
age-class fish.
```{r}
pars$`discard-all` <- 0
pars$`gtyp-ppn-fem-pre` <- "10-25 0.05 0.00 0.00 0.00 0.00"
pars$`gtyp-ppn-male-pre` <- "10-25 0.05 0.00 0.00 0.00 0.00"
pars$`gtyp-ppn-fem-post` <- "10-25 0.00 0.01 0.01 0.01 0.01"
pars$`gtyp-ppn-male-post` <- "10-25 0.00 0.01 0.01 0.01 0.01"
```


OK! At this point, we have set up the demography for a single population.  Let's just
show that we could simulate a single population with that demography and the sampling
like this:
```{r, eval=FALSE}
set.seed(15)
one_pop_dir <- run_spip(pars = pars, num_pops = 1)
```

But, of course, the question is, how do we simulate multiple, interacting populations?

## Simulating Multiple Populations

To simulate multiple populations in spip, from within CKMRpop, we set the
`num_pops` option of `run_spip()` to the number of populations.
And then, for the `pars` option we pass a _list_ of `num_pops`
different lists, each one holding the demography and sampling information
for each of the populations in the simulation.  In our case, since
we are simulating three populations with the same demography,
`num_pops = 3` and we can construct our `pars` argument like this:
```{r}
pars_list <- list(
  pars,
  pars,
  pars
)
```
because each population has the same demography.

Note that we have not included any migration between the the populations yet,
but we can still run a simulation with all of them like this:
```{r, echo=FALSE, results='hide', message=FALSE}
# NOTE the following if()...else() blocks are here
# to test whether spip has been installed yet.
# If spip is not available (for example, on CRAN's build machines) this
# is noted and stored package data are used for the variable
# "slurped" to build the remainder of the vignette.
if(spip_exists()) {
  message("spip is installed and will be used")
  set.seed(15)
  three_pop_dir <- run_spip(
    pars = pars_list,
    num_pops = 3
  )
  slurp3 <- slurp_spip(three_pop_dir, num_generations = 1)
} else {
  message("Using stored package data because spip is not installed")  
  slurp3 <- three_pops_no_mig_slurped_results
}
```
```{r, eval=FALSE}
set.seed(15)
three_pop_dir <- run_spip(
  pars = pars_list,
  num_pops = 3
)
slurp3 <- slurp_spip(three_pop_dir, num_generations = 2)
```


And you can investigate that output by making the plots and other
summaries shown in the main vignette.  Many of these summaries account
for the multiple populations like this:
```{r, fig.width=5, out.width = '100%'}
ggplot_census_by_year_age_sex(slurp3$census_postkill)
```

Those are the postkill census sizes in the three different populations, which get the labels
0, 1, and 2, according to the order in which they were passed to spip (i.e., their order
in the `pars_list`, above.)

## Migration between multiple populations

The above simulated multiple populations, but did not simulate any migration between them.

Here, we will imagine different rates of self-recruitment of the larvae, which means different rates
at which surviving larvae are going to be migrants.
The populations are labeled 0, 1, and 2, and, as stated above, we will simulate
somewhat unidirectional migration between them, like so:

- 20% of pop 0 individuals are migrants; 80% of those end up in pop 1 and 20% in pop 2
- 15% of pop 1 indivs are migrants; 95% of them end up in pop 2 and 5% in pop 0
- 5% of pop 2 are migrants; 90% of those end up in pop 1 and 10% make it to pop 0.

The way we implement this is by adding options to each population's demography parameters
like so:
```{r}
# out-migration rates pops 0, 1, 2
pars_list[[1]]$`fem-prob-mig-out` <- "5-25 1 .20"
pars_list[[1]]$`male-prob-mig-out` <- "5-25 1 .20"

pars_list[[2]]$`fem-prob-mig-out` <- "5-25 1 .15"
pars_list[[2]]$`male-prob-mig-out` <- "5-25 1 .15"

pars_list[[3]]$`fem-prob-mig-out` <- "5-25 1 .05"
pars_list[[3]]$`male-prob-mig-out` <- "5-25 1 .05"

# in-migration rates
pars_list[[1]]$`fem-prob-mig-in` <-  "5-25 1 0.00 0.80 0.20"
pars_list[[1]]$`male-prob-mig-in` <- "5-25 1 0.00 0.80 0.20"

pars_list[[2]]$`fem-prob-mig-in` <-  "5-25 1 0.05 0.00 0.95"
pars_list[[2]]$`male-prob-mig-in` <- "5-25 1 0.05 0.00 0.95"

pars_list[[3]]$`fem-prob-mig-in` <-  "5-25 1 0.10 0.90 0.00"
pars_list[[3]]$`male-prob-mig-in` <- "5-25 1 0.10 0.90 0.00"

```

Now, we can simulate this:
```{r, echo=FALSE, results='hide', message=FALSE}
# NOTE the following if()...else() blocks are here
# to test whether spip has been installed yet.
# If spip is not available (for example, on CRAN's build machines) this
# is noted and stored package data are used for the variable
# "slurped" to build the remainder of the vignette.
if(spip_exists()) {
  message("spip is installed and will be used")
  set.seed(15)
  mig_dir <- run_spip(
    pars = pars_list,
    num_pops = 3
  )
  slurp_mig <- slurp_spip(mig_dir, num_generations = 1)
} else {
  message("Using stored package data because spip is not installed")  
  slurp_mig <- three_pops_with_mig_slurped_results
}
```
```{r, eval=FALSE}
set.seed(15)
mig_dir <- run_spip(
  pars = pars_list,
  num_pops = 3
)
slurp_mig <- slurp_spip(mig_dir, num_generations = 1)
```


Now, the object `slurp_mig` holds information about census sizes of the populations
and the relationships between samples, etc.  That can be broken down by populations.

First, a record of every migration event is stored in `slurp_mig$migrants`. This can
be easily summarized:
```{r}
slurp_mig$migrants %>%
  count(age, from_pop, to_pop)
```
For a further example, to tabulate parent-offspring pairs found between different
populations, we can compile the relationships and summarize:
```{r}
# compile relationships
crel <- compile_related_pairs(slurp_mig$samples)

# count number of PO pairs by which populations
# the members were born in
crel %>%
  filter(dom_relat == "PO") %>%
  mutate(
    parent_born_pop = case_when(
      upper_member == 1 ~ born_pop_1,
      upper_member == 2 ~ born_pop_2,
      TRUE ~ NA_integer_
    ),
    child_born_pop = case_when(
      upper_member == 1 ~ born_pop_2,
      upper_member == 2 ~ born_pop_1,
      TRUE ~ NA_integer_
    )
  ) %>% 
  count(parent_born_pop, child_born_pop)
  
```


