---
title: "medicalrisk: Calculating risk and comorbidities from ICD-9-CM codes"
author: "Patrick McCormick <patrick.mccormick@alum.mit.edu>"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
bibliography:
   - medicalrisk.bib
vignette: >
  %\VignetteIndexEntry{medicalrisk: Calculating risk and comorbidities from ICD-9-CM codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteDepends{plyr}
---
Introduction
------------

The routines in the **medicalrisk** package [@R-medicalrisk]
are designed to help determine comorbidity and medical risk status of a given patient
using several popular models published in the peer-reviewed literature.

Administrative healthcare data is frequently the only available source for determining 
individual risk of mortality when looking at thousands or millions of patient records.
Medical chart abstraction just isn't feasible for projects of this scale.

In the United States, the records for every inpatient and outpatient encounter is 
reviewed by a qualified medical coder who assigns a set of diagnosis and procedural codes based
on phrases within the medical record.  The coding system currently in use is 
[ICD-9-CM](http://en.wikipedia.org/wiki/ICD-9-CM).
ICD-9-CM is an adaptation of the venerable ICD-9 standard which was developed in 1978.
The 
[U.S. National Center for Health Statistics (NCHS)](http://www.cdc.gov/nchs/) 
developed ICD-9-CM, which has been required for Medicare and Medicaid claims 
since 1979.  ICD-9-CM is updated annually.

At some point, perhaps as soon as October 2015, [ICD-10-CM codes](http://www.cdc.gov/nchs/icd/icd10cm.htm)
will need to be used instead.
It is likely that "dual coding" of claims in both sets will continue for some time.

In the meantime, there is a wealth of administrative data available within the ICD-9-CM 
diagnostic and procedural codes stored within US healthcare systems. 

Working with ICD-9-CM Data
------------

In order to demonstrate this package, this package includes data on 100 patients 
from the [Vermont Uniform Hospital Discharge Data Set for 2011, Inpatient](http://healthvermont.gov/research/hospital-utilization/RECENT_PU_FILES.aspx).

```{r}
library(medicalrisk)
library(plyr)
data(vt_inp_sample)
x <- count(vt_inp_sample, c('id'))
cat("average count of ICD codes per patient is: ", mean(x$freq))
y <- count(vt_inp_sample, c('icd9cm'))
```
```{r, results='asis'}
library(knitr)
kable(head(y[order(-y$freq),], n=5), row.names=F,
      caption='Top 5 most popular ICD-9-CM codes in this dataset')
```

Within this package, ICD-9-CM codes are presented as a string where the first letter 
is "P" or "D" depending on whether the code is Procedure or Diagnosis.  The rest of
the code is present as a string of numbers.  Periods are omitted.  In the list above, the
code "D4019" is diagnostic code 401.9 which corresponds to Hypertension.

Comorbidity Maps
--------------

The package includes a set of mapping functions that transform a list of ICD-9-CM
codes into a comorbidity matrix:

* icd9cm_charlson_deyo
* icd9cm_charlson_romano
* icd9cm_charlson_quan
* icd9cm_elixhauser_ahrq37
* icd9cm_elixhauser_quan
* icd9cm_rcri

"Charlson" refers to the Charlson Comorbidity Index[@Charlson1987].  
The names 
"Deyo"[@Deyo1992], 
"Romano"[@Romano1993], and 
"Quan"[@Quan2005]
refer to the primary authors of different methods of determining Charlson 
comorbidities from ICD-9-CM codes.

"Elixhauser" refers to the Elixhauser comorbidity map, 
which is a more detailed list than 
Charlson.  "AHRQ37" is an adapation of the 
AHRQ version 37 software[@AgencyforHealthcareResearchQuality2013].  
"Quan" refers to the same paper by Quan mentioned above.

"RCRI" is the Revised Cardiac Risk Index[@Lee1999] 
set of categories using a 
method published by Boersma[@Boersma2005].

For example, the #5 ICD-9-CM code above is D25000, or "250.00", which is for
"Diabetes Mellitus Unspecified Type".  Here's what happens when that code is passed
to a few of the mapping functions listed above:

```{r, results='asis'}
kable(
  icd9cm_charlson_quan(c('D25000')))
kable(
  icd9cm_elixhauser_ahrq37(c('D25000')))
kable(
  icd9cm_rcri(c('D25000')))
```

For each of these maps the "dm" column becomes TRUE.

The most efficient way to use these maps for a set of patients is to 
generate a single map for all ICD-9-CM codes in the set and then
apply that map to each patient.  Here's an example that
generates a comorbidity matrix for the first five patients in the Vermont dataset:

```{r, results='asis'}
cases <- vt_inp_sample[vt_inp_sample$id %in% 1:5, c('id','icd9cm')]
cases_with_cm <- merge(cases, icd9cm_charlson_quan(levels(cases$icd9cm)), 
   by.x="icd9cm", by.y="row.names", all.x=TRUE)
 
# generate crude comorbidity summary for each patient
kable(
  ddply(cases_with_cm, .(id), 
   function(x) { data.frame(lapply(x[,3:ncol(x)], any)) }),
  row.names=F)
```

The above process is encapsulated in a single function `generate_comorbidity_df`.
This function also includes an optimization from Van Walraven that
reduces `dmcx` to `dm` if the specific diabetic complication is separately coded.

```{r, results='asis'}
kable(
  generate_comorbidity_df(cases, icd9mapfn=icd9cm_charlson_quan))
```

This function only considers each ICD-9-CM code once and then merges the
resulting comorbidity flags together for each patient.  This makes the function 
quite fast for large data sets.

Given appropriate arguments, the `generate_comorbidity_df` function will 
use the parallel backend provided by `foreach` to improve performance.

Comorbidity Index
--------------

It is common in the medical literature to see a set of comorbidities reduced to an index.
When the Charlson Comorbidity Index was first published it had the following weights
for each comorbidity:

```{r, results='asis'}
data(charlson_weights_orig)
kable(
  t(charlson_weights_orig))
```

However, these weights have not stood the test of time.  For example, the
prognosis for HIV/AIDS has dramatically improved.  
The medicalrisk package offers the 
revised Charlson weights developed by Schneeweiss[@Schneeweiss2003]:

```{r, results='asis'}
data(charlson_weights)
kable(
  t(charlson_weights))
```

The `generate_charlson_index_df` function will sum the weights for each patient 
to generate a final index:

```{r, results='asis'}
kable(
  generate_charlson_index_df(generate_comorbidity_df(cases)), row.names=F)
```

Risk Stratification Index
--------------

The Risk Stratification Index uses ICD-9-CM codes to determine four risk estimates:

* 1 Year Mortality
* 30 Day Mortality
* In-Hospital Mortality
* 30 Day Length of Stay

The author of the paper (Sessler) published SPSS code to perform the calculation. 
The medicalrisk implements the RSi calculation using a method based on the SPSS code.

```{r}
ddply(cases, .(id), function(x) { icd9cm_sessler_rsi(x$icd9cm) } )
```

Conclusion
----------

The medicalrisk package can be used to generate risk data from ICD-9-CM codes 
in large datasets.  The above discussion describes basic use of the package.
There are some additional helper functions not described above which are included
in the per function documentation.  

The aim of this package is to include future medical risk estimation procedures as
they are published in the literature.

## References
