---
title: "Spatial Utility Functions"
author: "Rob Williams"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Utility Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3, 
  fig.align = 'center'
)

## old graphical parameters
oldpar <- par(mfrow = c(1,2))
```

This package contains convenience functions for carrying out GIS operations
that I have repeatedly encountered in my research. The following packages are
used in this vignette:

```{r setup, message = FALSE, warning = FALSE, results = 'hide'}
library(sf)
library(raster)
library(RWmisc)
```

# Weighting raster values by overlapping polygons

The `overlap.weight` function allows you to weight the values of a raster cell
by the inverse of the number of polygons that overlap the cell. This is useful
when, e.g., calculating the population of ethnic group settlement areas when
different group settlement areas can overlap one another. The `count` argument
allows the count of overlapping polygons to be returned instead of the weighted
cell values. Note that this converts any cells not covered by at least one
polygon to `NA`.

```{r overlap-weight}
## create three overlapping squares
polys_t <- st_sfc(list(st_polygon(list(rbind(c(2,2), c(2,6), c(6,6),
                                             c(6,2), c(2, 2)))),
                       st_polygon(list(rbind(c(8,8), c(4,8), c(4,4),
                                             c(8,4), c(8,8)))),
                       st_polygon(list(rbind(c(3,3), c(3,7), c(7,7),
                                             c(7,3), c(3,3))))),
                  crs = st_crs('OGC:CRS84'))

## create raster
raster_t <- raster(nrows = 6, ncols = 6, xmn = 2, xmx = 8, ymn = 2, ymx = 8,
                   vals = 1:36, crs = CRS(st_crs(polys_t)$proj4string))

## set plotting parameters
par(mfrow = c(1, 3))

## plot raw raster values
plot(raster_t, main = 'raw')
plot(polys_t, add = TRUE)

## plot count of overlapping polygons
plot(overlap.weight(raster_t, polys_t, count = TRUE), main = "count")
plot(polys_t, add = TRUE)

## plot overlap-weighted raster values
plot(overlap.weight(raster_t, polys_t), main = "weighted")
plot(polys_t, add = TRUE)
```

# Projecting spatial objects to UTM

The `projectUTM` function converts any `sf` or `sfc` objects in longitude,
latitude decimal degrees to the UTM zone where the majority of the data lie.
This function accounts for North and South UTM zones as well.

```{r projectUTM}
## read in North Carolina shapefile
nc <- st_read(system.file("shape/nc.shp", package="sf"))

## transform crs to WGS84 and inspect CRS
nc <- st_transform(nc, st_crs('OGC:CRS84'))
st_crs(nc)

## project to UTM and inspect CRS
st_crs(projectUTM(nc))
```

Projection of the North Carolina polygons can be further seen by plotting them.

```{r projectUTM-plot}
## set plotting parameters
par(mfrow = c(1, 2), mar = rep(0, 4))

## plot WGS84 and UTM projected North Carolina
plot(nc$geometry)
plot(projectUTM(nc)$geometry)
```

# Maximum and minimum distance from point(s) to a polygon

The `point.poly.dist` function computes the maximum or minimum distance from a
point or set of points to a polygon. It correctly calculates distances for both
geographic and projected data.

```{r point-poly-dist}
## create north carolina centroids
nc_centroids <- st_centroid(nc)

## calculate maximum distance
point.poly.dist(nc_centroids[53,]$geometry, nc[53,]$geometry, max = TRUE)
```

The following illustration depicts the line connecting the centroid of the
polygon to the farthest point on the polygon (red) and the nearest point on the
polygon (blue).

```{r point-poly-dist-illus}
nc_points <- st_geometry(nc[53,]) %>%
  st_cast('POINT')

farthest_ind <- st_distance(nc_points, nc_centroids[53,]) %>%
  which.max()

farthest_point <- rbind(st_coordinates(nc_points[farthest_ind]),
                        st_coordinates(nc_centroids[53,])) %>% 
  st_linestring()

nearest_ind <- st_distance(nc_points, nc_centroids[53,]) %>%
  which.min()

nearest_point <- rbind(st_coordinates(nc_points[nearest_ind]),
                        st_coordinates(nc_centroids[53,])) %>% 
  st_linestring()

## plot
par(mar = rep(0,4))
plot(nc[53,]$geometry)
plot(nc_centroids[53,]$geometry, add = TRUE)
plot(farthest_point, add = TRUE, col = 'red')
plot(nearest_point, add = TRUE, col = 'blue')
```

Carrying out the same calculations using built-in `sf` functions  takes roughly
twice as long to execute.

```{r point-poly-dist-benchmark}
microbenchmark::microbenchmark(pk = point.poly.dist(nc_centroids[53,]$geometry,
                                                    nc[53,]$geometry, max = TRUE),
                               sf = st_distance(st_cast(st_geometry(nc[53,]),
                                                        'POINT')[farthest_ind],
                                                nc_centroids[53,]),
                               times = 100)
```

```{r oldpar, echo = FALSE}
par(oldpar)
```

