---
title: "iemisc: Sound Frequencies & Nikola Tesla&#8217;s 3-6-9 Theory"
author: "Irucka Embry, E.I.T. (EcoC²S)"
date: "`r Sys.Date()`"
lang: en-us
urlcolor: blue
output:
  rmarkdown::pdf_document:
    highlight: kate
    toc: true
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{iemisc Sound Frequencies & Nikola Tesla&#8217;s 3-6-9 Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\bigskip

# All 9 Solfeggio Frequencies

```{r, warning = FALSE, message = FALSE, tidy = TRUE}

library(iemisc)

reduce_single_digit("174 Hz")

reduce_single_digit("285 Hz")

reduce_single_digit("396 Hz")

reduce_single_digit("417 Hz")

reduce_single_digit("528 Hz")

reduce_single_digit("639 Hz")

reduce_single_digit("741 Hz")

reduce_single_digit("852 Hz")

reduce_single_digit("963 Hz")

```

\bigskip
\bigskip

# A and C Musical Frequencies

```{r, warning = FALSE, message = FALSE, tidy = TRUE}

reduce_single_digit("432 Hz") # A = 432 Hertz

reduce_single_digit("440 Hz") # A = 440 Hertz

reduce_single_digit("444 Hz") # A = 444 Hertz

reduce_single_digit("128 Hz") # C = 128 Hertz

reduce_single_digit("256 Hz") # C = 256 Hertz

reduce_single_digit("512 Hz") # C = 512 Hertz

reduce_single_digit("528 Hz") # C = 528 Hertz

```

# R Help for iemisc Function

Please refer to the `iemisc` [https://CRAN.R-project.org/package=iemisc] help definition for the `reduce_single_digit` function below for more information on the function, including references for the significance of the individual numbers 1 - 9:  

\bigskip
\bigskip

```{r, echo = FALSE}

loadNamespace("printr")
```

```{r tidy = TRUE, comment = "", results = "asis"}

help(reduce_single_digit, package = "iemisc")

```

\bigskip
\bigskip
\bigskip
\bigskip

# Useful References

‪432 Hz vs 440 Hz Uploaded by WE VIBE HIGH on Dec 7, 2014 (YouTube video) [https://yewtu.be/watch?v=wZWDWXkVoOs]  

Brendan D. Murphy, Wake Up World: The A=432 Hz Frequency: DNA Tuning and the Bastardization of Music, [https://web.archive.org/web/20210526035709/https://wakeup-world.com/2015/08/26/the-a432-hz-frequency-dna-tuning-and-the-bastardization-music/]     {Recovered with the Internet Archive: Wayback Machine}  

Meditative Mind: Nicola Tesla&#8217;s 3-6-9 Theory: What You Need To Know￼ [https://meditativemind.org/nicola-teslas-3-6-9-theory-what-you-need-to-know%EF%BF%BC/]  

Schiller Institute and Fidelio Online: A Revolution in Musical Tuning: Return to Verdi&#8217;s Scientific Pitch C=256 Hertz [https://archive.schillerinstitute.com/music/revolution.html]  

Solfeggio Guide: Solfeggio Frequency Guide [https://solfeggioguide.com/solfeggio-frequency-guide/]  

Sondra Barrett. The Universal Law of 3 – it works! May 27, 2022 [https://sondrabarrett.com/2022/05/27/the-universal-law-of-3-it-works/]  

\bigskip
\bigskip

# EcoC^2^S Links

EcoC^2^S Home -- https://www.ecoccs.com/  
About EcoC^2^S -- https://www.ecoccs.com/about-ecoc2s.html  
Services -- https://www.ecoccs.com/services.html  
1 Stop Shop -- https://www.ecoccs.com/other-biz.html  
Products -- https://www.questionuniverse.com/products.html  
Media -- https://www.ecoccs.com/media.html  
Resources -- https://www.ecoccs.com/resources.html  
R Trainings and Resources provided by EcoC^2^S (Irucka Embry, EIT) -- https://www.ecoccs.com/rtraining.html  

\bigskip
\bigskip

# Donations accepted with Liberapay

If you would like to contribute to the continued development of [Irucka Embry&#8217;s R packages](https://www.ecoccs.com/rtraining.html#repos) and/or [Irucka Embry&#8217;s R Examples](https://www.ecoccs.com/rtraining.html#examp), please feel free to donate via the link below:  

[https://liberapay.com/iaembry/donate](https://liberapay.com/iaembry/donate)  

Please feel free to review [Irucka Embry (iaembry)&#8217;s profile](https://liberapay.com/iaembry/) on [Liberapay](https://liberapay.com).  

\bigskip
\bigskip

# Copyright and License

All R code written by Irucka Embry is distributed under the GPL-3 (or later) license, see the [GNU General Public License {GPL} page](https://www.gnu.org/licenses/gpl-3.0.html).  

All written content originally created by Irucka Embry is copyrighted under the Creative Commons Attribution-ShareAlike 4.0 International license. All other written content retains the copyright of the original author(s).  

This work is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International license](https://creativecommons.org/licenses/by-sa/4.0/).  
