---
title: "W4MRUtils"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{W4MRUtils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction: package history and purpose

Workflow4Metabolomics (W4M) provides tools to help people to process metabolomic datasets.
Its preferred Chanel is the web-based interface [Galaxy](https://galaxyproject.org/),
particularly suited for analysts with no skill regarding software programing nor programing language in general.
Galaxy also has the advantage to be compatible with a large variety of languages, enabling to combine
tools coded in various programing languages, which is a huge advantage regarding to the diversity of steps
(and thus tools) needed for comprehensive metabolomic data processing.

Nonetheless, a significant part of the tools provided by W4M are coded using R.
Due to the diversity of tools, W4M chose to adopt common standards across its tools,
in particular regarding data table formats and basics data checks.
It also intents to gather utility functions to help the construction of R-based Galaxy tools using W4M standards,
along with functions to easily handle basic actions given the W4M table data format.

The `W4MRUtils` package is meant to gather such functions, to enable W4M R-based Galaxy tools to
have easy access to these standardised approaches.

# How to use the package

## What you should know about the W4M standards

Some functions are closely linked to Galaxy integration as the `parse_args` function for example,
but a large variety of functions can be helpful independently of Galaxy.
To use these functions, it is useful to understand the choices made by W4M,
in particular regarding the formats, as these will be often used as input parameters for functions.

The main format aspect to consider is the W4M 3-tables format.
Metabolomic data, similarly to other types of 'omics data, can be divided in 3 main types of information:

- analysed samples' information (sample meta-data, `sampleMetadata` in short): it corresponds to information
regarding the samples that have been analysed with an analytical device; this covers any information about the samples
that is not the analytical measures themselves (examples: studied biological groups, batches of analysis)
- analytical measures (`dataMatrix` in short): it corresponds to the intensities measured
using the analytical device used; it is a table with intensity values for each sample and for each analytical variable
- analytical variables' information (variable meta-data, `variableMetadata` in short): it corresponds to
information regarding the variables that have been measured through the analytical device; this covers
any information about the variables that is not the analytical measures themselves (examples: m/z ratios for MS data,
chemical shift for NMR data)

These 3 types of information can be stocked in 3 distinct tables.
In the package, the chosen format is the following:

- sampleMetadata: the samples' identifiers are positioned as the first column in the table; the table can contain
as many columns as you want, each column being a specific information about samples (*e.g.* a header indicating 'Groups'
and the content being 'A' or 'B' depending on each sample's affiliation)
- variableMetadata: same idea than the sampleMetadata, but for analytical variables; thus the analytical variables'
identifiers are positioned as the first column in the table, and additional columns correspond to information about
the variables (*e.g.* the p-values associated to each variable regarding a given test applied on each variable)
- dataMatrix: the analytical variables' identifiers are positioned as the first column, and the samples' identifiers
are given as the header; the remaining cells are quantitative values corresponding to the measured intensities

These 3 tables are standard inputs and/or outputs for several of the functions found in this package.
Consequently, when you need to specify sample-wise or variable-wise information, you will generally be invited to
stock it in one of these tables (same goes for outputs of functions, where to find the generated information).

## About the types of functions in the package

There are several types of functions in the package.
Globally, you will find functions enabling the following:

- importing and/or checking parameters (in a Galaxy-oriented way)
- importing and/or checking data tables (in a W4M-oriented way)
- managing identifiers and/or formats
- managing computing information and/or error messages (in a user-targeted way)



# Usage example for galaxy tools

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, include = FALSE}

input <- file.path(tempdir(), "./input.csv")
commandArgs <- function() { #nolint
  list(
    "--input", input,
    "--output", file.path(tempdir(), "./output.csv"),
    "--threshold", "2"
  )
}
write.csv(
  data.frame(x = c(10, 2, 1, 4, 7, 9), y = c("a", "b", "c", "d", "e", "f")),
  input,
  row.names = FALSE
)
```

```{r setup}
library(W4MRUtils)

TOOL_NAME <- "A Test Tool" #nolint
```

```{r create a logger}
logger <- get_logger(TOOL_NAME)
```

```{r read parameters}
logger$info("Parsing parameters...")
args <- optparse_parameters(
  input = optparse_character(),
  output = optparse_character(),
  threshold = optparse_numeric(default = 10),
  args = commandArgs()
)
logger$info("Parsing parameters OK.")
show_galaxy_header(
  TOOL_NAME,
  "1.2.0",
  args = args,
  logger = logger,
  show_sys = FALSE
)
```

```{r check parameters}
check_parameters <- function(args, logger) {
  logger$info("Checking parameters...")
  check_one_character(args$input)
  check_one_character(args$output)
  check_one_numeric(args$threshold)
  if (args$threshold < 1) {
    logger$errorf(
      "The threshold is too low (%s). Cannot continue", args$threshold
    )
    stopf("Threshold = %s is tool low.", args$threshold)
  }
  if (args$threshold < 3) {
    logger$warningf(
      paste(
        "The threshold is very low (%s).",
        "This may lead to erroneous results."
      ),
      args$threshold
    )
  }
  logger$info("Parameters OK.")
}

read_input <- function(args, logger) {
  logger$infof("Reading file in %s ...", args$input)
  return(read.csv(args$input))
}

process_input <- function(args, logger, table) {
  filter <- table[1, ] > args$threshold
  logger$infof(
    "%s values filtered (%s%%)",
    length(filter),
    round(length(filter) / nrow(table) * 100, 2)
  )
  if (length(filter) == nrow(table)) {
    logger$warning("All values have been filtered.")
  }
  return(table[which(table[, 1] > args$threshold), ])
}

write_output <- function(args, logger, output) {
  logger$info("Writing output file...")
  write.table(output, args$output)
}

# check_parameters(args, logger$sublogger("Param Checking"))
# input <- read_input(args, logger$sublogger("Inputs Reader"))
# print(input)
# output <- process_input(args, logger$sublogger("Inputs Processing"), input)
# print(output)
# write_output(args, logger$sublogger("Output Writter"), output)
# show_galaxy_footer(TOOL_NAME, "1.2.0", logger = logger, show_packages = FALSE)
```
