---
title: "sgat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sgat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sgat

<!-- badges: start -->
<!-- badges: end -->

The goal of sgat package is to retrieve information from Google's searches' "Popular Times" and geolocation.

## Installation

You can install the released version of sgat package from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("sgat")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("matiaspoullain/sgat")
```
## Example

Once you've download the information, you can use it to compare the mos popular hours in the week between places and there relative poositions. Here is an example of its use in Pinamar city in Argentina:

https://matiaspoullain.shinyapps.io/Concurrencias_Pinamar/?_ga=2.194611503.386324231.1613673572-1914966656.1612880733

All that information was recolected using the following code:
```{r example pinamar, echo = TRUE, eval = FALSE, message=FALSE, warning=FALSE}
library(sgat)
initialization_sgat()

rest.pinamar <- restaurantes_general("Pinamar, argentina")

rest.pinamar <- paste(rest.pinamar, ", Pinamar, Argentina", sep = "")

for(i in rest.pinamar){
  sgat(i, tiempo.espera = 20, carpeta.guardado = "Pinamar")
}

```
