---
title: "An introduction to cesR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cesR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Standard Use

Load the package with the following code:

```{r setup}
library(cesR)
```


The purpose of `cesR` is to make loading and working with the Canadian Election Study (CES) datasets in R more efficient. This is done through the use of five built-in functions.


## Get download codes

In `cesR`, each CES dataset is associated with a code that can be used to download and load a dataset into the R workspace. The function for to get these codes is `get_cescodes()`, which prints a dataframe of the codes and the associated datasets.

```{r}
get_cescodes()
```


## Download and load CES datasets

Using the `get_ces()` function, it is possible to download and load a CES dataset using the values under the `get_ces_call_char` column in the printout provided by `get_cescodes()`. For the `get_ces()` function to work, it is necessary that the parameter passed to the function be a character string. Below, the 2019 CES online survey is being called.

```{r, eval = F}
get_ces("ces2019_web")

head(ces2019_web)
# A tibble: 6 x 620
  cps19_StartDate     cps19_EndDate       cps19_ResponseId            cps19_consent
  <dttm>              <dttm>              <chr>                           <dbl+lbl>
1 2019-09-13 08:09:44 2019-09-13 08:36:19 R_1OpYXEFGzHRUpjM 1 [I consent to partic~
2 2019-09-13 08:39:09 2019-09-13 08:57:06 R_2qdrL3J618rxYW0 1 [I consent to partic~
3 2019-09-13 10:01:19 2019-09-13 10:27:29 R_USWDAPcQEQiMmNb 1 [I consent to partic~
4 2019-09-13 10:05:37 2019-09-13 10:50:53 R_3IQaeDXy0tBzEry 1 [I consent to partic~
5 2019-09-13 10:05:52 2019-09-13 10:32:53 R_27WeMQ1asip2cMD 1 [I consent to partic~
6 2019-09-13 10:10:20 2019-09-13 10:29:45 R_3LiGZcCWJEcWV4P 1 [I consent to partic~
# ... with 616 more variables: cps19_citizenship <dbl+lbl>, cps19_yob <dbl+lbl>,
#   cps19_yob_2001_age <dbl+lbl>, cps19_gender <dbl+lbl>,
#   cps19_province <dbl+lbl>, cps19_education <dbl+lbl>, cps19_demsat <dbl+lbl>,
#   cps19_imp_iss <chr>, cps19_imp_iss_party <dbl+lbl>,
#   cps19_imp_iss_party_7_TEXT <chr>, cps19_imp_loc_iss <chr>,
#   cps19_imp_loc_iss_p <dbl+lbl>, cps19_imp_loc_iss_p_7_TEXT <chr>,
#   cps19_interest_gen_1 <dbl>, cps19_interest_elxn_1 <dbl>, ...
```


## Find a survey question for a variable

If you wish to know the survey question associated with a given variable for a CES dataset, the `get_question()` function will provide this information using a provided data object and column label. The function `colnames()` is useful to return the names of columns that can be used with `get_question()`.

```{r, eval = F}
colnames(ces2019_web)

get_question("ces2019_web", "cps19_province")
Which province or territory are you currently living in?
```

It should be noted that the `get_question()` function does not work with the 2019 CES phone dataset as it is downloaded in a `.tab` file type, and as such does not contain the necessary variable label information.


## Preview a dataset

If you don't wish to access an entire CES dataset, the `get_preview()` function provides a means of previewing a dataset for a desired number of rows. For this, you can specify the number of observations you wish returned, or leave the second parameter empty to return a default of 6 observations. The following would return a preview of 10 observations from the 2019 CES online survey.

```{r, eval = F}
get_preview("ces2019_web", 10)
# A tibble: 10 x 620
   cps19_StartDate     cps19_EndDate       cps19_ResponseId  cps19_consent         
   <dttm>              <dttm>              <chr>             <fct>                 
 1 2019-09-13 08:09:44 2019-09-13 08:36:19 R_1OpYXEFGzHRUpjM I consent to particip~
 2 2019-09-13 08:39:09 2019-09-13 08:57:06 R_2qdrL3J618rxYW0 I consent to particip~
 3 2019-09-13 10:01:19 2019-09-13 10:27:29 R_USWDAPcQEQiMmNb I consent to particip~
 4 2019-09-13 10:05:37 2019-09-13 10:50:53 R_3IQaeDXy0tBzEry I consent to particip~
 5 2019-09-13 10:05:52 2019-09-13 10:32:53 R_27WeMQ1asip2cMD I consent to particip~
 6 2019-09-13 10:10:20 2019-09-13 10:29:45 R_3LiGZcCWJEcWV4P I consent to particip~
 7 2019-09-13 10:14:47 2019-09-13 10:32:32 R_1Iu8R1UlYzVMycz I consent to particip~
 8 2019-09-13 10:15:39 2019-09-13 10:30:59 R_2EcS26hqrcVYlab I consent to particip~
 9 2019-09-13 10:15:48 2019-09-13 10:37:45 R_3yrt44wqQ1d4VRn I consent to particip~
10 2019-09-13 10:16:08 2019-09-13 10:40:14 R_10OBmXJyvn8feYQ I consent to particip~
# ... with 616 more variables: cps19_citizenship <fct>, cps19_yob <fct>,
#   cps19_yob_2001_age <fct>, cps19_gender <fct>, cps19_province <fct>,
#   cps19_education <fct>, cps19_demsat <fct>, cps19_imp_iss <chr>,
#   cps19_imp_iss_party <fct>, cps19_imp_iss_party_7_TEXT <chr>,
#   cps19_imp_loc_iss <chr>, cps19_imp_loc_iss_p <fct>,
#   cps19_imp_loc_iss_p_7_TEXT <chr>, cps19_interest_gen_1 <dbl>,
#   cps19_interest_elxn_1 <dbl>, cps19_v_likely <fct>, ...
```


## A prepared dataset

`cesR` also provides a prepared, non-exhaustive dataset for testing and teaching purposes. The provided function `get_decon()` calls and creates this dataset, which is made up of variables associated with demographics, economics, and voting intentions.

```{r, eval = F}
get_decon()

head(decon)
# A tibble: 6 x 52
  ces_code    citizenship      yob   gender  province_territory education     vote_likely
  <chr>       <fct>            <fct> <fct>   <fct>              <fct>         <fct>      
1 ces2019_web Canadian citizen 1989  A woman Quebec             Master's deg~ Certain to~
2 ces2019_web Canadian citizen 1998  A woman Quebec             Master's deg~ Certain to~
3 ces2019_web Canadian citizen 2000  A woman Ontario            Some univers~ Certain to~
4 ces2019_web Canadian citizen 1998  A man   Ontario            Some univers~ Certain to~
5 ces2019_web Canadian citizen 2000  A woman Ontario            Completed se~ Certain to~
6 ces2019_web Canadian citizen 1999  A woman Ontario            Some univers~ Certain to~
# ... with 45 more variables: vote_likely_ifable <fct>, votechoice <fct>,
#   votechoice_text <chr>, votechoice_couldvote <fct>, votechoice_couldvote_text <chr>,
#   vote_unlikely <fct>, vote_unlikely_text <chr>, vote_unlikely_couldvote <fct>,
#   vote_unlikely_couldvote_text <chr>, vote_advancevote_choice <fct>,
#   vote_advancevote_choice_text <chr>, vote_partylean <fct>, vote_partylean_text <chr>,
#   vote_partylean_couldvote <fct>, vote_partylean_couldvote_text <chr>,
#   votechoice_secondchoice <fct>, votechoice_secondchoice_text <chr>, ...
```

